### 2022-06-15 paper/writeup.tex
outline of paper

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{natbib}

\newcommand{\E}{\mathbb{E}}
\renewcommand{\P}{\mathbb{P}}

\begin{document}

% Name ideas:
% 
% extend edges
% bundle edges
% inflate edges
% bundle lines of descent
% longer ancestral haplotypes
% inflated ancestors
% compress paths
% optimizing edge tables
% reduce number of ancestors
% reduce ancestral paths

\section{Introduction}

% PETER
% * what's a tree sequence
% * why is a tree sequence (motivation)

The pedigree that describes how everyone of some species
are related to each other
is a large graph, where nodes are indexed by time.
Within this, one can trace back the paths along which segement of genome
have been inherited.
The \emph{succinct tree sequence} (or, tree sequence for short)
was introduced by \citet{kelleher2016efficient} to describe the subset of these paths
along which the genome of a sample of individuals was inherited.

\section{Motivation and statement of problem}

% PETER
% * minimize number of edges
% * gives extra info about shared haplotypes, reduces number of ancestral paths

\section{Algorithm}

% * description (HALLEY)
% * proof of something:
%     - arrives at a local minimum?
%     - guess at typical reduction?

\section{Results (how it works)}

% (ADAM)
% * reduction in edges ~ sequence length
% * speed increase ~ sequence length
% * proprotion of added edges that are true ~ sequeence length
% * apply to real data:
%     - % fewer edges
%     - % fewer distinct ancestors (ancestral paths?) of sample at some time

\bibliography{references}

\end{document}

### 2022-09-12 paper/writeup.tex
Update writeup.tex
some changes

@@
 \usepackage{natbib}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
+\usepackage{tikz, tikzcd}
 
@@
 \renewcommand{\P}{\mathbb{P}}
+\newcommand{\R}{\mathbb{R}}
 
@@
 
+\begin{section*}{Notation}
+    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
+    as a tuple $\left\( \left\{T_k\right\}, N, E, L\right\)$ where
+     $\left\{T_k\right\}_{k=1}^n is the set of trees,
+     $N$ and $E$ are the sets of nodes and edges respecitively,
+     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome. 
+    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
+    ie. only a single edge extending below it on the tree.
+    With these tree sequences our goal is to
+    minimize the number of edges
+    as well as infer coalescence of genes. 
+    To remove edges we consider the following example.
+    Suppose for a tree $T_k$ in the tree sequence $TS$ 
+    there exists a branch which contains a unary node $b\in N$.
+    Then the node $b$ has one parent, $a\in N$, 
+    and one child $c\in N$,
+    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
+    If there is an edge between the nodes $a$ and $c$ in trees 
+    $T_{k+1} (or $T_{k-1}$),
+    then we wish to extend the edges $e_{ab},e_{bc}$ 
+    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}. 
+    This action reduces the length of the edge $e_{ac}$ on the genome,
+    and in some cases, completely removes the edge. 
+    We now perfrom this action on all such unary nodes 
+    over the entire tree sequence.
+
+    In full detail, let $TS = \left\( \left\{T_k\right\}_{k=1}^n, N, E, L\right\)
+    be a tree sequence.
+
+\begin{algorithm}
+\caption{Extend Edges}\label{alg:cap}
+\begin{algorithmic}
+    N' <- N
+    e'\in E' <- 
+    T'_k <- T_{n-k}
+    T' <- (\{T'_k\}_{k=1}^n,N',E',L)
+
+\end{algorithmic}
+\end{algorithm}
+
 \section{Results (how it works)}
 
-% (ADAM)
+% (AVA)
 % * reduction in edges ~ sequence length

### 2022-09-14 paper/writeup.tex
minor fixes

@@
 \usepackage{algpseudocode}
-\usepackage{tikz, tikzcd}
+\usepackage{tikz}
 
@@
 
-\begin{section*}{Notation}
+\section{Notation}
     We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
-    as a tuple $\left\( \left\{T_k\right\}, N, E, L\right\)$ where
+    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
      $\left\{T_k\right\}_{k=1}^n is the set of trees,
@@
 
-    In full detail, let $TS = \left\( \left\{T_k\right\}_{k=1}^n, N, E, L\right\)
+    In full detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$
     be a tree sequence.

### 2022-09-15 algorithm/_metric_.py
discrepancy update

@@ -1,0 +1,32 @@
+import numpy as np
+import tskit
+
+def node_tree_discrepancy(x, ts1, ts2):
+    # here sim[y] is the span so far over which y has the same set of descendants as x
+    sim = np.zeros(ts2.num_nodes)
+   
+    for interval, t1, t2 in ts1.coiterate(ts2):
+        s = interval.span
+        Dx = set(t1.samples(x))
+
+        if len(Dx) > 0:
+            if len(Dx) == 1:
+                y = list(Dx)[0]
+                MRCA = list(Dx)[0] 
+            else:
+                MRCA = t1.mrca(*list(Dx))
+                y = t2.mrca(*list(Dx))
+            if set(t2.samples(y)) == Dx:
+                sim[y] += s 
+                y = t2.parent(y)
+                if x != MRCA:
+                    while y != tskit.NULL and set(t2.samples(y)) == Dx:
+                        sim[y] += s
+                        y = t2.parent(y)
+    return ts1.sequence_length - np.max(sim)
+
+def discrepancy(ts1, ts2):
+    dis = 0
+    for n in range(ts1.num_nodes):
+        dis += node_tree_discrepancy(n, ts1, ts2)
+    return dis

### 2022-09-23 paper/writeup.tex
updated to do list
readable now

@@
 \usepackage{natbib}
-\usepackage{algorithm}
+\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
 \usepackage{algpseudocode}
 \usepackage{tikz}
+\usetikzlibrary{arrows, snakes,backgrounds}
+\tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]
+
+\SetKwInput{input}{Input}
+\SetKwInput{output}{Output}
 
@@
     as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
-     $\left\{T_k\right\}_{k=1}^n is the set of trees,
-     $N$ and $E$ are the sets of nodes and edges respecitively,
+     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
+     $N$ and $E$ are the sets of nodes and edges respectively,
      and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome. 
@@
     If there is an edge between the nodes $a$ and $c$ in trees 
-    $T_{k+1} (or $T_{k-1}$),
+    $T_{k+1}$ (or $T_{k-1}$),
     then we wish to extend the edges $e_{ab},e_{bc}$ 
-    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}. 
+    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
     This action reduces the length of the edge $e_{ac}$ on the genome,
     and in some cases, completely removes the edge. 
-    We now perfrom this action on all such unary nodes 
+    We now perform this action on all such unary nodes 
     over the entire tree sequence.
+
+\begin{tikzpicture} %% Do this with phylogenetic trees using tikz (WILL NEED LUALATEX COMPILER) %%
+\node[place] (a)  {a};
+\node[place] (b)  [below of = a] {b};
+\node[place] (c)  [below of = b] {c};
+\draw [-] (a.south) -- (b.north);
+\draw [-] (b.south) -- (c.north);
+\end{tikzpicture}
 
@@
 
-\begin{algorithm}
+\begin{algorithm}  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
 \caption{Extend Edges}\label{alg:cap}
 \begin{algorithmic}
-    N' <- N
-    e'\in E' <- 
-    T'_k <- T_{n-k}
-    T' <- (\{T'_k\}_{k=1}^n,N',E',L)
-
+\SetKwFunction{mirror}{Mirror Coordinates}
+\State   $ N' \gets N$
+\State   $ e'\in E' \gets E $
+\State   $ T'_k \gets T_{n-k} $
+\State   $ T' \gets (\{T'_k\}_{k=1}^n,N',E',L) $ 
 \end{algorithmic}
 \end{algorithm}
+
+With this algorithm we only extend existing edges
+over a larger interval on the genome,
+and remove unnecessary edges in the process.
+%% Find counter example to global minimum.
 
@@
 
+\section{Discrepancy Function}
+% * How to measure agreement that includes haplotypes
+% * Definition and Algorithm
+% * Supp fig: runtime ~ # trees, samples
+% * fig: how add edges rueduces discrepancy
+%       - compare to Relate and tsinfer
+% * Show: how add edges reduces discrepancy
+% * THINK of a way to asynchronously computing discrepancy
+
+\begin{algorithm}[!ht]
+
+\input{$(x,ts_1,ts_2)$}
+\output{discrepancy number (some float value)}
+
+$sim \gets np.zeros(ts2.num_nodes)$\;
+
+\caption{Node Tree Discrepancy}\label{ndisc}
+\end{algorithm}
+
+\section{Data}
+% * 1. Simulate data
+% * 2. Run tsinfer
+% * 3. Compare d(x,truth) for x in tsinfer, tsinfer+add edge, tsinfer+add edge + simplify, tsinfer+simplify
+
+% * How much unary edges?
+% * How much can we add?
+
 \bibliography{references}

### 2022-10-24 paper/writeup.tex
edits

@@
 \section{Notation}
-    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
+    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$ (use Psi), 
     as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where

### 2022-12-07 paper/writeup.tex
algorithm section added

--- 2022-09-23 paper/writeup.tex
@@
 \usepackage[linesnumbered,ruled,vlined]{algorithm2e}
-\usepackage{algpseudocode}
 \usepackage{tikz}
@@
 \newcommand{\R}{\mathbb{R}}
+\newcommand{\tn}{\textnormal}
+\newcommand{\ov}{\overline}
 
@@
 
-\section{Notation}
+\subsection*{Notation}
     We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
@@
      and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
+     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
+     with the lengths of consecutive trees in the sequence 
+     only intersecting each intervals endpoints. 
     Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
@@
     over the entire tree sequence.
+\begin{center}
+\begin{tikzpicture} 
+	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
+	[growth parent anchor=south] 
+		child { node[circle,draw,inner sep=1pt] {b} 		
+	 			child {
+	 				node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
+	 					child{node (d) at (0,1) {$T_k$}
+	 						edge from parent [draw=none]}
+	 				edge from parent 
+	 					node[left] {$e_{b,c}$}
+	 			}	 		
+	 		edge from parent
+	 			node[left] {$e_{a,b}$}
+	 	};
 
-\begin{tikzpicture} %% Do this with phylogenetic trees using tikz (WILL NEED LUALATEX COMPILER) %%
-\node[place] (a)  {a};
-\node[place] (b)  [below of = a] {b};
-\node[place] (c)  [below of = b] {c};
-\draw [-] (a.south) -- (b.north);
-\draw [-] (b.south) -- (c.north);
+	\node[circle,draw,inner sep=1pt] (a') at (3,2) {a}
+	[growth parent anchor=south]
+		child {
+			child{
+				node[circle,draw,inner sep=1pt] (c') at (0,-0.25) {c}	
+					child{node (d) at (0,1) {$T_{k+1}$}
+						edge from parent [draw=none]}
+			}
+			node[left] {$e_{a,c}$}
+		};
+
+	\node at (6,0) {$\Longrightarrow$};
+	\node at (6,0.5) {\tn{Edge Extend}};
+
+	\node[circle,draw,inner sep=1pt] (a''') at (9,2) {a}
+	[growth parent anchor=south] 
+	child { node[circle,draw,inner sep=1pt]	 {b}	
+		child {
+			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
+			child{node (d) at (0,1) {$T_k$}
+				edge from parent [draw=none]}
+			edge from parent 
+			node[left] {$e_{b,c}$}
+		}
+		edge from parent
+		node[left] {$e_{a,b}$}
+	};
+
+	\node[circle,draw,inner sep=1pt] (a'v) at (12,2) {a}
+	[growth parent anchor=south] 
+	child { node[circle,draw,inner sep=1pt] {b}		
+		child {
+			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
+			child{node (d) at (0,1) {$T_{k+1}$}
+				edge from parent [draw=none]}
+			edge from parent 
+			node[left] {$e_{b,c}$}
+		}
+		edge from parent
+		node[left] {$e_{a,b}$}
+	};
+
 \end{tikzpicture}
+\end{center}
+    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
+    be a tree sequence.
+    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
+    This function tracks edges which are removed and added 
+    between consecutive trees in the sequence. 
+    As in \ref{} (tikzfigure), we can track if 
+    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
+    when creating $T_{k+1}$, 
+    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.
 
-    In full detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$
-    be a tree sequence.
+\begin{algorithm}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
 
-\begin{algorithm}  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
-\caption{Extend Edges}\label{alg:cap}
-\begin{algorithmic}
-\SetKwFunction{mirror}{Mirror Coordinates}
-\State   $ N' \gets N$
-\State   $ e'\in E' \gets E $
-\State   $ T'_k \gets T_{n-k} $
-\State   $ T' \gets (\{T'_k\}_{k=1}^n,N',E',L) $ 
-\end{algorithmic}
+\SetStartEndCondition{ }{}{}%
+\SetKwProg{Fn}{def}{\string:}{}
+\SetKwFunction{Range}{range}%%
+\SetKw{KwTo}{in}
+\SetKwFor{For}{for}{\string:}{}%
+\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+\SetKwFor{While}{while}{:}{fintq}%
+\SetKw{Break}{break}
+\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+
+\caption{Extends edges across a tree sequence $TS$.}\label{alg:edge}
+\SetKwFunction{ex}{Extend Edges}
+\SetKwFunction{fwd}{Forward Extend}
+\DontPrintSemicolon
+\LinesNotNumbered
+%\SetAlgoNoEnd
+\input{A tree sequence $TS$}
+\output{New tree sequence,$TS'$, with reduced number of edges}
+
+\BlankLine
+
+\Fn{\fwd (tree sequence $TS$, forwards = \tn{True})}{
+{	
+\eIf{forwards=\tn{True}}{EdgeDiffs $\gets TS$.edgediffs()\BlankLine}
+{\BlankLine
+	EdgeDiffs $\gets$ reversed($TS$.edgediffs())\;}
+}
+\Indm \For{(Interval, EdgesOut, EdgesIn) \KwTo EdgeDiffs}{
+	\For{e1 \KwTo EdgesOut}{
+		\For{e2 \KwTo EdgesOut}{
+			\If{e2.parent = e1.child}{
+				\For{e3 \KwTo EdgesIn}{
+					\If{e3.parent=e1.parent \tn{AND} e3.child=e2.child}{
+						e1.right $\gets$ Interval.right\;
+						e2.right $\gets$ Interval.right\;
+						e3.left $\gets$ Interval.right\;
+						\If{e3.left = e3.right}{
+							$E$.remove(e3)\;
+							}}}}}}}\;
+$TS'\gets TS$.update()\;						
+\KwRet{$TS'$}\;
+}
+\BlankLine						
+\Fn{\ex ($TS$, iter=100)}{
+	edges $\gets$ $TS$.numedges()\;
+	\BlankLine
+	\For{$*$ \KwTo \Range(iter)}{
+		$TS' \gets TS$.\fwd(fowards=\tn{True})\;
+		$TS' \gets TS$.\fwd(fowards=\tn{False})\;
+		\eIf{$TS$.numedges = edges}{
+			\Break\BlankLine}
+		{edges $\gets$ $TS'$.numedges()\;}
+		}
+}
+\KwRet{$TS'$}\;
 \end{algorithm}

### 2023-01-11 paper/writeup.tex
added discrepancy function

@@
 
+\par Now with an algorithm in place
+ we need a metric to discern
+ if trees modified with edge extend
+ in fact correctly infer haplotypes in our original tree sequence.
+ We have thus constructed a function which counts the differences 
+ between trees on a pair of tree sequences. 
+ To count all the discrepancies we must begin by counting differences for each node in the tree sequence.
+
+ Let $TS^1=(\{T_i^1\},N^1,E^1,L^1)$ and $TS^2=(\{T^2_j\},N^2,E^2,L^2)$ be tree sequences.
+ Suppose that $x\in N^1$ and $y\in N^2$ are nodes in each tree sequence.
+ We denote the \textit{set of descendant samples} of $x$ at position $i$ in the genome as $D_{TS^1}(x,i)$.
+ Now for a pair of nodes $(x,y)\in TS^1\times TS^2$ 
+ we can measure how much they agree along the genome.
+ We do so with the following function
+ $$\rho_{(TS^1,TS^2)}(x,y)=\frac{1}{L}\sum_{i=0}^{L-1}[D_{TS^1}(x,i)=D_{TS^2}(y,i)],$$
+ where $L=\min\{L^1,L^2\}$ and $[\cdot]$ is the characteristic function. 
+ We see $\rho$ measures the fraction of the tree sequence
+ along which $x$ and $y$ have exactly the same descendant set. 
+ For each node $x\in N^1$ we can use $\rho$ to find its global best matching node $y\in N^2$. 
+ This node $y$ should not just be a maximum of $\rho(x,\cdot)$ but it should also have a branch length similar to $x$. 
+ This means our node $y$ should minimize both $|t_x-t_y|$ and $1-\rho_{(TS^1,TS^2)}$.
+ We then use this $y$ to find count the length of genome which $x$ differs from its best matching node in $TS_2$ with the function
+ $$\hat{d}(x,TS^1,TS^2)=\min_y\{|t_x-t_y|(1-\rho_{(TS^1,TS^2)}(x,y))\}.$$
+ Here $t_x$ and $t_y$ denote the time of occurrences (ie. branch length) for nodes $x$ and $y$. 
+ The function $\hat{d}$ measures the proportion of the genome 
+ over which the descendant set of $x$ differs from its best matching node $y$.
+ We can now extend this from any node $x\in N^11$ to all of $TS_1$ with the \textbf{discrepancy function}
+ $$d(TS_1,TS_2)=\sum_{x\in N_1}\hat{d}(x,TS_1,TS_2).$$
+ This function, as we will see in the algorithm below,
+ is $|N^1||\{T^1\}|$ as for each tree in $TS_1$ and each node $x\in N^1$ we only need to look at $O(1)$ nodes rather than all nodes in $N^2$.
+
 \begin{algorithm}[!ht]
+\SetStartEndCondition{ }{}{}%
+\SetKwProg{Fn}{def}{\string:}{}
+\SetKwFunction{Range}{range}%%
+\SetKw{KwTo}{in}
+\SetKwFor{For}{for}{\string:}{}%
+\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+\SetKwFor{While}{while}{:}{fintq}%
+\SetKw{Break}{break}
+\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+\SetKwComment{Comment}{}{} 
+\DontPrintSemicolon
+\LinesNotNumbered
+\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
 
-\input{$(x,ts_1,ts_2)$}
+\input{$(TS_1,TS_2)$}
 \output{discrepancy number (some float value)}
 
-$sim \gets np.zeros(ts2.num_nodes)$\;
+\Fn{Node Discrepancy($x,TS_1,TS_2$)}{
+$s=0\in \mathbb{Z}^{|N^2|}$ \Comment*[f]{Similarity vector with each entry corresponds to a node in $TS_2$}
 
-\caption{Node Tree Discrepancy}\label{ndisc}
+\For{$I$, $T^1_k$, $T^2_j$ \KwTo $(L,TS_1,TS_2)$}{
+\Comment{Here $I=[a^1_k,a^1_{k+1})\cap [a^2_j,a^2_{j+1})$ is the interval which the two trees $T^1_k$, $T^2_j$ both span over. Here we find the best matching node $y$.}
+$D_{T^1_k}(x)\gets D_{TS^1}(x,i)$ where $i\in I$\;
+\If{$|D_{T^1_k}(x)| > 0$}{
+\If{$|D_{T^1_k}(x)|=1$}{
+$y \gets D_{T^1_k}(x)$\;
+$m \gets D_{T^1_k}(x)$\;}
+\Else{$m \gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^1_k$\;
+$y\gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^2_j$\;}}
+\If{$D_{T^2_j}(y)=D_{T^1_k}(x)$}{
+$s_y\gets s_y$\;}}
+\KwRet{node discrepancy}\;}
+\Fn{Discrepancy($TS_1,TS_2$)}{
+discrepancy = 0\;
+\For{$x\in N^1$}{discrepancy = $\sum_{x\in N^1}{$Node Discrepancy($x,TS_1,TS_2$)}}
+\BlankLine
+\KwRet{discrepancy}}
+\caption{Discrepancy Function}\label{ndisc}
 \end{algorithm}

### 2023-10-02 paper/writeup.tex
template; start at abstract/intro

@@
-\documentclass{article}
-\usepackage{fullpage}
+\documentclass[10pt,twoside,lineno]{gsajnl}
+\articletype{inv} % article type
+
 \usepackage{amsmath}
@@
 \usepackage{natbib}
-\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
+\usepackage[linesnumbered,ruled,vlined,algo2e]{algorithm2e}
 \usepackage{tikz}
@@
 \tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]
+
+\bibliographystyle{plainnat}
 
@@
 \newcommand{\ov}{\overline}
+\newcommand{\tskit}{\texttt{tskit}}
+
+\title{
+    Information in recombination junctions 
+    can both compress and improve inferred tree sequences
+}
+
+% NOTE: these are in alphabetical order for now; we can revisit that ordering
+\author[$\dagger$]{Ava Bamforth}
+\author[$\dagger$]{Halley Fritze}
+\author[$\ddagger$]{Jerome Kelleher}
+\author[$\dagger$]{Nathaniel Pope}
+\author[$\ast$,$\dagger$,1]{Peter Ralph}
+
+\affil[$\ast$]{Institute of Evolution and Ecology and Department of Biology, University of Oregon, Eugene, Oregon}
+\affil[$\dagger$]{Department of Mathematics, University of Oregon, Eugene, Oregon}
+\affil[$\ddagger$]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
+
+\keywords{genealogy, tree sequence, haplotypes}
+
+\runningtitle{Junctions in tree sequences}
+\runningauthor{Bamforth \textit{et al.}}
+
+%%%%%%%%%%
+\begin{abstract}
+    Foreshadowing haplotype-based methods of the genomics era,
+    it is an old observation that the ``junction'' between two distinct haplotypes
+    produced by recombination is inherited as a Mendelian marker.
+    In this paper, we describe how this recombination-mediated information
+    can in many cases be recovered from inference based solely on
+    polymorphic markers (i.e., information produced by mutation),
+    which produces unary nodes along the branches of the marginal genealogical trees.
+    The resulting inferred ARGs, represented as tree sequences,
+    are smaller, faster to compute with,
+    and potentially contain substantially more information
+    about the age of ancestral haplotypes,
+    than inferred ARGs without these unary nodes.
+    We provide efficient algorithms to identify this ancestral haplotype information,
+    new metrics of agreement/disagreement between inferred ARGs,
+    and explore some consequences for ARGs inferred from real data.
+\end{abstract}
 
 \begin{document}
+
+\maketitle
+\thispagestyle{firststyle}
+\marginmark
+\firstpagefootnote
+
+\correspondingauthoraffiliation{1}{Corresponding author: {plr@uoregon.edu}}
+\vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
 
@@
 
-The pedigree that describes how everyone of some species
-are related to each other
-is a large graph, where nodes are indexed by time.
-Within this, one can trace back the paths along which segement of genome
-have been inherited.
-The \emph{succinct tree sequence} (or, tree sequence for short)
-was introduced by \citet{kelleher2016efficient} to describe the subset of these paths
-along which the genome of a sample of individuals was inherited.
+There has been substantial recent progress
+in the problem of ``ARG inference'',
+which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
+that describes how a set of genotypes samples are related to each other
+at each position of the genome.
+For reviews of this progress, see CITE CITE CITE.
+Often such methods focus on the sequence of ``marginal trees'',
+i.e., the genealogical trees that describe how each portion of the genome
+was inherited by the focal genomes.
+For instance, the ``succinct tree sequence'',
+introduced by \citet{kelleher2016efficient},
+is a common format for describing these inferred ARGs,
+and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
+\tskit (CITE).
+
+However, the ARG is emphatically not merely a sequence of trees:
+since each segment of an ancestor's genome that has been inherited by the focal genomes
+may extend over a portion of the genome over which there are many distinct trees,
+a single ancestor should usually be represented by nodes in many distinct trees.
+In other words, one way to view an ARG is as a sequence of trees in which nodes may be shared across many trees,
+but another way to view the ARG is as a set of relationships between segments of ancestral haplotypes.
+
+Another reason we tend to focus on the trees is that
+much of our intuition about inference of relationships from genomic data
+comes from phylogenetics.
+Indeed, all methods might very roughly be summarized as
+``more similar sequences are more closely related''.
+For instance, two sequences that share a drerived mutation
+are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
+\citet{fisher} pointed out that not only mutations
+but also the ``junctions'' between distinct haplotypes,
+if they could be somehow identified,
+would be inherited as Mendelian markers.
 
@@
 
-\begin{algorithm}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
+\begin{algorithm2e}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
 
@@
 \KwRet{$TS'$}\;
-\end{algorithm}
+\end{algorithm2e}
 
@@
 
-\begin{algorithm}[!ht]
-\SetStartEndCondition{ }{}{}%
-\SetKwProg{Fn}{def}{\string:}{}
-\SetKwFunction{Range}{range}%%
-\SetKw{KwTo}{in}
-\SetKwFor{For}{for}{\string:}{}%
-\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-\SetKwFor{While}{while}{:}{fintq}%
-\SetKw{Break}{break}
-\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-\SetKwComment{Comment}{}{} 
-\DontPrintSemicolon
-\LinesNotNumbered
-\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-
-\input{$(TS_1,TS_2)$}
-\output{discrepancy number (some float value)}
-
-\Fn{Node Discrepancy($x,TS_1,TS_2$)}{
-$s=0\in \mathbb{Z}^{|N^2|}$ \Comment*[f]{Similarity vector with each entry corresponds to a node in $TS_2$}
-
-\For{$I$, $T^1_k$, $T^2_j$ \KwTo $(L,TS_1,TS_2)$}{
-\Comment{Here $I=[a^1_k,a^1_{k+1})\cap [a^2_j,a^2_{j+1})$ is the interval which the two trees $T^1_k$, $T^2_j$ both span over. Here we find the best matching node $y$.}
-$D_{T^1_k}(x)\gets D_{TS^1}(x,i)$ where $i\in I$\;
-\If{$|D_{T^1_k}(x)| > 0$}{
-\If{$|D_{T^1_k}(x)|=1$}{
-$y \gets D_{T^1_k}(x)$\;
-$m \gets D_{T^1_k}(x)$\;}
-\Else{$m \gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^1_k$\;
-$y\gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^2_j$\;}}
-\If{$D_{T^2_j}(y)=D_{T^1_k}(x)$}{
-$s_y\gets s_y$\;}}
-\KwRet{node discrepancy}\;}
-\Fn{Discrepancy($TS_1,TS_2$)}{
-discrepancy = 0\;
-\For{$x\in N^1$}{discrepancy = $\sum_{x\in N^1}{$Node Discrepancy($x,TS_1,TS_2$)}}
-\BlankLine
-\KwRet{discrepancy}}
-\caption{Discrepancy Function}\label{ndisc}
-\end{algorithm}
+% PLR: commented this out since it's not compiling
+% 
+% \begin{algorithm}[!ht]
+% \SetStartEndCondition{ }{}{}%
+% \SetKwProg{Fn}{def}{\string:}{}
+% \SetKwFunction{Range}{range}%%
+% \SetKw{KwTo}{in}
+% \SetKwFor{For}{for}{\string:}{}%
+% \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+% \SetKwFor{While}{while}{:}{fintq}%
+% \SetKw{Break}{break}
+% \newcommand\forcond{$i$ \KwTo\Range{$n$}}
+% \SetKwComment{Comment}{}{} 
+% \DontPrintSemicolon
+% \LinesNotNumbered
+% \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+% 
+% 
+% \input{$(TS_1,TS_2)$}
+% \output{discrepancy number (some float value)}
+% 
+% \Fn{Node Discrepancy($x,TS_1,TS_2$)}{
+% $s=0\in \mathbb{Z}^{|N^2|}$ \Comment*[f]{Similarity vector with each entry corresponds to a node in $TS_2$}
+% 
+% \For{$I$, $T^1_k$, $T^2_j$ \KwTo $(L,TS_1,TS_2)$}{
+% \Comment{Here $I=[a^1_k,a^1_{k+1})\cap [a^2_j,a^2_{j+1})$ is the interval which the two trees $T^1_k$, $T^2_j$ both span over. Here we find the best matching node $y$.}
+% $D_{T^1_k}(x)\gets D_{TS^1}(x,i)$ where $i\in I$\;
+% \If{$|D_{T^1_k}(x)| > 0$}{
+% \If{$|D_{T^1_k}(x)|=1$}{
+% $y \gets D_{T^1_k}(x)$\;
+% $m \gets D_{T^1_k}(x)$\;}
+% \Else{$m \gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^1_k$\;
+% $y\gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^2_j$\;}}
+% \If{$D_{T^2_j}(y)=D_{T^1_k}(x)$}{
+% $s_y\gets s_y$\;}}
+% \KwRet{node discrepancy}\;}
+% \Fn{Discrepancy($TS_1,TS_2$)}{
+% discrepancy = 0\;
+% \For{$x\in N^1$}{discrepancy = $\sum_{x\in N^1}{$Node Discrepancy($x,TS_1,TS_2$)}}
+% \BlankLine
+% \KwRet{discrepancy}}
+% \caption{Discrepancy Function}\label{ndisc}
+% \end{algorithm}
 

### 2023-11-15 paper/writeup.tex
figure!
a few more notes

@@
 \tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]
-
-\bibliographystyle{plainnat}
 
@@
 
-\section{Motivation and statement of problem}
+\subsection{Motivation and statement of problem}
 
@@
 
-\section{Algorithm}
+\begin{figure}
+    \caption{
+        Conceptual figure of what we're talking about.
+        \label{fig:conceptual}
+    }
+\end{figure}
+
+\section{Methods}
+
+\subsection{Algorithm}
 
@@
 %     - arrives at a local minimum?
-%     - guess at typical reduction?
+%     - guess at typical reduction? (reduces edges by 1/3?)
 
@@
     over the entire tree sequence.
+
+\begin{figure}
 \begin{center}
@@
 \end{center}
+\caption{
+    \label{fig:extending_diagram}
+}
+\end{figure}`
+
     In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
@@
 
-\section{Results (how it works)}
-
-% (AVA)
-% * reduction in edges ~ sequence length
-% * speed increase ~ sequence length
-% * proprotion of added edges that are true ~ sequeence length
-% * apply to real data:
-%     - % fewer edges
-%     - % fewer distinct ancestors (ancestral paths?) of sample at some time
-
-\section{Discrepancy Function}
+\subsection{Discrepancy Function}
 % * How to measure agreement that includes haplotypes
@@
  To count all the discrepancies we must begin by counting differences for each node in the tree sequence.
+
+\begin{figure}
+    \caption{
+        Conceptual figure displaying what discrepancy is?
+        TODO: brainstorm.
+        \label{fig:conceptual_discrepancy}
+    }
+\end{figure}
 
@@
 
-\section{Data}
+\section{Results}
+
+% (AVA)
+% * reduction in edges ~ sequence length
+% * speed increase ~ sequence length
+% * proprotion of added edges that are true ~ sequeence length
+
+\begin{figure}
+    \caption{
+        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
+        as a function of number of trees (modulated by sequence length),
+        on reasonably big simulated sequences.
+        For "speed": plot ratio of runtime before/after extending.
+        \label{fig:speed_and_edges}
+    }
+\end{figure}
+
+Simple experiment:
+(1) simulate, simplify, and extend;
+(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
+then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
+
+With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
+(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
+
+\begin{figure}
+    \caption{
+        Maybe: summaries of lengths and/or numbers of correctly and incorrectly added edges?
+        (See notebook.)
+        \label{fig:results_edges}
+    }
+\end{figure}
+
+\begin{figure}
+    \caption{
+        Summaries of correctly and incorrectly added spans
+        ... as a function of number of samples?
+        \label{fig:results_discrepancy}
+    }
+\end{figure}
+
+\subsection{Data}
 % * 1. Simulate data
@@
 
+% * apply to real data:
+%     - % fewer edges
+%     - % fewer distinct ancestors (ancestral paths?) of sample at some time
+
+\section{Discussion}
+
+TODO
+
 \bibliography{references}
 
+\appendix
+
+\begin{figure}
+    \caption{
+        Supplementary figure: runtime of discrepancy function as a function of number of trees
+        (which we modulate by changing sequence length).
+        \label{fig:speed_discrepancy}
+    }
+\end{figure}
+
 \end{document}

### 2023-12-04 paper/writeup.tex
more notes

@@
 }
+
+% Non-coalescing regions of dark matter in ARGs
+% Can we find them? Are they useful in inference, dating? 
+% Look, we can find them and they seem at least useful for compression.
+% Also, here's a way of measuring agreement that takes this sort of thing into account,
+%  which extends R-F to measure haplotypes
 
@@
 \vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
+
+\comment{idea: look at ratio of unary-to-not over time; does it change with
+    selection? or modes of recombination?}
+
+%%% OUTLINE
+% Intro: haplotypes, unary bits of non-coalescing nodes, ARGS, etcetera
+%    explain what tsinfer does to create unary regions
+%    statement of problem
+%    Yan: are breakpoints in true->simplify->extend the same as in original?
+%   Fig 1: conceptual figure
+%    Also IBD gets screwed up
+% 
+% Methods:
+%   edge extend algorithm
+%   discrepancy funciton algorithm
+%   Fig 2: Conceptual figure for discrepancy
+% 
+% Results:
+%   Fig 3:
+%     (a) reduction in number of edges and (b) speed change
+%
+%   Fig 4:
+%     Histograms of (a) total span added, (b) percent incorrect
+%
+%   Fig 5 and maybe 6: how it interacts with tsinfer
+%     (a) summarize total matching and unmatching span (or maybe percent matching?)
+%           across reps
+%     (b) discrepancy per node against depth or time or number of subtended samples
+%        describe percent span matched against true span or depth or something
+%
+%   Fig 7(?): compare IBD stats before/after
+% 
+% Supp:
+%   S1: runtime
+%%%
 
@@
 % * Supp fig: runtime ~ # trees, samples
-% * fig: how add edges rueduces discrepancy
-%       - compare to Relate and tsinfer
-% * Show: how add edges reduces discrepancy
-% * THINK of a way to asynchronously computing discrepancy
+% * fig: how add edges reduces discrepancy (RESULTS)
+%       - compare to tsinfer
+% * Show: how add edges reduces discrepancy (RESULTS)
+% * THINK of a way to asynchronously computing discrepancy (???)
 
@@
 
+\comment{
+    Idea: look at statistical properties of unary regions:
+    look at how much 'unary regions' there is in simulation
+    and how much we can put back in,
+    as a function of time.
+}
+
 % (AVA)
@@
         \label{fig:results_edges}
-    }
-\end{figure}
-
-\begin{figure}
-    \caption{
-        Summaries of correctly and incorrectly added spans
-        ... as a function of number of samples?
-        \label{fig:results_discrepancy}
     }

### 2023-12-11 paper/writeup.tex
conceptual figure

@@
 
+\usepackage{xcolor}
 \usepackage{amsmath}
@@
 \newcommand{\tskit}{\texttt{tskit}}
+\newcommand{\comment}[1]{{\color{blue} \it #1}}
 
@@
 
+Consider the (small portion) of a hypothetical tree sequence in Figure~\ref{fig:conceptual}A.
+On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
+coalesce into a small subtree: 1 and 2 find a common ancestor in ancestral node 3,
+which finds a common ancestor with node 0 in ancestral node 4.
+On the next portion of the genome (right-hand tree), sample node 2 has a different ancestor.
+This seems reasonable, and a method that infers trees separately on each portion of the genome
+could not be expected to produce anything different.
+However, things change once we think about what this implies about haplotype inheritance.
+Figure~\ref{fig:conceptual}B shows the implied inheritance of haplotypes,
+with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
+Here, sample node 2 has inherited the chunk of haplotype labeled $L$ from ancestral node 4 via 3,
+and the haplotype to the right of this from some other node (and so doesn't carry haplotype $R$).
+On the other hand, sample node 1 has inherited \emph{both} haplotypes $L$ and $R$
+from ancestral node 4, but the trees imply that only haplotype $L$ is inherited via ancestral node 3.
+This implies -- if taken literally -- that there must have been a recombination event
+at some point between node 1 and node 4 that separated the $L$ and $R$ haplotypes,
+and then these two ancestral (and nonoverlapping) haplotypes coalesced together in ancestral node 4.
+Although this is possible, it seems unlikely --
+a more parsimonious explanation is depicted in Figure~\ref{fig:conceptual}C,
+in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
+(and there is a recombination somewhere between node 3 and node 2).
+This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
+which is depicted in Figure~\ref{fig:conceptual}D --
+and so node 3 has become unary in this tree.
+Note that the more parsimonious tree sequence also includes fewer edges:
+the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in Figure~\ref{fig:conceptual}B
+have been reduced to the two edges
+$4 \to 3$ and $3 \to 1$ in Figure~\ref{fig:conceptual}D.
+
 \begin{figure}
+    \begin{center}
+    \includegraphics[width=0.9\textwidth]{conceptual_figure}
+    \end{center}
     \caption{
-        Conceptual figure of what we're talking about.
+        A simple example showing the basic idea
+        (described in more detail in the text):
+        \textbf{(A)} a small portion of a tree sequence without unary nodes;
+        \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
+        labeled $L$ and $R$;
+        \textbf{(C)} marginal trees with a unary node added,
+        which produces \textbf{(D)} a more parsimonious haplotype inheritance pattern
+        (that also includes fewer edges).
         \label{fig:conceptual}

### 2024-01-04 paper/writeup.tex
Some paper edits

@@
 
-\subsection{Algorithm}
+\subsection{Extend Edges}}
 
@@
 %     - guess at typical reduction? (reduces edges by 1/3?)
-
-\subsection*{Notation}
-    We begin with some notation. We define a \textit{tree sequence}, denoted $TS$, 
+    We define a \textit{tree sequence}, denoted $TS$, 
     as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
@@
      only intersecting each intervals endpoints.
-    Additionally, a node $x\in N$ is \textit{unary} if it only has one child, 
-    ie. only a single edge extending below it on the tree.
-    With these tree sequences our goal is to
-    minimize the number of edges
-    as well as infer coalescence of genes. 
-    To remove edges we consider the following example.
-    Suppose for a tree $T_k$ in the tree sequence $TS$ 
-    there exists a branch which contains a unary node $b\in N$.
-    Then the node $b$ has one parent, $a\in N$, 
-    and one child $c\in N$,
-    each with edges $e_{ab}, e_{bc}\in E$ connecting them.
+    Given any tree sequence, our goal is to
+    identify areas of implied inheritance of haplotypes.
+    We do so by identifying short paths (2 edges) in a tree.
+    If neighboring trees contain edges with 
+    identical parent and child nodes, we assume that there
+    is implied coalescence of haplotypes between nodes, and
+  	we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
+  	In more concrete terms, for a tree $T_k$ in the tree sequence $TS$ 
+    suppose there exists a path which contains a node $b\in N$,
+    with parent and child nodes $a, c\in N$, respectively.
     If there is an edge between the nodes $a$ and $c$ in trees 
@@
     to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
-    This action reduces the length of the edge $e_{ac}$ on the genome,
+    This action reduces the length of $e_{ac}$ on the genome,
     and in some cases, completely removes the edge. 
-    We now perform this action on all such unary nodes 
+    We now perform this action on all such ancestral paths 
     over the entire tree sequence.
 
-\begin{figure}
+\begin{figure}[!ht]
 \begin{center}
@@
 	 			child {
-	 				node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
-	 					child{node (d) at (0,1) {$T_k$}
+	 				node[circle,draw,inner sep=1pt] (c) at (-0.5,0) {c}
+	 				edge from parent 
+	 					node[left] {$e_{b,c}$}}
+ 				child {
+ 					node (d) at (0,-1) {$T_k$}
  					edge from parent [draw=none]}
+	 			child {
+	 				node[circle,draw,inner sep=1pt] (e) at (0.5,0)  {d}
  					edge from parent
-	 					node[left] {$e_{b,c}$}
-	 			}	 		
+ 						node[right] {$e_{b,d}$}} 		
 	 		edge from parent
@@
 	 	}; 	
-
-	\node[circle,draw,inner sep=1pt] (a') at (3,2) {a}
+	\node[circle,draw,inner sep=1pt] (a') at (4,2) {a}
 	[growth parent anchor=south]
@@
 
-	\node at (6,0) {$\Longrightarrow$};
-	\node at (6,0.5) {\tn{Edge Extend}};
+	\node at (7,0) {$\Longrightarrow$};
+	\node at (7,0.5) {\tn{Edge Extend}};
 
@@
 \end{center}
-\caption{
+\caption{A simple visualization of the Edge Extend method.
     \label{fig:extending_diagram}
 }
-\end{figure}`
+\end{figure}
 
@@
     between consecutive trees in the sequence. 
-    As in \ref{} (tikzfigure), we can track if 
+    As in \ref{fig:extending_diagram} (tikzfigure), we can track if 
     $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 

### 2024-01-04 paper/writeup.tex
added figures, extend edges section

@@
 \newcommand{\R}{\mathbb{R}}
+\newcommand{\T}{\mathbb{T}}
 \newcommand{\tn}{\textnormal}
@@
     \begin{center}
-    \includegraphics[width=0.9\textwidth]{conceptual_figure}
+    \includegraphics[width=0.9\textwidth]{conceptual_figure.pdf}
     \end{center}
@@
 %     - guess at typical reduction? (reduces edges by 1/3?)
-    We define a \textit{tree sequence}, denoted $TS$, 
-    as a tuple $\left( \left\{T_k\right\}, N, E, L\right)$ where
-     $\left\{T_k\right\}_{k=1}^n$ is the set of trees,
-     $N$ and $E$ are the sets of nodes and edges respectively,
-     and $L=[0,a)\subseteq \R$ is the length of the sequence on the genome.
-     The length of each tree $T_k$ is some subinterval $[a_k,a_{k+1})$ 
-     with the lengths of consecutive trees in the sequence 
-     only intersecting each intervals endpoints.
+    We define a \textit{tree sequence}, denoted $\T$, 
+    as a tuple $\left(T_1,...,T_{|\T|}}\right)$ with a sequence of points
+	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
+     each $T_k$ is a tree describing genealogical relationships between samples
+     over a distinct portion of the genome $[a_{k-1}, a_k)$.
     Given any tree sequence, our goal is to
@@
     If neighboring trees contain edges with 
-    identical parent and child nodes, we assume that there
-    is implied coalescence of haplotypes between nodes, and
-  	we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
-  	In more concrete terms, for a tree $T_k$ in the tree sequence $TS$ 
-    suppose there exists a path which contains a node $b\in N$,
-    with parent and child nodes $a, c\in N$, respectively.
-    If there is an edge between the nodes $a$ and $c$ in trees 
+    identical ancestor (parent) and descendant (child) nodes, 
+    we assume that there is implied coalescence of haplotypes between nodes, 
+    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
+  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
+    suppose there exists a path which contains a node $3$,
+    with parent and child nodes $4$ and $0$, respectively.
+    If there is an edge between the nodes $4$ and $0$ in trees 
     $T_{k+1}$ (or $T_{k-1}$),
-    then we wish to extend the edges $e_{ab},e_{bc}$ 
-    to $T_{k+1}$ and then remove the edge $e_{ac}$ from $T_{k+1}$. 
-    This action reduces the length of $e_{ac}$ on the genome,
+    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
+    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
+    This action reduces the length of $4 \to 0$ on the genome,
     and in some cases, completely removes the edge. 
@@
     over the entire tree sequence.
+    With this algorithm \ref{alg:edge} we only extend existing edges
+    over a larger interval on the genome,
+    and remove unnecessary edges in the process.
 
@@
 \begin{center}
-\begin{tikzpicture} 
-	\node[circle,draw,inner sep=1pt] (a) at (0,2) {a}
-	[growth parent anchor=south] 
-		child { node[circle,draw,inner sep=1pt] {b} 		
-	 			child {
-	 				node[circle,draw,inner sep=1pt] (c) at (-0.5,0) {c}
-	 				edge from parent 
-	 					node[left] {$e_{b,c}$}}
- 				child {
- 					node (d) at (0,-1) {$T_k$}
- 					edge from parent [draw=none]}
-	 			child {
-	 				node[circle,draw,inner sep=1pt] (e) at (0.5,0)  {d}
- 					edge from parent
- 						node[right] {$e_{b,d}$}} 		
-	 		edge from parent
-	 			node[left] {$e_{a,b}$}
-	 	}; 	
-	\node[circle,draw,inner sep=1pt] (a') at (4,2) {a}
-	[growth parent anchor=south]
-		child {
-			child{
-				node[circle,draw,inner sep=1pt] (c') at (0,-0.25) {c}	
-					child{node (d) at (0,1) {$T_{k+1}$}
-						edge from parent [draw=none]}
-			}
-			node[left] {$e_{a,c}$}
-		};
-
-	\node at (7,0) {$\Longrightarrow$};
-	\node at (7,0.5) {\tn{Edge Extend}};
-
-	\node[circle,draw,inner sep=1pt] (a''') at (9,2) {a}
-	[growth parent anchor=south] 
-	child { node[circle,draw,inner sep=1pt]	 {b}	
-		child {
-			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
-			child{node (d) at (0,1) {$T_k$}
-				edge from parent [draw=none]}
-			edge from parent 
-			node[left] {$e_{b,c}$}
-		}
-		edge from parent
-		node[left] {$e_{a,b}$}
-	};
-
-	\node[circle,draw,inner sep=1pt] (a'v) at (12,2) {a}
-	[growth parent anchor=south] 
-	child { node[circle,draw,inner sep=1pt] {b}		
-		child {
-			node[circle,draw,inner sep=1pt] (c) at (0,0) {c}
-			child{node (d) at (0,1) {$T_{k+1}$}
-				edge from parent [draw=none]}
-			edge from parent 
-			node[left] {$e_{b,c}$}
-		}
-		edge from parent
-		node[left] {$e_{a,b}$}
-	};
-
-\end{tikzpicture}
+	\includegraphics[width=5in]{edge_extend_method.pdf}
 \end{center}
-\caption{A simple visualization of the Edge Extend method.
+\caption{A visualization of the \textit{extend edges} method.
+Given two neighboring trees $T_1$ and $T_2$, the tree $T_1$ has a short path of edges $4 \to 3\to0$.
+The next tree contains a single edge $4 \to 0$.
+If node $3$ is not in $T_2$, then edge extend will extend edges $ 4\to 3$ and $3\to 0$ to tree $T_2$, hence removing the edge $4\to 0$.
     \label{fig:extending_diagram}
@@
 \end{figure}
-
-    In more detail, let $TS = \left( \left\{T_k\right\}_{k=1}^n, N, E, L\right)$ 
-    be a tree sequence.
-    Within tskit we can track a tree sequence using a function called `EdgeDiffs'.
-    This function tracks edges which are removed and added 
-    between consecutive trees in the sequence. 
-    As in \ref{fig:extending_diagram} (tikzfigure), we can track if 
-    $e_{a,b}$ and$e_{b,c}$ were removed from $T_k$ 
-    when creating $T_{k+1}$, 
-    and we can see if $e_{a,c}$ was just added in $T_{k+1}$.
 
@@
 \SetKwFunction{ex}{Extend Edges}
-\SetKwFunction{fwd}{Forward Extend}
+\SetKwFunction{fwd}{Extend}
 \DontPrintSemicolon
@@
 %\SetAlgoNoEnd
-\input{A tree sequence $TS$}
-\output{New tree sequence,$TS'$, with reduced number of edges}
+\input{A tree sequence $\T$}
+\output{New tree sequence,$\T'$}
 
@@
 
-\Fn{\fwd (tree sequence $TS$, forwards = \tn{True})}{
+\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
 {	
-\eIf{forwards=\tn{True}}{EdgeDiffs $\gets TS$.edgediffs()\BlankLine}
+\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
 {\BlankLine
-	EdgeDiffs $\gets$ reversed($TS$.edgediffs())\;}
+	$\T = \left(T_{|\T|},..., T_1\right)$\;}
 }
-\Indm \For{(Interval, EdgesOut, EdgesIn) \KwTo EdgeDiffs}{
-	\For{e1 \KwTo EdgesOut}{
-		\For{e2 \KwTo EdgesOut}{
-			\If{e2.parent = e1.child}{
-				\For{e3 \KwTo EdgesIn}{
-					\If{e3.parent=e1.parent \tn{AND} e3.child=e2.child}{
-						e1.right $\gets$ Interval.right\;
-						e2.right $\gets$ Interval.right\;
-						e3.left $\gets$ Interval.right\;
-						\If{e3.left = e3.right}{
-							$E$.remove(e3)\;
-							}}}}}}}\;
-$TS'\gets TS$.update()\;						
-\KwRet{$TS'$}\;
+\Indm \For{$T_k\in \T$}{
+	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
+	\For{edges $f$ in $T_{k+1}$}{
+		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
+		Remove $f$ from $T_{k+1}$.\;
+		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
+		Update $\T$.\;
+	}
+	}
+}\;						
+\KwRet{$\T$}\;
 }
 \BlankLine						
-\Fn{\ex ($TS$, iter=100)}{
-	edges $\gets$ $TS$.numedges()\;
-	\BlankLine
+\Fn{\ex ($\T$, iter=100)}{
 	\For{$*$ \KwTo \Range(iter)}{
-		$TS' \gets TS$.\fwd(fowards=\tn{True})\;
-		$TS' \gets TS$.\fwd(fowards=\tn{False})\;
-		\eIf{$TS$.numedges = edges}{
+		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
+		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
+		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
 			\Break\BlankLine}
-		{edges $\gets$ $TS'$.numedges()\;}
 		}
 }
-\KwRet{$TS'$}\;
+\KwRet{$\T'$}\;
 \end{algorithm2e}
 
-With this algorithm we only extend existing edges
-over a larger interval on the genome,
-and remove unnecessary edges in the process.
 %% Find counter example to global minimum.
@@
  in fact correctly infer haplotypes in our original tree sequence.
- We have thus constructed a function which counts the differences 
- between trees on a pair of tree sequences. 
+ One such function a so called \textit{discrepancy function}, can be called on a pair of tree 
  To count all the discrepancies we must begin by counting differences for each node in the tree sequence.
@@
 \begin{figure}
+
     \caption{

### 2024-01-04 paper/writeup.tex
added discrepancy concept figure

@@
 
-\subsection{Extend Edges}}
+\subsection{Extend Edges}
 
@@
     We define a \textit{tree sequence}, denoted $\T$, 
-    as a tuple $\left(T_1,...,T_{|\T|}}\right)$ with a sequence of points
+    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
 	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
@@
  in fact correctly infer haplotypes in our original tree sequence.
- One such function a so called \textit{discrepancy function}, can be called on a pair of tree 
+ One such function a so called \textit{discrepancy function}, can be called on a pair of tree sequences.
  To count all the discrepancies we must begin by counting differences for each node in the tree sequence.
 
-\begin{figure}
-
+\begin{figure}[!ht]
+	\begin{center}
+\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
+\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
     \caption{
@@
     }
+	\end{center}
 \end{figure}

### 2024-01-22 paper/writeup.tex

@@
 \newcommand{\comment}[1]{{\color{blue} \it #1}}
+\newcommand{\argmax}{\operatorname{argmax}}
 
@@
     We do so by identifying short paths (2 edges) in a tree.
+    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
     If neighboring trees contain edges with 
@@
     over the entire tree sequence.
-    With this algorithm \ref{alg:edge} we only extend existing edges
+    With algorithm \ref{alg:edge} we only extend existing edges
+    \comment{Note possible change pending algorithm change}
     over a larger interval on the genome,
@@
 The next tree contains a single edge $4 \to 0$.
-If node $3$ is not in $T_2$, then edge extend will extend edges $ 4\to 3$ and $3\to 0$ to tree $T_2$, hence removing the edge $4\to 0$.
+If node $3$ is not in $T_2$, then we will extend edges $ 4\to 3$ and $3\to 0$ to tree $T_2$, hence removing the edge $4\to 0$.
     \label{fig:extending_diagram}
@@
  we need a metric to discern
- if trees modified with edge extend
+ if trees modified with extend edges
  in fact correctly infer haplotypes in our original tree sequence.
@@
  To count all the discrepancies we must begin by counting differences for each node in the tree sequence.
+
+\comment{Note this is somewhat similar to the Robinson-Foulds metric in concept.}
 
@@
     \caption{
-        Conceptual figure displaying what discrepancy is?
-        TODO: brainstorm.
+        Something here \comment{something useful}
         \label{fig:conceptual_discrepancy}
@@
 % 
-% \begin{algorithm}[!ht]
-% \SetStartEndCondition{ }{}{}%
-% \SetKwProg{Fn}{def}{\string:}{}
-% \SetKwFunction{Range}{range}%%
-% \SetKw{KwTo}{in}
-% \SetKwFor{For}{for}{\string:}{}%
-% \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-% \SetKwFor{While}{while}{:}{fintq}%
-% \SetKw{Break}{break}
-% \newcommand\forcond{$i$ \KwTo\Range{$n$}}
-% \SetKwComment{Comment}{}{} 
-% \DontPrintSemicolon
-% \LinesNotNumbered
-% \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-% 
-% 
-% \input{$(TS_1,TS_2)$}
-% \output{discrepancy number (some float value)}
-% 
-% \Fn{Node Discrepancy($x,TS_1,TS_2$)}{
-% $s=0\in \mathbb{Z}^{|N^2|}$ \Comment*[f]{Similarity vector with each entry corresponds to a node in $TS_2$}
-% 
-% \For{$I$, $T^1_k$, $T^2_j$ \KwTo $(L,TS_1,TS_2)$}{
-% \Comment{Here $I=[a^1_k,a^1_{k+1})\cap [a^2_j,a^2_{j+1})$ is the interval which the two trees $T^1_k$, $T^2_j$ both span over. Here we find the best matching node $y$.}
-% $D_{T^1_k}(x)\gets D_{TS^1}(x,i)$ where $i\in I$\;
-% \If{$|D_{T^1_k}(x)| > 0$}{
-% \If{$|D_{T^1_k}(x)|=1$}{
-% $y \gets D_{T^1_k}(x)$\;
-% $m \gets D_{T^1_k}(x)$\;}
-% \Else{$m \gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^1_k$\;
-% $y\gets $ most recent common anscestor of $D_{T^1_k}(x)$ in $T^2_j$\;}}
-% \If{$D_{T^2_j}(y)=D_{T^1_k}(x)$}{
-% $s_y\gets s_y$\;}}
-% \KwRet{node discrepancy}\;}
-% \Fn{Discrepancy($TS_1,TS_2$)}{
-% discrepancy = 0\;
-% \For{$x\in N^1$}{discrepancy = $\sum_{x\in N^1}{$Node Discrepancy($x,TS_1,TS_2$)}}
-% \BlankLine
-% \KwRet{discrepancy}}
-% \caption{Discrepancy Function}\label{ndisc}
-% \end{algorithm}
+ \begin{algorithm2e}[!ht]
+ \SetStartEndCondition{ }{}{}%
+ \SetKwProg{Fn}{def}{\string:}{}
+ \SetKwFunction{Range}{range}%%
+ \SetKw{KwTo}{in}
+ \SetKwFor{For}{for}{\string:}{}%
+ \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+ \SetKwFor{While}{while}{:}{fintq}%
+ \SetKw{Break}{break}
+ \newcommand\forcond{$i$ \KwTo\Range{$n$}}
+ \SetKwComment{Comment}{}
+ \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+ \DontPrintSemicolon\LinesNotNumbered
+
+ \caption{Discrepancy Function}\label{alg:disc}
+ \input{$(T_1,T_2)$}
+ \output{tree discrepancy, root-mean-square error}
+
+ \Fn{Shared Spans($T_1,T_2$)}{
+ \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
+ 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
+ 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
+ 		Define $a_{ij}=|L_i\cap L_j|$.
+ 	}
+ 	\Else{Define $a_{ij}=0$.}
+ Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
+ }
+ \KwRet{$A$}
+ }
+\BlankLine
+\Fn{Discrepancy ($T_1,T_2$)}{
+\For{each node $i\in N_{T_1}$}{
+	Find nodes in $N_{T_2}$ which have the highest matching span.\;
+	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
+	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
+	Define $\alpha_i = a_{ij}$.\;
+}
+\begin{left} $$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
+\BlankLine\BlankLine
+\begin{left} $$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
+\BlankLine\BlankLine
+\KwRet{tree discrepancy, rmse}
+}
+\end{algorithm2e}
 

### 2024-01-22 paper/writeup.tex
discrepancy function write ups
Still under construction, but some progess

@@
 
-\par Now with an algorithm in place
+\par \comment{This needs work. Choose $n_1$ or $i$ for node not both. Do you want it through the perspective of TS or through nodes? You wording is mixed on this}
+ Now with an algorithm in place
  we need a metric to discern
@@
  in fact correctly infer haplotypes in our original tree sequence.
- One such function a so called \textit{discrepancy function}, can be called on a pair of tree sequences.
- To count all the discrepancies we must begin by counting differences for each node in the tree sequence.
+ One such function a so called \textit{discrepancy function} (see Algorithm \ref{alg:disc}), can be called on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
+ Let $N_1$ and $N_2$ be the set of nodes on $\T_1$ respectively $\T_2$.
+ For a pair of nodes $(n_1,n_2)\in N_1\times N_2$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
+ Each node $n\in\T$ has a given span along the genome and a time, 
+ so we can write each node $n = (t_n, L_n)$ where $L_n\subset \left[0,L\right)$ and $t_n\in \R^{\geq 0}$.
+ Clearly, computing the discrepancy between a random pair of nodes is not helpful, 
+ so we must first identify which pairs of nodes would be most identical to one another or a 'best match'.
+ When we find 'best matching' nodes we want our pair to be close in terms of time and have equal descendant sample sets,
+ ie. $s(n_1) = s(n_2)$ where $s$ is the function to compute the descendant sample set of $n_1$.
 
-\comment{Note this is somewhat similar to the Robinson-Foulds metric in concept.}
+ To find a best match for each node $n_1\in N_1$ \comment{(Should I do this by mentioning clades?)}
+ we compute the \textit{shared span} between $(n_1, m)$ for all $m\in N_2$.
+ The \textit{shared span} between two tree sequences, $ss(\T_1,\T_2)$, is an $|N_1|\times |N_2|$ matrix $A$
+ where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_1\times N_2$.
+ If a pair has equal descendant sample sets, $s(i) = s(j)$
+ then $a_{ij} = |L_i \cap L_j|$.
+ If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
+ Matrix $A$ then determines the \textit{best match} for every node $i\in N_1$
+ via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_2}a_{ij}$.
+ If this maximum is not unique, we use time difference to determine the best match pair.
+ We then define the \textit{discrepancy} between $\T_1$ and $\T_2$ as 
+ one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
+ This function is similar in idea to the Robinson-Foulds metric \comment{CITE}.
+ % Robinson-Foulds Paper: https://www.sciencedirect.com/science/article/abs/pii/0025556481900432?via%3Dihub 
+ However, where the Robinson-Foulds metric compares nodes between individual pairs of trees,
+ in an effort to infer coalescent haplotypes, our function chooses to compare nodes
+ across the entire genome. 
+ We think this helps track ancestry \comment{something something something...}
+
+\comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}
 
@@
     \caption{
-        Something here \comment{something useful}
+        For two tree sequences $T1$ and $T2$ the \textit{discrepancy function}, $d(T1,T2)$, matches nodes in $T1$ with nodes in $T2$
+        based on identical sample sets. We then measure each pair of nodes' \textit{shared span} across the entire tree sequence.
+        In this example we have nodes $0$ through $7$ in $T1$ match with their identical counterpart. 
+        Node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left{1,2,3\right}$.
+        \comment{make some reference to Discrepancy function section or appendix for more in depth calculation.}
         \label{fig:conceptual_discrepancy}
@@
 
- Let $TS^1=(\{T_i^1\},N^1,E^1,L^1)$ and $TS^2=(\{T^2_j\},N^2,E^2,L^2)$ be tree sequences.
- Suppose that $x\in N^1$ and $y\in N^2$ are nodes in each tree sequence.
- We denote the \textit{set of descendant samples} of $x$ at position $i$ in the genome as $D_{TS^1}(x,i)$.
- Now for a pair of nodes $(x,y)\in TS^1\times TS^2$ 
- we can measure how much they agree along the genome.
- We do so with the following function
- $$\rho_{(TS^1,TS^2)}(x,y)=\frac{1}{L}\sum_{i=0}^{L-1}[D_{TS^1}(x,i)=D_{TS^2}(y,i)],$$
- where $L=\min\{L^1,L^2\}$ and $[\cdot]$ is the characteristic function. 
- We see $\rho$ measures the fraction of the tree sequence
- along which $x$ and $y$ have exactly the same descendant set. 
- For each node $x\in N^1$ we can use $\rho$ to find its global best matching node $y\in N^2$. 
- This node $y$ should not just be a maximum of $\rho(x,\cdot)$ but it should also have a branch length similar to $x$. 
- This means our node $y$ should minimize both $|t_x-t_y|$ and $1-\rho_{(TS^1,TS^2)}$.
- We then use this $y$ to find count the length of genome which $x$ differs from its best matching node in $TS_2$ with the function
- $$\hat{d}(x,TS^1,TS^2)=\min_y\{|t_x-t_y|(1-\rho_{(TS^1,TS^2)}(x,y))\}.$$
- Here $t_x$ and $t_y$ denote the time of occurrences (ie. branch length) for nodes $x$ and $y$. 
- The function $\hat{d}$ measures the proportion of the genome 
- over which the descendant set of $x$ differs from its best matching node $y$.
- We can now extend this from any node $x\in N^11$ to all of $TS_1$ with the \textbf{discrepancy function}
- $$d(TS_1,TS_2)=\sum_{x\in N_1}\hat{d}(x,TS_1,TS_2).$$
- This function, as we will see in the algorithm below,
- is $|N^1||\{T^1\}|$ as for each tree in $TS_1$ and each node $x\in N^1$ we only need to look at $O(1)$ nodes rather than all nodes in $N^2$.
+% Let $TS^1=(\{T_i^1\},N^1,E^1,L^1)$ and $TS^2=(\{T^2_j\},N^2,E^2,L^2)$ be tree sequences.
+% Suppose that $x\in N^1$ and $y\in N^2$ are nodes in each tree sequence.
+% We denote the \textit{set of descendant samples} of $x$ at position $i$ in the genome as $D_{TS^1}(x,i)$.
+% Now for a pair of nodes $(x,y)\in TS^1\times TS^2$ 
+% we can measure how much they agree along the genome.
+% We do so with the following function
+% $$\rho_{(TS^1,TS^2)}(x,y)=\frac{1}{L}\sum_{i=0}^{L-1}[D_{TS^1}(x,i)=D_{TS^2}(y,i)],$$
+% where $L=\min\{L^1,L^2\}$ and $[\cdot]$ is the characteristic function. 
+% We see $\rho$ measures the fraction of the tree sequence
+% along which $x$ and $y$ have exactly the same descendant set. 
+% For each node $x\in N^1$ we can use $\rho$ to find its global best matching node $y\in N^2$. 
+% This node $y$ should not just be a maximum of $\rho(x,\cdot)$ but it should also have a branch length similar to $x$. 
+% This means our node $y$ should minimize both $|t_x-t_y|$ and $1-\rho_{(TS^1,TS^2)}$.
+% We then use this $y$ to find count the length of genome which $x$ differs from its best matching node in $TS_2$ with the function
+% $$\hat{d}(x,TS^1,TS^2)=\min_y\{|t_x-t_y|(1-\rho_{(TS^1,TS^2)}(x,y))\}.$$
+% Here $t_x$ and $t_y$ denote the time of occurrences (ie. branch length) for nodes $x$ and $y$. 
+% The function $\hat{d}$ measures the proportion of the genome 
+% over which the descendant set of $x$ differs from its best matching node $y$.
+% We can now extend this from any node $x\in N^11$ to all of $TS_1$ with the \textbf{discrepancy function}
+% $$d(TS_1,TS_2)=\sum_{x\in N_1}\hat{d}(x,TS_1,TS_2).$$
+% This function, as we will see in the algorithm below,
+% is $|N^1||\{T^1\}|$ as for each tree in $TS_1$ and each node $x\in N^1$ we only need to look at $O(1)$ nodes rather than all nodes in $N^2$.
+%
 
@@
 	Define $\alpha_i = a_{ij}$.\;
+	\BlankLine
 }
-\begin{left} $$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
+\begin{left}
+$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
 \BlankLine\BlankLine
-\begin{left} $$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
+\begin{left}
+$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
 \BlankLine\BlankLine
@@
 \begin{figure}
+	\includegraphics[width=0.9\textwidth]{Edge_reduc_comp_speed.pdf}
     \caption{

### 2024-08-20 paper/writeup.tex
new title

@@
 \title{
-    Information in recombination junctions 
-    can both compress and improve inferred tree sequences
+    A forest is more than its trees:
+    haplotypes and inferred ARGs
 }
 
+% Information in recombination junctions can both compress and improve inferred tree sequences
 % Non-coalescing regions of dark matter in ARGs
@@
         In this example we have nodes $0$ through $7$ in $T1$ match with their identical counterpart. 
-        Node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left{1,2,3\right}$.
+        Node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
         \comment{make some reference to Discrepancy function section or appendix for more in depth calculation.}

### 2024-08-21 COPYtest_extend_edges.py
Reorganized, newplots, saved data

@@ -1,1356 +1,0 @@
-import msprime
-import numpy as np
-import pytest
-
-import _tskit
-# import tests.test_wright_fisher as wf
-import tskit
-import COPYtsutil
-# from tests.test_highlevel import get_example_tree_sequences
-
-# ↑ See https://github.com/tskit-dev/tskit/issues/1804 for when
-# we can remove this.
-
-def extend_edges(ts, max_iter=10):
-    tables = ts.dump_tables()
-    mutations = tables.mutations.copy()
-    tables.mutations.clear()
-
-    last_num_edges = ts.num_edges
-    for _ in range(max_iter):
-        for forwards in [True, False]:
-            edges = _extend(ts, forwards=forwards)
-            tables.edges.replace_with(edges)
-            tables.build_index()
-            ts = tables.tree_sequence()
-        if ts.num_edges == last_num_edges:
-            break
-        else:
-            last_num_edges = ts.num_edges
-
-    tables = ts.dump_tables()
-    mutations = _slide_mutation_nodes_up(ts, mutations)
-    tables.mutations.replace_with(mutations)
-    ts = tables.tree_sequence()
-
-    return ts
-
-
-def _slide_mutation_nodes_up(ts, mutations):
-    # adjusts mutations' nodes to place each mutation on the correct edge given
-    # their time; requires mutation times be nonmissing and the mutation times
-    # be >= their nodes' times.
-
-    assert np.all(~tskit.is_unknown_time(mutations.time)), "times must be known"
-    new_nodes = mutations.node.copy()
-
-    mut = 0
-    for tree in ts.trees():
-        _, right = tree.interval
-        while (
-            mut < mutations.num_rows and ts.sites_position[mutations.site[mut]] < right
-        ):
-            t = mutations.time[mut]
-            c = mutations.node[mut]
-            p = tree.parent(c)
-            assert ts.nodes_time[c] <= t
-            while p != -1 and ts.nodes_time[p] <= t:
-                c = p
-                p = tree.parent(c)
-            assert ts.nodes_time[c] <= t
-            if p != -1:
-                assert t < ts.nodes_time[p]
-            new_nodes[mut] = c
-            mut += 1
-
-    # in C the node column can be edited in place
-    new_mutations = mutations.copy()
-    new_mutations.clear()
-    for mut, n in zip(mutations, new_nodes):
-        new_mutations.append(mut.replace(node=n))
-
-    return new_mutations
-
-
-def _extend(ts, forwards=True):
[…]

### 2024-08-21 COPYtsutil.py

@@ -1,2290 +1,0 @@
-# MIT License
-#
-# Copyright (c) 2018-2023 Tskit Developers
-# Copyright (C) 2017 University of Oxford
-#
-# Permission is hereby granted, free of charge, to any person obtaining a copy
-# of this software and associated documentation files (the "Software"), to deal
-# in the Software without restriction, including without limitation the rights
-# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell
-# copies of the Software, and to permit persons to whom the Software is
-# furnished to do so, subject to the following conditions:
-#
-# The above copyright notice and this permission notice shall be included in all
-# copies or substantial portions of the Software.
-#
-# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
-# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
-# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE
-# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
-# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM,
-# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
-# SOFTWARE.
-"""
-A collection of utilities to edit and construct tree sequences.
-"""
-import collections
-import dataclasses
-import functools
-import json
-import random
-import string
-import struct
-import typing
-
-import msprime
-import numpy as np
-
-import tskit
-import tskit.provenance as provenance
-
-
-def random_bytes(max_length):
-    """
-    Returns a random bytearray of the specified maximum length.
-    """
-    length = random.randint(0, max_length)
-    return bytearray(random.randint(0, 255) for _ in range(length))
-
-
-def random_strings(max_length):
-    """
-    Returns a random bytearray of the specified maximum length.
-    """
-    length = random.randint(0, max_length)
-    return "".join(random.choice(string.printable) for _ in range(length))
-
-
-def add_provenance(provenance_table, method_name):
-    d = provenance.get_provenance_dict({"command": f"tsutil.{method_name}"})
-    provenance_table.add_row(json.dumps(d))
-
-
-def subsample_sites(ts, num_sites):
-    """
[…]

### 2024-08-21 OldCode/COPYtest_extend_edges.py

@@ -1,0 +1,1357 @@
+import msprime
+import numpy as np
+import pytest
+
+import _tskit
+# import tests.test_wright_fisher as wf
+import tskit
+import COPYtsutil
+# from tests.test_highlevel import get_example_tree_sequences
+
+# ↑ See https://github.com/tskit-dev/tskit/issues/1804 for when
+# we can remove this.
+
+def extend_edges(ts, max_iter=10):
+    tables = ts.dump_tables()
+    mutations = tables.mutations.copy()
+    tables.mutations.clear()
+
+    last_num_edges = ts.num_edges
+    for _ in range(max_iter):
+        for forwards in [True, False]:
+            edges = _extend(ts, forwards=forwards)
+            tables.edges.replace_with(edges)
+            tables.build_index()
+            ts = tables.tree_sequence()
+        if ts.num_edges == last_num_edges:
+            break
+        else:
+            last_num_edges = ts.num_edges
+
+    tables = ts.dump_tables()
+    mutations = _slide_mutation_nodes_up(ts, mutations)
+    tables.mutations.replace_with(mutations)
+    ts = tables.tree_sequence()
+
+    return ts
+
+
+def _slide_mutation_nodes_up(ts, mutations):
+    # adjusts mutations' nodes to place each mutation on the correct edge given
+    # their time; requires mutation times be nonmissing and the mutation times
+    # be >= their nodes' times.
+
+    assert np.all(~tskit.is_unknown_time(mutations.time)), "times must be known"
+    new_nodes = mutations.node.copy()
+
+    mut = 0
+    for tree in ts.trees():
+        _, right = tree.interval
+        while (
+            mut < mutations.num_rows and ts.sites_position[mutations.site[mut]] < right
+        ):
+            t = mutations.time[mut]
+            c = mutations.node[mut]
+            p = tree.parent(c)
+            assert ts.nodes_time[c] <= t
+            while p != -1 and ts.nodes_time[p] <= t:
+                c = p
+                p = tree.parent(c)
+            assert ts.nodes_time[c] <= t
+            if p != -1:
+                assert t < ts.nodes_time[p]
+            new_nodes[mut] = c
+            mut += 1
+
+    # in C the node column can be edited in place
+    new_mutations = mutations.copy()
+    new_mutations.clear()
+    for mut, n in zip(mutations, new_nodes):
+        new_mutations.append(mut.replace(node=n))
+
+    return new_mutations
+
+
+def _extend(ts, forwards=True):
[…]

### 2024-08-21 OldCode/COPYtest_extend_paths.py

@@ -1,0 +1,1125 @@
+import msprime
+import numpy as np
+import pytest
+
+import _tskit
+# import tests.test_wright_fisher as wf
+import tskit
+import COPYtsutil as tsutil
+# from tests.test_highlevel import get_example_tree_sequences
+
+# ↑ See https://github.com/tskit-dev/tskit/issues/1804 for when
+# we can remove this.
+
+
+def _slide_mutation_nodes_up(ts, mutations):
+    # adjusts mutations' nodes to place each mutation on the correct edge given
+    # their time; requires mutation times be nonmissing and the mutation times
+    # be >= their nodes' times.
+
+    assert np.all(~tskit.is_unknown_time(mutations.time)), "times must be known"
+    new_nodes = mutations.node.copy()
+
+    mut = 0
+    for tree in ts.trees():
+        _, right = tree.interval
+        while (
+            mut < mutations.num_rows and ts.sites_position[mutations.site[mut]] < right
+        ):
+            t = mutations.time[mut]
+            c = mutations.node[mut]
+            p = tree.parent(c)
+            assert ts.nodes_time[c] <= t
+            while p != -1 and ts.nodes_time[p] <= t:
+                c = p
+                p = tree.parent(c)
+            assert ts.nodes_time[c] <= t
+            if p != -1:
+                assert t < ts.nodes_time[p]
+            new_nodes[mut] = c
+            mut += 1
+
+    # in C the node column can be edited in place
+    new_mutations = mutations.copy()
+    new_mutations.clear()
+    for mut, n in zip(mutations, new_nodes):
+        new_mutations.append(mut.replace(node=n))
+
+    return new_mutations
+
+
+"""
+Here is our new `extend_paths` algorithm.
+This handles some other tricker cases we want the
+`extend_edges` algorithm to succeed on.
+This algorithm can also extend edges, however
+its convergence is now not monotonically decreasing;
+this makes convergence INCREDIBLY SLOW.
+We think that we can combine `extend_paths` and `extend_edges`
+in a piece-meal way to speed up this convergence, but this requires
+further testing.
+"""
+
+
+class PathExtender:
+    def __init__(self, ts, forwards):
+        """
+        Below we will iterate through the trees, either to the left or the right,
+        keeping the following state consistent:
+        - we are moving from a previous tree, last_tree, to new one, next_tree
+        - here: the position that separates the last_tree from the next_tree
+        - (here, there): the segment covered by next_tree
+        - edges_out: edges to be removed from last_tree to get next_tree
[…]

### 2024-08-21 OldCode/match_mutations.py

@@ -1,0 +1,139 @@
+import msprime, tskit
+import numpy as np
+import collections
+
+ts = msprime.sim_ancestry(4, sequence_length=1e6, recombination_rate=1e-8, population_size=100, random_seed=123)
+ts2 = msprime.sim_ancestry(4, sequence_length=1e6, recombination_rate=1e-8, population_size=100, random_seed=123).extend_edges()
+
+class NodeMapper:
+
+    def __init__(self, ts, ts2, *args, **kwargs):
+        """
+        Identify, for each node in ts, matching nodes in ts2 as follows:
+            1. put down infinite-sites mutations at rate `mutation_rate` on ts,
+            2. map these to ts2 with `tskit.Tree.map_mutations`,
+            3. record "matches" between the nodes adjacent to the mutation in ts
+                and any mutations in ts2
+        The "matches" are defined as follows: if the mutation in ts is on an edge
+        from `p` to `c`, and a mapped mutation in ts2 is on a series of unary edges
+        from `p2` through more non-sample nodes `n` to `c`, then we record matches
+        between `p` and `p2`, `c` and `c2`, and both `p` and `c` to all nodes in
+        `n`.  (This reflects uncertainty about where in a series of edges separated
+        by unary nodes a mutation should be located.) Furthermore, samples always match,
+        and non-samples are not allowed to match to samples.
+
+        This is recorded in the `NodeMapper.node_map`, a dictionary with non-sample
+        nodes in `ts` as keys and a Counter of non-sample nodes in `ts2` as values,
+        recording the number of matches.
+        """
+        assert ts.num_samples == ts2.num_samples, "Numbers of samples must match."
+        self.ts = ts
+        self.ts2 = ts2
+        self.node_map = {}
+        self.extra_matches = 0
+        self.num_mutations = 0
+        self.num_uniquely_mapped = 0
+        self._map_nodes(*args, **kwargs)
+
+    def _add_to_node_map(self, n, n2, extra=False):
+        if not self.ts.node(n).is_sample() and not self.ts2.node(n2).is_sample():
+            if n not in self.node_map:
+                self.node_map[n] = collections.Counter()
+            self.node_map[n].update((n2,))
+            if extra:
+                self.extra_matches += 1
+
+    def _map_nodes(self, rate, random_seed=None):
+        mts = msprime.sim_mutations(self.ts, rate=rate, discrete_genome=False, keep=False, model=msprime.BinaryMutationModel(), random_seed=random_seed)
[…]

### 2024-08-21 edgewise_tally_unary_spans.py

@@ -1,0 +1,89 @@
+import msprime
+import tskit
+import numpy as np
+from itertools import groupby
+
+def edgewise_tally_unary_spans(ts):
+    """
+    Tally spans for each node into three classes, corresponding to the columns of the output:
+        0. Region is not part of a contiguous span in which it is coalescent somewhere.
+        1. Region is part of a contiguous span in which it is coalescent somewhere,
+            and is unary.
+        2. Region is coalescent.
+    """
+    out = np.zeros((ts.num_nodes, 3))
+    p = ts.edge(0).parent
+    pi = 0
+    for e in ts.edges():
+        if e.parent != p or e.id + 1 == ts.num_edges:
+            # done with parent p
+            pj = e.id + (e.id + 1 == ts.num_edges)
+            lefts = ts.edges_left[pi:pj]
+            rights = ts.edges_right[pi:pj]
+            breaks = np.sort(np.unique(np.concatenate([[0], lefts, rights])))
+            overlaps = np.zeros(len(breaks))
+            for j, x in enumerate(breaks):
+                overlaps[j] = np.sum(lefts <= x) - np.sum(rights <= x)
+            not_isolated = np.full(len(breaks) - 1, False)
+            coalescent_region = False
+            for j in range(0, len(breaks) - 1, 1):
+                if overlaps[j] == 0:
+                    coalescent_region = False
+                else:
+                    if overlaps[j] == 1:
+                        not_isolated[j] = coalescent_region
+                    else:
+                        coalescent_region = True
+            coalescent_region = False
+            for j in range(len(breaks) - 2, -1, -1):
+                if overlaps[j] == 0:
+                    coalescent_region = False
+                else:
+                    if overlaps[j] == 1:
+                        not_isolated[j] = not_isolated[j] or coalescent_region
+                    else:
+                        coalescent_region = True
+            for j in range(0, len(breaks) - 1, 1):
+                if overlaps[j] > 1:
+                    out[p, 2] += breaks[j + 1] - breaks[j]
+                elif overlaps[j] == 1:
+                    if not_isolated[j]:
+                        out[p, 1] += breaks[j + 1] - breaks[j]
+                    else:
[…]

### 2024-08-21 match_mutations.py

@@ -1,139 +1,0 @@
-import msprime, tskit
-import numpy as np
-import collections
-
-ts = msprime.sim_ancestry(4, sequence_length=1e6, recombination_rate=1e-8, population_size=100, random_seed=123)
-ts2 = msprime.sim_ancestry(4, sequence_length=1e6, recombination_rate=1e-8, population_size=100, random_seed=123).extend_edges()
-
-class NodeMapper:
-
-    def __init__(self, ts, ts2, *args, **kwargs):
-        """
-        Identify, for each node in ts, matching nodes in ts2 as follows:
-            1. put down infinite-sites mutations at rate `mutation_rate` on ts,
-            2. map these to ts2 with `tskit.Tree.map_mutations`,
-            3. record "matches" between the nodes adjacent to the mutation in ts
-                and any mutations in ts2
-        The "matches" are defined as follows: if the mutation in ts is on an edge
-        from `p` to `c`, and a mapped mutation in ts2 is on a series of unary edges
-        from `p2` through more non-sample nodes `n` to `c`, then we record matches
-        between `p` and `p2`, `c` and `c2`, and both `p` and `c` to all nodes in
-        `n`.  (This reflects uncertainty about where in a series of edges separated
-        by unary nodes a mutation should be located.) Furthermore, samples always match,
-        and non-samples are not allowed to match to samples.
-
-        This is recorded in the `NodeMapper.node_map`, a dictionary with non-sample
-        nodes in `ts` as keys and a Counter of non-sample nodes in `ts2` as values,
-        recording the number of matches.
-        """
-        assert ts.num_samples == ts2.num_samples, "Numbers of samples must match."
-        self.ts = ts
-        self.ts2 = ts2
-        self.node_map = {}
-        self.extra_matches = 0
-        self.num_mutations = 0
-        self.num_uniquely_mapped = 0
-        self._map_nodes(*args, **kwargs)
-
-    def _add_to_node_map(self, n, n2, extra=False):
-        if not self.ts.node(n).is_sample() and not self.ts2.node(n2).is_sample():
-            if n not in self.node_map:
-                self.node_map[n] = collections.Counter()
-            self.node_map[n].update((n2,))
-            if extra:
-                self.extra_matches += 1
-
-    def _map_nodes(self, rate, random_seed=None):
-        mts = msprime.sim_mutations(self.ts, rate=rate, discrete_genome=False, keep=False, model=msprime.BinaryMutationModel(), random_seed=random_seed)
[…]

### 2024-09-03 paper/writeup.tex
new algorithm description

@@
 % NOTE: these are in alphabetical order for now; we can revisit that ordering
-\author[$\dagger$]{Ava Bamforth}
+%\author[$\dagger$]{Ava Bamforth}
 \author[$\dagger$]{Halley Fritze}
@@
 
-\subsection{Extend Edges}
+\subsection{Extend Haplotypes}
 
@@
     identify areas of implied inheritance of haplotypes.
-    We do so by identifying short paths (2 edges) in a tree.
-    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
-    If neighboring trees contain edges with 
+    We do so by identifying paths in pairs of consecutive tree sequences
+    and merge them, given the correct condition.
+    For example, if neighboring trees contain edges with 
     identical ancestor (parent) and descendant (child) nodes, 
     we assume that there is implied coalescence of haplotypes between nodes, 
-    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
-  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
-    suppose there exists a path which contains a node $3$,
-    with parent and child nodes $4$ and $0$, respectively.
-    If there is an edge between the nodes $4$ and $0$ in trees 
-    $T_{k+1}$ (or $T_{k-1}$),
-    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
-    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
-    This action reduces the length of $4 \to 0$ on the genome,
-    and in some cases, completely removes the edge. 
-    We now perform this action on all such ancestral paths 
-    over the entire tree sequence.
-    With algorithm \ref{alg:edge} we only extend existing edges
-    \comment{Note possible change pending algorithm change}
-    over a larger interval on the genome,
-    and remove unnecessary edges in the process.
+    and we will extend the path's edges to the neighboring tree
+     \ref{fig:extending_diagram}.
+     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
+     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
+     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
+     and the subforest of edges in $T_{k+1}$ which were added to construct 
+     $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
+     and denoted $EO$ and $EI$. \comment{Need better notation. Currently cant think of anything better.}
+     For each edge $e \in EO$ with child $c$,
+     we construct a path $p_c^{EO}$ comprised of edges in the $EO$ subforest.
+     If there exists an edge $e'\in EI$ also with child $c$,
+     we construct path $p_c^{EI}$ from edges in the $EI$ subforest.
+     If there is a common node in both paths excluding $c$, ie.
+$(p_c^{EO}\cap p_c^{EI})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests.
+The merged path $\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EO}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
+We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $EI$, see Figure \ref{complicated extending diagram}.
+We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
+%    We do so by identifying short paths (2 edges) in a tree.
+%    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
+%    If neighboring trees contain edges with 
+%    identical ancestor (parent) and descendant (child) nodes, 
+%    we assume that there is implied coalescence of haplotypes between nodes, 
+%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
+%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
+%    suppose there exists a path which contains a node $3$,
+%    with parent and child nodes $4$ and $0$, respectively.
+%    If there is an edge between the nodes $4$ and $0$ in trees 
+%    $T_{k+1}$ (or $T_{k-1}$),
+%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
+%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
+%    This action reduces the length of $4 \to 0$ on the genome,
+%    and in some cases, completely removes the edge. 
+%    We now perform this action on all such ancestral paths 
+%    over the entire tree sequence.
+%    With algorithm \ref{alg:edge} we only extend existing edges
+%    \comment{Note possible change pending algorithm change}
+%    over a larger interval on the genome,
+%    and remove unnecessary edges in the process.
 
@@
 \end{center}
-\caption{A visualization of the \textit{extend edges} method.
-Given two neighboring trees $T_1$ and $T_2$, the tree $T_1$ has a short path of edges $4 \to 3\to0$.
-The next tree contains a single edge $4 \to 0$.
-If node $3$ is not in $T_2$, then we will extend edges $ 4\to 3$ and $3\to 0$ to tree $T_2$, hence removing the edge $4\to 0$.
+\caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method.
+Given two neighboring trees $T_1$ and $T_2$, $EO$ in the tree $T_1$ contains the edge $3\to 0$.
+Starting at $0$, there is a path of edges $0 \to 3 \to 4$.
+The tree $T_2$ has edge $4 \to 0$ in $EI$, and so there exists path $0 to 4$.
+These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
     \label{fig:extending_diagram}
@@
 \end{figure}
+
+\begin{figure}[!ht]
+	\begin{center}
+		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-Ex2.pdf}
+		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-extended-Ex2.pdf}
+	\end{center}
+	\caption{More complicated extension}
+	\label{fig:complicated_extending_diagram}
+\end{figure}
+
+\begin{algorithm2e}[!ht]
+	\SetStartEndCondition{ }{}{}%
+	\SetKwProg{Fn}{def}{\string:}{}
+	\SetKwFunction{Range}{range}%%
+	\SetKw{KwTo}{in}
+	\SetKwFor{For}{for}{\string:}{}%
+	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+	\SetKwFor{While}{while}{:}{fintq}%
+	\SetKw{Break}{break}
+	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+
+	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
+	\SetKwFunction{ex}{Extend Paths}
+	\SetKwFunction{fp}{Find Paths}
+	\DontPrintSemicolon
+	\LinesNotNumbered
+	%\SetAlgoNoEnd
+	\input{A tree sequence $\T$}
+	\output{New tree sequence,$\T'$}
+	\BlankLine
+	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
+	{
+		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
+		{\BlankLine
+			$\T = \left(T_{|\T|},..., T_1\right)$\;
+		}
+	}
+		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
+		Compute subforests $EO\subseteq T_k$ and $EI\subseteq\in T_{k+1}$.\;
+		\For{edges $e$ in $EO$}{
+			Let $c$ be child node.\;
+			$p_c^{EO} = [c]$\;
+			$p_c^{EI} = [c]$\;
+			$\bar{p}_c = []$ Path to be inserted\;
+			\While{$\exists f\in EO$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
+				$p = \tn{parent}(f)$\;
+				Append $p$ to $p_c^{EO}$\;
+				$c = p$\;
+			}
+			\While{$\exists f\in EI$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
+				$p =\tn{parent}(f)$\;
+				Append $p$ to $p_c^{EI}$\;
+				$c = p$\;
+			}
+			\If{($p_c^{EO}\cap p_c^{EI})\backslash\{c\}\neq\emptyset$}{
+				Merge paths up to intersection point\;
+				$\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EI}_c$\;
+				and sort by node time.\;
+			}
+			Insert $\bar{p}_c$ into $T_{k+1}$\;		
+		}
+	}\;
+	Update $\T$.\;
+	\KwRet{$\T$}\;
+}
+\BlankLine						
+\Fn{\ex ($\T$, iter=100)}{
+	\For{$*$ \KwTo \Range(iter)}{
+		$\T' \gets \T$.\fp(fowards=\tn{True})\;
+		$\T' \gets \T$.\fp(fowards=\tn{False})\;
+		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
+			\Break\BlankLine}
+	}
+	\KwRet{$\T'$}\;
+}
+\end{algorithm2e}
 
@@
 }
-\Indm \For{$T_k\in \T$}{
+	\For{$T_k\in \T$}{
 	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
@@
 }
-\begin{left}
-$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
+$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_1)\right)}$$\;
 \BlankLine\BlankLine
-\begin{left}
-$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
+$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_1)\right)}}$$\;
+\BlankLine
+$$true\ proportion = \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname(tr)\left( Shared Spans(T_2, T_2)\right)}$$\;
 \BlankLine\BlankLine
-\KwRet{tree discrepancy, rmse}
+\KwRet{tree discrepancy, rmse, true proportion}
 }
 \end{algorithm2e}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7.pdf}
+		\caption{Left: Discrepancy \newline Right: True Proportion }
+	\end{center}
+	\label{fig:discrepancy-over-sample}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s.pdf}
+		\caption{Left: Discrepancy \newline Right: True Proportion}
+	\end{center}
+\end{figure}
 
@@
 
+Notes on trapped unary spans:
+Simulate a TS with the following parameters:
+$msprime.sim_ancestry(1000, sequence_length=5e7, recombination_rate=1e-8, population_size=1e4, coalescing_segments_only=False, random_seed=1)$
+From the above output we see that ratio of unary span bordering coalescing between 
+- ExtendPaths/Truth = 0.4917
+- ExtendEdges/Truth = 0.27828
+
+The trapped unary span of our ground truth is 49\% of the TS.
+
+Total unary span bordering coalescing: 255151996720.0 (51.0\%)
+Total trapped unary span: 245038231157.0 (49.0\%)
+Total coalescing span: 99950000000.0
+Total unary span bordering coalescing: 125466159245.0 (100.0\%)
+Total trapped unary span: 0.0 (0.0\%)
+Total coalescing span: 99950000000.0
+
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
+	\caption{Correct Unary Span for Extend Haplotypes (log)}
+	\label{fig:correct-unary-span-log}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
+	\caption{Correct Unary Span for Extend Haplotypes}
+	\label{fig:correct-unary-span}
+\end{figure}
+
 % (AVA)
@@
 \begin{figure}
-	\includegraphics[width=0.9\textwidth]{Edge_reduc_comp_speed.pdf}
+	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
     \caption{
@@
 \begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/SL_vs_Num_edges_subplot.pdf}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_reduction_prop_vs_SL.pdf}
     \caption{

### 2024-09-03 paper/writeup.tex
notation change

@@
 % NOTE: these are in alphabetical order for now; we can revisit that ordering
-%\author[$\dagger$]{Ava Bamforth}
+\author[$\dagger$]{Ava Bamforth}
 \author[$\dagger$]{Halley Fritze}
@@
      $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
-     and denoted $EO$ and $EI$. \comment{Need better notation. Currently cant think of anything better.}
-     For each edge $e \in EO$ with child $c$,
-     we construct a path $p_c^{EO}$ comprised of edges in the $EO$ subforest.
-     If there exists an edge $e'\in EI$ also with child $c$,
-     we construct path $p_c^{EI}$ from edges in the $EI$ subforest.
+     and denoted $F_O$ and $F_O$. \comment{Need better notation. Currently cant think of anything better.}
+     For each edge $e \in F_O$ with child $c$,
+     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
+     If there exists an edge $e'\in F_I$ also with child $c$,
+     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
      If there is a common node in both paths excluding $c$, ie.
-$(p_c^{EO}\cap p_c^{EI})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests.
-The merged path $\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EO}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
-We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $EI$, see Figure \ref{complicated extending diagram}.
+$(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests.
+The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
+We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $F_I$, see Figure \ref{complicated extending diagram}.
 We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
@@
 \caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method.
-Given two neighboring trees $T_1$ and $T_2$, $EO$ in the tree $T_1$ contains the edge $3\to 0$.
+Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$.
 Starting at $0$, there is a path of edges $0 \to 3 \to 4$.
-The tree $T_2$ has edge $4 \to 0$ in $EI$, and so there exists path $0 to 4$.
+The tree $T_2$ has edge $4 \to 0$ in $F_I$, and so there exists path $0 to 4$.
 These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
@@
 		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
-		Compute subforests $EO\subseteq T_k$ and $EI\subseteq\in T_{k+1}$.\;
-		\For{edges $e$ in $EO$}{
+		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
+		\For{edges $e$ in $F_O$}{
 			Let $c$ be child node.\;
-			$p_c^{EO} = [c]$\;
-			$p_c^{EI} = [c]$\;
+			$p_c^{F_O} = [c]$\;
+			$p_c^{F_I} = [c]$\;
 			$\bar{p}_c = []$ Path to be inserted\;
-			\While{$\exists f\in EO$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
+			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
 				$p = \tn{parent}(f)$\;
-				Append $p$ to $p_c^{EO}$\;
+				Append $p$ to $p_c^{F_O}$\;
 				$c = p$\;
 			}
-			\While{$\exists f\in EI$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
+			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
 				$p =\tn{parent}(f)$\;
-				Append $p$ to $p_c^{EI}$\;
+				Append $p$ to $p_c^{F_I}$\;
 				$c = p$\;
 			}
-			\If{($p_c^{EO}\cap p_c^{EI})\backslash\{c\}\neq\emptyset$}{
+			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
 				Merge paths up to intersection point\;
-				$\bar{p}_c = \hat{p}_c^{EI} \cup \hat{p}^{EI}_c$\;
+				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
 				and sort by node time.\;

### 2024-09-30 paper/writeup.tex
added figures, new alg

@@
 
-\subsection{Extend Haplotypes}
-
-% * description (HALLEY)
-% * proof of something:
-%     - arrives at a local minimum?
-%     - guess at typical reduction? (reduces edges by 1/3?)
-    We define a \textit{tree sequence}, denoted $\T$, 
-    as a tuple $\left(T_1,...,T_{|\T|}\right)$ with a sequence of points
-	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$ where
-     each $T_k$ is a tree describing genealogical relationships between samples
-     over a distinct portion of the genome $[a_{k-1}, a_k)$.
-    Given any tree sequence, our goal is to
-    identify areas of implied inheritance of haplotypes.
-    We do so by identifying paths in pairs of consecutive tree sequences
-    and merge them, given the correct condition.
-    For example, if neighboring trees contain edges with 
-    identical ancestor (parent) and descendant (child) nodes, 
-    we assume that there is implied coalescence of haplotypes between nodes, 
-    and we will extend the path's edges to the neighboring tree
-     \ref{fig:extending_diagram}.
-     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
-     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
-     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
-     and the subforest of edges in $T_{k+1}$ which were added to construct 
-     $T_{k+1}$ from $T_k$. We call these subforest the out-forest and in-forest, respectively,
-     and denoted $F_O$ and $F_O$. \comment{Need better notation. Currently cant think of anything better.}
-     For each edge $e \in F_O$ with child $c$,
-     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
-     If there exists an edge $e'\in F_I$ also with child $c$,
-     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
-     If there is a common node in both paths excluding $c$, ie.
-$(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests.
-The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
-We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original inforest $F_I$, see Figure \ref{complicated extending diagram}.
-We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
+%\subsection{Extend Edges}
+%	\comment{Should we keep a brief section about the original extend edges? If yes, some notes are below.}
 %    We do so by identifying short paths (2 edges) in a tree.
-%    \comment{Change this if we append new Extend Edges Method to account for longer paths $p\to x \to y\to c$}
 %    If neighboring trees contain edges with 
@@
 %    and remove unnecessary edges in the process.
+
+\subsection{Extend Haplotypes}
+
+% * description (HALLEY)
+% * proof of something:
+%     - arrives at a local minimum?
+%     - guess at typical reduction? (reduces edges by 1/3?)
+    We define a \textit{tree sequence}, denoted $\T$, 
+    as a tuple $\left(T_1,...,T_{|\T|}\right)$ of $|\T|$ trees with a sequence of points
+	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$.
+     Each $T_k$ is a tree describing genealogical relationships between samples
+     over a distinct portion of the genome $[a_{k-1}, a_k)$.
+    Given any tree sequence, our goal is to
+    identify areas of implied inheritance of haplotypes.
+    We do so by identifying paths in pairs of consecutive tree sequences
+    and merging them, given the correct condition.
+    For example, if there exists paths neighboring trees which contain 
+    identical ancestor (parent) and descendant (child) nodes, 
+    we assume that there is implied coalescence of haplotypes between nodes, 
+    and we will extend the path's edges to the neighboring tree
+     \ref{fig:extending_diagram}.
+     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
+     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
+     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
+     and the subforest of edges in $T_{k+1}$ which were added to construct 
+     $T_{k+1}$ from $T_k$. We call these subforests the out-forest and in-forest, respectively,
+     and denoted $F_O$ and $F_O$.
+     For each edge $e \in F_O$ with child $c$,
+     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
+     If there exists an edge $e'\in F_I$ also with child $c$,
+     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
+     If there is a common node in both paths excluding $c$, ie.
+$(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests.
+The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
+We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original in-forest $F_I$, see Figure \ref{complicated extending diagram}.
+We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
 
@@
 \BlankLine						
-\Fn{\ex ($\T$, iter=100)}{
-	\For{$*$ \KwTo \Range(iter)}{
-		$\T' \gets \T$.\fp(fowards=\tn{True})\;
-		$\T' \gets \T$.\fp(fowards=\tn{False})\;
-		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
-			\Break\BlankLine}
+\Fn{\ex ($\T$)}{
+	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
+	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
+	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
+		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
+		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;
+
 	}
@@
 
+ We note that the discrepancy function is a metric for detecting non-similarity between two tree sequences.
+However, we note that this function is not a metric in the mathematical sense, as it is not symmetric, nor is it reflexive.
+For example, given a tree sequence $\T$ and its simplification $\mathbb{S}$ and $\T\neq\mathbb{S}$ we have $\operatorname{discrepancy}(\mathbb{S}, \T)=0$.
+Additionally, given Figure \ref{fig:conceptual_discrepancy}, $\operatorname{discrepancy}(T1, T2) \neq \operatorname{discrepancy}(T2, T1)$ \comment{actually compute this.
+We may just do it in this section instead of the appendix?}
+
 \comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}
@@
 	\begin{center}
-		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7.pdf}
+		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
 		\caption{Left: Discrepancy \newline Right: True Proportion }
@@
 	\begin{center}
-		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s.pdf}
+		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
 		\caption{Left: Discrepancy \newline Right: True Proportion}
 	\end{center}
+	\label{fig:discrepancy-over-seqlen}
 \end{figure}
@@
 }
+
+When observing our algorithm, we have to observe that for general tree sequences that have been initially simplified, there are portions of the tree sequence that the extend haplotypes method will not be able to recover.
+We call these portions the trapped unary spans.
+Most of these spans occur in the oldest portions of the ancestry near roots of the tree sequence.
+Thus we must disregard these unary spans when observings the effects of extend haplotypes.
+
+For a tree sequence $\T$ simulated in `msprime` of length $5\times 10^7$ with 1000 samples, recombination rate $10^{-8}$ and population size $10^4$, we found that there
 
@@
 
-\begin{figure}
-	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
-    \caption{
-        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
-        as a function of number of trees (modulated by sequence length),
-        on reasonably big simulated sequences.
-        For "speed": plot ratio of runtime before/after extending.
-        \label{fig:speed_and_edges}
-    }
-\end{figure}
+%\begin{figure}
+%	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
+%    \caption{
+%        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
+%        as a function of number of trees (modulated by sequence length),
+%        on reasonably big simulated sequences.
+%        For "speed": plot ratio of runtime before/after extending.
+%        \label{fig:speed_and_edges}
+%    }
+%\end{figure}
 
-Simple experiment:
-(1) simulate, simplify, and extend;
-(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
-then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
-
-With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
-(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
+%Simple experiment:
+%(1) simulate, simplify, and extend;
+%(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
+%then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
+%
+%With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
+%(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
 
@@
     }
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
+	\caption{Observation of Extendhapltypes with gene conversion = some number
+	Left: 
+	Middle:
+	Right: }
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
+	\caption{Observation of Extendhapltypes with recombination = some number
+		Left: 
+		Middle:
+		Right: }
 \end{figure}
@@
 \appendix
-
-\begin{figure}
-    \caption{
-        Supplementary figure: runtime of discrepancy function as a function of number of trees
-        (which we modulate by changing sequence length).
-        \label{fig:speed_discrepancy}
-    }
-\end{figure}
+%
+%\begin{figure}
+%    \caption{
+%        Supplementary figure: runtime of discrepancy function as a function of number of trees
+%        (which we modulate by changing sequence length).
+%        \label{fig:speed_discrepancy}
+%    }
+%\end{figure}
 

### 2024-10-07 paper/writeup.tex
methods section writing

@@
 
-\usepackage{xcolor}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{amsmath}
@@
 \tikzstyle{place}=[circle,draw=black,thick, inner sep=0pt, minimum size = 5mm]
+\usepackage{hyperref}
+\hypersetup{
+	pagebackref=true,
+	colorlinks = true,
+	allcolors = OliveGreen}
 
@@
 \newcommand{\tskit}{\texttt{tskit}}
-\newcommand{\comment}[1]{{\color{blue} \it #1}}
+\newcommand{\comment}[1]{{\color{violet} \it #1}}
 \newcommand{\argmax}{\operatorname{argmax}}
+
+\newcommand{\algorithmref}[2][]{%
+	\hyperref[{#2}]{%
+		Algorithm~\ref*{#2}%
+		\ifx\\#1\\%
+		\else
+		\,#1%
+		\fi
+	}%
+}
+
+\newcommand*{\figref}[2][]{%
+	\hyperref[{#2}]{%
+		Figure~\ref*{#2}%
+		\ifx\\#1\\%
+		\else
+		\,#1%
+		\fi
+	}%
+}
 
@@
 % reduce ancestral paths
+
+%%%% COLOR SCHEME FOR FIGURES %%%%
+% colors = {'blue': 'rgb(46,37,133)',
+%	'red': 'rgb(194,106,119)',
+%	'lgreen': 'rgb(93,168,153)',
+%	'gold': 'rgb(220,205,125)',
+%	'green': 'rgb(51, 117,56)',
+%	'lblue': 'rgb(148,203,236)',
+%	'magenta': 'rgb(159,74,150)',
+%	'wine': 'rgb(126,041,084)', 
+%}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     and merging them, given the correct condition.
-    For example, if there exists paths neighboring trees which contain 
+    For example, \figref{fig:extending_diagram}, if there exists paths neighboring trees which contain 
     identical ancestor (parent) and descendant (child) nodes, 
     we assume that there is implied coalescence of haplotypes between nodes, 
-    and we will extend the path's edges to the neighboring tree
-     \ref{fig:extending_diagram}.
-     More generally, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
+    and we will extend the path's edges to the neighboring tree.
+     More generally, see \algorithmref{alg:hap}, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
      to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
@@
      $T_{k+1}$ from $T_k$. We call these subforests the out-forest and in-forest, respectively,
-     and denoted $F_O$ and $F_O$.
+     and denote them as $F_O$ and $F_O$.
      For each edge $e \in F_O$ with child $c$,
@@
 The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
-We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original in-forest $F_I$, see Figure \ref{complicated extending diagram}.
+We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original in-forest $F_I$, see \figref{complicated_extending_diagram}.
 We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
@@
 	\end{center}
-	\caption{More complicated extension}
+	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes.
+Given two neighboring trees, there exists paths $4\to 6\to8\to10$ and $5\to 9\to 10$ in $F_O$, and in $F_I$ we have paths $4\to 7\to 11\to 10$ and $5\to 12\to 10$.
+Both the out-forest and in-forest both contain paths with the same initial and terminal nodes, $4\to 10$ and $5\to 10$ respectively, we merge both paths and output the tree sequence on the right.}
 	\label{fig:complicated_extending_diagram}
@@
 
-	\caption{Extends haplotypes across a tree sequence $TS$.}\label{alg:hap}
+	\caption{Extends haplotypes across a tree sequence $TS$.}
+	\label{alg:hap}
+
 	\SetKwFunction{ex}{Extend Paths}
@@
 
-\begin{algorithm2e}[!ht]  % How 'code-like' do we want the algorithm to be? Do we want to keep a lot of the syntax in layman's terms or do we want to use functions like ts1.coiterate(ts2)??
-
-\SetStartEndCondition{ }{}{}%
-\SetKwProg{Fn}{def}{\string:}{}
-\SetKwFunction{Range}{range}%%
-\SetKw{KwTo}{in}
-\SetKwFor{For}{for}{\string:}{}%
-\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-\SetKwFor{While}{while}{:}{fintq}%
-\SetKw{Break}{break}
-\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-
-\caption{Extends edges across a tree sequence $TS$.}\label{alg:edge}
-\SetKwFunction{ex}{Extend Edges}
-\SetKwFunction{fwd}{Extend}
-\DontPrintSemicolon
-\LinesNotNumbered
-%\SetAlgoNoEnd
-\input{A tree sequence $\T$}
-\output{New tree sequence,$\T'$}
-
-\BlankLine
-
-\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
-{	
-\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-{\BlankLine
-	$\T = \left(T_{|\T|},..., T_1\right)$\;}
-}
-	\For{$T_k\in \T$}{
-	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
-	\For{edges $f$ in $T_{k+1}$}{
-		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
-		Remove $f$ from $T_{k+1}$.\;
-		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
-		Update $\T$.\;
-	}
-	}
-}\;						
-\KwRet{$\T$}\;
-}
-\BlankLine						
-\Fn{\ex ($\T$, iter=100)}{
-	\For{$*$ \KwTo \Range(iter)}{
-		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
-		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
-		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
-			\Break\BlankLine}
-		}
-}
-\KwRet{$\T'$}\;
-\end{algorithm2e}
+%% Can most likely delete the extend-edges algorithm %%
+%\begin{algorithm2e}[!ht] 
+%	
+%\SetStartEndCondition{ }{}{}%
+%\SetKwProg{Fn}{def}{\string:}{}
+%\SetKwFunction{Range}{range}%%
+%\SetKw{KwTo}{in}
+%\SetKwFor{For}{for}{\string:}{}%
+%\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+%\SetKwFor{While}{while}{:}{fintq}%
+%\SetKw{Break}{break}
+%\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+%\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+%
+%\caption{Extends edges across a tree sequence $TS$.}
+%\label{alg:edge}
+%
+%\SetKwFunction{ex}{Extend Edges}
+%\SetKwFunction{fwd}{Extend}
+%\DontPrintSemicolon
+%\LinesNotNumbered
+%%\SetAlgoNoEnd
+%\input{A tree sequence $\T$}
+%\output{New tree sequence,$\T'$}
+%
+%\BlankLine 
+%
+%\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
+%{	
+%\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
+%{\BlankLine
+%	$\T = \left(T_{|\T|},..., T_1\right)$\;}
+%}
+%	\For{$T_k\in \T$}{
+%	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
+%	\For{edges $f$ in $T_{k+1}$}{
+%		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
+%		Remove $f$ from $T_{k+1}$.\;
+%		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
+%		Update $\T$.\;
+%	}
+%	}
+%}\;						
+%\KwRet{$\T$}\;
+%}
+%\BlankLine						
+%\Fn{\ex ($\T$, iter=100)}{
+%	\For{$*$ \KwTo \Range(iter)}{
+%		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
+%		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
+%		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
+%			\Break\BlankLine}
+%		}
+%}
+%\KwRet{$\T'$}\;
+%\end{algorithm2e}
 
@@
  in fact correctly infer haplotypes in our original tree sequence.
- One such function a so called \textit{discrepancy function} (see Algorithm \ref{alg:disc}), can be called on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
- Let $N_1$ and $N_2$ be the set of nodes on $\T_1$ respectively $\T_2$.
- For a pair of nodes $(n_1,n_2)\in N_1\times N_2$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
- Each node $n\in\T$ has a given span along the genome and a time, 
- so we can write each node $n = (t_n, L_n)$ where $L_n\subset \left[0,L\right)$ and $t_n\in \R^{\geq 0}$.
- Clearly, computing the discrepancy between a random pair of nodes is not helpful, 
- so we must first identify which pairs of nodes would be most identical to one another or a 'best match'.
- When we find 'best matching' nodes we want our pair to be close in terms of time and have equal descendant sample sets,
- ie. $s(n_1) = s(n_2)$ where $s$ is the function to compute the descendant sample set of $n_1$.
+  % Robinson-Foulds Paper: https://www.sciencedirect.com/science/article/abs/pii/0025556481900432?via%3Dihub 
+ An immediate choice of metric would be the Robinson-Foulds metric \comment{CITE} which compares nodes between individual pairs of trees.
+ However, in an effort to infer coalescent haplotypes, our function chooses to compare nodes
+ across the entire genome. 
+ \comment{something something something...}
 
- To find a best match for each node $n_1\in N_1$ \comment{(Should I do this by mentioning clades?)}
- we compute the \textit{shared span} between $(n_1, m)$ for all $m\in N_2$.
- The \textit{shared span} between two tree sequences, $ss(\T_1,\T_2)$, is an $|N_1|\times |N_2|$ matrix $A$
- where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_1\times N_2$.
- If a pair has equal descendant sample sets, $s(i) = s(j)$
+ Inspired by Robinson-Foulds we define a function, called a \textit{discrepancy function} (see \algorithmref{alg:disc}), on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
+ Let $N_{T_1}$ and $N_{T_2}$ be the set of nodes on $\T_1$ respectively $\T_2$.
+ For a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
+ Each node $n\in\T$ has a given position along the genome and time, 
+ so we can write each node as a tuple $n = (t_n, L_n)$ with position $L_n\subset \left[0,L\right)$ and time $t_n\in \R^{\geq 0}$.
+ Computing the discrepancy between a random pair of nodes is not helpful, 
+ so we must first identify which pairs of nodes would be most identical to one another or a `best match'.
+ These `best matching' nodes should be close in terms of time and have equal descendant sample sets,
+ ie. $s(i) = s(j)$ where $s$ is the function to compute the descendant sample set of $n_1$.
+
+ To find a best match for each node $i\in N_{T_1}$ \comment{(Should I do this by mentioning clades?)}
+ we compute the \textit{shared span} between node $i\in N_{T_1}$ and every node in $N_{T_2}$.
+ Then, the \textit{shared span} between two tree sequences, denoted $ss(\T_1,\T_2)$, is a $|N_{T_1}|\times |N_{T_2}|$ matrix $A$
+ where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$.
+ If a pair $(i,j)$ has equal descendant sample sets, $s(i) = s(j)$
  then $a_{ij} = |L_i \cap L_j|$.
  If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
- Matrix $A$ then determines the \textit{best match} for every node $i\in N_1$
- via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_2}a_{ij}$.
+ Matrix $A$ determines the \textit{best match} for every node $i\in N_{T_1}$
+ via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_{T_2}} a_{ij}$.
  If this maximum is not unique, we use time difference to determine the best match pair.
@@
  one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
- This function is similar in idea to the Robinson-Foulds metric \comment{CITE}.
- % Robinson-Foulds Paper: https://www.sciencedirect.com/science/article/abs/pii/0025556481900432?via%3Dihub 
- However, where the Robinson-Foulds metric compares nodes between individual pairs of trees,
- in an effort to infer coalescent haplotypes, our function chooses to compare nodes
- across the entire genome. 
- We think this helps track ancestry \comment{something something something...}
 
@@
 For example, given a tree sequence $\T$ and its simplification $\mathbb{S}$ and $\T\neq\mathbb{S}$ we have $\operatorname{discrepancy}(\mathbb{S}, \T)=0$.
-Additionally, given Figure \ref{fig:conceptual_discrepancy}, $\operatorname{discrepancy}(T1, T2) \neq \operatorname{discrepancy}(T2, T1)$ \comment{actually compute this.
+Additionally, given \figref{fig:conceptual_discrepancy}, $\operatorname{discrepancy}(T1, T2) \neq \operatorname{discrepancy}(T2, T1)$ \comment{actually compute this.
 We may just do it in this section instead of the appendix?}
@@
 
-% Let $TS^1=(\{T_i^1\},N^1,E^1,L^1)$ and $TS^2=(\{T^2_j\},N^2,E^2,L^2)$ be tree sequences.
-% Suppose that $x\in N^1$ and $y\in N^2$ are nodes in each tree sequence.
-% We denote the \textit{set of descendant samples} of $x$ at position $i$ in the genome as $D_{TS^1}(x,i)$.
-% Now for a pair of nodes $(x,y)\in TS^1\times TS^2$ 
-% we can measure how much they agree along the genome.
-% We do so with the following function
-% $$\rho_{(TS^1,TS^2)}(x,y)=\frac{1}{L}\sum_{i=0}^{L-1}[D_{TS^1}(x,i)=D_{TS^2}(y,i)],$$
-% where $L=\min\{L^1,L^2\}$ and $[\cdot]$ is the characteristic function. 
-% We see $\rho$ measures the fraction of the tree sequence
-% along which $x$ and $y$ have exactly the same descendant set. 
-% For each node $x\in N^1$ we can use $\rho$ to find its global best matching node $y\in N^2$. 
-% This node $y$ should not just be a maximum of $\rho(x,\cdot)$ but it should also have a branch length similar to $x$. 
-% This means our node $y$ should minimize both $|t_x-t_y|$ and $1-\rho_{(TS^1,TS^2)}$.
-% We then use this $y$ to find count the length of genome which $x$ differs from its best matching node in $TS_2$ with the function
-% $$\hat{d}(x,TS^1,TS^2)=\min_y\{|t_x-t_y|(1-\rho_{(TS^1,TS^2)}(x,y))\}.$$
-% Here $t_x$ and $t_y$ denote the time of occurrences (ie. branch length) for nodes $x$ and $y$. 
-% The function $\hat{d}$ measures the proportion of the genome 
-% over which the descendant set of $x$ differs from its best matching node $y$.
-% We can now extend this from any node $x\in N^11$ to all of $TS_1$ with the \textbf{discrepancy function}
-% $$d(TS_1,TS_2)=\sum_{x\in N_1}\hat{d}(x,TS_1,TS_2).$$
-% This function, as we will see in the algorithm below,
-% is $|N^1||\{T^1\}|$ as for each tree in $TS_1$ and each node $x\in N^1$ we only need to look at $O(1)$ nodes rather than all nodes in $N^2$.
-%
-
-% PLR: commented this out since it's not compiling
-% 
  \begin{algorithm2e}[!ht]
@@
 
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
-		\caption{Left: Discrepancy \newline Right: True Proportion }
-	\end{center}
-	\label{fig:discrepancy-over-sample}
-\end{figure}
-
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
-		\caption{Left: Discrepancy \newline Right: True Proportion}
-	\end{center}
-	\label{fig:discrepancy-over-seqlen}
-\end{figure}
-
 \section{Results}
@@
 
+%% TO DO:
+%% 1. Discuss trapped unary spans more
+%% 2. Discuss the comparison between tsinfer and extend haplotypes
+%% 3. Add more figures 
+%%	- (Edge counts between tsinfer and extend haplotypes)
+%%	- Average Edge counts per iterations
+%% 4. Discuss figures about total edge span added in both recombination and gene conversion events
+%% 5. Discuss the figures on edge reduction (include #3 part 1.)
+
 When observing our algorithm, we have to observe that for general tree sequences that have been initially simplified, there are portions of the tree sequence that the extend haplotypes method will not be able to recover.
-We call these portions the trapped unary spans.
+We call these portions \textit{trapped unary spans}.
 Most of these spans occur in the oldest portions of the ancestry near roots of the tree sequence.
-Thus we must disregard these unary spans when observings the effects of extend haplotypes.
+Thus we must disregard these unary spans when observing the effects of extend haplotypes.
+First we have to determine how much of a tree sequence is trapped?
+For a tree sequence $\T$ simulated in `msprime` with genomic length $5\times 10^7$ with 1000 samples, recombination rate $10^{-8}$ and population size $10^4$, we found that trapped unary spans represented 49\%.
+This means that extend haplotypes can only affect 51\% of this tree sequence.
+When we apply our algorithm to the simplified given tree sequence, the ratio of unary span bordering coalescing between the extend tree sequence and the original is 0.4917.
+This means that \comment{something something something...}
+See \figref{fig:correct-unary-span-log}, \figref{fig:correct-unary-span}, \figref{fig:node-spans}.
 
-For a tree sequence $\T$ simulated in `msprime` of length $5\times 10^7$ with 1000 samples, recombination rate $10^{-8}$ and population size $10^4$, we found that there
+%$msprime.sim_ancestry(1000, sequence_length=5e7, recombination_rate=1e-8, population_size=1e4, coalescing_segments_only=False, random_seed=1)$
+%From the above output we see that ratio of unary span bordering coalescing between 
+%- ExtendPaths/Truth = 0.4917
+%- ExtendEdges/Truth = 0.27828
+%
+%The trapped unary span of our ground truth is 49\% of the TS.
+%
+%Total unary span bordering coalescing: 255151996720.0 (51.0\%)
+%Total trapped unary span: 245038231157.0 (49.0\%)
+%Total coalescing span: 99950000000.0
+%Total unary span bordering coalescing: 125466159245.0 (100.0\%)
+%Total trapped unary span: 0.0 (0.0\%)
+%Total coalescing span: 99950000000.0
 
-Notes on trapped unary spans:
-Simulate a TS with the following parameters:
-$msprime.sim_ancestry(1000, sequence_length=5e7, recombination_rate=1e-8, population_size=1e4, coalescing_segments_only=False, random_seed=1)$
-From the above output we see that ratio of unary span bordering coalescing between 
-- ExtendPaths/Truth = 0.4917
-- ExtendEdges/Truth = 0.27828
+\begin{figure}[bht]
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions.pdf}
+	\caption{Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).}
+	\label{fig:node-spans}
+\end{figure}
 
-The trapped unary span of our ground truth is 49\% of the TS.
-
-Total unary span bordering coalescing: 255151996720.0 (51.0\%)
-Total trapped unary span: 245038231157.0 (49.0\%)
-Total coalescing span: 99950000000.0
-Total unary span bordering coalescing: 125466159245.0 (100.0\%)
-Total trapped unary span: 0.0 (0.0\%)
-Total coalescing span: 99950000000.0
-
-\begin{figure}
+\begin{figure}[bht]
 	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
@@
 	\label{fig:correct-unary-span}
+\end{figure}
+
+\begin{figure}
+%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
+	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
+	\label{fig:edges-per-iteration}
 \end{figure}
@@
 %(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
+		\caption{Left: Discrepancy \newline Right: True Proportion }
+	\end{center}
+	\label{fig:discrepancy-over-sample}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
+		\caption{Left: Discrepancy \newline Right: True Proportion}
+	\end{center}
+	\label{fig:discrepancy-over-seqlen}
+\end{figure}
 

### 2024-10-10 paper/writeup.tex
plot fixes
writing!

@@
 \newcommand{\T}{\mathbb{T}}
+\newcommand{\ind}{\mathbf{1}}
 \newcommand{\tn}{\textnormal}
@@
 \newcommand{\tskit}{\texttt{tskit}}
+\newcommand{\argmax}{\operatorname{argmax}}
+\newcommand{\dis}{\operatorname{dis}}
+\newcommand{\agree}{\operatorname{agree}}
 \newcommand{\comment}[1]{{\color{violet} \it #1}}
-\newcommand{\argmax}{\operatorname{argmax}}
 
@@
 \author[$\dagger$]{Nathaniel Pope}
-\author[$\ast$,$\dagger$,1]{Peter Ralph}
+\author[$\ast$,$\dagger$,S]{Peter Ralph}
 
@@
 \affil[$\ddagger$]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, University of Oxford}
+\affil[$\S$]{Department of Data Science, University of Oregon, Eugene, Oregon}
 
@@
 that describes how a set of genotypes samples are related to each other
-at each position of the genome.
-For reviews of this progress, see CITE CITE CITE.
-Often such methods focus on the sequence of ``marginal trees'',
+at each position of the genome \citep{XXX}.
+One way of viewing the ARG is as a sequence of ``marginal trees'',
 i.e., the genealogical trees that describe how each portion of the genome
 was inherited by the focal genomes.
-For instance, the ``succinct tree sequence'',
+This is reflected in methodology of some ARG inference methods \citep[e.g.,][]{Relate},
+in metrics used to assess inference accuracy \citep{XXX},
+as well as in basic terminology.
+For instance, 
+the ``succinct tree sequence'',
 introduced by \citet{kelleher2016efficient},
@@
 and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
-\tskit (CITE).
+\tskit \citep{XXX}.
 
 However, the ARG is emphatically not merely a sequence of trees:
-since each segment of an ancestor's genome that has been inherited by the focal genomes
-may extend over a portion of the genome over which there are many distinct trees,
-a single ancestor should usually be represented by nodes in many distinct trees.
-In other words, one way to view an ARG is as a sequence of trees in which nodes may be shared across many trees,
-but another way to view the ARG is as a set of relationships between segments of ancestral haplotypes.
+viewed another way, it describes inheritance relationships between ancestral haplotypes.
+These two points of view are related because a single haplotype
+may extend over many marginal trees;
+in other words, the internal nodes in the trees are labeled, and many of these labels
+are shared between adjacent trees.
 
@@
 are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
-\citet{fisher} pointed out that not only mutations
+It has long been observed 
+that not only mutations
 but also the ``junctions'' between distinct haplotypes,
 if they could be somehow identified,
-would be inherited as Mendelian markers.
+would be inherited as Mendelian markers
+\citet{fisher1954fuller,chapman2003model}.
+In more modern terminology, 
+even in the absence of new mutations,
+recombination between distinct haplotypes can create a novel haplotype
+whose relationships and origination time could be inferred.
+
+Haplotype identity has been relatively overlooked in the literature on ARG inference so far --
+all metrics that have been used so far to measure accuracy of inferred ARGs
+depend only on the sequence of marginal trees,
+not on sharing of ancestral haplotype identities between them.
+Suppose two methods are used to infer ARGs from the same set of genotyped samples.
+Although both ARGs describes relationships between many haplotypes,
+only those haplotypes that represent the focal set of genotyped samples
+can be unambiguously said to represent the same thing in both.
+So, most metrics only describe these relationships,
+which ignores the additional information carried by the structure of other haplotypes.
+For instance, \citet{XXX} compared true and inferred ARGs
+using average XXX distance between trees across a regular sequence of genomic positions,
+using sampled genotypes as labels,
+while \citet{brandt2022evaluation} compared times to most recent common ancestor
+between pairs of sampled genomes.
+Neither is affected by shared haplotype structure -
+two ARGs could be identiical by either measure
+but have completely different identity by descent (IBD) segment structure
+(as defined in the context of the ARG, \citep{XXX}).
+
+In this paper, we study various aspects of haplotype identity in ARGs.
+First, we describe a deterministic algorithm that
+extends the span over which ancestral haplotypes extend
+using the principle that intermediate nodes in inheritance paths
+should remain unchanged when possible.
+These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
+To quantify how accurate the new information is,
+we define and describe how to compute new measures of (dis-)agreement between ARGs
+that are motivated by the XXX distance between trees
+but account for haplotype identity.
+These measures show that the vast majority of these extended haplotypes are correct,
+and that substantial information about haplotypes is contained in these nodes
+in inferred trees as well.
 
@@
 
-Consider the (small portion) of a hypothetical tree sequence in Figure~\ref{fig:conceptual}A.
+Consider the (small portion) of a hypothetical tree sequence in \figref{fig:conceptual}A.
 On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
@@
 However, things change once we think about what this implies about haplotype inheritance.
-Figure~\ref{fig:conceptual}B shows the implied inheritance of haplotypes,
+\figref{fig:conceptual}B shows the implied inheritance of haplotypes,
 with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
@@
 Although this is possible, it seems unlikely --
-a more parsimonious explanation is depicted in Figure~\ref{fig:conceptual}C,
+a more parsimonious explanation is depicted in \figref{fig:conceptual}C,
 in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
@@
 This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
-which is depicted in Figure~\ref{fig:conceptual}D --
+which is depicted in \figref{fig:conceptual}D --
 and so node 3 has become unary in this tree.
 Note that the more parsimonious tree sequence also includes fewer edges:
-the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in Figure~\ref{fig:conceptual}B
+the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref{fig:conceptual}B
 have been reduced to the two edges
-$4 \to 3$ and $3 \to 1$ in Figure~\ref{fig:conceptual}D.
+$4 \to 3$ and $3 \to 1$ in \figref{fig:conceptual}D.
+
+So, given the ARG shown in \figref{fig:conceptual}A\&B,
+it should be possible to extend the ancestral haplotype represented by node 3
+to obtain the ARG shown in \figref{fig:conceptual}C\&D,
+thus adding additional information to the ARG.
+This might be surprising,
+as intuition from phylogenetics suggests we can only infer
+information about the branching points in the tree, not intermediate (unary) nodes.
+The goal of this paper is to answer:
+How can we do this, and how accurate is the resulting inference?
 
@@
 \end{figure}
-
-\section{Methods}
 
@@
 
-\subsection{Extend Haplotypes}
+\section{Methods}
 
-% * description (HALLEY)
-% * proof of something:
-%     - arrives at a local minimum?
-%     - guess at typical reduction? (reduces edges by 1/3?)
-    We define a \textit{tree sequence}, denoted $\T$, 
-    as a tuple $\left(T_1,...,T_{|\T|}\right)$ of $|\T|$ trees with a sequence of points
-	$0 = a_0 \leq a_1 \leq \cdots < a_{|\T|} = L$.
-     Each $T_k$ is a tree describing genealogical relationships between samples
-     over a distinct portion of the genome $[a_{k-1}, a_k)$.
-    Given any tree sequence, our goal is to
+\subsection{Notation and terminology}
+
+We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
+to take advantage of the extensive, stable, and well-documented set of open-source tools
+available in \tskit{} \citep{tskit}.
+So, our terminology and notation follows \citet{ralph2020efficiently}.
+For our purposes here,
+a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
+which represent ancestral segments of genome,
+and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
+Each node $n \in N$ has a \emph{time} $t_n$,
+which is the amount of time in the past that the individual who carried that segment of genome lived.
+Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
+over a segment of genome $[\ell_e, r_e)$.
+(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
+Suppose that the unique elements of the set of left and right edge endpoints
+are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
+Using this information, one can construct the sequence of 
+trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
+each $T_k$ is a tree whose nodes are in $N$
+and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
+Not all nodes appear in each tree,
+and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
+for node $n$.
+
+\subsection{An algorithm to extend haplotypes}
+
+% Description of algorithm
+Given a tree sequence, our goal is to
 identify areas of implied inheritance of haplotypes.
-    We do so by identifying paths in pairs of consecutive tree sequences
-    and merging them, given the correct condition.
-    For example, \figref{fig:extending_diagram}, if there exists paths neighboring trees which contain 
-    identical ancestor (parent) and descendant (child) nodes, 
-    we assume that there is implied coalescence of haplotypes between nodes, 
-    and we will extend the path's edges to the neighboring tree.
-     More generally, see \algorithmref{alg:hap}, for a pair of trees $T_k, T_{k+1}\in\mathbb{T}$,
-     to transition from tree $T_k$ to tree $T_{k+1}$ we identify the 
-     subforest of edges in $T_k$ that are removed to construct $T_{k+1}$
-     and the subforest of edges in $T_{k+1}$ which were added to construct 
-     $T_{k+1}$ from $T_k$. We call these subforests the out-forest and in-forest, respectively,
-     and denote them as $F_O$ and $F_O$.
-     For each edge $e \in F_O$ with child $c$,
-     we construct a path $p_c^{F_O}$ comprised of edges in the $F_O$ subforest.
-     If there exists an edge $e'\in F_I$ also with child $c$,
-     we construct path $p_c^{F_I}$ from edges in the $F_I$ subforest.
-     If there is a common node in both paths excluding $c$, ie.
-$(p_c^{F_O}\cap p_c^{F_I})\backslash \{c\}\neq\emptyset$, then there is a mergable path between both subforests.
-The merged path $\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_O}$ is the union of both paths truncated ($\hat{p}$) at the common intersection node.
-We then insert $\bar{p}$ into $T_2$, postponing or deleting edges in its original in-forest $F_I$, see \figref{complicated_extending_diagram}.
-We perform this action across all consecutive pairs of tree sequences, going both forward and backward along the genome, until there are no more edges extended in this way.
+Generalizing from \figref{fig:conceptual} (or \figref{fig:extending_diagram}?),
+we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
+but for which some of the intermediate nodes are missing.
+Concretely, suppose that 
+if in tree $T_k$ there is a chain of inheritance
+$p \to u_1 \to \cdots \to u_m \to c$
+(where $a \to b$ denotes a parent-child relationship)
+and in tree $T_{k+1}$ there is a chain of inheritance
+$p \to v_1 \to \cdots \to v_n \to c$,
+where $\{u_i\}_{i=1}^m$ and $\{v_j\}_{j=1}^n$ are disjoint.
+This situation implies that $c$ inherited from $p$ over the entire interval $[a_{k-1}, a_{k+1})$,
+so it seems reasonable to assume that $c$ has inherited from $p$ \emph{along the same path} for that entire interval.
+In other words, the intermediate nodes $\{u_i\}$ should also lie on the path from $c$ to $p$ in tree $T_{k+1}$,
+and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
+Of course, this does not always make sense --
+for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
+or if $t_{u_i} = t_{v_j}$ for some $i$ and $j$.
+for which the intermediate nodes are not in the other tree
+So, we restrict our attention to pairs of such paths in adjacent trees
+for which
+$u_i \notin T_{k+1}$ for all $1 \le i \le m$,
+$v_j \notin T_k$ for all $1 \le j \le n$,
+and the times of the nodes $\{u_i\}$ and $\{v_j\}$ are unique.
+Call a pair of such paths \emph{mergeable}.
+So, the goal of our algorithm is to iterate over trees,
+identify mergeable pairs of paths,
+and then extend the nodes $\{u_i\}$ to $T_{k+1}$.
+(We also extend $\{v_j\}$ to $T_k$, but on a backwards pass.)
 
@@
 \end{center}
-\caption{\comment{Change figure to say Extend halplotypes}A visualization of the \textit{extend haplotypes} method.
+\caption{\comment{Change figure to say Extend halplotypes} A visualization of the \textit{extend haplotypes} method.
 Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$.
@@
 
+A relatively efficient algorithm to do this is described in \algorithmref{alg:extend}, 
+and implemented in \tskit{} as \texttt{extend\_haplotypes}.
+The algorithm considers each tree transition from $T_k$ to $T_{k+1}$ in turn, updating its internal state
+(which includes possibly modifying $T_k$) as it goes.
+Suppose we are at the transition from tree $T_k$ to tree $T_{k+1}$,
+which is done by first removing a set of edges $O$
+and then adding another set of edges $I$.
+$O$ defines a sub-forest $F_O$ of $T_k$,
+and $I$ defines a sub-forest $F_I$ of $T_{k+1}$.
+The key step in the algorithm is to determine whether the pair of paths
+that terminate in a given node in two adjacent trees are mergeable,
+\algorithmref{alg:mergeable}, which works as follows.
+If a pair of paths is mergeable,
+then the edges of the two paths must lie in $O$ and $I$, respectively.
+Suppose an edge in $O$ has child $c$.
+To see if $c$ is the base of a pair of mergeable paths,
+the algorithm traverses up from $c$ in both $F_O$ and $F_I$;
+terminating if a node in the other tree is found
+(i.e., if the node traversed in $F_O$ is in $T_{k+1}$ or if the node traversed in $F_I$ is in $T_k$)
+or if a pair of traversed nodes have the same time.
+If these two traversals end in the same node $p$, the paths are mergeable.
+Iterating over all edges in $O$ will thus find all mergeable pairs of paths.
+There is often more than one pair of mergeable paths in a tree transition;
+so, the algorithm merges pairs of mergeable paths,
+starting with pairs that add the smallest number of new edges,
+until no more are found.
+\comment{Maybe we need to highlight in a figure how new edges can be added.}
+
+\algorithmref{alg:extend} simplifies the algorithm in several ways --
+for instance, the bookeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
+Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
+in practice we do repeated passes in both directions until no changes can be made.
+The main step that is omitted is a description of the \texttt{merge} operation,
+which performs the actual extending of haplotypes.
+This algorithm is essentially the same as \texttt{mergeable} in \algorithmref{alg:mergeable},
+except with additional bookkeeping.
+Roughly speaking, the algorithm
+traverses up from the shared base node $c$,
+doing the appropriate operations to insert the nodes along the path found in $T_k$
+into the path in $T_{k+1}$.
+To do this, some edges that end at $a_k$ will be extended to end at $a_{k+1}$;
+some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
+and some entirely new edges may be added.
+Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
+need to be updated.
+
 \begin{algorithm2e}[!ht]
@@
 	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-
-	\caption{Extends haplotypes across a tree sequence $TS$.}
-	\label{alg:hap}
-
-	\SetKwFunction{ex}{Extend Paths}
-	\SetKwFunction{fp}{Find Paths}
+    \caption{
+        Given a node $c$, trees $T_O$ and $T_I$,
+        and sub-forests $F_O$ and $F_I$
+        such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
+        checks to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
+        If the paths are mergeable then this returns the number of new edges
+        that would be added by extending the path from $T_O$ to $T_I$;
+        otherwise, returns $\infty$.
+        Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
+        in the set of edges to be removed and added, respectively
+        (i.e., in $F_O$ and $F_I$).
+    }
+	\label{alg:mergeable}
+	\SetKwFunction{mergeable}{Mergeable}
 	\DontPrintSemicolon
-	\LinesNotNumbered
-	%\SetAlgoNoEnd
-	\input{A tree sequence $\T$}
-	\output{New tree sequence,$\T'$}
-	\BlankLine
-	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
-	{
-		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-		{\BlankLine
-			$\T = \left(T_{|\T|},..., T_1\right)$\;
+    \Fn{\mergeable($c$, $T_O$, $T_I$)}{
+        $p_i = P_I[c]$\;
+        $t_i = t[p_i]$\;
+        $p_o = P_O[c]$\;
+        $t_o = t[p_o]$\;
+        $n_e = n = 0$\;
+        \While{\tn{True}}{
+            $y_i = (p_i \neq NULL)
+                \text{ and } (p_i \notin T_O)
+                \text{ and } (t_i < t_o)$ \;
+            $y_o = (p_o \neq NULL)
+                \text{ and } (p_o \notin T_I)
+                \text{ and } (t_o < t_i)$ \;
+            \If{not ($y_i$ or $y_o$)}{ \Break\; }
+            \eIf{$y_i$}{
+                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
+                    $n_e += 1$\;
+                }
+                $c = p_i$\;
+                $p_i = P_I[p_i]$\;
+                $t_i = (p_i = NULL) ? \infty : t[p_i]$\;
+            }{
+                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
+                    $n_e += 1$\;
+                }
+                $c = p_o$\;
+                $p_o = P_O[p_o]$\;
+                $t_o = (p_o = NULL) ? \infty : t[p_o]$\;
+                $n += 1$\;
             }
         }
-		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
-		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
-		\For{edges $e$ in $F_O$}{
-			Let $c$ be child node.\;
-			$p_c^{F_O} = [c]$\;
-			$p_c^{F_I} = [c]$\;
-			$\bar{p}_c = []$ Path to be inserted\;
-			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
-				$p = \tn{parent}(f)$\;
-				Append $p$ to $p_c^{F_O}$\;
-				$c = p$\;
+        \If{$n = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
+            $n_e = \infty$\;
         }
-			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
-				$p =\tn{parent}(f)$\;
-				Append $p$ to $p_c^{F_I}$\;
-				$c = p$\;
-			}
-			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
-				Merge paths up to intersection point\;
-				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
-				and sort by node time.\;
-			}
-			Insert $\bar{p}_c$ into $T_{k+1}$\;		
-		}
-	}\;
-	Update $\T$.\;
-	\KwRet{$\T$}\;
-}
-\BlankLine						
-\Fn{\ex ($\T$)}{
-	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
-	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
-	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
-		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
-		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;
-
-	}
-	\KwRet{$\T'$}\;
+        \KwRet{$n_e$}\;
     }
@@
 
-%% Can most likely delete the extend-edges algorithm %%
-%\begin{algorithm2e}[!ht] 
-%	
-%\SetStartEndCondition{ }{}{}%
-%\SetKwProg{Fn}{def}{\string:}{}
-%\SetKwFunction{Range}{range}%%
-%\SetKw{KwTo}{in}
-%\SetKwFor{For}{for}{\string:}{}%
-%\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-%\SetKwFor{While}{while}{:}{fintq}%
-%\SetKw{Break}{break}
-%\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-%\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-%
-%\caption{Extends edges across a tree sequence $TS$.}
-%\label{alg:edge}
-%
-%\SetKwFunction{ex}{Extend Edges}
-%\SetKwFunction{fwd}{Extend}
-%\DontPrintSemicolon
-%\LinesNotNumbered
-%%\SetAlgoNoEnd
-%\input{A tree sequence $\T$}
-%\output{New tree sequence,$\T'$}
-%
-%\BlankLine 
-%
-%\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
-%{	
-%\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-%{\BlankLine
-%	$\T = \left(T_{|\T|},..., T_1\right)$\;}
-%}
-%	\For{$T_k\in \T$}{
-%	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
-%	\For{edges $f$ in $T_{k+1}$}{
-%		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
-%		Remove $f$ from $T_{k+1}$.\;
-%		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
-%		Update $\T$.\;
-%	}
-%	}
-%}\;						
-%\KwRet{$\T$}\;
-%}
-%\BlankLine						
-%\Fn{\ex ($\T$, iter=100)}{
-%	\For{$*$ \KwTo \Range(iter)}{
-%		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
-%		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
-%		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
-%			\Break\BlankLine}
-%		}
-%}
-%\KwRet{$\T'$}\;
-%\end{algorithm2e}
+\begin{algorithm2e}[!ht]
+	\SetStartEndCondition{ }{}{}%
+	\SetKwProg{Fn}{def}{\string:}{}
+	\SetKwFunction{Range}{range}%%
+	\SetKw{KwTo}{in}
+	\SetKwFor{For}{for}{\string:}{}%
+	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+	\SetKwFor{While}{while}{:}{fintq}%
+	\SetKw{Break}{break}
+	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+    \caption{
+        Extend haplotypes.
+        \comment{TODO: include this in the previous algorithm?}
+    }
+	\label{alg:extend}
+	\SetKwFunction{mergeable}{Mergeable}
+	\SetKwFunction{merge}{Merge}
+	\SetKwFunction{extend}{ExtendHaplotypes}
+	\DontPrintSemicolon
+    \Fn{\extend($\T$)}{
+        \For{$k$ in $1 \ldots N$}{
+            $M = 0$\;
+            $M' = \infty$\;
+            \While{$M < \infty$}{
+                \For{$e \in O_k$}{
+                    $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$)\;
+                    \eIf{$m < M$}{
+                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
+                    }{
+                        $M' = \min(m, M)$\;
+                    }
+                }
+                $M = M'$\;
+                $M' = \infty$\;
+            }
+        }
+    }
+\end{algorithm2e}
+
+Since the algorithm needs to take multiple passes over the tree sequence
+in each direction,
+an important practical quesiton for this algorithm is:
+how passes do we need to do?
+The algorithm is monotone (spans of ancestral nodes only increase),
+so it is guaranteed to require only a finite number of passes,
+but it is also not hard to construct pathalogical cases that require an arbitrary number of passes.
+\figref{fig:num_passes} shows empirical results from simulated tree sequences,
+which suggest that \comment{XXX TODO}.
+
+\begin{figure}[!ht]
+    \caption{
+        Number of edges in the tree sequence (vertical axis)
+        after each pass of the algorithm (number of passes on the horizontal axis),
+        for tree sequences with different numbers of samples (different lines).
+        \comment{TODO}
+        \label{fig:num_passes}
+    }
+\end{figure}
 
@@
 % * Show: how add edges reduces discrepancy (RESULTS)
-% * THINK of a way to asynchronously computing discrepancy (???)
 
-\par \comment{This needs work. Choose $n_1$ or $i$ for node not both. Do you want it through the perspective of TS or through nodes? You wording is mixed on this}
- Now with an algorithm in place
- we need a metric to discern
- if trees modified with extend edges
- in fact correctly infer haplotypes in our original tree sequence.
-  % Robinson-Foulds Paper: https://www.sciencedirect.com/science/article/abs/pii/0025556481900432?via%3Dihub 
- An immediate choice of metric would be the Robinson-Foulds metric \comment{CITE} which compares nodes between individual pairs of trees.
- However, in an effort to infer coalescent haplotypes, our function chooses to compare nodes
- across the entire genome. 
- \comment{something something something...}
+If we begin with a tree sequence containing unary nodes,
+it is straightforward to remove the portions of each node's span on which it is unary,
+apply \algorithmref{alg:extend},
+and quantify how much node span was correctly or incorrectly added.
+However, we are also interested in whether \algorithmref{alg:extend}
+improves \emph{inferred} tree sequences.
+Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
+that take into account haplotype identity,
+we define a measure of \emph{discrepancy} to quantify this.
+
+It is helpful to first describe what we compute in the simple case.
+\figref{fig:node-spans}A shows the spans for each node in three tree sequences:
+for each node, the top point is the total span in a ``true'' tree sequence
+that was simulated with \texttt{msprime} \citep{msprime}
+in a way that included spans over which nodes are unary in the output
+\comment{(using which option?)}.
+The bottom point is the span of the node after \emph{simplifying} the tree sequence
+\citep{kelleher2018efficient},
+which removes all unary nodes from all trees.
+\comment{Usually (?)} between the two is the span of the node after applying \algorithmref{alg:extend}
+to the simplified tree sequence.
+So, the difference between the middle and bottom points is the amount by which
+the span of that node has been extended.
+\figref{fig:node-spans}B \comment{TODO} shows how much of this span is incorrect:
+for each node, it gives the percent of the span after running \algorithmref{alg:extend}
+that is incorrect (i.e., not present in the original tree sequence).
+The values in these plots (perhaps summed over nodes)
+therefore quantify how much the ancestral haplotypes represented by each node
+were correctly, and incorrectly, extended.
+They also tell us how much of the relationships in the original tree sequence
+are represented in the result.
+
+Now suppose that instead of comparing two tree sequences with the same set of nodes,
+we wish to compare two tree sequence for which we know the sample nodes are the same
+but are not given an identification between nodes otherwise
+(as for instance with a true, simulated tree sequence and one inferred from its genotypes).
+\comment{should have introduced ``samples'' before this}
+Call the two tree sequence $T_1$ and $T_2$, which should have the same genome length
+and the same set of sample nodes.
+We would like to measure (a) how much of $T_1$ is found in $T_2$;
+(b) how much of $T_1$ is \emph{not} found in $T_2$; and
+(c) how much of $T_2$ is not found in $T_1$.
+(Think of these three quantities as the size of the intersection
+and two differences between the tree sequences,
+thought of vaguely as sets.)
+Roughly speaking, we first identify matching nodes
+as those whose sets of descendant samples agree for the largest span along the genome,
+and then compute for how much of their spans do their descendant samples agree (or not).
+
+Concretely, suppose that the two tree sequences have the same set of breakpoints
+between trees,
+so that $T_1^{(1)}, \ldots, T_N^{(1)}$ are the trees in $\T_1$
+and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
+For a node $n$ and tree $T$
+let $S(T, n)$ denote the set of samples that inherit from $n$ in $T$,
+and for a pair of nodes $n_1$ and $n_2$ with $n_1$ in $\T_1$ and $n_2$ in $\T_2$,
+define
+\begin{align*}
+    m(n_1, n_2)
+    =
+    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
+\end{align*}
+which is the total span over which the samples below $n_1$ in $\T_1$
+matches the samples below $n_2$ in $\T_2$.
+Also let
+\begin{align*}
+    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
+\end{align*}
+the total span that node $n$ is present in the marginal trees.
+The \emph{agreement} between $\T_1$ and $\T_2$ is then defined to be
+\begin{align*}
+    \agree(\T_1, \T_2)
+    =
+    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
+\end{align*}
+where the maximum is over all mappings nodes in $\T_1$ to nodes in $\T_2$.
+(Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
+and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
+Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
+it's \emph{best matching node} in $\T_2$ as
+\begin{align*}
+    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
+\end{align*}
+so that
+\begin{align*}
+    \agree(\T_1, \T_2)
+    =
+    \sum_{n \in N_1} m(n, \alpha(n)) .
+\end{align*}
+If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
+out of those minimizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
+(and if \emph{this} is not unique, we pick an arbitrary one) --
+however, this potential ambiguity does not affect the definition of $\agree(\T_1, \T_2)$.
+We then define the \emph{discrepancy} of $\T_1$ from $\T_2$ by
+\begin{align*}
+    \dis(\T_1, \T_2)
+    =
+    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) ,
+\end{align*}
+which is the total span of all nodes in $\T_1$
+for which their descendant samples do \emph{not} match those of their best match in $\T_2$.
+
+Other authors \citep{XXX} have used
+an average Robinson-Foulds distance \citep{robinson1981comparison}
+as a measure of disagreement between tree sequences.
+This could be computed in a very similar way:
+instead of $m(n, \alpha(n))$ define
+\begin{align*}
+    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k, n_2) = s(T^{(1)}_k, n)} ,
+\end{align*}
+the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
+matches $s(T_1,n)$.
+Then the average Robinson-Foulds distance (averaged over locations in the genome)
+is
+\begin{align*}
+    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
+\end{align*}
+\comment{Check this!}
+In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
+across all trees, but average Robinson-Foulds distance allows a different node to match
+on each tree.
+The other differences are that \emph{average} Rbonson-Foulds distance
+normalizes by sequence length, and is symmetrized.
+Robinson-Foulds distance between two trees
+was defined by \citet{robinson1981comparison}
+to the minimum number of edge contraction/expansion operations needed to move
+from one tree to the other
+(which they then show is equal to the number of edges that induce different splits on the labels).
+It would be interesting to define a similar metric on tree sequences,
+perhaps using the Subgraph-Prune-and-Regraft moves of \citet{singer}.
+
+The discrepancy $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
+but not times.
+If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
+to also compare times,
+for instance by returning the weighted root-mean-squared error on inferred times as
+\begin{align*}
+    \text{wRMSE}_t(\T_1, \T_2)
+    = \sqrt{\frac{
+        \sum_{n \in N_1} s(\T_1,n) \left(t^{(1)}_n - t^{(2)}_{\alpha(n)} \right)^2 
+    }{
+        \sum_{n \in N_1} s(\T_1,n)
+    } } .
+\end{align*}
+The mean is computed weighting by node span, so that a dating error
+is more serious for a node with a longer span.
+\comment{TODO: change the definition?
+In practice, time estimation errors seem to be closer to homoskedastic on a log scale,
+so this would probably be more informative if errors are computed on the log sale.}
+
+\comment{PREVIOUS VERSION: retrieve bits from below?}
 
@@
 
+\subsection{Tree sequence compression and computation}
+
+In the simple example in \figref{fig:extending_diagram},
+extending haplotypes replaces three edges
+($0 \to 3$ and $3 \to 4$ on the left tree, and $0 \to 4$ on the right tree)
+by two edges ($0 \to 3$ and $3 \to 4$ on both trees).
+If all edge endpoints were unique, then we'd expect \emph{every} edge to be extendable
+on one of its ends
+(except those pendant to the root and some of those adjacent to chromosome ends),
+leading to a reduction in number of edges by 1/3.
+Experiments with an earlier version of the algorithm showed that
+if we only extend haplotypes on ``paths of length 1'' like this, 
+then this is achieved for long sequences.
+It is possible for \algorithmref{alg:extend} to add edges,
+but we still expect the number of edges to decrease by more than 1/3.
+Indeed,
+\figref{fig:num_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
+in half, as long as the sequence is long enough.
+
+One might expect that a reduction in edges would also lead to a reduction in computation time
+(for instance, of the class of statistics described in \citet{ralph2020efficiently}.
+This turns out not to be the case, as shown in \figref{fig:num_edges}.
+The reason for this is clear if we
+return to the algorithms in \citet{ralph2020efficiently}:
+each edge added or removed requires an update
+to the state of the parent node \emph{and every node above it in the tree}.
+So, even if an edge extends for a long segment of genome,
+something has to happen to it every time the subtree below it changes.
+Extending haplotypes does not change at all the locations in the tree affected by subtree changes.
+\comment{If the increase in computation time holds up:}
+In fact, introducing unary nodes to branches
+\emph{increases} the distance from a typical node up to the root,
+and so increases the work required (and also computation time, as seen in \figref{fig:num_edges}).
+
+\subsection{Reconstructing haplotypes of true trees}
+
 When observing our algorithm, we have to observe that for general tree sequences that have been initially simplified, there are portions of the tree sequence that the extend haplotypes method will not be able to recover.
@@
 
+\subsection{Extending haplotypes in inferred trees}
+
 \begin{figure}[bht]
 	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions.pdf}
-	\caption{Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).}
+	\caption{
+        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
+        \comment{TODO: nodes should be in the same order in all three lines,
+        using the ordering from the true tree sequence.}
+        \comment{TODO: add plot with percent wrongly added by node?}
+    }
 	\label{fig:node-spans}
@@
 \begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/SL_vs_Num_edges_subplot.pdf}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_reduction_prop_vs_SL.pdf}
+    \includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_reduction_prop_vs_SL}
+    \includegraphics{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_SL}
     \caption{
-        Maybe: summaries of lengths and/or numbers of correctly and incorrectly added edges?
-        (See notebook.)
-        \label{fig:results_edges}
+        \textbf{(A)} Proportion reduction in number of edges
+        achieved by extending haplotypes (\algorithmref{alg:extend})
+        on tree sequences of various sequence lengths and numbers of samples.
+        See Supplementary \figref{sfig:num_edges} for the raw number of edges.
+        \textbf{(B)} Runtime for computation of Tajima's $D$
+        (using the \tskit method) on the same tree sequences as in (A).
+        \comment{TODO: remove shortest lengths and double-check runtime.}
+        \label{fig:num_edges}
     }
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
 \end{figure}
@@
 
+In this paper, we have taken a ``historical'' view of the ARG --
+i.e., that each aspect of an inferred ARG is intended to represent
+a portion of some particular historical genome
+(for instance, the MRCA of some set of sampled genomes).
+Furthermore, \figref{fig:conceptual}A\&B implicitly takes the position
+that relationships \emph{not} depicted in the ARG are implied to not exist.
+This is slighly disingenuous -- an alternative and more accurate interpetation
+of the ARG of \figref{fig:conceptual}A\&B would be that we do not know
+how node 2 inherited from node 4 on the right-hand interval,
+rather than saying that it specifically did not pass through node 3.
+The ``simplification'' algorithm of \citet{XXX},
+or the Hudson algorithm for coalescent simulation
+as implemented in msprime \citep{kelleher2016efficient},
+each specifically discard information about any such ``non-coalescent'' nodes;
+so the correct interpretation of these ARGs is a lack of knowledge.
+In this paper, we have shown that this information can be largely replaced.
+
 \bibliography{references}
@@
 \appendix
-%
-%\begin{figure}
-%    \caption{
-%        Supplementary figure: runtime of discrepancy function as a function of number of trees
-%        (which we modulate by changing sequence length).
-%        \label{fig:speed_discrepancy}
-%    }
-%\end{figure}
+
+\begin{figure}
+    \caption{
+        Supplementary figure: runtime of discrepancy function as a function of number of trees
+        (which we modulate by changing sequence length).
+        \label{fig:speed_discrepancy}
+    }
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/num_trees_vs_Num_edges_subplot.pdf}
+    \caption{
+        Raw numbers of edges before and after extending haplotypes
+        for the tree sequences shown in \figref{fig:num_edges}
+        \label{sfig:num_edges}
+    }
+\end{figure}
+
+\begin{algorithm2e}[!ht]
+	\SetStartEndCondition{ }{}{}%
+	\SetKwProg{Fn}{def}{\string:}{}
+	\SetKwFunction{Range}{range}%%
+	\SetKw{KwTo}{in}
+	\SetKwFor{For}{for}{\string:}{}%
+	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+	\SetKwFor{While}{while}{:}{fintq}%
+	\SetKw{Break}{break}
+	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+
+	\caption{Extends haplotypes across a tree sequence $TS$.
+    \comment{I think we can remove this now?}}
+	\label{alg:hap}
+
+	\SetKwFunction{ex}{Extend Paths}
+	\SetKwFunction{fp}{Find Paths}
+	\DontPrintSemicolon
+	\LinesNotNumbered
+	%\SetAlgoNoEnd
+	\input{A tree sequence $\T$}
+	\output{New tree sequence,$\T'$}
+	\BlankLine
+	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
+	{
+		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
+		{\BlankLine
+			$\T = \left(T_{|\T|},..., T_1\right)$\;
+		}
+	}
+		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
+		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
+		\For{edges $e$ in $F_O$}{
+			Let $c$ be child node.\;
+			$p_c^{F_O} = [c]$\;
+			$p_c^{F_I} = [c]$\;
+			$\bar{p}_c = []$ Path to be inserted\;
+			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
+				$p = \tn{parent}(f)$\;
+				Append $p$ to $p_c^{F_O}$\;
+				$c = p$\;
+			}
+			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
+				$p =\tn{parent}(f)$\;
+				Append $p$ to $p_c^{F_I}$\;
+				$c = p$\;
+			}
+			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
+				Merge paths up to intersection point\;
+				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
+				and sort by node time.\;
+			}
+			Insert $\bar{p}_c$ into $T_{k+1}$\;		
+		}
+	}\;
+	Update $\T$.\;
+	\KwRet{$\T$}\;
+}
+\BlankLine						
+\Fn{\ex ($\T$)}{
+	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
+	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
+	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
+		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
+		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;
+
+	}
+	\KwRet{$\T'$}\;
+}
+\end{algorithm2e}
+
+% Can most likely delete the extend-edges algorithm %%
+\begin{algorithm2e}[!ht]
+
+\SetStartEndCondition{ }{}{}%
+\SetKwProg{Fn}{def}{\string:}{}
+\SetKwFunction{Range}{range}%%
+\SetKw{KwTo}{in}
+\SetKwFor{For}{for}{\string:}{}%
+\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+\SetKwFor{While}{while}{:}{fintq}%
+\SetKw{Break}{break}
+\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+
+\caption{Extends edges across a tree sequence $TS$. 
+\comment{Can be removed, probably -- unless this seems a lot simpler?}}
+\label{alg:edge}
+
+\SetKwFunction{ex}{Extend Edges}
+\SetKwFunction{fwd}{Extend}
+\DontPrintSemicolon
+\LinesNotNumbered
+%\SetAlgoNoEnd
+\input{A tree sequence $\T$}
+\output{New tree sequence,$\T'$}
+
+\BlankLine
+
+\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
+{	
+\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
+{\BlankLine
+	$\T = \left(T_{|\T|},..., T_1\right)$\;}
+}
+	\For{$T_k\in \T$}{
+	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
+	\For{edges $f$ in $T_{k+1}$}{
+		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
+		Remove $f$ from $T_{k+1}$.\;
+		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
+		Update $\T$.\;
+	}
+	}
+}\;						
+\KwRet{$\T$}\;
+}
+\BlankLine						
+\Fn{\ex ($\T$, iter=100)}{
+	\For{$*$ \KwTo \Range(iter)}{
+		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
+		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
+		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
+			\Break\BlankLine}
+		}
+}
+\KwRet{$\T'$}\;
+\end{algorithm2e}
 

### 2024-10-11 paper/writeup.tex
citations

@@
 \vspace{-33pt}% Only used for adjusting extra space in the left column of the first page
-
-\comment{idea: look at ratio of unary-to-not over time; does it change with
-    selection? or modes of recombination?}
 
@@
 that describes how a set of genotypes samples are related to each other
-at each position of the genome \citep{XXX}.
+at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
 One way of viewing the ARG is as a sequence of ``marginal trees'',
@@
 was inherited by the focal genomes.
-This is reflected in methodology of some ARG inference methods \citep[e.g.,][]{Relate},
+This is reflected in methodology of some ARG inference methods \citep[e.g., Relate][]{speidel2019method},
 in metrics used to assess inference accuracy \citep{XXX},
@@
 which ignores the additional information carried by the structure of other haplotypes.
-For instance, \citet{XXX} compared true and inferred ARGs
-using average XXX distance between trees across a regular sequence of genomic positions,
+For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
+compared true and inferred ARGs
+using average Robinson-Foulds and Kendall-Colijn distances between trees
+across a regular sequence of genomic positions,
 using sampled genotypes as labels,
@@
 (as defined in the context of the ARG, \citep{XXX}).
+\citet{deng2021distribution} evaluated agreement of distributions of
+distances along the genome between tree topology changes,
+and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
+that is the total variation distance between the induced distribution on genotypes;
+neither of which measure sharing of haplotypes between adjacent trees.
 
@@
 for each node, the top point is the total span in a ``true'' tree sequence
-that was simulated with \texttt{msprime} \citep{msprime}
+that was simulated with \texttt{msprime} \citep{kelleher2016efficient}
 in a way that included spans over which nodes are unary in the output
@@
 
-Other authors \citep{XXX} have used
+Other authors \citep[e.g.,]{kelleher2019inferring} have used
 an average Robinson-Foulds distance \citep{robinson1981comparison}
@@
 It would be interesting to define a similar metric on tree sequences,
-perhaps using the Subgraph-Prune-and-Regraft moves of \citet{singer}.
+perhaps using the Subgraph-Prune-and-Regraft moves used by Singer \citet{deng2024robust}.
 
@@
 rather than saying that it specifically did not pass through node 3.
-The ``simplification'' algorithm of \citet{XXX},
+The ``simplification'' algorithm of \citet{kelleher2018efficient},
 or the Hudson algorithm for coalescent simulation

### 2024-10-12 paper/writeup.tex
more on tree metrics

@@
 
-Haplotype identity has been relatively overlooked in the literature on ARG inference so far --
+Haplotype identity has been nearly overlooked in the literature on ARG inference so far --
 all metrics that have been used so far to measure accuracy of inferred ARGs
@@
 we define a measure of \emph{discrepancy} to quantify this.
+The method is implemented in the \texttt{tsdate} package \citep{wohns2022unified}
+as \comment{XXX}.
 
@@
 
+%%%%%%%%%%%%
 \comment{PREVIOUS VERSION: retrieve bits from below?}
@@
 
+\paragraph{Metrics on ARGs}
+\comment{collecting random notes here; TODO is tie them together}
+The discrepancy $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
+(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
+This is by design: in practice it is not possible to infer all aspects
+of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
+and so we wanted to quantify
+``How much of the true relationships does this ARG represent?''
+However, the symmetrized version
+$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
+To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
+and view each ARG as a subset of the space $[0,L) \times N \times N$,
+where $N$ is the shared set of nodes.
+Then, discrepancy is the Lesbegue measure of the relative difference of the two sets:
+$|\T_1 \setminus \T_2|$,
+and so the symmetrized version is the measure of the symmetric difference,
+which is well-known to be a metric.
+If the two ARGs have the same number of nodes,
+then we take the minimum over the all bijections between their nodes,
+and minima over a finite number of metrics is also a metric.
+Finally, if they have a different number of nodes,
+then \comment{???}.
+
+The Robinson-Foulds distance \citep{robinson1981comparison}
+essentially counts the number of differing branches between two trees;
+the averaged Robinson-Foulds distance \citep{kelleher2019inferring} % TODO: first usage?
+combines this distance across marginal trees, weighted by span along the genome.
+The method we present here for measuring discrepancy between topologies of ARGs
+is a straightforward generalization
+that takes into account span along the genome of inferred ancestral haplotypes
+(and separates the metric into two pieces as discussed in the previous paragraph).
+However, the Robinson-Foulds metric has many undesireable properties --
+for instance, moving a single tip can result in a tree with maximum distance to the original --
+and there is a substantial literature giving generalizations
+\citep[reviewed by][]{llabres2021generalized}.
+Many of these generalizations \citep[e.g.,][]{bocker2013generalized}
+relax the requirement that the match between subtended sample sets be exact,
+and weight matches in some way by the size of the discrepancy.
+We considered such definitions as well, but kept to the simple case
+for computational tractability --
+the generalization of \citet{bocker2013generalized} is NP-hard to compute, even for a single tree.
+In the ARG literature, \citet{zhang2023biobankscale}
+defines a metric (called ``ARG total variation distance'') that includes branch lengths,
+in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
+however, it is still applied to ARGs as an average over marginal trees,
+without enforcement of identity across haplotypes;
+it would be useful to extend our discrepancy to include branch lengths.
+
 \bibliography{references}

### 2024-10-21 paper/writeup.tex
new figures p2

@@
 identify areas of implied inheritance of haplotypes.
-Generalizing from \figref{fig:conceptual} (or \figref{fig:extending_diagram}?),
+Generalizing from \figref{fig:extending_diagram},
 we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
@@
 \end{center}
-\caption{\comment{Change figure to say Extend halplotypes} A visualization of the \textit{extend haplotypes} method.
-Given two neighboring trees $T_1$ and $T_2$, $F_O$ in the tree $T_1$ contains the edge $3\to 0$.
-Starting at $0$, there is a path of edges $0 \to 3 \to 4$.
-The tree $T_2$ has edge $4 \to 0$ in $F_I$, and so there exists path $0 to 4$.
-These two paths share common end points, $0$ and $4$, and so we will extend path $0 \to 3 \to 4$ onto tree $T_2$, hence removing the edge $4\to 0$.
+\caption{A visualization of the \textit{extend haplotypes} method.
+Given two neighboring trees $T_1$ and $T_2$, there exists paths from root node $6$ to sample node $0$.
+$T_1$ contains the path $6\to 4\to 0$ while $T_2$ has path $6\to 5\to 0$.
+The intermediate nodes $4$ and $5$ do not appear in $T_2$ respectively $T_1$, and so we consider these two paths \textit{mergeable}.
+The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$.
     \label{fig:extending_diagram}
@@
 	\end{center}
-	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes.
+	\caption{\comment{DELETE}}
+%	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes.
 Given two neighboring trees, there exists paths $4\to 6\to8\to10$ and $5\to 9\to 10$ in $F_O$, and in $F_I$ we have paths $4\to 7\to 11\to 10$ and $5\to 12\to 10$.
@@
 	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
-	\caption{Correct Unary Span for Extend Haplotypes}
+	\caption{\comment{DELETE}}
+%	\caption{Correct Unary Span for Extend Haplotypes}
 	\label{fig:correct-unary-span}

### 2024-10-22 paper/writeup.tex
more writeup, sample figure with "correct dpi"
It doesn't work

@@
 The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$.
+This new path extends the edge $4\to 0$ to $T_2$ as well as constructs a new edge $5\to 4$ in the tree sequence.
     \label{fig:extending_diagram}
@@
 until no more are found.
-\comment{Maybe we need to highlight in a figure how new edges can be added.}
 
@@
 some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
-and some entirely new edges may be added.
+and some entirely new edges may be added, see \figref{fig:extending_diagram}.
 Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
@@
 % * Show: how add edges reduces discrepancy (RESULTS)
+
+\comment{Rename Dissimilarity and Similarity}
 
@@
 
- \begin{algorithm2e}[!ht]
- \SetStartEndCondition{ }{}{}%
- \SetKwProg{Fn}{def}{\string:}{}
- \SetKwFunction{Range}{range}%%
- \SetKw{KwTo}{in}
- \SetKwFor{For}{for}{\string:}{}%
- \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
- \SetKwFor{While}{while}{:}{fintq}%
- \SetKw{Break}{break}
- \newcommand\forcond{$i$ \KwTo\Range{$n$}}
- \SetKwComment{Comment}{}
- \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
- \DontPrintSemicolon\LinesNotNumbered
-
- \caption{Discrepancy Function}\label{alg:disc}
- \input{$(T_1,T_2)$}
- \output{tree discrepancy, root-mean-square error}
-
- \Fn{Shared Spans($T_1,T_2$)}{
- \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
- 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
- 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
- 		Define $a_{ij}=|L_i\cap L_j|$.
- 	}
- 	\Else{Define $a_{ij}=0$.}
- Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
- }
- \KwRet{$A$}
- }
-\BlankLine
-\Fn{Discrepancy ($T_1,T_2$)}{
-\For{each node $i\in N_{T_1}$}{
-	Find nodes in $N_{T_2}$ which have the highest matching span.\;
-	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
-	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
-	Define $\alpha_i = a_{ij}$.\;
-	\BlankLine
-}
-$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_1)\right)}$$\;
-\BlankLine\BlankLine
-$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_1)\right)}}$$\;
-\BlankLine
-$$true\ proportion = \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname(tr)\left( Shared Spans(T_2, T_2)\right)}$$\;
-\BlankLine\BlankLine
-\KwRet{tree discrepancy, rmse, true proportion}
-}
-\end{algorithm2e}
+%\comment{Delete Discrepancy Algorithm!!}
+% 
+% \begin{algorithm2e}[!ht]
+% \SetStartEndCondition{ }{}{}%
+% \SetKwProg{Fn}{def}{\string:}{}
+% \SetKwFunction{Range}{range}%%
+% \SetKw{KwTo}{in}
+% \SetKwFor{For}{for}{\string:}{}%
+% \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+% \SetKwFor{While}{while}{:}{fintq}%
+% \SetKw{Break}{break}
+% \newcommand\forcond{$i$ \KwTo\Range{$n$}}
+% \SetKwComment{Comment}{}
+% \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+% \DontPrintSemicolon\LinesNotNumbered
+% 
+% \caption{Discrepancy Function}\label{alg:disc}
+% \input{$(T_1,T_2)$}
+% \output{tree discrepancy, root-mean-square error}
+% 
+% \Fn{Shared Spans($T_1,T_2$)}{
+% \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
+% 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
+% 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
+% 		Define $a_{ij}=|L_i\cap L_j|$.
+% 	}
+% 	\Else{Define $a_{ij}=0$.}
+% Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
+% }
+% \KwRet{$A$}
+% }
+%\BlankLine
+%\Fn{Discrepancy ($T_1,T_2$)}{
+%\For{each node $i\in N_{T_1}$}{
+%	Find nodes in $N_{T_2}$ which have the highest matching span.\;
+%	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
+%	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
+%	Define $\alpha_i = a_{ij}$.\;
+%	\BlankLine
+%}
+%$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_1)\right)}$$\;
+%\BlankLine\BlankLine
+%$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_1)\right)}}$$\;
+%\BlankLine
+%$$true\ proportion = \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname(tr)\left( Shared Spans(T_2, T_2)\right)}$$\;
+%\BlankLine\BlankLine
+%\KwRet{tree discrepancy, rmse, true proportion}
+%}
+%\end{algorithm2e}
 
@@
 \begin{figure}[bht]
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions.pdf}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
 	\caption{
@@
 \begin{figure}
+	\includegraphics{dissimilarity.pdf}
+\end{figure}
+
+\begin{figure}
     \caption{
@@
 
-\begin{algorithm2e}[!ht]
-	\SetStartEndCondition{ }{}{}%
-	\SetKwProg{Fn}{def}{\string:}{}
-	\SetKwFunction{Range}{range}%%
-	\SetKw{KwTo}{in}
-	\SetKwFor{For}{for}{\string:}{}%
-	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-	\SetKwFor{While}{while}{:}{fintq}%
-	\SetKw{Break}{break}
-	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-
-	\caption{Extends haplotypes across a tree sequence $TS$.
-    \comment{I think we can remove this now?}}
-	\label{alg:hap}
-
-	\SetKwFunction{ex}{Extend Paths}
-	\SetKwFunction{fp}{Find Paths}
-	\DontPrintSemicolon
-	\LinesNotNumbered
-	%\SetAlgoNoEnd
-	\input{A tree sequence $\T$}
-	\output{New tree sequence,$\T'$}
-	\BlankLine
-	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
-	{
-		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-		{\BlankLine
-			$\T = \left(T_{|\T|},..., T_1\right)$\;
-		}
-	}
-		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
-		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
-		\For{edges $e$ in $F_O$}{
-			Let $c$ be child node.\;
-			$p_c^{F_O} = [c]$\;
-			$p_c^{F_I} = [c]$\;
-			$\bar{p}_c = []$ Path to be inserted\;
-			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
-				$p = \tn{parent}(f)$\;
-				Append $p$ to $p_c^{F_O}$\;
-				$c = p$\;
-			}
-			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
-				$p =\tn{parent}(f)$\;
-				Append $p$ to $p_c^{F_I}$\;
-				$c = p$\;
-			}
-			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
-				Merge paths up to intersection point\;
-				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
-				and sort by node time.\;
-			}
-			Insert $\bar{p}_c$ into $T_{k+1}$\;		
-		}
-	}\;
-	Update $\T$.\;
-	\KwRet{$\T$}\;
-}
-\BlankLine						
-\Fn{\ex ($\T$)}{
-	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
-	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
-	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
-		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
-		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;
-
-	}
-	\KwRet{$\T'$}\;
-}
-\end{algorithm2e}
-
-% Can most likely delete the extend-edges algorithm %%
-\begin{algorithm2e}[!ht]
-
-\SetStartEndCondition{ }{}{}%
-\SetKwProg{Fn}{def}{\string:}{}
-\SetKwFunction{Range}{range}%%
-\SetKw{KwTo}{in}
-\SetKwFor{For}{for}{\string:}{}%
-\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-\SetKwFor{While}{while}{:}{fintq}%
-\SetKw{Break}{break}
-\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-
-\caption{Extends edges across a tree sequence $TS$. 
-\comment{Can be removed, probably -- unless this seems a lot simpler?}}
-\label{alg:edge}
-
-\SetKwFunction{ex}{Extend Edges}
-\SetKwFunction{fwd}{Extend}
-\DontPrintSemicolon
-\LinesNotNumbered
-%\SetAlgoNoEnd
-\input{A tree sequence $\T$}
-\output{New tree sequence,$\T'$}
-
-\BlankLine
-
-\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
-{	
-\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-{\BlankLine
-	$\T = \left(T_{|\T|},..., T_1\right)$\;}
-}
-	\For{$T_k\in \T$}{
-	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
-	\For{edges $f$ in $T_{k+1}$}{
-		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
-		Remove $f$ from $T_{k+1}$.\;
-		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
-		Update $\T$.\;
-	}
-	}
-}\;						
-\KwRet{$\T$}\;
-}
-\BlankLine						
-\Fn{\ex ($\T$, iter=100)}{
-	\For{$*$ \KwTo \Range(iter)}{
-		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
-		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
-		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
-			\Break\BlankLine}
-		}
-}
-\KwRet{$\T'$}\;
-\end{algorithm2e}
+\begin{tabular}[c|c|c|c|c|]
+	Initial & Iteration 1 & Iteration 2 & Iteration 3 & Iteration 4 \\
+	113463	&	76301	&	76059	&	76057	&	76057 \\
+	113266 & 76084	&	75835	&	75833	&	75833 \\
+	114489 & 76703 & 76436 & 76434 & 76434 \\
+	114086	& 76550	& 76294	& 76291	& 76291
+\end{tabular}
+
+%\begin{algorithm2e}[!ht]
+%	\SetStartEndCondition{ }{}{}%
+%	\SetKwProg{Fn}{def}{\string:}{}
+%	\SetKwFunction{Range}{range}%%
+%	\SetKw{KwTo}{in}
+%	\SetKwFor{For}{for}{\string:}{}%
+%	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+%	\SetKwFor{While}{while}{:}{fintq}%
+%	\SetKw{Break}{break}
+%	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+%	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+%	
+%	\caption{Extends haplotypes across a tree sequence $TS$.
+%    \comment{I think we can remove this now?}}
+%	\label{alg:hap}
+%	
+%	\SetKwFunction{ex}{Extend Paths}
+%	\SetKwFunction{fp}{Find Paths}
+%	\DontPrintSemicolon
+%	\LinesNotNumbered
+%	%\SetAlgoNoEnd
+%	\input{A tree sequence $\T$}
+%	\output{New tree sequence,$\T'$}
+%	\BlankLine
+%	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
+%	{
+%		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
+%		{\BlankLine
+%			$\T = \left(T_{|\T|},..., T_1\right)$\;
+%		}
+%	}
+%		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
+%		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
+%		\For{edges $e$ in $F_O$}{
+%			Let $c$ be child node.\;
+%			$p_c^{F_O} = [c]$\;
+%			$p_c^{F_I} = [c]$\;
+%			$\bar{p}_c = []$ Path to be inserted\;
+%			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
+%				$p = \tn{parent}(f)$\;
+%				Append $p$ to $p_c^{F_O}$\;
+%				$c = p$\;
+%			}
+%			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
+%				$p =\tn{parent}(f)$\;
+%				Append $p$ to $p_c^{F_I}$\;
+%				$c = p$\;
+%			}
+%			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
+%				Merge paths up to intersection point\;
+%				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
+%				and sort by node time.\;
+%			}
+%			Insert $\bar{p}_c$ into $T_{k+1}$\;		
+%		}
+%	}\;
+%	Update $\T$.\;
+%	\KwRet{$\T$}\;
+%}
+%\BlankLine						
+%\Fn{\ex ($\T$)}{
+%	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
+%	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
+%	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
+%		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
+%		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;
+%		
+%	}
+%	\KwRet{$\T'$}\;
+%}
+%\end{algorithm2e}
+%
+%% Can most likely delete the extend-edges algorithm %%
+%\begin{algorithm2e}[!ht] 
+%	
+%\SetStartEndCondition{ }{}{}%
+%\SetKwProg{Fn}{def}{\string:}{}
+%\SetKwFunction{Range}{range}%%
+%\SetKw{KwTo}{in}
+%\SetKwFor{For}{for}{\string:}{}%
+%\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+%\SetKwFor{While}{while}{:}{fintq}%
+%\SetKw{Break}{break}
+%\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+%\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+%
+%\caption{Extends edges across a tree sequence $TS$. 
+%\comment{Can be removed, probably -- unless this seems a lot simpler?}}
+%\label{alg:edge}
+%
+%\SetKwFunction{ex}{Extend Edges}
+%\SetKwFunction{fwd}{Extend}
+%\DontPrintSemicolon
+%\LinesNotNumbered
+%%\SetAlgoNoEnd
+%\input{A tree sequence $\T$}
+%\output{New tree sequence,$\T'$}
+%
+%\BlankLine 
+%
+%\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
+%{	
+%\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
+%{\BlankLine
+%	$\T = \left(T_{|\T|},..., T_1\right)$\;}
+%}
+%	\For{$T_k\in \T$}{
+%	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
+%	\For{edges $f$ in $T_{k+1}$}{
+%		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
+%		Remove $f$ from $T_{k+1}$.\;
+%		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
+%		Update $\T$.\;
+%	}
+%	}
+%}\;						
+%\KwRet{$\T$}\;
+%}
+%\BlankLine						
+%\Fn{\ex ($\T$, iter=100)}{
+%	\For{$*$ \KwTo \Range(iter)}{
+%		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
+%		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
+%		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
+%			\Break\BlankLine}
+%		}
+%}
+%\KwRet{$\T'$}\;
+%\end{algorithm2e}
 

### 2024-10-24 paper/writeup.tex
minor edits

@@
 \newcommand{\tskit}{\texttt{tskit}}
+\newcommand{\msprime}{\texttt{msprime}}
 \newcommand{\argmax}{\operatorname{argmax}}
@@
 for each node, the top point is the total span in a ``true'' tree sequence
-that was simulated with \texttt{msprime} \citep{kelleher2016efficient}
+that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
 in a way that included spans over which nodes are unary in the output
-\comment{(using which option?)}.
+(by setting \texttt{coalescing\_segments\_only} to False).
 The bottom point is the span of the node after \emph{simplifying} the tree sequence
@@
 which removes all unary nodes from all trees.
-\comment{Usually (?)} between the two is the span of the node after applying \algorithmref{alg:extend}
+\comment{Usually (?)} between the two points
+is the span of the node after applying \algorithmref{alg:extend}
 to the simplified tree sequence.
@@
 
+\subsection{Simulations}
+
+To do this, we simulate ARGs containing full haplotypes using \msprime's 
+option \texttt{coalescing\_segments\_only} set to False.
+Although \msprime simulates many events that do not create a coalescence in some marginal tree,
+it only outputs information for nodes which contain a coalescence
+(i.e., are the MRCA of some pair of samples at some point on the genome).
+Furthermore, by default it only outputs those segments of the genome
+on which there is a coalescence.
+Said another way, by default all ancestral nodes in the ARG
+output by \msprime are the MRCA of some pair of samples at every point in the genome
+on which they are represented.
+However, here we are interested in those segments of genome
+on which the nodes are \emph{not} coalescent;
+i.e., where they are unary in the marginal trees.
+More concretely, we hope to recover those portions of ancestral haplotypes
+on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
+For instance, if a lineage carrying an ancestral segment of genome that spans $[a, c)$
+coalesces with another spanning $[b, d)$, with $a < b < c < d$,
+then the resulting node is only coalescent on $[b, c)$ but we hope using this algorithm
+to extend the node's span to $[a, b)$ and $[c, d)$
+(on which the node is unary).
+However, following this example, the first lineage might carry a segment $[x, y)$
+that is disjoint from $[a, d)$.
+We call these segments ``isolated non-coalescent segments'';
+they have also been called ``trapped unary spans'' \citep[by][]{XXX}.
+Such segments will not be recovered by our algorithm,
+and it seems likely that most such segments are unrecoverable by any method.
+So, when evaluating what proportion of the true ARG is represented in an inferred ARG
+we omit these isolated non-coalescent segments.
+To give an idea of what proportion of the full spans of ancestral nodes
+these isolated non-coalescent segments represent,
+a simulation of 1000 samples
+with genome length $5\times 10^7$, recombination rate $10^{-8}$, and population size $10^4$
+has about half the total span of all nodes in isolated, non-coalescent segments.
+For more discussion of these segments, see \citet{baumdicker2021efficient}.
+
 \section{Results}
@@
 
-When observing our algorithm, we have to observe that for general tree sequences that have been initially simplified, there are portions of the tree sequence that the extend haplotypes method will not be able to recover.
-We call these portions \textit{trapped unary spans}.
-Most of these spans occur in the oldest portions of the ancestry near roots of the tree sequence.
-Thus we must disregard these unary spans when observing the effects of extend haplotypes.
-First we have to determine how much of a tree sequence is trapped?
-For a tree sequence $\T$ simulated in `msprime` with genomic length $5\times 10^7$ with 1000 samples, recombination rate $10^{-8}$ and population size $10^4$, we found that trapped unary spans represented 49\%.
-This means that extend haplotypes can only affect 51\% of this tree sequence.
+Our first task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
+are indeed correct.
+To do this, we simulate ARGs containing full haplotypes using \msprime
+as described in the Methods.
+
 When we apply our algorithm to the simplified given tree sequence, the ratio of unary span bordering coalescing between the extend tree sequence and the original is 0.4917.
@@
 	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
-	\caption{Observation of Extendhapltypes with gene conversion = some number
-	Left: 
-	Middle:
-	Right: }
+	\caption{
+        Accuracy of \algorithmref{alg:extend}
+        on simulated ARGs with \comment{TODO, including gene conversion}.
+        The same quantities for simulated ARGs without gene conversion
+        are shown in Supplementary \figref{fig:accuracy_on_truth_no_gc}.
+    \label{fig:accuracy_on_truth}
+}
 \end{figure}
@@
 	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
-	\caption{Observation of Extendhapltypes with recombination = some number
-		Left: 
-		Middle:
-		Right: }
+	\caption{Accuracy of \algorithmref{alg:extend}
+        as in \figref{fig:accuracy_on_truth} except that simulations were done
+        without gene conversion.
+        \label{fig:accuracy_on_truth_no_gc}
+    }
 \end{figure}

### 2024-10-24 benchmarks/do_predictions.sh
draft of benchmarking figures

@@ -1,0 +1,24 @@
+#!/bin/bash
+
+OUTDIR="one_pop_results"
+if [ ! -e $OUTDIR ]
+then
+    mkdir -p $OUTDIR
+
+    for L in 1e6 5e6 1e7 5e7 1e8
+    do
+        for N in 10 50 100 500 1000
+        do
+            SEED=$RANDOM$RANDOM
+            OUT=$OUTDIR/run_${L}_${N}_${SEED}
+            echo "$OUT  ....."
+            python run_experiment.py --num_samples $N --length $L \
+                --seed $SEED one_pop.trees $OUT
+            SEED=$((SEED + 1))
+        done
+    done
+else
+    echo "${OUTDIR} already exists; doing nothing."
+fi
+
+python plot_results.py $OUTDIR

### 2024-10-24 benchmarks/plot_results.py

@@ -1,0 +1,71 @@
+import sys
+import json
+import glob
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+import sys, os
+
+assert len(sys.argv) == 2, "Usage: python plot_results.py <directory name>"
+basedir = sys.argv[1]
+
+jj = []
+for jf in glob.glob(os.path.join(basedir, "*.json")):
+    with open(jf, "r") as f:
+        jj.append(json.load(f))
+
+df = pd.DataFrame(jj)
+
+# combined number of edges & runtime for Tajimas D
+fname = os.path.join(basedir, "absolute_values.pdf")
+fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(9, 5))
+
+ax0.set_xlabel("number of samples")
+ax1.set_xlabel("number of samples")
+ax0.set_ylabel("number of edges")
+ax1.set_ylabel("runtime")
+for L, sub_df in df.groupby("length"):
+    ax0.plot(sub_df.num_samples, sub_df.num_edges_before, marker=".", label="before")
+    ax0.plot(sub_df.num_samples, sub_df.num_edges_after, marker="o", label="after")
+    ax1.plot(sub_df.num_samples, sub_df.runtime_before, marker=".", label="before")
+    ax1.plot(sub_df.num_samples, sub_df.runtime_after, marker=".", label="after")
+
+ax0.legend()
+
+plt.tight_layout()
+plt.savefig(fname)
+
+
+# combined number of edges & runtime for Tajimas D
+fname = os.path.join(basedir, "ratios.pdf")
+fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(9, 5))
+
+ax0.set_xlabel("number of samples")
+ax1.set_xlabel("number of samples")
+ax0.set_ylabel("number of edges, after/before")
+ax1.set_ylabel("runtime, after/before")
+for L, sub_df in df.groupby("length"):
+    ax0.plot(sub_df.num_samples, sub_df.num_edges_after / sub_df.num_edges_before, marker=".", label="before")
+    ax1.plot(sub_df.num_samples, sub_df.runtime_after / sub_df.runtime_before, marker=".", label="before")
+
+ax0.legend()
+
+plt.tight_layout()
+plt.savefig(fname)
+
+
+# timing plots
+fname = os.path.join(basedir, "timing.pdf")
[…]

### 2024-10-24 benchmarks/run_experiment.py

@@ -1,0 +1,72 @@
+import argparse
+import sys
+import json
+
+import tskit
+import numpy as np
+
+import time
+
+def parse_args():
+    dstring = "Run extend haplotypes, comparing numbers of edges and runtime for Tajima's D.",
+    parser = argparse.ArgumentParser(description=dstring,
+                                     formatter_class=argparse.ArgumentDefaultsHelpFormatter)
+    parser.add_argument('input', type=str,
+                        help="Tree sequence filename")
+    parser.add_argument('output', type=str,
+                        help="Output file base name")
+    parser.add_argument('--num_samples', '-N', type=int, required=True,
+                        help="Number of samples")
+    parser.add_argument('--length', '-L', type=float,
+                        help="Genome length")
+    parser.add_argument('--seed', '-S', type=int, required=True,
+                        help="Random seed")
+    return parser
+
+if __name__ == "__main__":
+    parser = parse_args()
+    args = parser.parse_args(sys.argv[1:])
+
+    num_samples = args.num_samples
+
+    rng = np.random.default_rng(seed=args.seed)
+
+    ts = tskit.load(args.input)
+
+    L = args.length
+    if L is None:
+        L = ts.sequence_length
+
+    if L < ts.sequence_length:
+        left = rng.choice(int(ts.sequence_length - L))
+        ts = ts.keep_intervals([[left, left + L]]).trim()
+
+    samples = rng.choice(list(ts.samples()), size=num_samples, replace=False)
+    ts = ts.simplify(samples)
+    num_edges_before = ts.num_edges
+
+    start_time = time.time()
+    _ = ts.Tajimas_D(mode="branch")
+    runtime_before = time.time() - start_time
+
+    start_time = time.time()
+    ets = ts.extend_haplotypes()
+    num_edges_after = ets.num_edges
+    extend_time = time.time() - start_time
+    print(f"Extending ts done in {extend_time:.2f} seconds, reducing {num_edges_before} to {num_edges_after} edges.")
+
+    start_time = time.time()
+    _ = ets.Tajimas_D(mode="branch")
+    runtime_after = time.time() - start_time
+
+    results = vars(args)
+    results["L"] = L
+    results.update({
+        "num_edges_before": num_edges_before,
+        "num_edges_after": num_edges_after,
+        "extend_time": extend_time,
+        "runtime_before": runtime_before,
+        "runtime_after": runtime_after,
+    })
+    with open(f"{args.output}.json", "w") as f:
+        json.dump(results, f)

### 2024-10-24 benchmarks/one_pop.py
benchmarking figures
constant pop sim as well

@@ -1,0 +1,19 @@
+import stdpopsim
+import tskit
+
+engine = stdpopsim.get_engine("slim")
+
+species = stdpopsim.get_species("CanFam")
+contig = species.get_contig('1', genetic_map="Campbell2016_CanFam3_1")
+
+popsize_changes = [
+        # (time-ago, size)
+        (k, 2**(10-k-1) * 1000) for k in range(10)
+] + [(100, 100), ]
+model = stdpopsim.PiecewiseConstantSize(species.population_size, *popsize_changes)
+
+samples = {"pop_0": 100000}
+
+ts = engine.simulate(model, contig, samples, slim_burn_in=1000)
+
+ts.dump("one_pop.trees")

### 2024-10-24 benchmarks/do_predictions.sh
plots good now

@@ -1,6 +1,7 @@
 #!/bin/bash
 
 OUTDIR="one_pop_results"
+CSVFILE=${OUTDIR}.csv
 if [ ! -e $OUTDIR ]
 then
     mkdir -p $OUTDIR
@@ -21,4 +22,5 @@ else
     echo "${OUTDIR} already exists; doing nothing."
 fi
 
-python plot_results.py $OUTDIR
+python jsons-to-csv.py $OUTDIR $CSVFILE
+python plot_results.py $CSVFILE

### 2024-10-24 benchmarks/plot_results.py

@@ -1,53 +1,67 @@
-import sys
+import sys, os
 import json
 import glob
 import pandas as pd
 import numpy as np
 import matplotlib.pyplot as plt
 import matplotlib
-import sys, os
 
-assert len(sys.argv) == 2, "Usage: python plot_results.py <directory name>"
-basedir = sys.argv[1]
+assert len(sys.argv) == 2, "Usage: python plot_results.py <csv file>"
+infile = sys.argv[1]
 
-jj = []
-for jf in glob.glob(os.path.join(basedir, "*.json")):
-    with open(jf, "r") as f:
-        jj.append(json.load(f))
+df = pd.read_csv(infile)
+basename = ".".join(infile.split(".")[:-1])
 
-df = pd.DataFrame(jj)
+length_vals = np.unique(df.length)
+length_cols = plt.colormaps["viridis"](np.linspace(0, 1, len(length_vals)))
+ns_vals = np.unique(df.num_samples)
+ns_cols = plt.colormaps["viridis"](np.linspace(0, 1, len(ns_vals)))
 
 # combined number of edges & runtime for Tajimas D
-fname = os.path.join(basedir, "absolute_values.pdf")
+fname = f"{basename}_absolute_values.pdf"
 fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(9, 5))
 
-ax0.set_xlabel("number of samples")
-ax1.set_xlabel("number of samples")
+ax0.set_xlabel("sequence length")
+ax1.set_xlabel("sequence length")
 ax0.set_ylabel("number of edges")
-ax1.set_ylabel("runtime")
-for L, sub_df in df.groupby("length"):
-    ax0.plot(sub_df.num_samples, sub_df.num_edges_before, marker=".", label="before")
-    ax0.plot(sub_df.num_samples, sub_df.num_edges_after, marker="o", label="after")
-    ax1.plot(sub_df.num_samples, sub_df.runtime_before, marker=".", label="before")
-    ax1.plot(sub_df.num_samples, sub_df.runtime_after, marker=".", label="after")
+ax1.set_ylabel("runtime, Tajima's D")
+for ns, this_df in df.groupby("num_samples"):
+    sub_df = this_df.sort_values("length")
+    j,  = np.where(ns == ns_vals)
+    ax0.plot(sub_df.length, sub_df.num_edges_before, marker=".",
+             c=ns_cols[j])
[…]

### 2024-10-24 benchmarks/do_experiments.sh
.

@@ -1,5 +1,7 @@
 #!/bin/bash
 
+set -eo pipefail
+
 OUTDIR="one_pop_results"
 CSVFILE=${OUTDIR}.csv
 if [ ! -e $OUTDIR ]
@@ -8,7 +10,7 @@ then
 
     for L in 1e6 5e6 1e7 5e7 1e8
     do
-        for N in 10 50 100 500 1000
+        for N in 100 500 1000 5000 10000
         do
             SEED=$RANDOM$RANDOM
             OUT=$OUTDIR/run_${L}_${N}_${SEED}

### 2024-10-24 benchmarks/run_experiment.py

@@ -31,7 +31,7 @@ if __name__ == "__main__":
 
     rng = np.random.default_rng(seed=args.seed)
 
-    ts = tskit.load(args.input)
+    ts = tskit.load(args.input).trim()
 
     L = args.length
     if L is None:

### 2024-10-24 benchmarks/do_experiments.sh
.

@@ -1,6 +1,13 @@
 #!/bin/bash
 
 set -eo pipefail
+if [ -e "run_py.srun" ]
+then
+    PYTHON="sbatch run_py.srun"
+else
+    PYTHON="python"
+fi
+
 
 OUTDIR="one_pop_results"
 CSVFILE=${OUTDIR}.csv
@@ -15,14 +22,13 @@ then
             SEED=$RANDOM$RANDOM
             OUT=$OUTDIR/run_${L}_${N}_${SEED}
             echo "$OUT  ....."
-            python run_experiment.py --num_samples $N --length $L \
+            $PYTHON run_experiment.py --num_samples $N --length $L \
                 --seed $SEED one_pop.trees $OUT
             SEED=$((SEED + 1))
         done
     done
+
 else
     echo "${OUTDIR} already exists; doing nothing."
 fi
 
-python jsons-to-csv.py $OUTDIR $CSVFILE
-python plot_results.py $CSVFILE

### 2024-10-24 benchmarks/plot_experiments.sh

@@ -1,0 +1,8 @@
+#!/bin/bash
+
+set -eo pipefail
+
+OUTDIR="one_pop_results"
+CSVFILE=${OUTDIR}.csv
+python jsons-to-csv.py $OUTDIR $CSVFILE
+python plot_results.py $CSVFILE

### 2024-10-24 benchmarks/do_experiments.sh
constant pop sim as well

@@ -9,26 +9,28 @@ else
 fi
 
 
-OUTDIR="one_pop_results"
-CSVFILE=${OUTDIR}.csv
-if [ ! -e $OUTDIR ]
-then
-    mkdir -p $OUTDIR
+for OUTBASE in one_pop constant_pop
+do
+    OUTDIR="${OUTBASE}_results"
+    CSVFILE=${OUTDIR}.csv
+    if [ ! -e $OUTDIR ]
+    then
+        mkdir -p $OUTDIR
 
-    for L in 1e6 5e6 1e7 5e7 1e8
-    do
-        for N in 100 500 1000 5000 10000
+        for L in 1e6 5e6 1e7 5e7 1e8
         do
-            SEED=$RANDOM$RANDOM
-            OUT=$OUTDIR/run_${L}_${N}_${SEED}
-            echo "$OUT  ....."
-            $PYTHON run_experiment.py --num_samples $N --length $L \
-                --seed $SEED one_pop.trees $OUT
-            SEED=$((SEED + 1))
+            for N in 100 500 1000 5000 10000
+            do
+                SEED=$RANDOM$RANDOM
+                OUT=$OUTDIR/run_${L}_${N}_${SEED}
+                echo "$OUT  ....."
+                $PYTHON run_experiment.py --num_samples $N --length $L \
+                    --seed $SEED ${OUTBASE}.trees $OUT
+                SEED=$((SEED + 1))
+            done
         done
-    done
 
-else
-    echo "${OUTDIR} already exists; doing nothing."
-fi
-
+    else
+        echo "${OUTDIR} already exists; doing nothing."
+    fi
+done

### 2024-10-24 benchmarks/plot_experiments.sh

@@ -2,7 +2,10 @@
 
 set -eo pipefail
 
-OUTDIR="one_pop_results"
-CSVFILE=${OUTDIR}.csv
-python jsons-to-csv.py $OUTDIR $CSVFILE
-python plot_results.py $CSVFILE
+for OUTBASE in one_pop constant_pop
+do
+    OUTDIR="${OUTBASE}_results"
+    CSVFILE=${OUTDIR}.csv
+    python jsons-to-csv.py $OUTDIR $CSVFILE
+    python plot_results.py $CSVFILE
+done

### 2024-10-26 paper/writeup.tex
edits to dissimilarity
also comment out unnecessary things

@@
 \newcommand{\dis}{\operatorname{dis}}
-\newcommand{\agree}{\operatorname{agree}}
+\newcommand{\similarity}{\operatorname{sim}}
 \newcommand{\comment}[1]{{\color{violet} \it #1}}
@@
 \figref{fig:num_passes} shows empirical results from simulated tree sequences,
-which suggest that \comment{XXX TODO}.
+which suggest that only at most five iterations are needed before the algorithm terminates.
+Indeed, for even larger sequences \ref{tab:edge-counts} shows that 99\% of all changes to an ARG occur after the first iteration with the algorithm completing after four iterations.
+\comment{ XXX Do we want to add more here?
+XXX}
 
@@
 
-\subsection{Discrepancy Function}
+\subsection{Dissimilarity Function}
 % * How to measure agreement that includes haplotypes
@@
 that take into account haplotype identity,
-we define a measure of \emph{discrepancy} to quantify this.
+we define a measure of \emph{dissimilarity} to quantify this.
 The method is implemented in the \texttt{tsdate} package \citep{wohns2022unified}
@@
 Now suppose that instead of comparing two tree sequences with the same set of nodes,
-we wish to compare two tree sequence for which we know the sample nodes are the same
+we wish to compare two tree sequences for which we know the sample nodes are the same
 but are not given an identification between nodes otherwise
@@
 and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
+\comment{Should we include an addendum that we can do this for tree sequences with different break points its just a little more tedious?}
 For a node $n$ and tree $T$
@@
 the total span that node $n$ is present in the marginal trees.
-The \emph{agreement} between $\T_1$ and $\T_2$ is then defined to be
+The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
 \begin{align*}
-    \agree(\T_1, \T_2)
+    \similarity(\T_1, \T_2)
     =
@@
 \end{align*}
-where the maximum is over all mappings nodes in $\T_1$ to nodes in $\T_2$.
+where the maximum is over all mappings of nodes in $\T_1$ to nodes in $\T_2$.
 (Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
@@
 \begin{align*}
-    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
+    \alpha(n_1) = \argmax_{n_2 \in N_2} \frac{1}{1+|t_{n_1}-t_{n_2}|}\max_{n_2\in N_2} m(n_1, n_2) ,
 \end{align*}
@@
 \begin{align*}
-    \agree(\T_1, \T_2)
+    \similarity(\T_1, \T_2)
     =
@@
 If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
-out of those minimizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
+out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
 (and if \emph{this} is not unique, we pick an arbitrary one) --
-however, this potential ambiguity does not affect the definition of $\agree(\T_1, \T_2)$.
-We then define the \emph{discrepancy} of $\T_1$ from $\T_2$ by
+however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
+We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
 \begin{align*}
@@
     =
-    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) ,
+    \frac{\sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n)))}{\sum_{n\in N_1} s(\T_1,n)} ,
 \end{align*}
-which is the total span of all nodes in $\T_1$
+which is the proportion of total span of all nodes in $\T_1$
 for which their descendant samples do \emph{not} match those of their best match in $\T_2$.
@@
 \end{align*}
-\comment{Check this!}
+\comment{Check this!}\comment{Not true. I believe you can have the problem again of two nodes in TS1 match with node in TS2}
 In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
 across all trees, but average Robinson-Foulds distance allows a different node to match
-on each tree.
-The other differences are that \emph{average} Rbonson-Foulds distance
+on each tree. \comment{I don't think this is true nor do I understand how $m'$ is truely different than $m$. It seems like $m'$ is just the extension of $m$ to the level of tree sequences.}
+The other differences are that \emph{average} Robinson-Foulds distance
 normalizes by sequence length, and is symmetrized.
@@
 
-The discrepancy $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
+The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
 but not times.
@@
 
+We compute the dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.
+
 %%%%%%%%%%%%
-\comment{PREVIOUS VERSION: retrieve bits from below?}
-
-Inspired by Robinson-Foulds we define a function, called a \textit{discrepancy function} (see \algorithmref{alg:disc}), on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
-Let $N_{T_1}$ and $N_{T_2}$ be the set of nodes on $\T_1$ respectively $\T_2$.
-For a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
-Each node $n\in\T$ has a given position along the genome and time, 
-so we can write each node as a tuple $n = (t_n, L_n)$ with position $L_n\subset \left[0,L\right)$ and time $t_n\in \R^{\geq 0}$.
-Computing the discrepancy between a random pair of nodes is not helpful, 
-so we must first identify which pairs of nodes would be most identical to one another or a `best match'.
-These `best matching' nodes should be close in terms of time and have equal descendant sample sets,
-ie. $s(i) = s(j)$ where $s$ is the function to compute the descendant sample set of $n_1$.
-
-To find a best match for each node $i\in N_{T_1}$ \comment{(Should I do this by mentioning clades?)}
-we compute the \textit{shared span} between node $i\in N_{T_1}$ and every node in $N_{T_2}$.
-Then, the \textit{shared span} between two tree sequences, denoted $ss(\T_1,\T_2)$, is a $|N_{T_1}|\times |N_{T_2}|$ matrix $A$
-where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$.
-If a pair $(i,j)$ has equal descendant sample sets, $s(i) = s(j)$
-then $a_{ij} = |L_i \cap L_j|$.
-If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
-Matrix $A$ determines the \textit{best match} for every node $i\in N_{T_1}$
-via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_{T_2}} a_{ij}$.
-If this maximum is not unique, we use time difference to determine the best match pair.
-We then define the \textit{discrepancy} between $\T_1$ and $\T_2$ as 
-one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
-
-We note that the discrepancy function is a metric for detecting non-similarity between two tree sequences.
+%\comment{PREVIOUS VERSION: retrieve bits from below?}
+%
+%Inspired by Robinson-Foulds we define a function, called a \textit{discrepancy function} (see \algorithmref{alg:disc}), on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
+%Let $N_{T_1}$ and $N_{T_2}$ be the set of nodes on $\T_1$ respectively $\T_2$.
+%For a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
+%Each node $n\in\T$ has a given position along the genome and time, 
+%so we can write each node as a tuple $n = (t_n, L_n)$ with position $L_n\subset \left[0,L\right)$ and time $t_n\in \R^{\geq 0}$.
+%Computing the discrepancy between a random pair of nodes is not helpful, 
+%so we must first identify which pairs of nodes would be most identical to one another or a `best match'.
+%These `best matching' nodes should be close in terms of time and have equal descendant sample sets,
+%ie. $s(i) = s(j)$ where $s$ is the function to compute the descendant sample set of $n_1$.
+%
+%To find a best match for each node $i\in N_{T_1}$ \comment{(Should I do this by mentioning clades?)}
+%we compute the \textit{shared span} between node $i\in N_{T_1}$ and every node in $N_{T_2}$.
+%Then, the \textit{shared span} between two tree sequences, denoted $ss(\T_1,\T_2)$, is a $|N_{T_1}|\times |N_{T_2}|$ matrix $A$
+%where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$.
+%If a pair $(i,j)$ has equal descendant sample sets, $s(i) = s(j)$
+%then $a_{ij} = |L_i \cap L_j|$.
+%If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
+%Matrix $A$ determines the \textit{best match} for every node $i\in N_{T_1}$
+%via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_{T_2}} a_{ij}$.
+%If this maximum is not unique, we use time difference to determine the best match pair.
+%We then define the \textit{discrepancy} between $\T_1$ and $\T_2$ as 
+%one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
+%
+%We note that the discrepancy function is a metric for detecting non-similarity between two tree sequences.
 However, we note that this function is not a metric in the mathematical sense, as it is not symmetric, nor is it reflexive.
@@
 We may just do it in this section instead of the appendix?}
-
-\comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}
+% 
+%\comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}
 
@@
     \caption{
-        For two tree sequences $T1$ and $T2$ the \textit{discrepancy function}, $d(T1,T2)$, matches nodes in $T1$ with nodes in $T2$
-        based on identical sample sets. We then measure each pair of nodes' \textit{shared span} across the entire tree sequence.
-        In this example we have nodes $0$ through $7$ in $T1$ match with their identical counterpart. 
-        Node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
-        \comment{make some reference to Discrepancy function section or appendix for more in depth calculation.}
+        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity function}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
+        based on identical sample sets.
+        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
+        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. On the rest of the genome, all of nodes match with their identical counterpart.
+        This makes the dissimilarity $\dis(T1,T2)=\frac{3}{46}$.
         \label{fig:conceptual_discrepancy}
@@
 
-In the simple example in \figref{fig:extending_diagram},
+In the simple example in \figref{fig:conceptual},
 extending haplotypes replaces three edges
@@
 then this is achieved for long sequences.
-It is possible for \algorithmref{alg:extend} to add edges,
+It is possible for \algorithmref{alg:extend} to add edges, as in \figref{extending_diagram},
 but we still expect the number of edges to decrease by more than 1/3.
@@
 		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
-		\caption{Left: Discrepancy \newline Right: True Proportion }
+		\caption{Left: Dissimilarity \newline Right: True Proportion }
 	\end{center}
@@
 		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
-		\caption{Left: Discrepancy \newline Right: True Proportion}
+		\caption{Left: Dissimilarity \newline Right: True Proportion}
 	\end{center}
@@
         Accuracy of \algorithmref{alg:extend}
-        on simulated ARGs with \comment{TODO, including gene conversion}.
+        on simulated ARGs with sample size $10^2$, length $10^7$, gene conversion rate $10^{-8}$ and gene tract length = $10^3$.
         The same quantities for simulated ARGs without gene conversion
@@
         as in \figref{fig:accuracy_on_truth} except that simulations were done
-        without gene conversion.
+        without gene conversion. \comment{Add to supplement.}
         \label{fig:accuracy_on_truth_no_gc}
@@
 \comment{collecting random notes here; TODO is tie them together}
-The discrepancy $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
+The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
 (i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
@@
 where $N$ is the shared set of nodes.
-Then, discrepancy is the Lesbegue measure of the relative difference of the two sets:
+Then, dissimilarity is the Lesbegue measure of the relative difference of the two sets:
 $|\T_1 \setminus \T_2|$,
@@
 combines this distance across marginal trees, weighted by span along the genome.
-The method we present here for measuring discrepancy between topologies of ARGs
+The method we present here for measuring dissimilarity between topologies of ARGs
 is a straightforward generalization
@@
 relax the requirement that the match between subtended sample sets be exact,
-and weight matches in some way by the size of the discrepancy.
+and weight matches in some way by the size of the dissimilarity.
 We considered such definitions as well, but kept to the simple case
@@
 without enforcement of identity across haplotypes;
-it would be useful to extend our discrepancy to include branch lengths.
+it would be useful to extend our dissimilarity to include branch lengths.
 
@@
 
-\begin{figure}
-    \caption{
-        Supplementary figure: runtime of discrepancy function as a function of number of trees
-        (which we modulate by changing sequence length).
-        \label{fig:speed_discrepancy}
-    }
-\end{figure}
+%\begin{figure}
+%    \caption{
+%        Supplementary figure: runtime of discrepancy function as a function of number of trees
+%        (which we modulate by changing sequence length).
+%        \label{fig:speed_discrepancy}
+%    }
+%\end{figure}
 
@@
 
-\begin{tabular}[c|c|c|c|c|]
+\begin{center}
+\begin{table}[hbt]
+\begin{tabular}{|c|c|c|c|c|}
+	\hline
 	Initial & Iteration 1 & Iteration 2 & Iteration 3 & Iteration 4 \\
+	\hline
+	\hline
 	113463	&	76301	&	76059	&	76057	&	76057 \\
+	\hline
 	113266 & 76084	&	75835	&	75833	&	75833 \\
+	\hline
 	114489 & 76703 & 76436 & 76434 & 76434 \\
-	114086	& 76550	& 76294	& 76291	& 76291
+	\hline
+	114086	& 76550	& 76294	& 76291	& 76291\\
+	\hline
 \end{tabular}
+\caption{Number of edges for each iteration of \textit{extend haplotypes} applied to 
+simulated tree sequences with sample size $10^4$ and length $5\times 10^7$. 
+Each of the simulations terminated by the fourth iteration, however 99\% of the edges
+are removed within the first iteration.}
+\label{tab:edge-counts}
+\end{table}
+\end{center}
 

### 2024-10-28 paper/writeup.tex
.

@@
 \newcommand{\tskit}{\texttt{tskit}}
+\newcommand{\tsinfer}{\texttt{tsinfer}}
 \newcommand{\msprime}{\texttt{msprime}}
@@
 
-\comment{
-    Idea: look at statistical properties of unary regions:
-    look at how much 'unary regions' there is in simulation
-    and how much we can put back in,
-    as a function of time.
-}
+% Outline:
+% 1. benchmarking: edge reduction and runtime
+% 2. correctness:
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Tree sequence compression and computation}
@@
 
-One might expect that a reduction in edges would also lead to a reduction in computation time
-(for instance, of the class of statistics described in \citet{ralph2020efficiently}.
-This turns out not to be the case, as shown in \figref{fig:num_edges}.
-The reason for this is clear if we
-return to the algorithms in \citet{ralph2020efficiently}:
+This reduction in edges can also lead to a reduction in computation time.
+Indeed, \figref{fig:num_edges} shows that computation time is reduced by 10--20\%
+for a typical statistic (here, Tajima's $D$),
+computed in an efficient iterative manner along the genome as implemented in \tskit.
+This is actually somewhat surprising:
+as described in \citet{ralph2020efficiently},
 each edge added or removed requires an update
@@
 So, even if an edge extends for a long segment of genome,
-something has to happen to it every time the subtree below it changes.
-Extending haplotypes does not change at all the locations in the tree affected by subtree changes.
-\comment{If the increase in computation time holds up:}
-In fact, introducing unary nodes to branches
-\emph{increases} the distance from a typical node up to the root,
-and so increases the work required (and also computation time, as seen in \figref{fig:num_edges}).
-
-\subsection{Reconstructing haplotypes of true trees}
-
-Our first task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
-are indeed correct.
-To do this, we simulate ARGs containing full haplotypes using \msprime
-as described in the Methods.
+something has to happen to it every time the subtree below it changes,
+which is not affected by extending haplotypes.
+
+%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Accuracy with true trees}
+
+Our next task is to confirm that the haplotypes extended by \algorithmref{alg:extend}
+are indeed correct -- i.e., that in addition to compression, we are also gaining information.
+To do this, we simulate ARGs containing full haplotypes using \msprime{}
+(as described in the Methods).
 
@@
 
-\subsection{Extending haplotypes in inferred trees}
+%%%%%%%%%%%%%%%%%%
+\subsection{Inferred ARGs}
+
+So far, we have demonstrated that there is potentially ample information
+in the coalescent-only trees to extend haplotypes.
+Does this work with \emph{inferred} ARGs?
+Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes;
+to what degree are they taking advantage of this information already?
+Here, we study the behavior of \tsinfer{},
+leaving a comparison of different ARG inference methods for future work.
+To study this, we simulated ARGs containing unary-spanning haplotypes (as above),
+then performed various operations to them,
+reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:discrepancy}.
+
+The first thing we see is that coalescent nodes are unary
+over a substantial portion of their spans:
+since the \texttt{simplify} operations removes these unary spans,
+the ``simplified'' line on the right of Figure~\ref{fig:discrepancy} shows
+the proportion of nodes' spans on which they are not unary.
+For instance, on an ARG of 10 samples on a sequence of length $5 \times 10^7$,
+only about 15\% of the total spans are retained after simplification;
+this rises to around 50\% for 1,000 samples (top right plot)
 
@@
 	\label{fig:correct-unary-span-log}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
-	\caption{\comment{DELETE}}
-%	\caption{Correct Unary Span for Extend Haplotypes}
-	\label{fig:correct-unary-span}
 \end{figure}
@@
 	\begin{center}
-		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
-		\caption{Left: Dissimilarity \newline Right: True Proportion }
+		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
+		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
 	\end{center}
-	\label{fig:discrepancy-over-sample}
-\end{figure}
-
-\begin{figure}
-	\begin{center}
-		\includegraphics[width=0.9\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
-		\caption{Left: Dissimilarity \newline Right: True Proportion}
-	\end{center}
-	\label{fig:discrepancy-over-seqlen}
+    \caption{
+        Accuracy and sensitivity of extended haplotypes
+        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
+        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
+        For each, a simulated ARG containing unary haplotype spans
+        was (i) \emph{simplified}, removing the unary spans,
+        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
+        then each of these had its haplotypes \emph{extended}.
+        The inferred, then simplified, ARG is also shown.
+        \textbf{Left:} Discrepancy to the true ARG,
+        as proportion of haplotypes that are not represented in the true ARG
+        (equation~\eqref{eq:discrepancy}).
+        \textbf{Right:} Agreement to the true ARG,
+        as proportion of the true ARG that is represented.
+        The largest number of trees in the bottom row corresponds
+        to the sequence length in the top row;
+        see text for more details.
+    }
+    \label{fig:discrepancy}
 \end{figure}
@@
 
+\paragraph{Inference methods}
+\comment{Possibly an overview of what the different inference methods do with unary spans?}
+
 \bibliography{references}
@@
 
+\begin{table}[hbt]
 \begin{center}
-\begin{table}[hbt]
 \begin{tabular}{|c|c|c|c|c|}
@@
 \label{tab:edge-counts}
+\end{center}
 \end{table}
-\end{center}
 
@@
 
+\begin{figure}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
+	\caption{\comment{DELETE}}
+%	\caption{Correct Unary Span for Extend Haplotypes}
+	\label{fig:correct-unary-span}
+\end{figure}
+
 \end{document}

### 2024-11-01 paper/writeup.tex

@@
 % NOTE: these are in alphabetical order for now; we can revisit that ordering
+\author[$\dagger$]{Halley Fritze}
+\author[$\dagger$]{Nathaniel Pope}
 \author[$\dagger$]{Ava Bamforth}
-\author[$\dagger$]{Halley Fritze}
 \author[$\ddagger$]{Jerome Kelleher}
-\author[$\dagger$]{Nathaniel Pope}
 \author[$\ast$,$\dagger$,S]{Peter Ralph}
@@
 
+\comment{TODO: put in benchmarking figures}
+
 In the simple example in \figref{fig:conceptual},
@@
 then this is achieved for long sequences.
-It is possible for \algorithmref{alg:extend} to add edges, as in \figref{extending_diagram},
+It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
 but we still expect the number of edges to decrease by more than 1/3.
@@
 are indeed correct -- i.e., that in addition to compression, we are also gaining information.
-To do this, we simulate ARGs containing full haplotypes using \msprime{}
-(as described in the Methods).
-
-When we apply our algorithm to the simplified given tree sequence, the ratio of unary span bordering coalescing between the extend tree sequence and the original is 0.4917.
-This means that \comment{something something something...}
-See \figref{fig:correct-unary-span-log}, \figref{fig:correct-unary-span}, \figref{fig:node-spans}.
+To do this, 
+we simulate ARGs containing full haplotypes using \msprime{},
+apply the simplification algorithm to reduce these so that there are no unary nodes
+(i.e., any node present in a marginal tree is a coalescent node or a sample),
+and then apply \algorithmref{alg:extend} to the result
+(see Methods for more detail).
+The method can potentially extend the spans of each node
+(spans on which the node will be unary);
+and we can quantify how much of these extended spans were in the original ARG
+(and thus correctly extended).
+
+\comment{TODO: figure out if we can remove this from figures. If so, move this to Methods.}
+Some stuff can't be inferred since it's "isolated unary";
+for instance with 1,000 samples, 5e7 that's about half the ARG.
+Etcetera.
 
@@
 
-%%%%%%%%%%%%%%%%%%
-\subsection{Inferred ARGs}
-
-So far, we have demonstrated that there is potentially ample information
-in the coalescent-only trees to extend haplotypes.
-Does this work with \emph{inferred} ARGs?
-Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes;
-to what degree are they taking advantage of this information already?
-Here, we study the behavior of \tsinfer{},
-leaving a comparison of different ARG inference methods for future work.
-To study this, we simulated ARGs containing unary-spanning haplotypes (as above),
-then performed various operations to them,
-reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:discrepancy}.
-
+As seen in \figref{fig:node-spans},
+Look, we add lots of stuff
+and most most of it is right;
+this applies to both recent and ancient nodes;
+we're typically extending the span of a node by like maybe 50\%.
+
+\begin{figure}[bht]
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
+	\caption{
+        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
+        \comment{TODO: nodes should be in the same order in all three lines,
+        using the ordering from the true tree sequence.}
+        \comment{TODO: somehow communicate per-node amount/proportion of correctly and incorrectly added span}
+    }
+	\label{fig:node-spans}
+\end{figure}
+
+\comment{merge this in}
 The first thing we see is that coalescent nodes are unary
@@
 
-\begin{figure}[bht]
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
-	\caption{
-        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
-        \comment{TODO: nodes should be in the same order in all three lines,
-        using the ordering from the true tree sequence.}
-        \comment{TODO: add plot with percent wrongly added by node?}
-    }
-	\label{fig:node-spans}
-\end{figure}
-
-\begin{figure}[bht]
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
-	\caption{Correct Unary Span for Extend Haplotypes (log)}
-	\label{fig:correct-unary-span-log}
-\end{figure}
-
-\begin{figure}
-%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
-	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
-	\label{fig:edges-per-iteration}
-\end{figure}
-
-% (AVA)
-% * reduction in edges ~ sequence length
-% * speed increase ~ sequence length
-% * proprotion of added edges that are true ~ sequeence length
-
-%\begin{figure}
-%	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
-%    \caption{
-%        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
-%        as a function of number of trees (modulated by sequence length),
-%        on reasonably big simulated sequences.
-%        For "speed": plot ratio of runtime before/after extending.
-%        \label{fig:speed_and_edges}
-%    }
-%\end{figure}
-
-%Simple experiment:
-%(1) simulate, simplify, and extend;
-%(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
-%then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
-%
-%With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
-%(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
+We can also use measures of dissimiliarity and similarity to quantify summaries across nodes:
+these are shown in \figref{} (purple lines only)
+Points:
+proportion left after smiplify varies a LOT: from 15\% to 95\%.
+    (goes up with L: with longer sequence length, greater opportunity for distant chunks to be on the same lineage)
+extending adds hardly any wrong stuff (left plots); and adds a substantial chunk: with large samples, adds like half of the missing stuff
+
+It's also interesting to  note that inference
+(see \figref{fig:inferred_edge_counts})
+(a) increases number of edges;
+(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
+(c) but then extending reduces to below the original.
 
@@
 	\begin{center}
-		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
-		\includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
+		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
+		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
+        \includegraphics[width=\textwidth]{dissimilarity.pdf}
 	\end{center}
@@
         see text for more details.
+        \comment{TODO: make "inferred simplified" green (dotted) not red}
+        \comment{TODO: label "inferred extended" "inferred simplfieid extended"??}
+        \comment{TODO: also add a "inferred extended" line}
     }
@@
 \begin{figure}
-    \includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_reduction_prop_vs_SL}
-    \includegraphics{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_SL}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
     \caption{
-        \textbf{(A)} Proportion reduction in number of edges
-        achieved by extending haplotypes (\algorithmref{alg:extend})
-        on tree sequences of various sequence lengths and numbers of samples.
-        See Supplementary \figref{sfig:num_edges} for the raw number of edges.
-        \textbf{(B)} Runtime for computation of Tajima's $D$
-        (using the \tskit method) on the same tree sequences as in (A).
-        \comment{TODO: remove shortest lengths and double-check runtime.}
-        \label{fig:num_edges}
+        number of edges, 5e7, 1,000 samples
+        \label{fig:inferred_edge_counts}
     }
@@
 
+%%%%%%%%%%%%%%%%%%
+\subsection{Inferred ARGs}
+
+So far, we have demonstrated that there is potentially ample information
+in the coalescent-only trees to extend haplotypes.
+Does this work with \emph{inferred} ARGs?
+Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes;
+to what degree are they taking advantage of this information already?
+Here, we study the behavior of \tsinfer{},
+leaving a comparison of different ARG inference methods for future work.
+To study this, we simulated ARGs containing unary-spanning haplotypes (as above),
+then performed various operations to them,
+reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:discrepancy}.
+
+Main points: in \figref{fig:discrepancy}
+(green line) tsinfer gets a lot right (above 60\% with 1000 samples; bottom-right plot)
+    and does a lot better with more samples;
+    also, gets 10\%--25\% wrong (left panels)
+    with longer sequence, larger proportion is wrong (bottom-left panel)
+(red dotted)
+    simplifying removes a lot of the wrong stuff
+    but also a lot of the right stuff
+(red solid)
+    similarly extending the simplified one adds a bunch of right and also wrong stuff
+
+\begin{figure}[bht]
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
+	\caption{Correct Unary Span for Extend Haplotypes (log)}
+	\label{fig:correct-unary-span-log}
+\end{figure}
+
 \begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
+%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
+	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
+	\label{fig:edges-per-iteration}
 \end{figure}
+
+% (AVA)
+% * reduction in edges ~ sequence length
+% * speed increase ~ sequence length
+% * proprotion of added edges that are true ~ sequeence length
+
+%\begin{figure}
+%	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
+%    \caption{
+%        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
+%        as a function of number of trees (modulated by sequence length),
+%        on reasonably big simulated sequences.
+%        For "speed": plot ratio of runtime before/after extending.
+%        \label{fig:speed_and_edges}
+%    }
+%\end{figure}
+
+%Simple experiment:
+%(1) simulate, simplify, and extend;
+%(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
+%then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
+%
+%With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
+%(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
 
@@
 
-\begin{figure}
-	\includegraphics{dissimilarity.pdf}
-\end{figure}
-
 %\begin{figure}
@@
 %\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/num_trees_vs_Num_edges_subplot.pdf}
-    \caption{
-        Raw numbers of edges before and after extending haplotypes
-        for the tree sequences shown in \figref{fig:num_edges}
-        \label{sfig:num_edges}
-    }
-\end{figure}
 

### 2024-11-02 remove_isolated_unary.py
scriptified it

@@ -1,6 +1,14 @@
+#!/usr/bin/env python
+
+import sys
 import tskit, msprime
 import numpy as np
 
+usage = f"""
+Usage:
+    python {sys.argv[0]} (input .trees filename) (output .trees filename)
+
+"""
 
 def get_intervals(ts):
     last_start = np.full(ts.num_nodes, 0.0)
@@ -130,3 +138,13 @@ if False: # for testing
             for a in ts.samples():
                 for b in ts.samples():
                     assert t1.mrca(a, b) == t2.mrca(a, b), f"{a}, {b}: {t1.tmrca(a, b)} != {t2.tmrca(a, b)}"
+
+if __name__ == "__main__":
+    if len(sys.argv) != 3:
+        print(usage)
+        exit(0)
+    infile = sys.argv[1]
+    outfile = sys.argv[2]
+    ts = tskit.load(infile)
+    rts = remove_isolated_unary(ts)
+    rts.dump(outfile)

### 2024-11-02 remove_isolated_unary.py
bugfix!

@@ -81,13 +81,14 @@ def get_node_map(ts, remove_intervals):
     for n in to_remap:
         if parent[n] != tskit.NULL and last_start[n] < pos:
             # parent changed!
-            node_map[n].append(([float(last_start[n]), pos], int(u)))
+            node_map[n].append(([float(last_start[n]), pos], int(parent[n])))
     return node_map
 
-def check_node_map(node_map):
+def check_node_map(ts, node_map):
     for n in node_map:
         for (a, b), x in node_map[n]:
             assert a < b, f"{n}: ({a}, {b}) -> {x}"
+            assert ts.nodes_time[x] > ts.nodes_time[n], f"{n} (t={ts.nodes_time[n]}): ({a}, {b}) -> {x} (t={ts.nodes_time[x]})"
 
 def overlaps(a, b):
     return b[1] > a[0] and a[1] > b[0]
@@ -108,7 +109,7 @@ def remove_isolated_unary(ts, debug=False):
     remove_intervals = get_intervals(ts)
     node_map = get_node_map(ts, remove_intervals)
     if debug:
-        check_node_map(node_map)
+        check_node_map(ts, node_map)
     tables = ts.dump_tables()
     edges = tables.edges
     edges.clear()
@@ -146,5 +147,5 @@ if __name__ == "__main__":
     infile = sys.argv[1]
     outfile = sys.argv[2]
     ts = tskit.load(infile)
-    rts = remove_isolated_unary(ts)
+    rts = remove_isolated_unary(ts, debug=True)
     rts.dump(outfile)

### 2024-11-02 remove_isolated_unary.py
another bugfix

@@ -119,7 +119,8 @@ def remove_isolated_unary(ts, debug=False):
             for (left, right), new_parent in overlaps_map((e.left, e.right), node_map[e.parent]):
                 remapped = True
                 assert max(e.left, left) < min(e.right, right), f"max({e.left}, {left}) < min({e.right}, {right}) "
-                edges.add_row(parent=new_parent, child=e.child, left=max(e.left, left), right=min(e.right, right))
+                if new_parent != tskit.NULL:
+                    edges.add_row(parent=new_parent, child=e.child, left=max(e.left, left), right=min(e.right, right))
             if not remapped:
                 edges.append(e)
     tables.sort()
@@ -147,5 +148,5 @@ if __name__ == "__main__":
     infile = sys.argv[1]
     outfile = sys.argv[2]
     ts = tskit.load(infile)
-    rts = remove_isolated_unary(ts, debug=True)
+    rts = remove_isolated_unary(ts)
     rts.dump(outfile)

### 2024-11-07 COPYevaluation.py
new figures, with data, etc.

@@ -336,6 +336,7 @@ def tree_discrepancy(ts, other):
     ts_times = ts.nodes_time[row_ind[match]]
     other_times = other.nodes_time[col_ind[match]]
     time_difference = np.absolute(np.asarray(ts_times - other_times))
+    # log_time_difference = np.log(np.asarray(ts_times)) - np.log(np.asarray(other_times))
     # If a node x in `ts` has no match then we set time_difference to zero
     # This node then does not effect the rmse
     for j in range(len(shared_spans.data[match])):

### 2024-11-08 paper/writeup.tex
added new shiny figures

@@
 		Figure~\ref*{#2}%
+		\ifx\\#1\\%
+		\else
+		\,#1%
+		\fi
+	}%
+}
+
+\newcommand*{\tabref}[2][]{%
+	\hyperref[{#2}]{%
+		Table~\ref*{#2}%
 		\ifx\\#1\\%
@@
 
-\begin{figure}[!ht]
-	\begin{center}
-		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-Ex2.pdf}
-		\includegraphics[width=0.4\linewidth, height=1.2in]{EH-extended-Ex2.pdf}
-	\end{center}
-	\caption{\comment{DELETE}}
-%	\caption{\comment{Change figure to have the same arrow with 'Extend Haplotypes' in inkscape for unity's sake.}The extend haplotypes method between tree with more unary nodes.
-Given two neighboring trees, there exists paths $4\to 6\to8\to10$ and $5\to 9\to 10$ in $F_O$, and in $F_I$ we have paths $4\to 7\to 11\to 10$ and $5\to 12\to 10$.
-Both the out-forest and in-forest both contain paths with the same initial and terminal nodes, $4\to 10$ and $5\to 10$ respectively, we merge both paths and output the tree sequence on the right.}
-	\label{fig:complicated_extending_diagram}
-\end{figure}
-
 A relatively efficient algorithm to do this is described in \algorithmref{alg:extend}, 
@@
 in each direction,
-an important practical quesiton for this algorithm is:
+an important practical question for this algorithm is:
 how passes do we need to do?
@@
 so it is guaranteed to require only a finite number of passes,
-but it is also not hard to construct pathalogical cases that require an arbitrary number of passes.
-\figref{fig:num_passes} shows empirical results from simulated tree sequences,
-which suggest that only at most five iterations are needed before the algorithm terminates.
-Indeed, for even larger sequences \ref{tab:edge-counts} shows that 99\% of all changes to an ARG occur after the first iteration with the algorithm completing after four iterations.
-\comment{ XXX Do we want to add more here?
-XXX}
-
-\begin{figure}[!ht]
-    \caption{
-        Number of edges in the tree sequence (vertical axis)
-        after each pass of the algorithm (number of passes on the horizontal axis),
-        for tree sequences with different numbers of samples (different lines).
-        \comment{TODO}
-        \label{fig:num_passes}
-    }
-\end{figure}
-
+but it is also not hard to construct pathological cases that require an arbitrary number of passes.
+Counting edges for every iteration suggest that only at most five iterations are needed before the algorithm terminates.
+Indeed, for even larger sequences \tabref{tab:edge-counts} shows that 99\% of all changes to an ARG occur after the first iteration with the algorithm completing after four iterations.
 %% Find counter example to global minimum.
@@
 % * Show: how add edges reduces discrepancy (RESULTS)
-
-\comment{Rename Dissimilarity and Similarity}
 
@@
 
-Concretely, suppose that the two tree sequences have the same set of breakpoints
+Concretely, our method can be applied to any two tree sequences,
+ as in the case of \figref{fig:conceptual_discrepancy},
+ where we consider the pair of trees which exist at that break point.
+However, suppose that the two tree sequences have the same set of breakpoints
 between trees,
@@
 and $T_1^{(2)}, \ldots, T_N^{(2)}$ are the trees in $\T_2$.
-\comment{Should we include an addendum that we can do this for tree sequences with different break points its just a little more tedious?}
 For a node $n$ and tree $T$
@@
 \begin{align*}
-    \alpha(n_1) = \argmax_{n_2 \in N_2} \frac{1}{1+|t_{n_1}-t_{n_2}|}\max_{n_2\in N_2} m(n_1, n_2) ,
+    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
 \end{align*}
@@
     =
-    \frac{\sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n)))}{\sum_{n\in N_1} s(\T_1,n)} ,
+    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))),
 \end{align*}
@@
 \end{align*}
-\comment{Check this!}\comment{Not true. I believe you can have the problem again of two nodes in TS1 match with node in TS2}
 In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
 across all trees, but average Robinson-Foulds distance allows a different node to match
-on each tree. \comment{I don't think this is true nor do I understand how $m'$ is truely different than $m$. It seems like $m'$ is just the extension of $m$ to the level of tree sequences.}
+on each tree.
 The other differences are that \emph{average} Robinson-Foulds distance
@@
     = \sqrt{\frac{
-        \sum_{n \in N_1} s(\T_1,n) \left(t^{(1)}_n - t^{(2)}_{\alpha(n)} \right)^2 
+        \sum_{n \in N_1} s(\T_1,n) \left(\log t^{(1)}_n - \log t^{(2)}_{\alpha(n)} \right)^2 
     }{
@@
 is more serious for a node with a longer span.
-\comment{TODO: change the definition?
-In practice, time estimation errors seem to be closer to homoskedastic on a log scale,
-so this would probably be more informative if errors are computed on the log sale.}
 
@@
 \begin{figure}[bht]
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/node_spans_wo_trapped_unary_regions_CDF.pdf}
+	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/figure-4-2.pdf}
 	\caption{
@@
 	\label{fig:node-spans}
+\end{figure}
+
+\begin{figure}
+	\includegraphics{newplots_wo_ee/figure-4.pdf}
 \end{figure}
@@
 		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
-        \includegraphics[width=\textwidth]{dissimilarity.pdf}
+        \includegraphics[width=\textwidth]{newplots_wo_ee/dissimilarity.pdf}
 	\end{center}
@@
 \begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
-    \caption{
-        number of edges, 5e7, 1,000 samples
-        \label{fig:inferred_edge_counts}
-    }
+	\includegraphics{newplots_wo_ee/figure5.pdf}
 \end{figure}
+
+%\begin{figure}
+%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
+%    \caption{
+%        number of edges, 5e7, 1,000 samples
+%        \label{fig:inferred_edge_counts}
+%    }
+%\end{figure}
 
@@
 
-\begin{figure}[bht]
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
-	\caption{Correct Unary Span for Extend Haplotypes (log)}
-	\label{fig:correct-unary-span-log}
-\end{figure}
-
-\begin{figure}
-%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
-	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
-	\label{fig:edges-per-iteration}
-\end{figure}
+%
+%\begin{figure}[bht]
+%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
+%	\caption{Correct Unary Span for Extend Haplotypes (log)}
+%	\label{fig:correct-unary-span-log}
+%\end{figure}
+%
+%\begin{figure}
+%%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
+%	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
+%	\label{fig:edges-per-iteration}
+%\end{figure}
 
@@
 
-\begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
-	\caption{
-        Accuracy of \algorithmref{alg:extend}
-        on simulated ARGs with sample size $10^2$, length $10^7$, gene conversion rate $10^{-8}$ and gene tract length = $10^3$.
-        The same quantities for simulated ARGs without gene conversion
-        are shown in Supplementary \figref{fig:accuracy_on_truth_no_gc}.
-    \label{fig:accuracy_on_truth}
-}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
-	\caption{Accuracy of \algorithmref{alg:extend}
-        as in \figref{fig:accuracy_on_truth} except that simulations were done
-        without gene conversion. \comment{Add to supplement.}
-        \label{fig:accuracy_on_truth_no_gc}
-    }
-\end{figure}
+%\begin{figure}
+%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
+%	\caption{
+%        Accuracy of \algorithmref{alg:extend}
+%        on simulated ARGs with sample size $10^2$, length $10^7$, gene conversion rate $10^{-8}$ and gene tract length = $10^3$.
+%        The same quantities for simulated ARGs without gene conversion
+%        are shown in Supplementary \figref{fig:accuracy_on_truth_no_gc}.
+%    \label{fig:accuracy_on_truth}
+%}
+%\end{figure}
+%
+%\begin{figure}
+%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
+%	\caption{Accuracy of \algorithmref{alg:extend}
+%        as in \figref{fig:accuracy_on_truth} except that simulations were done
+%        without gene conversion. \comment{Add to supplement.}
+%        \label{fig:accuracy_on_truth_no_gc}
+%    }
+%\end{figure}
 
@@
 
-\begin{figure}
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_notlog.pdf}
-	\caption{\comment{DELETE}}
-%	\caption{Correct Unary Span for Extend Haplotypes}
-	\label{fig:correct-unary-span}
-\end{figure}
-
 \end{document}

### 2024-11-08 benchmarks/plot_experiments.sh
color fixes

@@ -6,6 +6,11 @@ for OUTBASE in one_pop constant_pop
 do
     OUTDIR="${OUTBASE}_results"
     CSVFILE=${OUTDIR}.csv
-    python jsons-to-csv.py $OUTDIR $CSVFILE
+    if [ ! -e $CSVFILE ]
+    then
+        python jsons-to-csv.py $OUTDIR $CSVFILE
+    else
+        echo "$CSVFILE already exists, not regenerating."
+    fi
     python plot_results.py $CSVFILE
 done

### 2024-11-08 benchmarks/plot_results.py

@@ -24,11 +24,25 @@ sdf = df.groupby(["length", "num_samples"]).agg({
 }).astype({'length': int, 'num_samples': int})
 sdf.reset_index(drop=True, inplace=True)
 
+# from Halley
+colors = {'blue': (46,37,133),
+          'red': (194,106,119),
+          'lgreen': (93,168,153),
+          'gold': (220,205,125),
+          'green': (51, 117,56),
+          'lblue': (148,203,236),
+          'magenta': (159,74,150),
+          'wine': (126,41,84),
+         }
+for x in colors:
+    colors[x] = [u/256 for u in colors[x]] + [1.0]
 
 length_vals = np.unique(df.length)
-length_cols = plt.colormaps["viridis"](np.linspace(0, 1, len(length_vals)))
+# length_cols = plt.colormaps["viridis"](np.linspace(0, 1, len(length_vals)))
+length_cols = np.array([colors[n] for n in ['blue', 'green', 'lblue', 'lgreen', 'gold']])
 ns_vals = np.unique(df.num_samples)
-ns_cols = plt.colormaps["plasma"](np.linspace(0, 1, len(ns_vals)))
+# ns_cols = plt.colormaps["plasma"](np.linspace(0, 1, len(ns_vals)))
+ns_cols = np.array([colors[n] for n in ['wine', 'magenta', 'red', 'gold', 'blue']])
 
 # combined number of edges & runtime for Tajimas D
 fname = f"{basename}_absolute_values.pdf"
@@ -133,7 +147,7 @@ for L, this_df in df.groupby("length"):
     sub_df = this_df.sort_values("num_samples")
     j,  = np.where(L == length_vals)
     ax2.scatter(sub_df.num_edges_before, sub_df.extend_time, marker="o", label=f"length={L:g}",
-                c=ns_cols[j])
+                c=length_cols[j])
 
 for L, this_df in sdf.groupby("length"):
     sub_df = this_df.sort_values("num_samples")

### 2024-11-08 paper/writeup.tex
added benchmarkingn figs

@@
 
-%\begin{figure}
-%	\includegraphics[width=0.9\textwidth]{newplots_wo_ee/simp_vs_ext_tajimasD_runtime_numtrees.pdf}
-%    \caption{
-%        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
-%        as a function of number of trees (modulated by sequence length),
-%        on reasonably big simulated sequences.
-%        For "speed": plot ratio of runtime before/after extending.
-%        \label{fig:speed_and_edges}
-%    }
-%\end{figure}
+\begin{figure}
+    \includegraphics{benchmarks/one_pop_results_ratios}
+    \caption{
+        Ratio of
+        \textbf{(A)} number of edges, and
+        \textbf{(B)} runtime for computing Tajima's $D$;
+        before and after extending haplotypes.
+        For instance, extending haplotypes reduces number of edges by about 50\%
+        and statistic computation runtime by about 20\%
+        for long sequences.
+        Horizontal axis shows sequence length;
+        colors show numbers of samples;
+        with lines showing averages across replicates.
+        The original tree sequence was simulated with an expanding population;
+        see text for details. \comment{TODO}
+        \label{fig:speed_and_edges}
+    }
+\end{figure}
 
@@
 
+\begin{figure}
+    \includegraphics{benchmarks/one_pop_results_absolute_values}
+    \caption{
+        \label{fig:speed_and_edges_absolute}
+    }
+\end{figure}
+
+\begin{figure}
+    \includegraphics{benchmarks/one_pop_results_timing}
+    \caption{
+        \label{fig:speed_and_edges_timing}
+    }
+\end{figure}
+
+\begin{figure}
+    \includegraphics{benchmarks/constant_pop_results_ratios}
+    \caption{
+        As in \figref{fig:speed_and_edges} except that
+        the original tree sequence was simulated with an constant population;
+        see text for details. \comment{TODO}
+        \label{fig:speed_and_edges_constant}
+    }
+\end{figure}
+
+\begin{figure}
+    \includegraphics{benchmarks/constant_pop_results_absolute_values}
+    \caption{
+        \label{fig:speed_and_edges_absolute_constant}
+    }
+\end{figure}
+
+\begin{figure}
+    \includegraphics{benchmarks/constant_pop_results_timing}
+    \caption{
+        \label{fig:speed_and_edges_timing_constant}
+    }
+\end{figure}
+
 %\begin{figure}

### 2024-11-12 paper/writeup.tex
first final draft of results

@@
 
-% NOTE: these are in alphabetical order for now; we can revisit that ordering
 \author[$\dagger$]{Halley Fritze}
 \author[$\dagger$]{Nathaniel Pope}
-\author[$\dagger$]{Ava Bamforth}
 \author[$\ddagger$]{Jerome Kelleher}
@@
 
-\runningtitle{Junctions in tree sequences}
-\runningauthor{Bamforth \textit{et al.}}
+\runningtitle{Haplotypes in ARGs}
+\runningauthor{Fritze \textit{et al.}}
 
@@
     In this paper, we describe how this recombination-mediated information
-    can in many cases be recovered from inference based solely on
-    polymorphic markers (i.e., information produced by mutation),
+    can in many cases be recovered from inference based solely on polymorphic markers,
     which produces unary nodes along the branches of the marginal genealogical trees.
-    The resulting inferred ARGs, represented as tree sequences,
-    are smaller, faster to compute with,
-    and potentially contain substantially more information
-    about the age of ancestral haplotypes,
+    The resulting inferred ARGs are smaller, faster to compute with,
+    and contain substantially more information
     than inferred ARGs without these unary nodes.
     We provide efficient algorithms to identify this ancestral haplotype information,
-    new metrics of agreement/disagreement between inferred ARGs,
+    new metrics of agreement and disagreement between inferred ARGs,
     and explore some consequences for ARGs inferred from real data.
@@
 Counting edges for every iteration suggest that only at most five iterations are needed before the algorithm terminates.
-Indeed, for even larger sequences \tabref{tab:edge-counts} shows that 99\% of all changes to an ARG occur after the first iteration with the algorithm completing after four iterations.
-%% Find counter example to global minimum.
+Indeed, for even larger sequences \tabref{tab:edge-counts}
+shows that 99\% of all changes to an ARG occur after the first iteration,
+with the algorithm completing after four iterations.
 
@@
 
+We used simulations of several scenarios.
+To include the effects of heterogeneous recombination rate,
+in some we used stdpopsim \citep{stdpopsim} 
+to simulate chromosome 1 of \textit{Canis familiaris}
+using the CanFam3 genetic map from \citet{campell2016}.
+\emph{``Constant dog''} simulations simulated this chromosome in a 
+population of (constant) size $10^4$.
+\emph{``Expanding dog''} simulations were similar, but
+used a discrete-time Wright--Fisher model to simulate
+a small population of 100 that expanded to 1,000 individuals ten generations ago,
+which then doubled every generation to reach 512,000 individuals.
+All jobs used for recording runtime were run on an Intel Xeon Gold 6148 processor.
+\comment{TODO: describe sims for inference?}
+
 \section{Results}
@@
 \subsection{Tree sequence compression and computation}
-
-\comment{TODO: put in benchmarking figures}
 
@@
 Indeed,
-\figref{fig:num_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
+\figref{fig:speed_and_edges} shows that \algorithmref{alg:extend} nearly cuts the number of edges
 in half, as long as the sequence is long enough.
 
+\begin{figure}
+    \centering
+    \includegraphics{benchmarks/one_pop_results_ratios}
+    \caption{
+        Ratio of
+        \textbf{(A)} number of edges, and
+        \textbf{(B)} runtime for computing Tajima's $D$;
+        before and after extending haplotypes.
+        For instance, extending haplotypes reduces number of edges by about 50\%
+        and statistic computation runtime by about 20\%
+        for long sequences.
+        Horizontal axis shows sequence length;
+        colors show numbers of samples;
+        with lines showing averages across replicates.
+        The original tree sequence was simulated with the ``expanding dog'' expanding population
+        and subset to various sizes;
+        see Methods for details. 
+        Absolute values are shown in Supplementary \figref{fig:speed_and_edges_other}.
+        \label{fig:speed_and_edges}
+    }
+\end{figure}
+
 This reduction in edges can also lead to a reduction in computation time.
-Indeed, \figref{fig:num_edges} shows that computation time is reduced by 10--20\%
+Indeed, \figref{fig:speed_and_edges} shows that computation time is reduced by 10--20\%
 for a typical statistic (here, Tajima's $D$),
@@
 which is not affected by extending haplotypes.
+
+Supplementary \figref{fig:speed_and_edges_constant} shows these results are not specific
+to the demographic scenario.
+Supplementary Figures~\ref{fig:speed_and_edges_other}
+and~\ref{fig:speed_and_edges_other_constant}
+also show that our implementation of \algorithmref{alg:extend} is quite efficient,
+running at chromosome scale in seconds to minutes for hundreds or thousands of samples,
+or minutes to hours for tens of thousands of samples
 
@@
 
-\comment{TODO: figure out if we can remove this from figures. If so, move this to Methods.}
-Some stuff can't be inferred since it's "isolated unary";
-for instance with 1,000 samples, 5e7 that's about half the ARG.
-Etcetera.
-
 %$msprime.sim_ancestry(1000, sequence_length=5e7, recombination_rate=1e-8, population_size=1e4, coalescing_segments_only=False, random_seed=1)$
@@
 As seen in \figref{fig:node-spans},
-Look, we add lots of stuff
-and most most of it is right;
-this applies to both recent and ancient nodes;
-we're typically extending the span of a node by like maybe 50\%.
+the vast majority of span added by extending haplotypes is correct.
+In this example (which is typical),
+99\% of all added span is correct;
+95\% of nodes have no incorrectly added span;
+and those incorrectly added spans are nearly always a small fraction of the original span.
+The added information is significant:
+the algorithm typically increases spans (i.e., lengths of ancestral haplotypes)
+by around 50\%.
 
 \begin{figure}[bht]
-	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/figure-4-2.pdf}
+	\includegraphics{newplots_wo_ee/figure-4.pdf}
 	\caption{
-        Node spans of simulated tree sequence (blue), simplified tree sequence (green), and extended tree sequence (red).
-        \comment{TODO: nodes should be in the same order in all three lines,
-        using the ordering from the true tree sequence.}
-        \comment{TODO: somehow communicate per-node amount/proportion of correctly and incorrectly added span}
+        The effect of extending haplotypes on per-node spans
+        in an ARG simulated \comment{TODO HOW}.
+        \textbf{(A)}
+        Distribution of total amount of span added across nodes by \algorithmref{alg:extend};
+        note the log scale on the $y$ axis.
+        \textbf{(B)}
+        Amount of incorrectly added span, plotted against total span, by node.
+        95\% of nodes have no incorrect span; of the remainder,
+        nearly all have less than 5\% incorrectly added;
+        see Supplementary \figref{fig:incorrect_ratio}.
+        Plots \textbf{(C)} and \textbf{(D)} show total spans per node,
+        ordered by total span in the original ARG (which includes unary nodes).
+        Dark blue dots in both show the spans in this original ARG.
+        In \textbf{(C)}, lighter green dots show
+        span after removing unary spans using \textit{simplify},
+        while in \textbf{(D)},
+        lighter red dots show span after extending haplotypes,
+        i.e., applying \algorithmref{alg:extend} to the simplified ARG.
     }
@@
 
-\begin{figure}
-	\includegraphics{newplots_wo_ee/figure-4.pdf}
-\end{figure}
-
-\comment{merge this in}
-The first thing we see is that coalescent nodes are unary
-over a substantial portion of their spans:
-since the \texttt{simplify} operations removes these unary spans,
-the ``simplified'' line on the right of Figure~\ref{fig:discrepancy} shows
-the proportion of nodes' spans on which they are not unary.
-For instance, on an ARG of 10 samples on a sequence of length $5 \times 10^7$,
-only about 15\% of the total spans are retained after simplification;
-this rises to around 50\% for 1,000 samples (top right plot)
-
-We can also use measures of dissimiliarity and similarity to quantify summaries across nodes:
-these are shown in \figref{} (purple lines only)
-Points:
-proportion left after smiplify varies a LOT: from 15\% to 95\%.
-    (goes up with L: with longer sequence length, greater opportunity for distant chunks to be on the same lineage)
-extending adds hardly any wrong stuff (left plots); and adds a substantial chunk: with large samples, adds like half of the missing stuff
-
-It's also interesting to  note that inference
-(see \figref{fig:inferred_edge_counts})
-(a) increases number of edges;
-(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
-(c) but then extending reduces to below the original.
+These statistics are also reflected at the genomic scale, using measures of dissimilarity.
+\figref{fig:dissimilarity} shows in purple total amounts of span removed by simplification
+and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
+(Lines in green involve re-inference of the ARG; discussed next.)
+The top row shows the proportion of the given ARG that does not match the original,
+showing that the total amount of mis-matching span
+produced by extending haplotypes is very small ($\approx 1\%$).
+(Simplification does not produce dissimilarity.)
+The bottom row shows the proprtional of the original ARG that is represented in the given ARG.
+This shows that a large proportion of the ARG can be removed by simplifiction,
+indicating that coalescent nodes are unary
+over a substantial portion of their spans.
+(Since the \texttt{simplify} operations removes these unary portions of haplotypes,
+the ``simplified'' line (``SS'') on the right of Figure~\ref{fig:dissimilarity} shows
+the proportion of nodes' spans on which they are not unary.)
+However, \algorithmref{alg:extend} can correctly replace most of these portions of haplotypes,
+especially with larger sample sizes and sequence lengths
+(the ``simplified--extended'' line; ``SSE'').
+For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
+haplotypes are unary on about half their spans (on average),
+and extending haplotypes can infer more than half of this span from coalescent information only.
 
@@
         \comment{TODO: make "inferred simplified" green (dotted) not red}
-        \comment{TODO: label "inferred extended" "inferred simplfieid extended"??}
         \comment{TODO: also add a "inferred extended" line}
-    }
-    \label{fig:discrepancy}
+        \comment{TODO: instead of `SS` have `S` ; and instead of `SSE` have `SE`.}
+    }
+    \label{fig:dissimilarity}
 \end{figure}
@@
 	\includegraphics{newplots_wo_ee/figure5.pdf}
+    \caption{I think this is being removed?}
 \end{figure}
@@
 To study this, we simulated ARGs containing unary-spanning haplotypes (as above),
-then performed various operations to them,
-reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:discrepancy}.
-
-Main points: in \figref{fig:discrepancy}
+then re-inferred ARGs from their genotypes and performed various operations on the results.
+
+First, \figref{fig:dissimilarity} shows that \tsinfer{}
+has a substantial portion of already--extended haplotypes:
+comparing the ``inferred'' (``I''; dotted) line to the ``inferred-simplified'' (``IS''; dashed) line
+we see that inclusion of these unary spans increases the amount of correctly inferred material
+by around 20\% (bottom panels),
+but that roughly 10\% of the unary spans in the inferred ARG are incorrect (top panels).
+Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
+(``ISE''; solid green line)
+produces an ARG with both less correct and incorrect span.
+It is also helpful to note that accuracy
+(i.e., proportion of the true ARG that is inferred; bottom panels)
+greatly increases with larger sample sizes,
+possibly due to resolution of polytomies.
+(Recall that due to computational constraints,
+``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
+of subtending samples.)
+
+reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:dissimilarity}.
+Main points: in \figref{fig:dissimilarity}
 (green line) tsinfer gets a lot right (above 60\% with 1000 samples; bottom-right plot)
@@
 
+It's also interesting to  note that inference
+(see \figref{fig:inferred_edge_counts})
+(a) increases number of edges;
+(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
+(c) but then extending reduces to below the original.
+
 %
@@
 % * proprotion of added edges that are true ~ sequeence length
-
-\begin{figure}
-    \includegraphics{benchmarks/one_pop_results_ratios}
-    \caption{
-        Ratio of
-        \textbf{(A)} number of edges, and
-        \textbf{(B)} runtime for computing Tajima's $D$;
-        before and after extending haplotypes.
-        For instance, extending haplotypes reduces number of edges by about 50\%
-        and statistic computation runtime by about 20\%
-        for long sequences.
-        Horizontal axis shows sequence length;
-        colors show numbers of samples;
-        with lines showing averages across replicates.
-        The original tree sequence was simulated with an expanding population;
-        see text for details. \comment{TODO}
-        \label{fig:speed_and_edges}
-    }
-\end{figure}
 
@@
 
-TODO
-
-In this paper, we have taken a ``historical'' view of the ARG --
+We began work on this paper with the simple observation that the simple transformation
+of \figref{fig:extending_diagram} would reduce the number of edges
+in the succinct tree sequence representation of the ARG.
+This is essentially a recombination-based parsimony argument,
+and we have shown that this line of reasoning
+leads to ARGs that are substantially more compact and faster to operate on,
+and that contain more accurate information about true ancestral relationships.
+These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
+Although a number of ARG inference methods may be taking advantage of this information,
+it is our impression that this source of information is not widely appreciated in the community.
+In fact, the field's focus on marginal trees rather than haplotypes
+meant that to study accuracy of this method,
+we had to invent new haplotype-aware measures of (dis)agreement between ARGs.
+
+There are good reasons to think that lengthening the spans of ancestral haplotypes
+could lead to substantial gains in accuracy of ARG inference.
+For instance, information about inferring the age of a particular mutation
+derives almost entirely from constraints at nearby, linked sites.
+Extending ancestral haplotypes from one site into neighboring regions
+conceptually allows information from those marginal trees
+to inform age inference at that site as well.
+
+We have also explored the degree to which \tsinfer{} already makes use of this information,
+and whether this algorithm can be used to improve inference.
+The results do not provide a clear ordering:
+for instance, although \tsinfer-produced ARGs
+have a substantial portion of correctly inferred unary haplotypes,
+removing this with simplification decreases both dissimilarity
+(i.e., proportion of ``wrong'' haplotypes)
+and the proportion of the truth that is correctly inferred.
+Extending haplotypes restores a large amount of this correctly inferred span,
+but also introduces incorrect spans.
+Further work is needed to determine how the balance of ``true and false positive rates''
+affects downstream uses.
+The efficient computational tools we have implemented (in \tskit{})
+should facilitate this exploration.
+
+\paragraph{Ignorance and omission in the ARG}
+As motivation, we presented above a ``historical'' view of the ARG --
 i.e., that each aspect of an inferred ARG is intended to represent
@@
 that relationships \emph{not} depicted in the ARG are implied to not exist.
-This is slighly disingenuous -- an alternative and more accurate interpetation
+An alternative interpetation
 of the ARG of \figref{fig:conceptual}A\&B would be that we do not know
@@
 \appendix
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
 
@@
     \includegraphics{benchmarks/one_pop_results_absolute_values}
-    \caption{
-        \label{fig:speed_and_edges_absolute}
-    }
-\end{figure}
-
-\begin{figure}
     \includegraphics{benchmarks/one_pop_results_timing}
     \caption{
-        \label{fig:speed_and_edges_timing}
+        \textbf{(Above:)}
+        Absolute values for ratios shown in \figref{fig:speed_and_edges}:
+        \textbf{(left)} numbers of edges; and
+        \textbf{(right)} runtime.
+        \textbf{(Below:)} runtime for the \texttt{extend\_haplotypes}
+        implementation of \algorithmref{alg:extend} provided in the \tskit{} library.
+        For other details, see \figref{fig:speed_and_edges}.
+        \label{fig:speed_and_edges_other}
     }
@@
         As in \figref{fig:speed_and_edges} except that
-        the original tree sequence was simulated with an constant population;
-        see text for details. \comment{TODO}
+        the original tree sequence was simulated with an constant population
+        of size $10^4$ (using the ``constant dog'' scenario);
+        see Methods for details.
+        Absolute values are shown in Supplementary \figref{fig:speed_and_edges_constant_other}.
         \label{fig:speed_and_edges_constant}
@@
     \includegraphics{benchmarks/constant_pop_results_absolute_values}
-    \caption{
-        \label{fig:speed_and_edges_absolute_constant}
-    }
-\end{figure}
-
-\begin{figure}
     \includegraphics{benchmarks/constant_pop_results_timing}
     \caption{
-        \label{fig:speed_and_edges_timing_constant}
+        As in \figref{fig:speed_and_edges_other} except that
+        the original tree sequence was simulated with an constant population
+        of size $10^4$ (using the ``constant dog'' scenario).
+        \label{fig:speed_and_edges_other_constant}
     }

### 2024-11-12 benchmarks/plot_results.py

@@ -46,7 +46,7 @@ ns_cols = np.array([colors[n] for n in ['wine', 'magenta', 'red', 'gold', 'blue'
 
 # combined number of edges & runtime for Tajimas D
 fname = f"{basename}_absolute_values.pdf"
-fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(9, 5))
+fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(6.5, 3))
 
 ax0.set_xlabel("sequence length")
 ax1.set_xlabel("sequence length")
@@ -84,7 +84,7 @@ plt.savefig(fname)
 
 # combined number of edges & runtime for Tajimas D
 fname = f"{basename}_ratios.pdf"
-fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(9, 5))
+fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(6.5, 3))
 
 ax0.set_xlabel("sequence length")
 ax1.set_xlabel("sequence length")
@@ -120,7 +120,7 @@ plt.savefig(fname)
 
 # timing plots
 fname = f"{basename}_timing.pdf"
-fig, (ax0, ax1, ax2) = plt.subplots(1, 3, figsize=(9, 4), sharey=True)
+fig, (ax0, ax1, ax2) = plt.subplots(1, 3, figsize=(6.5, 3), sharey=True)
 
 ax0.set_xlabel("number of samples")
 ax0.set_ylabel("runtime (s)")

### 2024-11-12 benchmarks/plot_results.py
legend size

@@ -76,7 +76,7 @@ for ns, this_df in sdf.groupby("num_samples"):
     ax1.plot(sub_df.length, sub_df.runtime_after,
              c=ns_cols[j])
 
-ax1.legend()
+ax1.legend(fontsize="x-small")
 
 plt.tight_layout()
 plt.savefig(fname)
@@ -112,7 +112,7 @@ for ns, this_df in sdf.groupby("num_samples"):
 
 ax0.set_ylim((0, max(1, 1.05 * np.max(df.num_edges_after / df.num_edges_before))))
 ax1.set_ylim((0, max(1, 1.05 * np.max(df.runtime_after / df.runtime_before))))
-ax0.legend()
+ax0.legend(fontsize="x-small")
 
 plt.tight_layout()
 plt.savefig(fname)
@@ -161,9 +161,9 @@ for ns, this_df in sdf.groupby("num_samples"):
     ax1.plot(sub_df.length, sub_df.extend_time, label=f"samples={ns}",
              c=ns_cols[j])
 
-ax0.legend()
-ax1.legend()
-ax2.legend()
+ax0.legend(fontsize='xx-small')
+ax1.legend(fontsize='xx-small')
+ax2.legend(fontsize='xx-small')
 
 plt.tight_layout()
 plt.savefig(fname)

### 2024-11-20 paper/writeup.tex
some edits

@@
 }
+\newcommand*{\eqnref}[2][]{%
+	\hyperref[{#2}]{%
+		Equation~\ref*{#2}%
+		\ifx\\#1\\%
+		\else
+		\,#1%
+		\fi
+	}%
+}
 
@@
 \author[$\dagger$]{Halley Fritze}
-\author[$\dagger$]{Nathaniel Pope}
+\author[$\ast$]{Nathaniel Pope}
 \author[$\ddagger$]{Jerome Kelleher}
@@
 Neither is affected by shared haplotype structure -
-two ARGs could be identiical by either measure
+two ARGs could be identical by either measure
 but have completely different identity by descent (IBD) segment structure
@@
 
-Consider the (small portion) of a hypothetical tree sequence in \figref{fig:conceptual}A.
+Consider the (small portion) of a hypothetical tree sequence in \figref[A]{fig:conceptual}.
 On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
@@
 However, things change once we think about what this implies about haplotype inheritance.
-\figref{fig:conceptual}B shows the implied inheritance of haplotypes,
+\figref[B]{fig:conceptual} shows the implied inheritance of haplotypes,
 with the haplotypes carried by 4 to the left and right of the recombination breakpoint labeled $L$ and $R$.
@@
 Although this is possible, it seems unlikely --
-a more parsimonious explanation is depicted in \figref{fig:conceptual}C,
+a more parsimonious explanation is depicted in \figref[C]{fig:conceptual},
 in which sample node 1 inherits the entire $LR$ haplotype from ancestral node 4 through node 3
@@
 This implies that ancestral node 3 inherits from node 4 on the right-hand tree as well,
-which is depicted in \figref{fig:conceptual}D --
+which is depicted in \figref[D]{fig:conceptual} --
 and so node 3 has become unary in this tree.
 Note that the more parsimonious tree sequence also includes fewer edges:
-the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref{fig:conceptual}B
+the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref[B]{fig:conceptual}
 have been reduced to the two edges
-$4 \to 3$ and $3 \to 1$ in \figref{fig:conceptual}D.
-
-So, given the ARG shown in \figref{fig:conceptual}A\&B,
+$4 \to 3$ and $3 \to 1$ in \figref[D]{fig:conceptual}.
+
+So, given the ARG shown in \figref[A\&B]{fig:conceptual},
 it should be possible to extend the ancestral haplotype represented by node 3
-to obtain the ARG shown in \figref{fig:conceptual}C\&D,
+to obtain the ARG shown in \figref[C\&D]{fig:conceptual},
 thus adding additional information to the ARG.
@@
 which is the amount of time in the past that the individual who carried that segment of genome lived.
+Some nodes are \emph{samples}, meaning that we know them from data. 
+Typically, samples are exclusively leaves of $T$, however it is not the case that all samples are leaves.
 Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
@@
 for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
-or if $t_{u_i} = t_{v_j}$ for some $i$ and $j$.
-for which the intermediate nodes are not in the other tree
+or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$,
+for which the intermediate nodes are not in the other tree.
 So, we restrict our attention to pairs of such paths in adjacent trees
@@
 $T_1$ contains the path $6\to 4\to 0$ while $T_2$ has path $6\to 5\to 0$.
-The intermediate nodes $4$ and $5$ do not appear in $T_2$ respectively $T_1$, and so we consider these two paths \textit{mergeable}.
+The intermediate nodes $4$ and $5$ do not appear in $T_2$ and $T_1$ respectively, and so we consider these two paths \textit{mergeable}.
 The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$.
@@
         such that removing $F_O$ and adding $F_I$ turns $T_O$ into $T_I$,
-        checks to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
+        check to see if the paths upwards from $c$ in $T_O$ and $T_I$ are mergeable.
         If the paths are mergeable then this returns the number of new edges
@@
 	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-    \caption{
-        Extend haplotypes.
-        \comment{TODO: include this in the previous algorithm?}
-    }
+    \caption{Extend haplotypes.}
 	\label{alg:extend}
@@
 It is helpful to first describe what we compute in the simple case.
-\figref{fig:node-spans}A shows the spans for each node in three tree sequences:
-for each node, the top point is the total span in a ``true'' tree sequence
+\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
+for each node, the dark blue point is the total span in a ``true'' tree sequence
 that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
@@
 (by setting \texttt{coalescing\_segments\_only} to False).
-The bottom point is the span of the node after \emph{simplifying} the tree sequence
+The green point is the span of the node after \emph{simplifying} the tree sequence
 \citep{kelleher2018efficient},
 which removes all unary nodes from all trees.
-\comment{Usually (?)} between the two points
+The red point
 is the span of the node after applying \algorithmref{alg:extend}
 to the simplified tree sequence.
-So, the difference between the middle and bottom points is the amount by which
-the span of that node has been extended.
-\figref{fig:node-spans}B \comment{TODO} shows how much of this span is incorrect:
-for each node, it gives the percent of the span after running \algorithmref{alg:extend}
-that is incorrect (i.e., not present in the original tree sequence).
+\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
+from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
+\figref[B]{fig:node-spans} shows how much of this span is incorrect:
+for each node, it gives the amount of span
+that is incorrect (i.e., not present in the original tree sequence)
+over the total amount of span added after running \algorithmref{alg:extend}.
 The values in these plots (perhaps summed over nodes)
@@
 (as for instance with a true, simulated tree sequence and one inferred from its genotypes).
-\comment{should have introduced ``samples'' before this}
-Call the two tree sequence $T_1$ and $T_2$, which should have the same genome length
+Call the two tree sequences $T_1$ and $T_2$, which should have the same genome length
 and the same set of sample nodes.
@@
 for instance by returning the weighted root-mean-squared error on inferred times as
-\begin{align*}
+\begin{align}\label{eqn:wrmse}
     \text{wRMSE}_t(\T_1, \T_2)
     = \sqrt{\frac{
-        \sum_{n \in N_1} s(\T_1,n) \left(\log t^{(1)}_n - \log t^{(2)}_{\alpha(n)} \right)^2 
+        \sum_{n \in N_1} s(\T_1,n) \left(\log\left(1+ t^{(1)}_n\right) - \log\left(1+ t^{(2)}_{\alpha(n)}\right) \right)^2 
     }{
@@
     } } .
-\end{align*}
+\end{align}
 The mean is computed weighting by node span, so that a dating error
@@
 
-We compute the dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.
+The implementation of this method, called \emph{tree\_dissimilarity} \comment{change?}, in \texttt{tsdate}
+produces three values after computation: 
+the relative dissimilarity, weighted root-mean-square-error, and true proportion represented. 
+The \emph{relative dissimilarity} is defined to be the dissimilarity proportional
+ to the total span of nodes in $\T_1$,
+ \begin{align}\label{eqn:dissimilarity}
+ 	\overline{\dis}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}
+ 	=1-\frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}.
+\end{align}
+The \emph{weighted root-mean-square-error} is given in \eqnref{eqn:wrmse}.
+The \emph{true proportion represented} is similarity between two trees relative 
+to the total span of nodes in $\T_2$,
+\begin{align}\label{eqn:true-proportion}
+	\overline{\similarity}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_2}s(\T_2,n)}.
+\end{align}
+The outputs framed as proportions relative to one of the given tree sequences is more easily understood
+for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
+We compute the tree\_dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.
 
@@
     \caption{
-        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity function}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
+        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
         based on identical sample sets.
@@
         Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. On the rest of the genome, all of nodes match with their identical counterpart.
-        This makes the dissimilarity $\dis(T1,T2)=\frac{3}{46}$.
+        This makes the dissimilarity $\dis(T1,T2)=3$ and the \emph{relative similarity} $\overline{\dis}(T1,T2)=\frac{3}{46}$.
         \label{fig:conceptual_discrepancy}
@@
 option \texttt{coalescing\_segments\_only} set to False.
-Although \msprime simulates many events that do not create a coalescence in some marginal tree,
+Although \msprime\ simulates many events that do not create a coalescence in some marginal tree,
 it only outputs information for nodes which contain a coalescence
@@
 Said another way, by default all ancestral nodes in the ARG
-output by \msprime are the MRCA of some pair of samples at every point in the genome
+output by \msprime\ are the MRCA of some pair of samples at every point in the genome
 on which they are represented.
@@
 All jobs used for recording runtime were run on an Intel Xeon Gold 6148 processor.
-\comment{TODO: describe sims for inference?}
+To observe the accuracy of our method, we compute node spans for a correctly and 
+incorrectly added for a tree sequence simulated with
+ 1,000 samples over a genome of length $10^8$.
+ Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
+ using our dissimilarity method. 
+ These tree sequences range from 10 to 1,000 samples with genome length $5\times 10^7$
+ as well as having 1,000 samples and ranging the genome length from $10^6$ to $5\times 10^7$.
+All jobs used for recording node span and (dis)similarity for true and inferred ARGs
+ were run on an Intel Core i5-1035G4 processor.
 
@@
         The effect of extending haplotypes on per-node spans
-        in an ARG simulated \comment{TODO HOW}.
+        in an ARG simulated with $10^4$ \comment{diploid?} samples on a sequence of length $10^8$.
         \textbf{(A)}
@@
 (Since the \texttt{simplify} operations removes these unary portions of haplotypes,
-the ``simplified'' line (``SS'') on the right of Figure~\ref{fig:dissimilarity} shows
+the ``simplified'' line (``S'') on the right of Figure~\ref{fig:dissimilarity} shows
 the proportion of nodes' spans on which they are not unary.)
@@
 especially with larger sample sizes and sequence lengths
-(the ``simplified--extended'' line; ``SSE'').
+(the ``simplified--extended'' line; ``SE'').
 For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
@@
 	\begin{center}
-		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_sample_5e7_new.pdf}
-		% \includegraphics[width=0.8\linewidth]{newplots_wo_ee/discrepancy_over_seqlen_1000s_new_5e7.pdf}
         \includegraphics[width=\textwidth]{newplots_wo_ee/dissimilarity.pdf}
@@
         as proportion of haplotypes that are not represented in the true ARG
-        (equation~\eqref{eq:discrepancy}).
+        (\eqnref{eqn:dissimilarity}).
         \textbf{Right:} Agreement to the true ARG,
@@
         to the sequence length in the top row;
-        see text for more details.
-        \comment{TODO: make "inferred simplified" green (dotted) not red}
-        \comment{TODO: also add a "inferred extended" line}
-        \comment{TODO: instead of `SS` have `S` ; and instead of `SSE` have `SE`.}
-    }
+        see text for more details.  }     
     \label{fig:dissimilarity}
 \end{figure}
-
-\begin{figure}
-	\includegraphics{newplots_wo_ee/figure5.pdf}
-    \caption{I think this is being removed?}
-\end{figure}
-
-%\begin{figure}
-%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/edge_counts_5e7.pdf}
-%    \caption{
-%        number of edges, 5e7, 1,000 samples
-%        \label{fig:inferred_edge_counts}
-%    }
-%\end{figure}
 
@@
 (c) but then extending reduces to below the original.
+\comment{Also say something about IE line...}
 
@@
 (for instance, the MRCA of some set of sampled genomes).
-Furthermore, \figref{fig:conceptual}A\&B implicitly takes the position
+Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
 that relationships \emph{not} depicted in the ARG are implied to not exist.
 An alternative interpetation
-of the ARG of \figref{fig:conceptual}A\&B would be that we do not know
+of the ARG of \figref[A\&B]{fig:conceptual} would be that we do not know
 how node 2 inherited from node 4 on the right-hand interval,
@@
 Finally, if they have a different number of nodes,
-then \comment{???}.
+say $|N_1|<|N_2|$ we can take matching maps $\varphi\colon\T_1\to\T_2$ (not surjective)
+and $\psi\colon T_2\to T_1$ (not injective)
+and compute the dissimilarity as symmetric difference between $|\T_1\setminus \T_2|_{\varphi}|$
+and $|\T_2\setminus \T_1|_\psi|$. This should still be a metric \comment{?????}
+then \comment{???} .
 
@@
 \setcounter{figure}{0}
+
+\begin{figure}
+	\includegraphics{newplots_wo_ee/figure-4b-supplement}
+	\caption{Distribution of incorrectly added span percentages
+		 for the simulated tree sequence in \figref{fig:node-spans}.}
+	\label{fig:incorrect_ratio}
+\end{figure}
 

### 2024-11-20 paper/benchmarks/do_experiments.sh

@@ -1,0 +1,29 @@
+#!/bin/bash
+
+set -eo pipefail
+if [ -e "run_py.srun" ]
+then
+    PYTHON="sbatch run_py.srun"
+else
+    PYTHON="python"
+fi
+
+
+for OUTBASE in one_pop constant_pop
+do
+    OUTDIR="${OUTBASE}_results"
+    mkdir -p $OUTDIR
+
+    for L in 1e6 5e6 1e7 5e7 1e8
+    do
+        for N in 100 500 1000 5000 10000
+        do
+            SEED=$RANDOM$RANDOM
+            OUT=$OUTDIR/run_${L}_${N}_${SEED}
+            echo "$OUT  ....."
+            $PYTHON run_experiment.py --num_samples $N --length $L \
+                --seed $SEED ${OUTBASE}.trees $OUT
+            SEED=$((SEED + 1))
+        done
+    done
+done

### 2024-11-20 paper/benchmarks/plot_experiments.sh

@@ -1,0 +1,16 @@
+#!/bin/bash
+
+set -eo pipefail
+
+for OUTBASE in one_pop constant_pop
+do
+    OUTDIR="${OUTBASE}_results"
+    CSVFILE=${OUTDIR}.csv
+    if [ ! -e $CSVFILE ]
+    then
+        python jsons-to-csv.py $OUTDIR $CSVFILE
+    else
+        echo "$CSVFILE already exists, not regenerating."
+    fi
+    python plot_results.py $CSVFILE
+done

### 2024-11-20 paper/benchmarks/plot_results.py

@@ -1,0 +1,169 @@
+import sys, os
+import json
+import glob
+import pandas as pd
+import numpy as np
+import matplotlib.pyplot as plt
+import matplotlib
+
+assert len(sys.argv) == 2, "Usage: python plot_results.py <csv file>"
+infile = sys.argv[1]
+
+df = pd.read_csv(infile)
+basename = ".".join(infile.split(".")[:-1])
+
+# average across replicates:
+sdf = df.groupby(["length", "num_samples"]).agg({
+        "length": "mean",
+        "num_samples": "mean",
+        "num_edges_before": "mean",
+        "num_edges_after": "mean",
+        "runtime_before": "mean",
+        "runtime_after": "mean",
+        "extend_time": "mean"
+}).astype({'length': int, 'num_samples': int})
+sdf.reset_index(drop=True, inplace=True)
+
+# from Halley
+colors = {'blue': (46,37,133),
+          'red': (194,106,119),
+          'lgreen': (93,168,153),
+          'gold': (220,205,125),
+          'green': (51, 117,56),
+          'lblue': (148,203,236),
+          'magenta': (159,74,150),
+          'wine': (126,41,84),
+         }
+for x in colors:
+    colors[x] = [u/256 for u in colors[x]] + [1.0]
+
+length_vals = np.unique(df.length)
+# length_cols = plt.colormaps["viridis"](np.linspace(0, 1, len(length_vals)))
+length_cols = np.array([colors[n] for n in ['blue', 'green', 'lblue', 'lgreen', 'gold']])
+ns_vals = np.unique(df.num_samples)
+# ns_cols = plt.colormaps["plasma"](np.linspace(0, 1, len(ns_vals)))
+ns_cols = np.array([colors[n] for n in ['wine', 'magenta', 'red', 'gold', 'blue']])
+
+# combined number of edges & runtime for Tajimas D
+fname = f"{basename}_absolute_values.pdf"
+fig, (ax0, ax1) = plt.subplots(1, 2, figsize=(6.5, 3))
+
+ax0.set_xlabel("sequence length")
+ax1.set_xlabel("sequence length")
+ax0.set_ylabel("number of edges")
+ax1.set_ylabel("runtime, Tajima's D")
+for ns, this_df in df.groupby("num_samples"):
+    sub_df = this_df.sort_values("length")
+    j,  = np.where(ns == ns_vals)
[…]

### 2024-11-22 paper/writeup.tex
edits

@@
 we define and describe how to compute new measures of (dis-)agreement between ARGs
-that are motivated by the XXX distance between trees
+that are motivated by the Robinson-Foulds distance between trees
 but account for haplotype identity.
@@
 Some nodes are \emph{samples}, meaning that we know them from data. 
-Typically, samples are exclusively leaves of $T$, however it is not the case that all samples are leaves.
+Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
 Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
@@
 
-\subsection{Dissimilarity Function}
+\subsection{Dissimilarity between ARGs}
 % * How to measure agreement that includes haplotypes
@@
 we define a measure of \emph{dissimilarity} to quantify this.
-The method is implemented in the \texttt{tsdate} package \citep{wohns2022unified}
-as \comment{XXX}.
+The method is implemented in the \texttt{tscompare} package.
 
@@
 \figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
-for each node, the dark blue point is the total span in a ``true'' tree sequence
+for each node the total span in a ``true'' tree sequence, 
+given by the line,
 that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
@@
 (by setting \texttt{coalescing\_segments\_only} to False).
-The green point is the span of the node after \emph{simplifying} the tree sequence
+\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
 \citep{kelleher2018efficient},
-which removes all unary nodes from all trees.
-The red point
+which removes all unary nodes from all trees
+thus reducing the node spans for some nodes in the sequence.
+In \figref[D]{fig:node-spans}
 is the span of the node after applying \algorithmref{alg:extend}
@@
 however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
+Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
 We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
@@
     =
-    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))),
+    \sum_{n \in N_1} (s(\T_1, n) - m(n, \alpha(n))) 
+    = 
+    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
 \end{align*}
@@
     }{
-        \sum_{n \in N_1} s(\T_1,n)
+        \Vert{\T_1}\Vert
     } } .
@@
 
-The implementation of this method, called \emph{tree\_dissimilarity} \comment{change?}, in \texttt{tsdate}
-produces three values after computation: 
-the relative dissimilarity, weighted root-mean-square-error, and true proportion represented. 
-The \emph{relative dissimilarity} is defined to be the dissimilarity proportional
+The implementation of this method in \texttt{tscompare}
+additionally produces relative values:
+the ancestral recombination graph robinson-foulds value (relative dissimilarity)  and true proportion represented. 
+The \emph{ancestral recombination graph robinson-foulds} (ARF) is defined to be the dissimilarity proportional
  to the total span of nodes in $\T_1$,
  \begin{align}\label{eqn:dissimilarity}
- 	\overline{\dis}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}
- 	=1-\frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_1} s(\T_1,n)}.
+ 	\operatorname{ARF}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\Vert \T_1 \Vert}
+ 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert}.
 \end{align}
-The \emph{weighted root-mean-square-error} is given in \eqnref{eqn:wrmse}.
-The \emph{true proportion represented} is similarity between two trees relative 
+The \emph{true proportion represented} (TPR) is similarity between two trees relative 
 to the total span of nodes in $\T_2$,
 \begin{align}\label{eqn:true-proportion}
-	\overline{\similarity}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}{\sum_{n\in N_2}s(\T_2,n)}.
+	\operatorname{TPR}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}
+	{\Vert \T_2\Vert}.
 \end{align}
@@
 for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
-We compute the tree\_dissimilarity of two tree sequences in \figref{fig:conceptual_discrepancy}.
-
-%%%%%%%%%%%%
-%\comment{PREVIOUS VERSION: retrieve bits from below?}
-%
-%Inspired by Robinson-Foulds we define a function, called a \textit{discrepancy function} (see \algorithmref{alg:disc}), on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
-%Let $N_{T_1}$ and $N_{T_2}$ be the set of nodes on $\T_1$ respectively $\T_2$.
-%For a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
-%Each node $n\in\T$ has a given position along the genome and time, 
-%so we can write each node as a tuple $n = (t_n, L_n)$ with position $L_n\subset \left[0,L\right)$ and time $t_n\in \R^{\geq 0}$.
-%Computing the discrepancy between a random pair of nodes is not helpful, 
-%so we must first identify which pairs of nodes would be most identical to one another or a `best match'.
-%These `best matching' nodes should be close in terms of time and have equal descendant sample sets,
-%ie. $s(i) = s(j)$ where $s$ is the function to compute the descendant sample set of $n_1$.
-%
-%To find a best match for each node $i\in N_{T_1}$ \comment{(Should I do this by mentioning clades?)}
-%we compute the \textit{shared span} between node $i\in N_{T_1}$ and every node in $N_{T_2}$.
-%Then, the \textit{shared span} between two tree sequences, denoted $ss(\T_1,\T_2)$, is a $|N_{T_1}|\times |N_{T_2}|$ matrix $A$
-%where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_{T_1}\times N_{T_2}$.
-%If a pair $(i,j)$ has equal descendant sample sets, $s(i) = s(j)$
-%then $a_{ij} = |L_i \cap L_j|$.
-%If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
-%Matrix $A$ determines the \textit{best match} for every node $i\in N_{T_1}$
-%via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_{T_2}} a_{ij}$.
-%If this maximum is not unique, we use time difference to determine the best match pair.
-%We then define the \textit{discrepancy} between $\T_1$ and $\T_2$ as 
-%one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
-%
-%We note that the discrepancy function is a metric for detecting non-similarity between two tree sequences.
-However, we note that this function is not a metric in the mathematical sense, as it is not symmetric, nor is it reflexive.
-For example, given a tree sequence $\T$ and its simplification $\mathbb{S}$ and $\T\neq\mathbb{S}$ we have $\operatorname{discrepancy}(\mathbb{S}, \T)=0$.
-Additionally, given \figref{fig:conceptual_discrepancy}, $\operatorname{discrepancy}(T1, T2) \neq \operatorname{discrepancy}(T2, T1)$ \comment{actually compute this.
-We may just do it in this section instead of the appendix?}
-% 
-%\comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}
+We compute dissimilarity and ARF of two tree sequences in \figref{fig:conceptual_discrepancy}.
 
@@
         In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
-        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. On the rest of the genome, all of nodes match with their identical counterpart.
-        This makes the dissimilarity $\dis(T1,T2)=3$ and the \emph{relative similarity} $\overline{\dis}(T1,T2)=\frac{3}{46}$.
+        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$.
+Thus the maximal mapping for 4 should be to itself.
+On the rest of the genome, all of nodes match with their identical counterpart.
+        This makes the dissimilarity $\dis(T1,T2)=3$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
         \label{fig:conceptual_discrepancy}
@@
 
-%\comment{Delete Discrepancy Algorithm!!}
-% 
-% \begin{algorithm2e}[!ht]
-% \SetStartEndCondition{ }{}{}%
-% \SetKwProg{Fn}{def}{\string:}{}
-% \SetKwFunction{Range}{range}%%
-% \SetKw{KwTo}{in}
-% \SetKwFor{For}{for}{\string:}{}%
-% \SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-% \SetKwFor{While}{while}{:}{fintq}%
-% \SetKw{Break}{break}
-% \newcommand\forcond{$i$ \KwTo\Range{$n$}}
-% \SetKwComment{Comment}{}
-% \AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-% \DontPrintSemicolon\LinesNotNumbered
-% 
-% \caption{Discrepancy Function}\label{alg:disc}
-% \input{$(T_1,T_2)$}
-% \output{tree discrepancy, root-mean-square error}
-% 
-% \Fn{Shared Spans($T_1,T_2$)}{
-% \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
-% 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
-% 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
-% 		Define $a_{ij}=|L_i\cap L_j|$.
-% 	}
-% 	\Else{Define $a_{ij}=0$.}
-% Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
-% }
-% \KwRet{$A$}
-% }
-%\BlankLine
-%\Fn{Discrepancy ($T_1,T_2$)}{
-%\For{each node $i\in N_{T_1}$}{
-%	Find nodes in $N_{T_2}$ which have the highest matching span.\;
-%	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
-%	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
-%	Define $\alpha_i = a_{ij}$.\;
-%	\BlankLine
-%}
-%$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_1)\right)}$$\;
-%\BlankLine\BlankLine
-%$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_1)\right)}}$$\;
-%\BlankLine
-%$$true\ proportion = \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname(tr)\left( Shared Spans(T_2, T_2)\right)}$$\;
-%\BlankLine\BlankLine
-%\KwRet{tree discrepancy, rmse, true proportion}
-%}
-%\end{algorithm2e}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Simulations}
@@
 All jobs used for recording runtime were run on an Intel Xeon Gold 6148 processor.
-To observe the accuracy of our method, we compute node spans for a correctly and 
-incorrectly added for a tree sequence simulated with
- 1,000 samples over a genome of length $10^8$.
  Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
- using our dissimilarity method. 
- These tree sequences range from 10 to 1,000 samples with genome length $5\times 10^7$
- as well as having 1,000 samples and ranging the genome length from $10^6$ to $5\times 10^7$.
-All jobs used for recording node span and (dis)similarity for true and inferred ARGs
- were run on an Intel Core i5-1035G4 processor.
+ using our dissimilarity methods. 
+ These tree sequences have a population of 1,000 and recombination rate of $10^{-8}$ with samples ranging from 10 to 1,000 and genome length ranging from $10^6$ to $5\times 10^7$.
 
@@
 computed in an efficient iterative manner along the genome as implemented in \tskit.
-This is actually somewhat surprising:
+This may at first be somewhat surprising:
+This makes sense because
 as described in \citet{ralph2020efficiently},
 each edge added or removed requires an update
-to the state of the parent node \emph{and every node above it in the tree}.
-So, even if an edge extends for a long segment of genome,
-something has to happen to it every time the subtree below it changes,
-which is not affected by extending haplotypes.
+to the state of the parent node and nodes above it,
+and extending haplotypes yields a tree sequence with fewer edge removals and insertions.
 
@@
 
-\begin{figure}[bht]
+\begin{figure}[!bht]
 	\includegraphics{newplots_wo_ee/figure-4.pdf}
@@
         The effect of extending haplotypes on per-node spans
-        in an ARG simulated with $10^4$ \comment{diploid?} samples on a sequence of length $10^8$.
+        in an ARG simulated with $10^4$ diploid samples, $10^4$ population, and 
+        recombination rate of $10^{-8}$ on a sequence of length $10^8$.
         \textbf{(A)}
@@
 These statistics are also reflected at the genomic scale, using measures of dissimilarity.
-\figref{fig:dissimilarity} shows in purple total amounts of span removed by simplification
+Line ``SE'' in \figref{fig:dissimilarity} shows total amounts of span removed by simplification
 and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
-(Lines in green involve re-inference of the ARG; discussed next.)
-The top row shows the proportion of the given ARG that does not match the original,
+(Lines labeled with ``I'' involve re-inference of the ARG; discussed next.)
+The top row shows the proportion of the given ARG that does not match the original (ARF),
 showing that the total amount of mis-matching span
@@
 (Simplification does not produce dissimilarity.)
-The bottom row shows the proprtional of the original ARG that is represented in the given ARG.
+The bottom row shows the proprtional of the original ARG that is represented in the given ARG (TRP).
 This shows that a large proportion of the ARG can be removed by simplifiction,
@@
 
-\begin{figure}
+\begin{figure}[!hbt]
 	\begin{center}
-        \includegraphics[width=\textwidth]{newplots_wo_ee/dissimilarity.pdf}
+        \includegraphics[width=\textwidth]{figure5.pdf}
 	\end{center}
@@
         across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
-        and sequence lengths, with 1,000 \comment{diploid?} samples (bottom).
+        and sequence lengths, with 1,000 diploid samples (bottom).
         For each, a simulated ARG containing unary haplotype spans
@@
         The inferred, then simplified, ARG is also shown.
-        \textbf{Left:} Discrepancy to the true ARG,
+        \textbf{Top row:} ARF, the dissimilarity to the true ARG,
         as proportion of haplotypes that are not represented in the true ARG
         (\eqnref{eqn:dissimilarity}).
-        \textbf{Right:} Agreement to the true ARG,
+        \textbf{Bottom row:} TPR, agreement to the true ARG
         as proportion of the true ARG that is represented.
-        The largest number of trees in the bottom row corresponds
-        to the sequence length in the top row;
-        see text for more details.  }     
+        }     
     \label{fig:dissimilarity}
@@
 has a substantial portion of already--extended haplotypes:
-comparing the ``inferred'' (``I''; dotted) line to the ``inferred-simplified'' (``IS''; dashed) line
+comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
 we see that inclusion of these unary spans increases the amount of correctly inferred material
@@
 but that roughly 10\% of the unary spans in the inferred ARG are incorrect (top panels).
+Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span.
 Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
@@
 produces an ARG with both less correct and incorrect span.
+Additionally, the ``IE'' and ``ISE'' ARGs 
+contain approximately the same number of edges (difference of $\approx 1\%$).
 It is also helpful to note that accuracy
@@
 
-reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:dissimilarity}.
-Main points: in \figref{fig:dissimilarity}
-(green line) tsinfer gets a lot right (above 60\% with 1000 samples; bottom-right plot)
-    and does a lot better with more samples;
-    also, gets 10\%--25\% wrong (left panels)
-    with longer sequence, larger proportion is wrong (bottom-left panel)
-(red dotted)
-    simplifying removes a lot of the wrong stuff
-    but also a lot of the right stuff
-(red solid)
-    similarly extending the simplified one adds a bunch of right and also wrong stuff
-
-It's also interesting to  note that inference
-(see \figref{fig:inferred_edge_counts})
-(a) increases number of edges;
-(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
-(c) but then extending reduces to below the original.
-\comment{Also say something about IE line...}
-
+%reporting proportions of correct and incorrect haplotype spans in Figure~\ref{fig:dissimilarity}.
+%Main points: in \figref{fig:dissimilarity}
+%(green line) tsinfer gets a lot right (above 60\% with 1000 samples; bottom-right plot)
+%    and does a lot better with more samples;
+%    also, gets 10\%--25\% wrong (left panels)
+%    with longer sequence, larger proportion is wrong (bottom-left panel)
+%(red dotted)
+%    simplifying removes a lot of the wrong stuff
+%    but also a lot of the right stuff
+%(red solid)
+%    similarly extending the simplified one adds a bunch of right and also wrong stuff
 %    
-%\begin{figure}[bht]
-%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/better_unary_spans_ep_log.pdf}
-%	\caption{Correct Unary Span for Extend Haplotypes (log)}
-%	\label{fig:correct-unary-span-log}
-%\end{figure}
 %
-%\begin{figure}
-%%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/avg_edge_count_over_iters_5e7_1000s.pdf}
-%	\caption{The average number of edges for an extended tree sequence per iteration of the extend haplotypes algorithm.}
-%	\label{fig:edges-per-iteration}
-%\end{figure}
-
-% (AVA)
-% * reduction in edges ~ sequence length
-% * speed increase ~ sequence length
-% * proprotion of added edges that are true ~ sequeence length
-
-%Simple experiment:
-%(1) simulate, simplify, and extend;
-%(2) let $x$ be total matching span of simplified tree (here equal to the total span); $y$ total matching span of extended tree in truth and $z$ total not-matching span;
-%then $z$ is "total wrongly added stuff" and $y-x$ is "total rightly added stuff".
-%
-%With tsinfer: (1) simulate and tsinfer; (2) do the above either with or without simplifying the tsinfer'ed trees.
-%(but also take into account that there's wrong stuff in tsinfer'ed tree: look at whether it reduces discrepancy!)
-
-%\begin{figure}
-%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_geneconversion.pdf}
-%	\caption{
-%        Accuracy of \algorithmref{alg:extend}
-%        on simulated ARGs with sample size $10^2$, length $10^7$, gene conversion rate $10^{-8}$ and gene tract length = $10^3$.
-%        The same quantities for simulated ARGs without gene conversion
-%        are shown in Supplementary \figref{fig:accuracy_on_truth_no_gc}.
-%    \label{fig:accuracy_on_truth}
-%}
-%\end{figure}
-%
-%\begin{figure}
-%	\includegraphics[width=0.9\linewidth]{newplots_wo_ee/sampling_dist_recomb.pdf}
-%	\caption{Accuracy of \algorithmref{alg:extend}
-%        as in \figref{fig:accuracy_on_truth} except that simulations were done
-%        without gene conversion. \comment{Add to supplement.}
-%        \label{fig:accuracy_on_truth_no_gc}
-%    }
-%\end{figure}
-
-\subsection{Data}
-% * 1. Simulate data
-% * 2. Run tsinfer
-% * 3. Compare d(x,truth) for x in tsinfer, tsinfer+add edge, tsinfer+add edge + simplify, tsinfer+simplify
-
-% * How much unary edges?
-% * How much can we add?
-
-% * apply to real data:
-%     - % fewer edges
-%     - % fewer distinct ancestors (ancestral paths?) of sample at some time
+%It's also interesting to  note that inference
+%(see \figref{fig:inferred_edge_counts})
+%(a) increases number of edges;
+%(b) but inferred-simplifies have even more suggesting unary stuff is parsimonious;
+%(c) but then extending reduces to below the original.
 
@@
 have a substantial portion of correctly inferred unary haplotypes,
-removing this with simplification decreases both dissimilarity
+removing this with simplification decreases both ARF
 (i.e., proportion of ``wrong'' haplotypes)
-and the proportion of the truth that is correctly inferred.
+and TPR (the proportion of the truth that is correctly inferred).
 Extending haplotypes restores a large amount of this correctly inferred span,
@@
 \paragraph{Metrics on ARGs}
-\comment{collecting random notes here; TODO is tie them together}
 The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
@@
 $|\T_1 \setminus \T_2|$,
-and so the symmetrized version is the measure of the symmetric difference,
+and so the symmetrized version is the measure of the symmetric difference
+$|\T_1\Delta\T_2|$,
 which is well-known to be a metric.
@@
 and minima over a finite number of metrics is also a metric.
-Finally, if they have a different number of nodes,
-say $|N_1|<|N_2|$ we can take matching maps $\varphi\colon\T_1\to\T_2$ (not surjective)
-and $\psi\colon T_2\to T_1$ (not injective)
-and compute the dissimilarity as symmetric difference between $|\T_1\setminus \T_2|_{\varphi}|$
-and $|\T_2\setminus \T_1|_\psi|$. This should still be a metric \comment{?????}
-then \comment{???} .
+This also holds for two tree sequences with different sizes of nodes, $|N_1|\neq|N_2|$ as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
 
@@
 \paragraph{Inference methods}
-\comment{Possibly an overview of what the different inference methods do with unary spans?}
+Clearly, there is ample information in genomic data to infer the extent of ancestral haplotypes even where they are unary,
+and this has the potential to significantly inform downstream methods, especially dating.
+However, currently not all ARG inference methods return ARGs with unary nodes.
+For instance, we have seen that tsinfer returns ARGs to which the method presented here adds little -
+so, seems to be already taking advantage of this source of information.
+Singer [cite] returns ARGs without any unary nodes,
+although these are reconstructable using the times of the recombination nodes.
+On the other hand, locally building trees separately in each segment of genome
+simply does not provide this sort of haplotype-based information (although it might be added using a method like ours).
 
@@
 \setcounter{figure}{0}
-
-\begin{figure}
-	\includegraphics{newplots_wo_ee/figure-4b-supplement}
-	\caption{Distribution of incorrectly added span percentages
-		 for the simulated tree sequence in \figref{fig:node-spans}.}
-	\label{fig:incorrect_ratio}
-\end{figure}
 
@@
 
+\begin{figure}[!hbt]
+	\includegraphics{newplots_wo_ee/figure-4b-supplement}
+	\caption{Distribution of incorrectly added span percentages
+		for the simulated tree sequence in \figref{fig:node-spans}.}
+	\label{fig:incorrect_ratio}
+\end{figure}
+
 %\begin{figure}
@@
 
-\begin{table}[hbt]
+\begin{table}[!hbt]
 \begin{center}
@@
 
-%\begin{algorithm2e}[!ht]
-%	\SetStartEndCondition{ }{}{}%
-%	\SetKwProg{Fn}{def}{\string:}{}
-%	\SetKwFunction{Range}{range}%%
-%	\SetKw{KwTo}{in}
-%	\SetKwFor{For}{for}{\string:}{}%
-%	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-%	\SetKwFor{While}{while}{:}{fintq}%
-%	\SetKw{Break}{break}
-%	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-%	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-%	
-%	\caption{Extends haplotypes across a tree sequence $TS$.
-%    \comment{I think we can remove this now?}}
-%	\label{alg:hap}
-%	
-%	\SetKwFunction{ex}{Extend Paths}
-%	\SetKwFunction{fp}{Find Paths}
-%	\DontPrintSemicolon
-%	\LinesNotNumbered
-%	%\SetAlgoNoEnd
-%	\input{A tree sequence $\T$}
-%	\output{New tree sequence,$\T'$}
-%	\BlankLine
-%	\Fn{\fp (tree sequence $\T$, forwards = \tn{True})}{
-%	{
-%		\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-%		{\BlankLine
-%			$\T = \left(T_{|\T|},..., T_1\right)$\;
-%		}
-%	}
-%		\For{pairs $(T_k, T_{k+1})\in \T\times\T$}{
-%		Compute subforests $F_O\subseteq T_k$ and $F_I\subseteq\in T_{k+1}$.\;
-%		\For{edges $e$ in $F_O$}{
-%			Let $c$ be child node.\;
-%			$p_c^{F_O} = [c]$\;
-%			$p_c^{F_I} = [c]$\;
-%			$\bar{p}_c = []$ Path to be inserted\;
-%			\While{$\exists f\in F_O$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))=0$}{
-%				$p = \tn{parent}(f)$\;
-%				Append $p$ to $p_c^{F_O}$\;
-%				$c = p$\;
-%			}
-%			\While{$\exists f\in F_I$ such that $\tn{child}(f)=c$ and $\tn{deg}(\tn{parent}(f))<3$}{
-%				$p =\tn{parent}(f)$\;
-%				Append $p$ to $p_c^{F_I}$\;
-%				$c = p$\;
-%			}
-%			\If{($p_c^{F_O}\cap p_c^{F_I})\backslash\{c\}\neq\emptyset$}{
-%				Merge paths up to intersection point\;
-%				$\bar{p}_c = \hat{p}_c^{F_I} \cup \hat{p}^{F_I}_c$\;
-%				and sort by node time.\;
-%			}
-%			Insert $\bar{p}_c$ into $T_{k+1}$\;		
-%		}
-%	}\;
-%	Update $\T$.\;
-%	\KwRet{$\T$}\;
-%}
-%\BlankLine						
-%\Fn{\ex ($\T$)}{
-%	$\T' \gets $ \fp($\T$, forwards =\tn{True})\;
-%	$\T' \gets $ \fp($\T$, forwards = \tn{False})\;
-%	\While{Number of edges in $\T$ $\neq$ Number of edges in $\T'$}{
-%		$\T' \gets $ \fp($\T$, fowards=\tn{True})\;
-%		$\T' \gets $ \fp($\T$, fowards=\tn{False})\;
-%		
-%	}
-%	\KwRet{$\T'$}\;
-%}
-%\end{algorithm2e}
-%
-%% Can most likely delete the extend-edges algorithm %%
-%\begin{algorithm2e}[!ht] 
-%	
-%\SetStartEndCondition{ }{}{}%
-%\SetKwProg{Fn}{def}{\string:}{}
-%\SetKwFunction{Range}{range}%%
-%\SetKw{KwTo}{in}
-%\SetKwFor{For}{for}{\string:}{}%
-%\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-%\SetKwFor{While}{while}{:}{fintq}%
-%\SetKw{Break}{break}
-%\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-%\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-%
-%\caption{Extends edges across a tree sequence $TS$. 
-%\comment{Can be removed, probably -- unless this seems a lot simpler?}}
-%\label{alg:edge}
-%
-%\SetKwFunction{ex}{Extend Edges}
-%\SetKwFunction{fwd}{Extend}
-%\DontPrintSemicolon
-%\LinesNotNumbered
-%%\SetAlgoNoEnd
-%\input{A tree sequence $\T$}
-%\output{New tree sequence,$\T'$}
-%
-%\BlankLine 
-%
-%\Fn{\fwd (tree sequence $\T$, forwards = \tn{True})}{
-%{	
-%\eIf{forwards=\tn{True}}{$\T = \left(T_1,...,T_{|\T|}\right)$\BlankLine}
-%{\BlankLine
-%	$\T = \left(T_{|\T|},..., T_1\right)$\;}
-%}
-%	\For{$T_k\in \T$}{
-%	Find all edge pairs $(e_1,e_2)$ such that $e_1 = a\to b$ and $e_2 = b\to c$ for some nodes $a,b,c$.\;
-%	\For{edges $f$ in $T_{k+1}$}{
-%		\If{$f = a \to c$ and $b$ is not in $T_{k+1}$}{
-%		Remove $f$ from $T_{k+1}$.\;
-%		Add $e_1$ and $e_2$ to $T_{k+1}$.\;
-%		Update $\T$.\;
-%	}
-%	}
-%}\;						
-%\KwRet{$\T$}\;
-%}
-%\BlankLine						
-%\Fn{\ex ($\T$, iter=100)}{
-%	\For{$*$ \KwTo \Range(iter)}{
-%		$\T' \gets \T$.\fwd(fowards=\tn{True})\;
-%		$\T' \gets \T$.\fwd(fowards=\tn{False})\;
-%		\If{Number of edges in $\T$ = Number of edges in $\T'$}{
-%			\Break\BlankLine}
-%		}
-%}
-%\KwRet{$\T'$}\;
-%\end{algorithm2e}
-
 \end{document}

### 2024-11-24 paper/writeup.tex
Edits

@@
     In this paper, we describe how this recombination-mediated information
-    can in many cases be recovered from inference based solely on polymorphic markers,
-    which produces unary nodes along the branches of the marginal genealogical trees.
+    can in many cases be recovered from inference based solely on polymorphic markers.
+    % NSP-22NOV: "unary node" and "ARG" are not defined so I tried to provide better context here
+    In a genealogical context, this information reflects the persistence
+    of ancestral haplotypes across marginal genealogical trees where they
+    are not coalescence events.
+    We show how these non-coalescing haplotypes (``unary nodes'') may be
+    inserted into the ancestral recombination graph (ARG), a compact
+    but information-rich data structure describing the genealogical
+    relationships among recombinant sequences.
     The resulting inferred ARGs are smaller, faster to compute with,
-    and contain substantially more information
-    than inferred ARGs without these unary nodes.
-    We provide efficient algorithms to identify this ancestral haplotype information,
+    and contain substantially more information % NSP-22NOV: "substantially more information" in what regard?
+    than inferred ARGs without unary nodes.
+    We provide efficient algorithms to infer unary nodes within existing ARGs,
     new metrics of agreement and disagreement between inferred ARGs,
-    and explore some consequences for ARGs inferred from real data.
+    and explore some consequences of unary nodes for ARGs inferred from real data.
 \end{abstract}
@@
 ``more similar sequences are more closely related''.
-For instance, two sequences that share a drerived mutation
+For instance, two sequences that share a derived mutation
 are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
@@
 
-Haplotype identity has been nearly overlooked in the literature on ARG inference so far --
-all metrics that have been used so far to measure accuracy of inferred ARGs
+Haplotype identity has been largely overlooked in the literature on ARG inference so far --
+in that the metrics that have been used to measure accuracy of inferred ARGs
 depend only on the sequence of marginal trees,
-not on sharing of ancestral haplotype identities between them.
+not on the sharing of ancestral haplotype identities between them.
+% NSP-22NOV: what does "sharing of ancestral haplotype identities" mean, concretely?
+Would "...
+depend only on the sequence of marginal trees, but not on the ancestral haplotypes that span across these trees" be a clearer description?
 Suppose two methods are used to infer ARGs from the same set of genotyped samples.
-Although both ARGs describes relationships between many haplotypes,
-only those haplotypes that represent the focal set of genotyped samples
+Although both ARGs describe the relationships between many haplotypes,
+only those haplotypes that represent the focal set of genotyped samples % NSP-22NOV: the meaning here isn't clear to me-- what does "represent the focal set of genotyped samples" mean, concretely?
 can be unambiguously said to represent the same thing in both.
-So, most metrics only describe these relationships,
-which ignores the additional information carried by the structure of other haplotypes.
+Thus, metrics that only describe these relationships
+ignore the additional information carried by the structure of other haplotypes. % NSP-22NOV: again, "additional information carried by the structure of other haplotypes" is vague
 For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
@@
 between pairs of sampled genomes.
-Neither is affected by shared haplotype structure -
+Neither is affected by shared haplotype structure --
 two ARGs could be identical by either measure
@@
 that is the total variation distance between the induced distribution on genotypes;
-neither of which measure sharing of haplotypes between adjacent trees.
+neither of which measure the sharing of haplotypes between adjacent trees.
 
@@
 but account for haplotype identity.
-These measures show that the vast majority of these extended haplotypes are correct,
+These measures show that the vast majority of these extended haplotypes are correct, % NSP-22NOV: "vast majority" is correct in simulated (but not inferred) ARGs, right?
 and that substantial information about haplotypes is contained in these nodes
@@
 could not be expected to produce anything different.
-However, things change once we think about what this implies about haplotype inheritance.
+However, the example becomes more complicated once we consider what these marginal genealogies imply about haplotype inheritance.
 \figref[B]{fig:conceptual} shows the implied inheritance of haplotypes,
@@
 
-We use work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
+We work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
 to take advantage of the extensive, stable, and well-documented set of open-source tools
-available in \tskit{} \citep{tskit}.
-So, our terminology and notation follows \citet{ralph2020efficiently}.
+available in \tskit{} \citep{tskit}, 
+and our terminology and notation follows \citet{ralph2020efficiently}.
 For our purposes here,
@@
 over a segment of genome $[\ell_e, r_e)$.
-(The tree sequence includes substantially more information, including genotypes, but we will not use that in this paper.)
+(The tree sequence includes substantially more information, including genotypes, that we will not use in this paper.)
 Suppose that the unique elements of the set of left and right edge endpoints
@@
 
-\algorithmref{alg:extend} simplifies the algorithm in several ways --
+\algorithmref{alg:extend} simplifies the algorithm in several ways for the sake of clarity --
 for instance, the bookeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
@@
                 $p_i = P_I[p_i]$\;
-                $t_i = (p_i = NULL) ? \infty : t[p_i]$\;
+                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$\;
             }{
@@
                 $p_o = P_O[p_o]$\;
-                $t_o = (p_o = NULL) ? \infty : t[p_o]$\;
+                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$\;
                 $n += 1$\;
@@
 for each node, it gives the amount of span
-that is incorrect (i.e., not present in the original tree sequence)
+that is incorrect (i.e., not present in the original tree sequence) % NSP-22NOV: it seems a bit odd that this panel (B) is not on a log scale while all the others are
 over the total amount of span added after running \algorithmref{alg:extend}.
-The values in these plots (perhaps summed over nodes)
+The values in these plots
 therefore quantify how much the ancestral haplotypes represented by each node
@@
 They also tell us how much of the relationships in the original tree sequence
-are represented in the result.
+are represented in the result. % NSP-22NOV: "relationships" of what exactly?
 
@@
 we wish to compare two tree sequences for which we know the sample nodes are the same
-but are not given an identification between nodes otherwise
-(as for instance with a true, simulated tree sequence and one inferred from its genotypes).
+but are otherwise unclear as to the equivalency of nodes across sequences
+(as for instance with a true, simulated tree sequence and one inferred from its genotypes;
+in the former nodes represent actual ancestral haplotypes, and in the latter
+represent hypothetical ancestors which may or may not resemble the truth).
 Call the two tree sequences $T_1$ and $T_2$, which should have the same genome length
@@
  where we consider the pair of trees which exist at that break point.
-However, suppose that the two tree sequences have the same set of breakpoints
+However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
 between trees,
@@
 define
+% NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
+% \mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]
+% and also in next equation for consistency.
 \begin{align*}
@@
     =
-    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k(n_1) = S(T^{(2)}_k(n_2)} ,
+    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k, n_1) = S(T^{(2)}_k, n_2)}  
+    % NSP-22NOV: Added a missing comma. Also the comma at the very end looks like a prime, which is confusing, so I'm removing grammatical punctuation from these non-inline equations.
 \end{align*}
@@
 \begin{align*}
-    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} ,
+    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} 
 \end{align*}
@@
     =
-    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
+    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) 
 \end{align*}
-where the maximum is over all mappings of nodes in $\T_1$ to nodes in $\T_2$.
+where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$.
 (Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
@@
 \begin{align*}
-    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
+    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) 
 \end{align*}
@@
     =
-    \sum_{n \in N_1} m(n, \alpha(n)) .
+    \sum_{n \in N_1} m(n, \alpha(n)) 
 \end{align*}
@@
 instead of $m(n, \alpha(n))$ define
+% NSP-22NOV: as before I'd suggest using \mathbb{I}[...] for the indicator function
 \begin{align*}
@@
 (which they then show is equal to the number of edges that induce different splits on the labels).
-It would be interesting to define a similar metric on tree sequences,
-perhaps using the Subgraph-Prune-and-Regraft moves used by Singer \citet{deng2024robust}.
+A similar metric on tree sequences could possibly be defined
+using subgraph-prune-and-regraft moves \citet{deng2024robust}.
 
@@
 The mean is computed weighting by node span, so that a dating error
-is more serious for a node with a longer span.
+is more impactful for a node with a longer span.
 
@@
 additionally produces relative values:
-the ancestral recombination graph robinson-foulds value (relative dissimilarity)  and true proportion represented. 
-The \emph{ancestral recombination graph robinson-foulds} (ARF) is defined to be the dissimilarity proportional
+the ancestral recombination graph Robinson-Foulds value (relative dissimilarity) and true proportion represented. 
+The \emph{ARG Robinson-Foulds} (ARF) is defined to be the dissimilarity proportional
  to the total span of nodes in $\T_1$,
@@
 However, following this example, the first lineage might carry a segment $[x, y)$
-that is disjoint from $[a, d)$.
+that is disjoint from the coalescent segment $[a, d)$.
 We call these segments ``isolated non-coalescent segments'';
 they have also been called ``trapped unary spans'' \citep[by][]{XXX}.
-Such segments will not be recovered by our algorithm,
-and it seems likely that most such segments are unrecoverable by any method.
+Such isolated segments will not be recovered by our algorithm,
+and would likely be unrecoverable by any other method.
 So, when evaluating what proportion of the true ARG is represented in an inferred ARG
@@
 which then doubled every generation to reach 512,000 individuals.
-All jobs used for recording runtime were run on an Intel Xeon Gold 6148 processor.
+All jobs for which runtime was recorded were executed on an Intel Xeon Gold 6148 processor.
  Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
@@
 (except those pendant to the root and some of those adjacent to chromosome ends),
-leading to a reduction in number of edges by 1/3.
+leading to a reduction in number of edges by a third.
 Experiments with an earlier version of the algorithm showed that
-if we only extend haplotypes on ``paths of length 1'' like this, 
-then this is achieved for long sequences.
+if we only extend haplotypes on ``paths of length 1'', 
+then the hypothesized reduction of 1/3 is achieved for long sequences. % NSP-22NOV: is this sentence mentioning the earlier algorithm necessary?
 It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
@@
 
-This reduction in edges can also lead to a reduction in computation time.
+This reduction in edges can also lead to a reduction in computation time for
+algorithms using the succinct tree sequence data structure.
 Indeed, \figref{fig:speed_and_edges} shows that computation time is reduced by 10--20\%
 for a typical statistic (here, Tajima's $D$),
-computed in an efficient iterative manner along the genome as implemented in \tskit.
-This may at first be somewhat surprising:
-This makes sense because
-as described in \citet{ralph2020efficiently},
-each edge added or removed requires an update
-to the state of the parent node and nodes above it,
-and extending haplotypes yields a tree sequence with fewer edge removals and insertions.
+computed in an efficient incremental manner along the genome as implemented in \tskit.
+As described in \citet{ralph2020efficiently}, for these incremental algorithms
+the addition or removal of an edge requires updates
+to the state of the parent node and all nodes above it.
+Extending haplotypes yields a tree sequence with fewer edge removals and insertions, 
+and thus requires fewer traversals to the roots.
 
@@
 The method can potentially extend the spans of each node
-(spans on which the node will be unary);
+(additional span over which the node will be unary);
 and we can quantify how much of these extended spans were in the original ARG
@@
 (Simplification does not produce dissimilarity.)
-The bottom row shows the proprtional of the original ARG that is represented in the given ARG (TRP).
-This shows that a large proportion of the ARG can be removed by simplifiction,
+The bottom row shows the proportion of the original ARG that is represented in the given ARG (TRP).
+This shows that a large proportion of the ARG can be removed by simplification,
 indicating that coalescent nodes are unary
@@
 Does this work with \emph{inferred} ARGs?
-Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes;
+Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes; % NSP-22NOV: this isn't true I don't think-- isn't it just tsinfer?
 to what degree are they taking advantage of this information already?
-Here, we study the behavior of \tsinfer{},
-leaving a comparison of different ARG inference methods for future work.
-To study this, we simulated ARGs containing unary-spanning haplotypes (as above),
-then re-inferred ARGs from their genotypes and performed various operations on the results.
+Here, we study the behavior of \tsinfer{}, which produces unary nodes as a
+byproduct of its inference algorithm,
+and leave a comparison across other ARG inference methods for future work.
+We simulated ARGs containing unary-spanning haplotypes (as above), then
+re-inferred ARGs from the associated genotypes and performed various operations
+on the results.
+% NSP-22NOV: do we need to define TS simplification here (or earlier?)
 
@@
 
-We began work on this paper with the simple observation that the simple transformation
+We began this study with the observation that the simple transformation
 of \figref{fig:extending_diagram} would reduce the number of edges
@@
 leads to ARGs that are substantially more compact and faster to operate on,
-and that contain more accurate information about true ancestral relationships.
+and that contain more complete information about true ancestral relationships. % NSP-22NOV: 'accurate' ==> 'complete' as otherwise it sounds like we're changing the topologies
 These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
 Although a number of ARG inference methods may be taking advantage of this information,
-it is our impression that this source of information is not widely appreciated in the community.
-In fact, the field's focus on marginal trees rather than haplotypes
-meant that to study accuracy of this method,
-we had to invent new haplotype-aware measures of (dis)agreement between ARGs.
+it is our impression that this source of information is not widely appreciated.
+In fact, due to the field's focus on marginal trees rather than haplotypes,
+we had to develop a haplotype-aware measure of (dis)agreement between ARGs in order to study the accuracy of the proposed algorithm.
 
@@
 have a substantial portion of correctly inferred unary haplotypes,
-removing this with simplification decreases both ARF
+removing these with simplification decreases both ARF
 (i.e., proportion of ``wrong'' haplotypes)
@@
 that relationships \emph{not} depicted in the ARG are implied to not exist.
-An alternative interpetation
-of the ARG of \figref[A\&B]{fig:conceptual} would be that we do not know
+An alternative interpretation
+of the ARG depicted in \figref[A\&B]{fig:conceptual} would be that we have no information as to 
 how node 2 inherited from node 4 on the right-hand interval,
-rather than saying that it specifically did not pass through node 3.
+rather than saying that the line of transmission specifically did not pass
+through node 3.
 The ``simplification'' algorithm of \citet{kelleher2018efficient},
@@
 each specifically discard information about any such ``non-coalescent'' nodes;
-so the correct interpretation of these ARGs is a lack of knowledge.
-In this paper, we have shown that this information can be largely replaced.
+so for ARGs produced by these algorithms, the correct interpretation is that
+the omission of unary spans reflects a lack of knowledge.
+In this paper, we have shown that, for the most part, this missing information can be imputed.
 
@@
 ``How much of the true relationships does this ARG represent?''
-However, the symmetrized version
+However, it is worth noting that the symmetrized version
 $d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
@@
 where $N$ is the shared set of nodes.
-Then, dissimilarity is the Lesbegue measure of the relative difference of the two sets:
+Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
 $|\T_1 \setminus \T_2|$,
@@
 $|\T_1\Delta\T_2|$,
-which is well-known to be a metric.
+which is well-known to be a metric. % NSP-22NOV: "well-known" ==> give a citation here
 If the two ARGs have the same number of nodes,
 then we take the minimum over the all bijections between their nodes,
-and minima over a finite number of metrics is also a metric.
-This also holds for two tree sequences with different sizes of nodes, $|N_1|\neq|N_2|$ as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
+and minima over a finite number of metrics is also a metric. % NSP-22NOV: I don't understand this sentence, can you rewrite to be clearer?
+This also extends to two tree sequences with different numbers of nodes, $|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
 
@@
 the averaged Robinson-Foulds distance \citep{kelleher2019inferring} % TODO: first usage?
-combines this distance across marginal trees, weighted by span along the genome.
+averages this distance across marginal trees, weighted by span along the genome.
 The method we present here for measuring dissimilarity between topologies of ARGs
@@
 (and separates the metric into two pieces as discussed in the previous paragraph).
-However, the Robinson-Foulds metric has many undesireable properties --
+However, the Robinson-Foulds metric has many undesirable properties --
 for instance, moving a single tip can result in a tree with maximum distance to the original --
-and there is a substantial literature giving generalizations
+and there is a substantial literature giving more robust generalizations
 \citep[reviewed by][]{llabres2021generalized}.
@@
 However, currently not all ARG inference methods return ARGs with unary nodes.
-For instance, we have seen that tsinfer returns ARGs to which the method presented here adds little -
+For instance, we have seen that tsinfer returns ARGs to which the method presented here adds little --
 so, seems to be already taking advantage of this source of information.
-Singer [cite] returns ARGs without any unary nodes,
-although these are reconstructable using the times of the recombination nodes.
-On the other hand, locally building trees separately in each segment of genome
+To our knowledge, methods based on threading (sequentially adding or removing haplotypes) such as
+SINGER \cite{deng2024robust} and ARGNeedle \cite{zhang2023biobankscale} return ARGs without any unary nodes,
+although there is no conceptual issue with incorporating unary haplotypes into threading algorithms.
+On the other hand, locally building trees separately in each segment of genome as is done by Relate \cite{speidel2019method}
 simply does not provide this sort of haplotype-based information (although it might be added using a method like ours).

### 2024-11-25 paper/writeup.tex
edits on edits

@@
     The resulting inferred ARGs are smaller, faster to compute with,
-    and contain substantially more information % NSP-22NOV: "substantially more information" in what regard?
+    and contain substantially more information about inheritance % NSP-22NOV: "substantially more information" in what regard?
     than inferred ARGs without unary nodes.
@@
 depend only on the sequence of marginal trees,
-not on the sharing of ancestral haplotype identities between them.
+not on the sharing of ancestral haplotypes that span across these trees.
 % NSP-22NOV: what does "sharing of ancestral haplotype identities" mean, concretely?
@@
 depend only on the sequence of marginal trees, but not on the ancestral haplotypes that span across these trees" be a clearer description?
+% HF-24NOV: Noted and changed
 Suppose two methods are used to infer ARGs from the same set of genotyped samples.
@@
 only those haplotypes that represent the focal set of genotyped samples % NSP-22NOV: the meaning here isn't clear to me-- what does "represent the focal set of genotyped samples" mean, concretely?
+% HF-24NOV : I am not exactly sure, I will ping Peter
 can be unambiguously said to represent the same thing in both.
 Thus, metrics that only describe these relationships
-ignore the additional information carried by the structure of other haplotypes. % NSP-22NOV: again, "additional information carried by the structure of other haplotypes" is vague
+ignore the additional ancestral information carried by the structure of other haplotypes. % NSP-22NOV: again, "additional information carried by the structure of other haplotypes" is vague
+% HF-24NOV : Heard
 For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
@@
 These measures show that the vast majority of these extended haplotypes are correct, % NSP-22NOV: "vast majority" is correct in simulated (but not inferred) ARGs, right?
+% HF-24NOV : It should be both, as the IE line in Figure 6D shows that inferred trees match the TPR of the ground truth by at least 60% for inferred trees.
 and that substantial information about haplotypes is contained in these nodes
@@
 that is incorrect (i.e., not present in the original tree sequence) % NSP-22NOV: it seems a bit odd that this panel (B) is not on a log scale while all the others are
+% HF-24NOV Peter and I looked at both options and we thought the log scale on the Y-axis looked a bit misleading against our method.
+We instead provide our suplemental figure to provide more information about Figure 5b
 over the total amount of span added after running \algorithmref{alg:extend}.
@@
 were correctly, and incorrectly, extended.
-They also tell us how much of the relationships in the original tree sequence
-are represented in the result. % NSP-22NOV: "relationships" of what exactly?
-
+They also tell us how the node spans represented in the result compare to those in the original tree sequence. % NSP-22NOV: "relationships" of what exactly?
+% HF -24NOV: Noted and changed
 Now suppose that instead of comparing two tree sequences with the same set of nodes,
@@
 % NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
-% \mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]
+% $\mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]$
 % and also in next equation for consistency.
+% HF24-NOV: I think I disagree but we can get a third opinion.
 \begin{align*}
@@
     =
-    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k, n_1) = S(T^{(2)}_k, n_2)}  
+    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k,\ n_1) = S(T^{(2)}_k,\ n_2)}  
     % NSP-22NOV: Added a missing comma. Also the comma at the very end looks like a prime, which is confusing, so I'm removing grammatical punctuation from these non-inline equations.
+    % HF-24NOV: Noted
 \end{align*}
@@
 \begin{align*}
-    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k, n_2) = s(T^{(1)}_k, n)} ,
+    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k,\ n_2) = s(T^{(1)}_k,\ n)} 
 \end{align*}
@@
 then the hypothesized reduction of 1/3 is achieved for long sequences. % NSP-22NOV: is this sentence mentioning the earlier algorithm necessary?
+% HF-24NOV: No, but I kind of like the homage to our own earlier work and it sort of makes a benchmark for edge reduction proportions.
 It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
@@
 Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes; % NSP-22NOV: this isn't true I don't think-- isn't it just tsinfer?
+% HF-24NOV: I am unsure. 
 to what degree are they taking advantage of this information already?
@@
 % NSP-22NOV: do we need to define TS simplification here (or earlier?)
-
+% HF-24NOV: We define in the previous section line 960.
 First, \figref{fig:dissimilarity} shows that \tsinfer{}
@@
 and that contain more complete information about true ancestral relationships. % NSP-22NOV: 'accurate' ==> 'complete' as otherwise it sounds like we're changing the topologies
+% HF-24NOV: Heard.
 These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
@@
 which is well-known to be a metric. % NSP-22NOV: "well-known" ==> give a citation here
+% HF-24NOV: Peter would we cite Rudin?
 If the two ARGs have the same number of nodes,
-then we take the minimum over the all bijections between their nodes,
-and minima over a finite number of metrics is also a metric. % NSP-22NOV: I don't understand this sentence, can you rewrite to be clearer?
+we can consider all bijections between their nodes.
+The symmetric difference between $\T_1$ and $\T_2$,
+related through each bijection is a metric.
+Then the minimum over all such metrics will still be a metric
+since the minimum over a finite number of metrics is also a metric.
+ % NSP-22NOV: I don't understand this sentence, can you rewrite to be clearer?
+ % HF-24NOV: Hopefully it is clearer now
 This also extends to two tree sequences with different numbers of nodes, $|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.

### 2024-11-26 paper/writeup.tex
Update paper/writeup.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>

@@
 as implemented in msprime \citep{kelleher2016efficient},
-each specifically discard information about any such ``non-coalescent'' nodes;
+each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;
 so for ARGs produced by these algorithms, the correct interpretation is that

### 2024-11-26 paper/writeup.tex
Update paper/writeup.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>

@@
 Does this work with \emph{inferred} ARGs?
-Indeed, most ARG inference methods provide inferred ARGS already containing unary nodes; % NSP-22NOV: this isn't true I don't think-- isn't it just tsinfer?
-% HF-24NOV: I am unsure. 
+Indeed, \tsinfer{} provides inferred ARGS already containing unary nodes;
 to what degree are they taking advantage of this information already?

### 2024-11-26 paper/writeup.tex
Update paper/writeup.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>

@@
 but account for haplotype identity.
-These measures show that the vast majority of these extended haplotypes are correct, % NSP-22NOV: "vast majority" is correct in simulated (but not inferred) ARGs, right?
-% HF-24NOV : It should be both, as the IE line in Figure 6D shows that inferred trees match the TPR of the ground truth by at least 60% for inferred trees.
+These measures show that the vast majority of these extended haplotypes are correct if the trees are correct,
 and that substantial information about haplotypes is contained in these nodes

### 2024-11-26 paper/writeup.tex
Update paper/writeup.tex
Co-authored-by: Peter Ralph <petrel.harp@gmail.com>
edits^3

@@
     are not coalescence events.
-    We show how these non-coalescing haplotypes (``unary nodes'') may be
+    We show how these non-coalescing haplotypes (``unary regions'') may be
     inserted into the ancestral recombination graph (ARG), a compact
@@
 define
+\begin{align*}
+	\mathcal{M}(\T_1, n1; \T_2, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
+\end{align*}
+to be the subset of all places $k$ where $n_1$ and $n_2$ share the same sample set in $T_k^{(1)}$ and $T_k^{(2)}$, respectively.
+Further, we define
 % NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
@@
     =
-    \sum_{k=1}^N (a_k - a_{k-1}) \ind_{S(T^{(1)}_k,\ n_1) = S(T^{(2)}_k,\ n_2)}  
-    % NSP-22NOV: Added a missing comma. Also the comma at the very end looks like a prime, which is confusing, so I'm removing grammatical punctuation from these non-inline equations.
-    % HF-24NOV: Noted
+    \sum_{k\in \mathcal{M}(\T_1, n1; \T_2, n2)} (a_k - a_{k-1}) ,  
 \end{align*}
@@
 \begin{align*}
-    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} 
+    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
 \end{align*}
@@
     =
-    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) 
+    \max_{\beta:N_1 \to N_2} \sum_{n \in N_1} m(n, \beta(n)) ,
 \end{align*}
-where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$.
+where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
 (Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
@@
 \begin{align*}
-    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) 
+    \alpha(n_1) = \argmax_{n_2 \in N_2} m(n_1, n_2) ,
 \end{align*}
@@
     =
-    \sum_{n \in N_1} m(n, \alpha(n)) 
+    \sum_{n \in N_1} m(n, \alpha(n)) .
 \end{align*}
@@
 \end{align*}
-the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
+as the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
 matches $s(T_1,n)$.
@@
 A similar metric on tree sequences could possibly be defined
-using subgraph-prune-and-regraft moves \citet{deng2024robust}.
+using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.
 
@@
 $|\T_1\Delta\T_2|$,
-which is well-known to be a metric. % NSP-22NOV: "well-known" ==> give a citation here
-% HF-24NOV: Peter would we cite Rudin?
+which is well-known to be a metric \citep{rudin1976principles}. 
 If the two ARGs have the same number of nodes,
@@
 
-%\begin{figure}
-%    \caption{
-%        Supplementary figure: runtime of discrepancy function as a function of number of trees
-%        (which we modulate by changing sequence length).
-%        \label{fig:speed_discrepancy}
-%    }
-%\end{figure}
-
 \begin{table}[!hbt]

### 2024-11-28 paper/writeup.tex
added missing caption
peter edits to the edits

@@
     can in many cases be recovered from inference based solely on polymorphic markers.
-    % NSP-22NOV: "unary node" and "ARG" are not defined so I tried to provide better context here
     In a genealogical context, this information reflects the persistence
-    of ancestral haplotypes across marginal genealogical trees where they
-    are not coalescence events.
+    of ancestral haplotypes across marginal genealogical trees in which they
+    do not represent coalescences.
     We show how these non-coalescing haplotypes (``unary regions'') may be
@@
     relationships among recombinant sequences.
-    The resulting inferred ARGs are smaller, faster to compute with,
-    and contain substantially more information about inheritance % NSP-22NOV: "substantially more information" in what regard?
-    than inferred ARGs without unary nodes.
-    We provide efficient algorithms to infer unary nodes within existing ARGs,
-    new metrics of agreement and disagreement between inferred ARGs,
+    The resulting ARGs are smaller, faster to compute with,
+    and the additional ancestral information that is inserted
+    is nearly always correct where the initial ARG is correct.
+    We provide efficient algorithms to infer unary regions within existing ARGs,
     and explore some consequences of unary nodes for ARGs inferred from real data.
+    To do this, we introduce new metrics of agreement and disagreement between inferred ARGs
+    that, unlike previous methods, consider the ARG as describing relationships between haplotypes
+    rather than just a collection of trees.
 \end{abstract}
@@
 
-Haplotype identity has been largely overlooked in the literature on ARG inference so far --
-in that the metrics that have been used to measure accuracy of inferred ARGs
+Haplotype identity has been largely overlooked in the literature on ARG inference --
+to our knowledge, all metrics that have been used so far to measure accuracy of inferred ARGs
 depend only on the sequence of marginal trees,
-not on the sharing of ancestral haplotypes that span across these trees.
-% NSP-22NOV: what does "sharing of ancestral haplotype identities" mean, concretely?
-Would "...
-depend only on the sequence of marginal trees, but not on the ancestral haplotypes that span across these trees" be a clearer description?
-% HF-24NOV: Noted and changed
-Suppose two methods are used to infer ARGs from the same set of genotyped samples.
-Although both ARGs describe the relationships between many haplotypes,
-only those haplotypes that represent the focal set of genotyped samples % NSP-22NOV: the meaning here isn't clear to me-- what does "represent the focal set of genotyped samples" mean, concretely?
-% HF-24NOV : I am not exactly sure, I will ping Peter
-can be unambiguously said to represent the same thing in both.
-Thus, metrics that only describe these relationships
-ignore the additional ancestral information carried by the structure of other haplotypes. % NSP-22NOV: again, "additional information carried by the structure of other haplotypes" is vague
-% HF-24NOV : Heard
+not on how ancestral haplotypes span across these trees.
 For instance, \citet{kelleher2019inferring} and \citet{zhang2023biobankscale}
@@
 two ARGs could be identical by either measure
-but have completely different identity by descent (IBD) segment structure
-(as defined in the context of the ARG, \citep{XXX}).
+but have completely different identity by descent (IBD) segment structure.
+% (as defined in the context of the ARG, \citep{XXX}).
 \citet{deng2021distribution} evaluated agreement of distributions of
@@
 neither of which measure the sharing of haplotypes between adjacent trees.
+However, the additional information provided by haplotype structure can be important:
+for instance, haplotypes that extend over many marginal genealogies
+``tie together'' those genealogies, allowing estimates of times of particular ancestors
+to be informed by larger portions of the genome on which there are many genealogies.
 
@@
 \algorithmref{alg:extend} simplifies the algorithm in several ways for the sake of clarity --
-for instance, the bookeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
+for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
 Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
@@
 for each node, it gives the amount of span
-that is incorrect (i.e., not present in the original tree sequence) % NSP-22NOV: it seems a bit odd that this panel (B) is not on a log scale while all the others are
-% HF-24NOV Peter and I looked at both options and we thought the log scale on the Y-axis looked a bit misleading against our method.
-We instead provide our suplemental figure to provide more information about Figure 5b
+that is incorrect (i.e., not present in the original tree sequence)
 over the total amount of span added after running \algorithmref{alg:extend}.
@@
 were correctly, and incorrectly, extended.
-They also tell us how the node spans represented in the result compare to those in the original tree sequence. % NSP-22NOV: "relationships" of what exactly?
-% HF -24NOV: Noted and changed
+They also tell us how the node spans represented in the result compare to those in the original tree sequence.
 Now suppose that instead of comparing two tree sequences with the same set of nodes,
@@
 \end{align*}
-to be the subset of all places $k$ where $n_1$ and $n_2$ share the same sample set in $T_k^{(1)}$ and $T_k^{(2)}$, respectively.
+to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
 Further, we define
-% NSP-22NOV: The notation here is hard to read when subscripted. IMO it'd be better to use bracket notation here for the indicator function,
-% $\mathbb{I}[ S(T_k^{(1)}, n_1) = S(T_k^{(2)}, n_2) ]$
-% and also in next equation for consistency.
-% HF24-NOV: I think I disagree but we can get a third opinion.
 \begin{align*}
@@
     =
-    \sum_{k\in \mathcal{M}(\T_1, n1; \T_2, n2)} (a_k - a_{k-1}) ,  
+    \sum_{k\in \mathcal{M}(\T_1, n_1; \T_2, n_2)} (a_k - a_{k-1}) ,  
 \end{align*}
@@
 matches the samples below $n_2$ in $\T_2$.
-Also let
+Also let $s(\T,n)$ denote
+the total span that node $n$ is present in the marginal trees,
 \begin{align*}
@@
 \end{align*}
-the total span that node $n$ is present in the marginal trees.
+where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
 The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
@@
 (Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
-and that there may be no nodes in $\T_1$ that are mapped to some nodes in $\T_2$.)
+and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
 Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
@@
 
-Other authors \citep[e.g.,]{kelleher2019inferring} have used
+Other authors \citep[e.g.,][]{kelleher2019inferring} have used
 an average Robinson-Foulds distance \citep{robinson1981comparison}
@@
 instead of $m(n, \alpha(n))$ define
-% NSP-22NOV: as before I'd suggest using \mathbb{I}[...] for the indicator function
 \begin{align*}
-    m'(n, \T_2) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{\exists n_2: s(T^{(2)}_k,\ n_2) = s(T^{(1)}_k,\ n)} 
+    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
 \end{align*}
-as the total span over which there is \emph{some} node in $\T_2$ whose set of descendant samples
-matches $s(T_1,n)$.
+the indices of all trees on which there is \emph{some} node in $\T_2$
+whose set of descendant samples matches those of $n$, and
+\begin{align*}
+    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)1}^N (a_k - a_{k-1})
+\end{align*}
+the total span over which $n$ finds a match.
 Then the average Robinson-Foulds distance (averaged over locations in the genome)
@@
 (which they then show is equal to the number of edges that induce different splits on the labels).
-A similar metric on tree sequences could possibly be defined
+A similar metric on tree sequences could be defined
 using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.
@@
 the ancestral recombination graph Robinson-Foulds value (relative dissimilarity) and true proportion represented. 
-The \emph{ARG Robinson-Foulds} (ARF) is defined to be the dissimilarity proportional
+The \emph{ARG Robinson-Foulds} (which we call ``ARF'') is defined to be the dissimilarity proportional
  to the total span of nodes in $\T_1$,
@@
 option \texttt{coalescing\_segments\_only} set to False.
-Although \msprime\ simulates many events that do not create a coalescence in some marginal tree,
+Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
 it only outputs information for nodes which contain a coalescence
@@
 Said another way, by default all ancestral nodes in the ARG
-output by \msprime\ are the MRCA of some pair of samples at every point in the genome
+output by \msprime{} are the MRCA of some pair of samples at every point in the genome
 on which they are represented.
@@
 However, following this example, the first lineage might carry a segment $[x, y)$
-that is disjoint from the coalescent segment $[a, d)$.
+that is disjoint from the segment $[a, d)$.
 We call these segments ``isolated non-coalescent segments'';
@@
 if we only extend haplotypes on ``paths of length 1'', 
-then the hypothesized reduction of 1/3 is achieved for long sequences. % NSP-22NOV: is this sentence mentioning the earlier algorithm necessary?
-% HF-24NOV: No, but I kind of like the homage to our own earlier work and it sort of makes a benchmark for edge reduction proportions.
+then the hypothesized reduction of 1/3 is achieved for long sequences.
 It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
@@
 on the results.
-% NSP-22NOV: do we need to define TS simplification here (or earlier?)
-% HF-24NOV: We define in the previous section line 960.
 First, \figref{fig:dissimilarity} shows that \tsinfer{}
@@
 leads to ARGs that are substantially more compact and faster to operate on,
-and that contain more complete information about true ancestral relationships. % NSP-22NOV: 'accurate' ==> 'complete' as otherwise it sounds like we're changing the topologies
-% HF-24NOV: Heard.
+and that contain more complete information about true ancestral relationships.
 These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
@@
 or the Hudson algorithm for coalescent simulation
-as implemented in msprime \citep{kelleher2016efficient},
+as implemented in \msprime{} \citep{kelleher2016efficient},
 each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;
@@
 since the minimum over a finite number of metrics is also a metric.
- % NSP-22NOV: I don't understand this sentence, can you rewrite to be clearer?
- % HF-24NOV: Hopefully it is clearer now
-This also extends to two tree sequences with different numbers of nodes, $|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
+This also extends to two tree sequences with different numbers of nodes,
+$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
 
@@
 To our knowledge, methods based on threading (sequentially adding or removing haplotypes) such as
-SINGER \cite{deng2024robust} and ARGNeedle \cite{zhang2023biobankscale} return ARGs without any unary nodes,
-although there is no conceptual issue with incorporating unary haplotypes into threading algorithms.
+SINGER \citep{deng2024robust} and ARGNeedle \citep{zhang2023biobankscale} return ARGs without any unary nodes,
+although there is no conceptual issue with incorporating unary haplotypes into threading algorithms,
+and at least SINGER contains information about unary nodes in its internal format.
 On the other hand, locally building trees separately in each segment of genome as is done by Relate \cite{speidel2019method}
@@
         see Methods for details.
-        Absolute values are shown in Supplementary \figref{fig:speed_and_edges_constant_other}.
+        Absolute values are shown in Supplementary \figref{fig:speed_and_edges_other_constant}.
         \label{fig:speed_and_edges_constant}

### 2024-11-28 paper/writeup.tex
citations

@@
 was inherited by the focal genomes.
-This is reflected in methodology of some ARG inference methods \citep[e.g., Relate][]{speidel2019method},
-in metrics used to assess inference accuracy \citep{XXX},
+This is reflected in methodology of some ARG inference methods
+and in metrics used to assess inference accuracy,
 as well as in basic terminology.
@@
 and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
-\tskit \citep{XXX}.
+\tskit{} \citep{tskit}.
 
@@
 We call these segments ``isolated non-coalescent segments'';
-they have also been called ``trapped unary spans'' \citep[by][]{XXX}.
+they have also been called ``trapped unary spans'' \citep[by][]{wong2024general}.
 Such isolated segments will not be recovered by our algorithm,
@@
 To include the effects of heterogeneous recombination rate,
-in some we used stdpopsim \citep{stdpopsim} 
+in some we used stdpopsim \citep{adrion2020stdpopsim} 
 to simulate chromosome 1 of \textit{Canis familiaris}
-using the CanFam3 genetic map from \citet{campell2016}.
+using the CanFam3 genetic map from \citet{campbell2016}.
 \emph{``Constant dog''} simulations simulated this chromosome in a 

### 2024-11-28 paper/writeup.tex
peter edits through p6

@@
 \newcommand{\comment}[1]{{\color{violet} \it #1}}
+\newcommand{\pluseq}{\mathrel{{+}{=}}}
 
@@
 
-\keywords{genealogy, tree sequence, haplotypes}
+\keywords{genealogy, tree sequence, haplotypes, ancestral recombination graph}
 
@@
     We provide efficient algorithms to infer unary regions within existing ARGs,
-    and explore some consequences of unary nodes for ARGs inferred from real data.
+    and explore some consequences for ARGs inferred from real data.
     To do this, we introduce new metrics of agreement and disagreement between inferred ARGs
@@
 in the problem of ``ARG inference'',
-which seeks to infer (portions of) the ``ancestral recombination graph'' (or, ARG)
+which seeks to infer (portions of) the ancestral recombination graph (or, ARG)
 that describes how a set of genotypes samples are related to each other
 at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
-One way of viewing the ARG is as a sequence of ``marginal trees'',
+One way of viewing the ARG is as a sequence of marginal trees,
 i.e., the genealogical trees that describe how each portion of the genome
@@
 and is seeing wide use thanks in part to its efficiency and accompanying reliable toolkit,
-\tskit{} \citep{tskit}.
+\tskit{} \citep{tskit,ralph2020efficiently}.
 
@@
 For instance, two sequences that share a derived mutation
-are (probably) more closely related over some span of genome surrounding the location where the mutation occurs.
+are probably more closely related over some span of genome surrounding the location where the mutation occurs.
 It has long been observed 
@@
 would be inherited as Mendelian markers
-\citet{fisher1954fuller,chapman2003model}.
+\citep{fisher1954fuller,chapman2003model}.
 In more modern terminology, 
@@
 two ARGs could be identical by either measure
-but have completely different identity by descent (IBD) segment structure.
+but imply completely different patterns of haplotype sharing and inheritance.
 % (as defined in the context of the ARG, \citep{XXX}).
@@
 First, we describe a deterministic algorithm that
-extends the span over which ancestral haplotypes extend
+extends the genomic region spanned by ancestral haplotypes
 using the principle that intermediate nodes in inheritance paths
 should remain unchanged when possible.
-These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
+These extended portions of ancestral haplotypes manifest as unary nodes in the marginal trees.
 To quantify how accurate the new information is,
@@
 
-Consider the (small portion) of a hypothetical tree sequence in \figref[A]{fig:conceptual}.
+Consider the (small portion) of a hypothetical ARG in \figref[A]{fig:conceptual}.
 On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
@@
 and so node 3 has become unary in this tree.
-Note that the more parsimonious tree sequence also includes fewer edges:
+Note that the more parsimonious ARG also includes fewer edges:
 the three distinct edges $4 \to 3$, $3 \to 1$, and $4 \to 1$ in \figref[B]{fig:conceptual}
@@
         (described in more detail in the text):
-        \textbf{(A)} a small portion of a tree sequence without unary nodes;
+        \textbf{(A)} a small portion of a ARG without unary nodes;
         \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
@@
 \end{figure}
-
-%\subsection{Extend Edges}
-%	\comment{Should we keep a brief section about the original extend edges? If yes, some notes are below.}
-%    We do so by identifying short paths (2 edges) in a tree.
-%    If neighboring trees contain edges with 
-%    identical ancestor (parent) and descendant (child) nodes, 
-%    we assume that there is implied coalescence of haplotypes between nodes, 
-%    and we will extend the path's edges to the neighboring tree \ref{fig:extending_diagram}.
-%  	In more concrete terms, for a tree $T_k$ in the tree sequence $\T$ 
-%    suppose there exists a path which contains a node $3$,
-%    with parent and child nodes $4$ and $0$, respectively.
-%    If there is an edge between the nodes $4$ and $0$ in trees 
-%    $T_{k+1}$ (or $T_{k-1}$),
-%    then we wish to extend the edges $4\to 3$ and $3\to 0$ 
-%    into tree $T_{k+1}$ and then remove the edge $4 \to 0$ from $T_{k+1}$. 
-%    This action reduces the length of $4 \to 0$ on the genome,
-%    and in some cases, completely removes the edge. 
-%    We now perform this action on all such ancestral paths 
-%    over the entire tree sequence.
-%    With algorithm \ref{alg:edge} we only extend existing edges
-%    \comment{Note possible change pending algorithm change}
-%    over a larger interval on the genome,
-%    and remove unnecessary edges in the process.
 
@@
 which is the amount of time in the past that the individual who carried that segment of genome lived.
-Some nodes are \emph{samples}, meaning that we know them from data. 
-Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
+Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
+% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
 Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
@@
 and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
+Nodes represent (portions of) ancestral haplotypes, and so
+we will use the terms interchangeably.
 Not all nodes appear in each tree,
@@
 for instance, if $u_i$ is already represented somewhere else in $T_{k+1}$,
-or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$,
-for which the intermediate nodes are not in the other tree.
+or if $t_{u_i} = t_{v_j}$, for some $i$ and $j$.
 So, we restrict our attention to pairs of such paths in adjacent trees
@@
 \caption{A visualization of the \textit{extend haplotypes} method.
-Given two neighboring trees $T_1$ and $T_2$, there exists paths from root node $6$ to sample node $0$.
+    In both trees $T_1$ and $T_2$,
+    node 0 inherits from node 6, the root:
     $T_1$ contains the path $6\to 4\to 0$ while $T_2$ has path $6\to 5\to 0$.
-The intermediate nodes $4$ and $5$ do not appear in $T_2$ and $T_1$ respectively, and so we consider these two paths \textit{mergeable}.
-The extend haplotypes method joins these two paths, constructing the merged path $6\to 5\to 4\to 0$ for both $T_1$ and $T_2$.
-This new path extends the edge $4\to 0$ to $T_2$ as well as constructs a new edge $5\to 4$ in the tree sequence.
+    The intermediate nodes $4$ and $5$ do not appear in $T_2$ and $T_1$ respectively, and so the paths are \textit{mergeable}.
+    The ``extend haplotypes'' method joins these two paths,
+    inserting the merged path $6\to 5\to 4\to 0$ into both $T_1$ and $T_2$.
+    % This new path extends the edge $4\to 0$ to $T_2$ as well as constructs a new edge $5\to 4$ in the tree sequence.
     \label{fig:extending_diagram}
@@
 The key step in the algorithm is to determine whether the pair of paths
-that terminate in a given node in two adjacent trees are mergeable,
-\algorithmref{alg:mergeable}, which works as follows.
+that terminate in a given node in two adjacent trees are mergeable.
+The algorithm we use to do this is given as \algorithmref{alg:mergeable},
+and works as follows.
 If a pair of paths is mergeable,
@@
 
-\algorithmref{alg:extend} simplifies the algorithm in several ways for the sake of clarity --
+\algorithmref{alg:extend} simplifies the algorithm in \tskit{}
+in several ways for the sake of clarity --
 for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
@@
 some edges that begin at $a_k$ will be postponed to begin at $a_{k+1}$,
-and some entirely new edges may be added, see \figref{fig:extending_diagram}.
+and some entirely new edges may be added, as in \figref{fig:extending_diagram}.
 Furthermore, the trees $T_k$ and $T_{k+1}$ (and corresponding forests $F_O$ and $F_I$)
@@
         that would be added by extending the path from $T_O$ to $T_I$;
-        otherwise, returns $\infty$.
+        otherwise, this returns $\infty$.
         Let $P_O[n]$ and $P_I[n]$ be the parents of node $n$
@@
         (i.e., in $F_O$ and $F_I$).
+        The variable $m_e$ will record the number of new edges to be added,
+        and $m$ will record the number of extended haplotypes.
     }
@@
     \Fn{\mergeable($c$, $T_O$, $T_I$)}{
-        $p_i = P_I[c]$\;
-        $t_i = t[p_i]$\;
-        $p_o = P_O[c]$\;
-        $t_o = t[p_o]$\;
-        $n_e = n = 0$\;
+        \text{Let } $p_i = P_I[c]$, 
+        $t_i = t[p_i]$,
+        $p_o = P_O[c]$,
+        $t_o = t[p_o]$,
+        \text{ and }$m_e = m = 0$.\;
         \While{\tn{True}}{
-            $y_i = (p_i \neq NULL)
-                \text{ and } (p_i \notin T_O)
-                \text{ and } (t_i < t_o)$ \;
-            $y_o = (p_o \neq NULL)
-                \text{ and } (p_o \notin T_I)
-                \text{ and } (t_o < t_i)$ \;
+            \text{Set } $y_i = (p_i \neq NULL)
+                \& (p_i \notin T_O)
+                \& (t_i < t_o)$ \;
+            \text{and } $y_o = (p_o \neq NULL)
+                \& (p_o \notin T_I)
+                \& (t_o < t_i)$ \;
             \If{not ($y_i$ or $y_o$)}{ \Break\; }
@@
                 \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
-                    $n_e += 1$\;
-                }
-                $c = p_i$\;
-                $p_i = P_I[p_i]$\;
-                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$\;
+                    \text{Set } $m_e = m_e + 1$.\;
+                }
+                \text{Set } $c = p_i$,
+                $p_i = P_I[p_i]$, \text{ and}\;
+                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
             }{
                 \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
-                    $n_e += 1$\;
-                }
-                $c = p_o$\;
-                $p_o = P_O[p_o]$\;
-                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$\;
-                $n += 1$\;
-            }
-        }
-        \If{$n = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
-            $n_e = \infty$\;
-        }
-        \KwRet{$n_e$}\;
+                    \text{Set } $m_e = m_e + 1$.\;
+                }
+                \text{Set }$c = p_o$,
+                $p_o = P_O[p_o]$, \;
+                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
+                \text{and } $m = m + 1$ .\;
+            }
+        }
+        \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
+            \text{Set } $m_e = \infty$. \;
+        }
+        \KwRet{$m_e$}\;
     }
@@
         \For{$k$ in $1 \ldots N$}{
-            $M = 0$\;
-            $M' = \infty$\;
+            \text{Set } $M = 0$ \text{ and }
+            $M' = \infty$.\;
             \While{$M < \infty$}{
                 \For{$e \in O_k$}{
-                    $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$)\;
+                    \text{Set } $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\;
                     \eIf{$m < M$}{
@@
                     }{
-                        $M' = \min(m, M)$\;
-                    }
-                }
-                $M = M'$\;
-                $M' = \infty$\;
+                        \text{Set } $M' = \min(m, M)$. \;
+                    }
+                }
+                \text{Set } $M = M'$ \text{ and }
+                $M' = \infty$ .\;
             }
@@
 an important practical question for this algorithm is:
-how passes do we need to do?
+how many passes do we need to do?
 The algorithm is monotone (spans of ancestral nodes only increase),
-so it is guaranteed to require only a finite number of passes,
+so it is guaranteed to terminate in a finite number of passes,
 but it is also not hard to construct pathological cases that require an arbitrary number of passes.
-Counting edges for every iteration suggest that only at most five iterations are needed before the algorithm terminates.
-Indeed, for even larger sequences \tabref{tab:edge-counts}
+However, experimentation suggests that 
+in practice only at most five iterations are needed before the algorithm terminates.
+Indeed, for even large sequences \tabref{tab:edge-counts}
 shows that 99\% of all changes to an ARG occur after the first iteration,
-with the algorithm completing after four iterations.
+with the algorithm always completing after four iterations.
 
@@
 However, we are also interested in whether \algorithmref{alg:extend}
-improves \emph{inferred} tree sequences.
-Since we are not aware of any current methods for measuring (dis)agreement between tree sequences
+improves \emph{inferred} ARG.
+Since we are not aware of any current methods for measuring (dis)agreement between ARGs
 that take into account haplotype identity,
@@
 It is helpful to first describe what we compute in the simple case.
-\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
-for each node the total span in a ``true'' tree sequence, 
-given by the line,
-that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
-in a way that included spans over which nodes are unary in the output
-(by setting \texttt{coalescing\_segments\_only} to False).
-\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
-\citep{kelleher2018efficient},
-which removes all unary nodes from all trees
-thus reducing the node spans for some nodes in the sequence.
-In \figref[D]{fig:node-spans}
-is the span of the node after applying \algorithmref{alg:extend}
-to the simplified tree sequence.
-\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
-from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
-\figref[B]{fig:node-spans} shows how much of this span is incorrect:
-for each node, it gives the amount of span
-that is incorrect (i.e., not present in the original tree sequence)
-over the total amount of span added after running \algorithmref{alg:extend}.
-The values in these plots
-therefore quantify how much the ancestral haplotypes represented by each node
-were correctly, and incorrectly, extended.
-They also tell us how the node spans represented in the result compare to those in the original tree sequence.
+We will first simulate tree sequences in such a way that
+the unary regions of coalescent haplotypes are retained.
+Then, each node is present in both tree sequences,
+and we can quantify, for each node, how much of their span is correct or incorrect
+by comparing to the original, true tree sequence.
+
 Now suppose that instead of comparing two tree sequences with the same set of nodes,
 we wish to compare two tree sequences for which we know the sample nodes are the same
-but are otherwise unclear as to the equivalency of nodes across sequences
-(as for instance with a true, simulated tree sequence and one inferred from its genotypes;
-in the former nodes represent actual ancestral haplotypes, and in the latter
-represent hypothetical ancestors which may or may not resemble the truth).
-Call the two tree sequences $T_1$ and $T_2$, which should have the same genome length
+but are otherwise unclear as to the equivalency of nodes across sequences.
+(For instance, with a simulated tree sequence and one inferred from its genotypes;
+nodes in the former represent actual ancestral haplotypes, and in the latter
+represent hypothetical ancestors which may or may not resemble the truth.)
+Call the two tree sequences $\T_1$ and $\T_2$, which should have the same genome length
 and the same set of sample nodes.
-We would like to measure (a) how much of $T_1$ is found in $T_2$;
-(b) how much of $T_1$ is \emph{not} found in $T_2$; and
-(c) how much of $T_2$ is not found in $T_1$.
+We would like to measure (a) how much of $\T_1$ is found in $\T_2$;
+(b) how much of $\T_1$ is \emph{not} found in $\T_2$; and
+(c) how much of $\T_2$ is not found in $\T_1$.
 (Think of these three quantities as the size of the intersection
@@
 Concretely, our method can be applied to any two tree sequences,
-as in the case of \figref{fig:conceptual_discrepancy},
-where we consider the pair of trees which exist at that break point.
+as illustrated in \figref{fig:conceptual_discrepancy}.
 However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
@@
 \begin{align*}
-	\mathcal{M}(\T_1, n1; \T_2, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
+	\mathcal{M}(n1, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
 \end{align*}
@@
     =
-    \sum_{k\in \mathcal{M}(\T_1, n_1; \T_2, n_2)} (a_k - a_{k-1}) ,  
+    \sum_{k\in \mathcal{M}(n_1, n_2)} (a_k - a_{k-1}) ,  
 \end{align*}
@@
 matches the samples below $n_2$ in $\T_2$.
-Also let $s(\T,n)$ denote
-the total span that node $n$ is present in the marginal trees,
-\begin{align*}
-    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
-\end{align*}
-where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
 The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
@@
 Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
-it's \emph{best matching node} in $\T_2$ as
+its' \emph{best matching node} in $\T_2$ as
 \begin{align*}
@@
 \end{align*}
-If the best-matching node is unique, we define $\alpha(n_1)$ to be the node in $T_2$
+If the best-matching node is not unique, we define $\alpha(n_1)$ to be the node in $T_2$
 out of those maximizing $m(n_1, n_2)$ that minimizes $|t^{(1)}_{n_1} - t^{(2)}_{n_2}|$
@@
 however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
-Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
+Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$,
+and let $s(\T,n)$ denote
+the total span that node $n$ is present in the marginal trees,
+\begin{align*}
+    s(\T, n) = \sum_{k=1}^N (a_k - a_{k-1}) \ind_{n \in T_k} \quad ,
+\end{align*}
+where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
 We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
@@
 \end{align*}
-which is the proportion of total span of all nodes in $\T_1$
-for which their descendant samples do \emph{not} match those of their best match in $\T_2$.
+which is the total span for all nodes in $\T_1$
+over which their descendant samples do \emph{not} match those of their best match in $\T_2$.
 
@@
 
-To do this, we simulate ARGs containing full haplotypes using \msprime's 
-option \texttt{coalescing\_segments\_only} set to False.
+To do this, we simulate ARGs containing full haplotypes using \msprime \citep{kelleher2016efficient,baumdicker2021efficient},
+with the \texttt{coalescing\_segments\_only} option set to False.
 Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
@@
 
-% Outline:
-% 1. benchmarking: edge reduction and runtime
-% 2. correctness:
-
-%% TO DO:
-%% 1. Discuss trapped unary spans more
-%% 2. Discuss the comparison between tsinfer and extend haplotypes
-%% 3. Add more figures 
-%%	- (Edge counts between tsinfer and extend haplotypes)
-%%	- Average Edge counts per iterations
-%% 4. Discuss figures about total edge span added in both recombination and gene conversion events
-%% 5. Discuss the figures on edge reduction (include #3 part 1.)
-
 %%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Accuracy with true trees}
+
+\comment{Moved from above: don't need?}
+\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
+for each node the total span in a ``true'' tree sequence, 
+given by the line,
+that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
+in a way that included spans over which nodes are unary in the output
+(by setting \texttt{coalescing\_segments\_only} to False).
+\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
+\citep{kelleher2018efficient},
+which removes all unary nodes from all trees
+thus reducing the node spans for some nodes in the sequence.
+In \figref[D]{fig:node-spans}
+is the span of the node after applying \algorithmref{alg:extend}
+to the simplified tree sequence.
+\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
+from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
+\figref[B]{fig:node-spans} shows how much of this span is incorrect:
+for each node, it gives the amount of span
+that is incorrect (i.e., not present in the original tree sequence)
+over the total amount of span added after running \algorithmref{alg:extend}.
+The values in these plots
+therefore quantify how much the ancestral haplotypes represented by each node
+were correctly, and incorrectly, extended.
+They also tell us how the node spans represented in the result compare to those in the original tree sequence.
 
@@
 essentially counts the number of differing branches between two trees;
-the averaged Robinson-Foulds distance \citep{kelleher2019inferring} % TODO: first usage?
+the averaged Robinson-Foulds distance \citep{kelleher2019inferring}
 averages this distance across marginal trees, weighted by span along the genome.
@@
 it would be useful to extend our dissimilarity to include branch lengths.
+
+\paragraph{Parsimony}
+Much of the early work on ARG inference aimed to extract as much information as possible
+out of the small datasets of the time,
+and so, roughly speaking, integrated over possible ARGs
+with the goal of inferring higher-level parameters:
+mostly, scaled mutation rate and recombination rate
+\citep[e.g.,][]{hudson1985statistical,griffiths1996ancestral,kuhner2000maximum,stephens2000inference,fearnhead2001estimating}.
+However, the space of possible ARGs for a given dataset is extremely large,
+and other work aimed to identify the minimum number of recombinations
+needed to explain a given dataset under the infinite alleles model of mutation
+\citep[e.g.,][]{hein1990reconstructing,myers2003bounds,song2005constructing},
+which turns out to be NP-complete \citep{wang2001perfect}.
+For instance, \citet{minichiello2006mapping}
+used a heuristic algorithm to produce ``plausible'' ARGs
+(i.e., those that explained the data with few mutations and recombinations),
+and searched for associations with traits in the resulting ensemble of ARGs.
+Our approach for extending haplotypes follows the same logic,
+that an ARG with fewer recombination events is more parsimonious,
+and thus more likely.
+For this reason, it will occasionally be wrong even if the trees are correct,
+although in practice this source of error is likely much smaller than error in tree inference itself.
 

### 2024-11-29 paper/writeup.tex

@@
 \begin{align*}
-    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)1}^N (a_k - a_{k-1})
+    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)}^N (a_k - a_{k-1})
 \end{align*}
@@
 If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
-to also compare times,
-for instance by returning the weighted root-mean-squared error on inferred times as
+to also compare times.
+Empirically, dating error seems to be more or less homoskedastic on a log scale,
+so we recommend 
+using the weighted root-mean-squared error of log(times), computed as
 \begin{align}\label{eqn:wrmse}
@@
         \Vert{\T_1}\Vert
-    } } .
+    } } ,
 \end{align}
+where the transformation is $t \mapsto \log(1 + t)$ to avoid $\log(0)$.
 The mean is computed weighting by node span, so that a dating error
@@
 additionally produces relative values:
-the ancestral recombination graph Robinson-Foulds value (relative dissimilarity) and true proportion represented. 
+the ancestral recombination graph Robinson-Foulds value (dissimilarity relative to $\T_1$)
+and true proportion represented (similarity relative to $\T_2$).
 The \emph{ARG Robinson-Foulds} (which we call ``ARF'') is defined to be the dissimilarity proportional
@@
  	\operatorname{ARF}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\Vert \T_1 \Vert}
- 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert}.
+ 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert} ,
 \end{align}
+and so if $\T_2$ represents the truth, is analogous to a false positive rate.
 The \emph{true proportion represented} (TPR) is similarity between two trees relative 
@@
 	\operatorname{TPR}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}
-	{\Vert \T_2\Vert}.
+	{\Vert \T_2\Vert} ,
 \end{align}
+and is analogous to statistical power.
 The outputs framed as proportions relative to one of the given tree sequences is more easily understood
@@
 
-To do this, we simulate ARGs containing full haplotypes using \msprime \citep{kelleher2016efficient,baumdicker2021efficient},
+To do this, we simulate ARGs containing full haplotypes using \msprime{}
+\citep{kelleher2016efficient,baumdicker2021efficient},
 with the \texttt{coalescing\_segments\_only} option set to False.
@@
 i.e., where they are unary in the marginal trees.
-More concretely, we hope to recover those portions of ancestral haplotypes
+Setting \texttt{coalescing\_segments\_only} to False includes just this information:
+any ancestral segments for which these coalescent nodes
+are ancestral to any samples -- so, the unary portions of their spans as well.
+However, this includes more information than we want:
+we hope to recover those portions of ancestral haplotypes
 on which the nodes are unary, but adjacent to a region of the genome where the node is not unary.
@@
 and would likely be unrecoverable by any other method.
-So, when evaluating what proportion of the true ARG is represented in an inferred ARG
-we omit these isolated non-coalescent segments.
+So, after simulation, we first remove these isolated, non-coalescent segments.
 To give an idea of what proportion of the full spans of ancestral nodes
@@
 \subsection{Accuracy with true trees}
-
-\comment{Moved from above: don't need?}
-\figref[C\&D]{fig:node-spans} shows the spans for each node in three tree sequences:
-for each node the total span in a ``true'' tree sequence, 
-given by the line,
-that was simulated with \msprime \citep{kelleher2016efficient,baumdicker2021efficient}
-in a way that included spans over which nodes are unary in the output
-(by setting \texttt{coalescing\_segments\_only} to False).
-\figref[C]{fig:node-spans} shows the span of each node after \emph{simplifying} the tree sequence 
-\citep{kelleher2018efficient},
-which removes all unary nodes from all trees
-thus reducing the node spans for some nodes in the sequence.
-In \figref[D]{fig:node-spans}
-is the span of the node after applying \algorithmref{alg:extend}
-to the simplified tree sequence.
-\algorithmref{alg:extend} cannot reduce node span, resulting in node spans 
-from the extended tree sequence that are closer in value to the ``true'' tree sequence. 
-\figref[B]{fig:node-spans} shows how much of this span is incorrect:
-for each node, it gives the amount of span
-that is incorrect (i.e., not present in the original tree sequence)
-over the total amount of span added after running \algorithmref{alg:extend}.
-The values in these plots
-therefore quantify how much the ancestral haplotypes represented by each node
-were correctly, and incorrectly, extended.
-They also tell us how the node spans represented in the result compare to those in the original tree sequence.
 

### 2024-11-29 paper/writeup.tex
Fixup imports to compile on my system
Minor rework of the intro
More refs

@@
 
-\usepackage[dvipsnames]{xcolor}
+% JK Had to change the xcolor import here and turn off
+% the OliveGreen bit below to get this to compile
+% \usepackage[dvipsnames]{xcolor}
+\usepackage[]{xcolor}
 \usepackage{amsmath}
@@
 	colorlinks = true,
-	allcolors = OliveGreen}
+}
+	% allcolors = OliveGreen}
 
@@
 
-There has been substantial recent progress
-in the problem of ``ARG inference'',
-which seeks to infer (portions of) the ancestral recombination graph (or, ARG)
-that describes how a set of genotypes samples are related to each other
-at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
-One way of viewing the ARG is as a sequence of marginal trees,
+Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
+are related to each other at each position of the genome in a recombining
+species~\citep{brandt2024promise,lewanski2024introduction,nielsen2024inference,
+wong2024general}, and there has been significant recent progress 
+on inference through a range of approaches~\citep{rasmussen2014genome,
+speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust}.
+One way of viewing ARGs is as a sequence of marginal trees,
 i.e., the genealogical trees that describe how each portion of the genome
@@
 
-However, the ARG is emphatically not merely a sequence of trees:
+However, an ARG is emphatically not merely a sequence of trees:
 viewed another way, it describes inheritance relationships between ancestral haplotypes.
@@
 in other words, the internal nodes in the trees are labeled, and many of these labels
-are shared between adjacent trees.
+are shared between adjacent trees~\citep{wong2024general}.
 
@@
 compared true and inferred ARGs
-using average Robinson-Foulds and Kendall-Colijn distances between trees
+using average Robinson-Foulds~\citep{robinson1981comparison} and 
+Kendall-Colijn~\citep{kendall2016mapping} distances between trees
 across a regular sequence of genomic positions,

### 2024-11-29 paper/writeup.tex
Tweaks

@@
         (described in more detail in the text):
-        \textbf{(A)} a small portion of a ARG without unary nodes;
+        \textbf{(A)} a small portion of an ARG without unary nodes;
         \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
@@
 
-We work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
+We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
 to take advantage of the extensive, stable, and well-documented set of open-source tools
@@
 over a segment of genome $[\ell_e, r_e)$.
-(The tree sequence includes substantially more information, including genotypes, that we will not use in this paper.)
+(The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
 Suppose that the unique elements of the set of left and right edge endpoints
@@
 
-A relatively efficient algorithm to do this is described in \algorithmref{alg:extend}, 
+% JK note: earlier version said "relatively" - relative to what?
+An efficient algorithm to do this is described in \algorithmref{alg:extend}, 
 and implemented in \tskit{} as \texttt{extend\_haplotypes}.
@@
 
-Since the algorithm needs to take multiple passes over the tree sequence
+Because the algorithm needs to take multiple passes over the tree sequence
 in each direction,
@@
 However, experimentation suggests that 
-in practice only at most five iterations are needed before the algorithm terminates.
+in practice at most five iterations are needed before the algorithm terminates.
 Indeed, for even large sequences \tabref{tab:edge-counts}
@@
 \begin{align*}
-	\mathcal{M}(n1, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
+	\mathcal{M}(n_1, n_2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
 \end{align*}

### 2024-11-29 paper/writeup.tex
Updates to RF section. Mainly, use RF.

@@
 
-Other authors \citep[e.g.,][]{kelleher2019inferring} have used
-an average Robinson-Foulds distance \citep{robinson1981comparison}
-as a measure of disagreement between tree sequences.
+First proposed by~\cite{kuhner2015assessing}, a common approach 
+is to use a weighted average Robinson-Foulds (RF) distance
+% Other authors \citep[e.g.,][]{kelleher2019inferring} have used
+% an average Robinson-Foulds distance \citep{robinson1981comparison}
+as a measure of disagreement between ARGs.
 This could be computed in a very similar way:
@@
 the total span over which $n$ finds a match.
-Then the average Robinson-Foulds distance (averaged over locations in the genome)
+Then the average RF distance (averaged over locations in the genome)
 is
@@
 In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
-across all trees, but average Robinson-Foulds distance allows a different node to match
+across all trees, but average RF distance allows a different node to match
 on each tree.
-The other differences are that \emph{average} Robinson-Foulds distance
+The other differences are that \emph{average} RF distance
 normalizes by sequence length, and is symmetrized.
-Robinson-Foulds distance between two trees
+The RF distance between two trees
 was defined by \citet{robinson1981comparison}
-to the minimum number of edge contraction/expansion operations needed to move
+% JK - change edge->branch here to avoid confusion with tskit/ARG edge
+to be the minimum number of branch contraction/expansion operations needed to move
 from one tree to the other
-(which they then show is equal to the number of edges that induce different splits on the labels).
-A similar metric on tree sequences could be defined
+(which they then show is equal to the number of edges 
+that induce different splits on the labels).
+A similar metric on ARGs could be defined
 using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.
@@
 but not times.
-If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
+If the ARG is dated~\citep{wohns2022unified}, 
+we can naturally use the ``best match'' $\alpha$
 to also compare times.
@@
 additionally produces relative values:
-the ancestral recombination graph Robinson-Foulds value (dissimilarity relative to $\T_1$)
+the ARG RF value (dissimilarity relative to $\T_1$)
 and true proportion represented (similarity relative to $\T_2$).
-The \emph{ARG Robinson-Foulds} (which we call ``ARF'') is defined to be the dissimilarity proportional
+The ARG RF (which we call ``ARF'') is defined to be the dissimilarity proportional
  to the total span of nodes in $\T_1$,

### 2024-11-29 paper/writeup.tex
Tweak

@@
 
-First proposed by~\cite{kuhner2015assessing}, a common approach 
-is to use a weighted average Robinson-Foulds (RF) distance
-% Other authors \citep[e.g.,][]{kelleher2019inferring} have used
-% an average Robinson-Foulds distance \citep{robinson1981comparison}
-as a measure of disagreement between ARGs.
+A common measure of disagreement between ARGs, 
+first proposed by~\cite{kuhner2015assessing}, 
+is to use a weighted average Robinson-Foulds (RF) distance.
 This could be computed in a very similar way:

### 2024-11-29 paper/writeup.tex
Update to inferred ARGs section

@@
 i.e., where they are unary in the marginal trees.
-Setting \texttt{coalescing\_segments\_only} to False includes just this information:
+Setting \texttt{coalescing\_segments\_only} to False includes this information:
 any ancestral segments for which these coalescent nodes
@@
 the addition or removal of an edge requires updates
-to the state of the parent node and all nodes above it.
+to the state of the parent node and all nodes ancestral to it.
 Extending haplotypes yields a tree sequence with fewer edge removals and insertions, 
@@
 we simulate ARGs containing full haplotypes using \msprime{},
-apply the simplification algorithm to reduce these so that there are no unary nodes
+apply the simplification algorithm~\citep{kelleher2018efficient,wong2024general}
+to reduce these so that there are no unary nodes
 (i.e., any node present in a marginal tree is a coalescent node or a sample),
@@
 Does this work with \emph{inferred} ARGs?
-Indeed, \tsinfer{} provides inferred ARGS already containing unary nodes;
-to what degree are they taking advantage of this information already?
-Here, we study the behavior of \tsinfer{}, which produces unary nodes as a
-byproduct of its inference algorithm,
-and leave a comparison across other ARG inference methods for future work.
+As illustrated by~\cite{wong2024general}, there is a significant diversity
+in the structures inferred by current methods,
+and here we focus on \tsinfer{}~\cite{kelleher2019inferring} which 
+infers ARGs containing unary nodes as a byproduct of its inference algorithm.
+% Note this seems pretty weak and is sure to be dinged. We could at least 
+% comment on what we might observe with other inference methods? Would we 
+% recommend that people extend_edges() for their SINGER or Relate ARGs?
+A comparison across other ARG inference methods is left for future work.
 We simulated ARGs containing unary-spanning haplotypes (as above), then

### 2024-11-29 paper/writeup.tex
Tweaks to ignorance and omission section

@@
 of \figref{fig:extending_diagram} would reduce the number of edges
-in the succinct tree sequence representation of the ARG.
+in the succinct tree sequence representation ARGs.
 This is essentially a recombination-based parsimony argument,
@@
 
-\paragraph{Ignorance and omission in the ARG}
-As motivation, we presented above a ``historical'' view of the ARG --
+\paragraph{Ignorance and omission in an ARG}
+As motivation, we presented above a ``historical'' view of ARGs --
 i.e., that each aspect of an inferred ARG is intended to represent
@@
 Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
-that relationships \emph{not} depicted in the ARG are implied to not exist.
+that relationships \emph{not} depicted in an ARG are implied to not exist.
 An alternative interpretation
@@
 through node 3.
-The ``simplification'' algorithm of \citet{kelleher2018efficient},
-or the Hudson algorithm for coalescent simulation
-as implemented in \msprime{} \citep{kelleher2016efficient},
+The ``simplification'' algorithm~\citep{kelleher2018efficient}
+and the Hudson algorithm for coalescent
+simulation~\citep{hudson1983properties,kelleher2016efficient}
 each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;

### 2024-11-29 paper/writeup.tex
Tweaks

@@
 
-The Robinson-Foulds distance \citep{robinson1981comparison}
+The RF distance \citep{robinson1981comparison}
 essentially counts the number of differing branches between two trees;
-the averaged Robinson-Foulds distance \citep{kelleher2019inferring}
+the averaged RF distance~\citep{kuhner2015assessing}
 averages this distance across marginal trees, weighted by span along the genome.
@@
 (and separates the metric into two pieces as discussed in the previous paragraph).
-However, the Robinson-Foulds metric has many undesirable properties --
+However, the RF metric has many undesirable properties --
 for instance, moving a single tip can result in a tree with maximum distance to the original --

### 2024-11-29 paper/writeup.tex
Finish The ARG -> an ARG updates
Closes #11

@@
     We show how these non-coalescing haplotypes (``unary regions'') may be
-    inserted into the ancestral recombination graph (ARG), a compact
+    inserted into ancestral recombination graphs (ARGs), a compact
     but information-rich data structure describing the genealogical
@@
     and explore some consequences for ARGs inferred from real data.
-    To do this, we introduce new metrics of agreement and disagreement between inferred ARGs
-    that, unlike previous methods, consider the ARG as describing relationships between haplotypes
+    To do this, we introduce new metrics of agreement and disagreement between ARGs
+    that, unlike previous methods, consider ARGs as describing relationships between haplotypes
     rather than just a collection of trees.
@@
 on which there is a coalescence.
-Said another way, by default all ancestral nodes in the ARG
+Said another way, by default all ancestral nodes in an ARG
 output by \msprime{} are the MRCA of some pair of samples at every point in the genome
@@
 The bottom row shows the proportion of the original ARG that is represented in the given ARG (TRP).
-This shows that a large proportion of the ARG can be removed by simplification,
+This shows that a large proportion of an ARG can be removed by simplification,
 indicating that coalescent nodes are unary

### 2024-11-29 paper/writeup.tex
Final tweaks
added citation; closes #12

--- 2024-11-29 paper/writeup.tex
@@
 
-% JK Had to change the xcolor import here and turn off
-% the OliveGreen bit below to get this to compile
-% \usepackage[dvipsnames]{xcolor}
-\usepackage[]{xcolor}
+\usepackage[dvipsnames]{xcolor}
 \usepackage{amsmath}
@@
 	colorlinks = true,
-}
-	% allcolors = OliveGreen}
+	allcolors = OliveGreen}
 
@@
 
-Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
-are related to each other at each position of the genome in a recombining
-species~\citep{brandt2024promise,lewanski2024introduction,nielsen2024inference,
-wong2024general}, and there has been significant recent progress 
-on inference through a range of approaches~\citep{rasmussen2014genome,
-speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust}.
-One way of viewing ARGs is as a sequence of marginal trees,
+There has been substantial recent progress
+in the problem of ``ARG inference'',
+which seeks to infer (portions of) the ancestral recombination graph (or, ARG)
+that describes how a set of genotypes samples are related to each other
+at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
+One way of viewing the ARG is as a sequence of marginal trees,
 i.e., the genealogical trees that describe how each portion of the genome
@@
 
-However, an ARG is emphatically not merely a sequence of trees:
+However, the ARG is emphatically not merely a sequence of trees:
 viewed another way, it describes inheritance relationships between ancestral haplotypes.
@@
 in other words, the internal nodes in the trees are labeled, and many of these labels
-are shared between adjacent trees~\citep{wong2024general}.
+are shared between adjacent trees.
 
@@
 Haplotype identity has been largely overlooked in the literature on ARG inference --
-to our knowledge, all metrics that have been used so far to measure accuracy of inferred ARGs
+most methods that have been used so far to measure accuracy of inferred ARGs
 depend only on the sequence of marginal trees,
@@
 compared true and inferred ARGs
-using average Robinson-Foulds~\citep{robinson1981comparison} and 
-Kendall-Colijn~\citep{kendall2016mapping} distances between trees
+using average Robinson-Foulds and Kendall-Colijn distances between trees
 across a regular sequence of genomic positions,
@@
 % (as defined in the context of the ARG, \citep{XXX}).
-\citet{deng2021distribution} evaluated agreement of distributions of
-distances along the genome between tree topology changes,
-and \citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
+There has been some recent attention to haplotypes:
+\citet{ignatieva2024length} compared distributions of haplotype spans in true and inferred ARGs,
+as well as more sophisticated summaries of edges.
+Also, \citet{deng2021distribution} evaluated agreement of distributions of
+distances along the genome between tree topology changes, and
+\citet{zhang2023biobankscale} defined a generalization of Robinson-Foulds distance
 that is the total variation distance between the induced distribution on genotypes;
@@
         (described in more detail in the text):
-        \textbf{(A)} a small portion of an ARG without unary nodes;
+        \textbf{(A)} a small portion of a ARG without unary nodes;
         \textbf{(B)} the implied inheritance pattern of the two portions of the haplotype carried by ancestral node 4,
@@
 
-We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
+We work with the \emph{succinct tree sequence} representation of the ARG (henceforth, ``tree sequence''),
 to take advantage of the extensive, stable, and well-documented set of open-source tools
@@
 over a segment of genome $[\ell_e, r_e)$.
-(The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
+(The tree sequence includes substantially more information, including genotypes, that we will not use in this paper.)
 Suppose that the unique elements of the set of left and right edge endpoints
@@
 
-% JK note: earlier version said "relatively" - relative to what?
-An efficient algorithm to do this is described in \algorithmref{alg:extend}, 
+A relatively efficient algorithm to do this is described in \algorithmref{alg:extend}, 
 and implemented in \tskit{} as \texttt{extend\_haplotypes}.
@@
 
-Because the algorithm needs to take multiple passes over the tree sequence
+Since the algorithm needs to take multiple passes over the tree sequence
 in each direction,
@@
 However, experimentation suggests that 
-in practice at most five iterations are needed before the algorithm terminates.
+in practice only at most five iterations are needed before the algorithm terminates.
 Indeed, for even large sequences \tabref{tab:edge-counts}
@@
 \begin{align*}
-	\mathcal{M}(n_1, n_2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
+	\mathcal{M}(n1, n2) = \left\{ k : S\left(T^{(1)}_k,\ n_1\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
 \end{align*}
@@
 
-A common measure of disagreement between ARGs, 
-first proposed by~\cite{kuhner2015assessing}, 
-is to use a weighted average Robinson-Foulds (RF) distance.
+Other authors \citep[e.g.,][]{kelleher2019inferring} have used
+an average Robinson-Foulds distance \citep{robinson1981comparison}
+as a measure of disagreement between tree sequences.
 This could be computed in a very similar way:
@@
 the total span over which $n$ finds a match.
-Then the average RF distance (averaged over locations in the genome)
+Then the average Robinson-Foulds distance (averaged over locations in the genome)
 is
@@
 In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
-across all trees, but average RF distance allows a different node to match
+across all trees, but average Robinson-Foulds distance allows a different node to match
 on each tree.
-The other differences are that \emph{average} RF distance
+The other differences are that \emph{average} Robinson-Foulds distance
 normalizes by sequence length, and is symmetrized.
-The RF distance between two trees
+Robinson-Foulds distance between two trees
 was defined by \citet{robinson1981comparison}
-% JK - change edge->branch here to avoid confusion with tskit/ARG edge
-to be the minimum number of branch contraction/expansion operations needed to move
+to the minimum number of edge contraction/expansion operations needed to move
 from one tree to the other
-(which they then show is equal to the number of edges 
-that induce different splits on the labels).
-A similar metric on ARGs could be defined
+(which they then show is equal to the number of edges that induce different splits on the labels).
+A similar metric on tree sequences could be defined
 using the subgraph-prune-and-regraft moves used by \citet{deng2024robust}.
@@
 but not times.
-If the ARG is dated~\citep{wohns2022unified}, 
-we can naturally use the ``best match'' $\alpha$
+If the tree sequence is dated, we can naturally use the ``best match'' $\alpha$
 to also compare times.
@@
 additionally produces relative values:
-the ARG RF value (dissimilarity relative to $\T_1$)
+the ancestral recombination graph Robinson-Foulds value (dissimilarity relative to $\T_1$)
 and true proportion represented (similarity relative to $\T_2$).
-The ARG RF (which we call ``ARF'') is defined to be the dissimilarity proportional
+The \emph{ARG Robinson-Foulds} (which we call ``ARF'') is defined to be the dissimilarity proportional
  to the total span of nodes in $\T_1$,
@@
 i.e., where they are unary in the marginal trees.
-Setting \texttt{coalescing\_segments\_only} to False includes this information:
+Setting \texttt{coalescing\_segments\_only} to False includes just this information:
 any ancestral segments for which these coalescent nodes
@@
 the addition or removal of an edge requires updates
-to the state of the parent node and all nodes ancestral to it.
+to the state of the parent node and all nodes above it.
 Extending haplotypes yields a tree sequence with fewer edge removals and insertions, 
@@
 we simulate ARGs containing full haplotypes using \msprime{},
-apply the simplification algorithm~\citep{kelleher2018efficient,wong2024general}
-to reduce these so that there are no unary nodes
+apply the simplification algorithm to reduce these so that there are no unary nodes
 (i.e., any node present in a marginal tree is a coalescent node or a sample),
@@
 Does this work with \emph{inferred} ARGs?
-As illustrated by~\cite{wong2024general}, there is a significant diversity
-in the structures inferred by current methods,
-and here we focus on \tsinfer{}~\cite{kelleher2019inferring} which 
-infers ARGs containing unary nodes as a byproduct of its inference algorithm.
-% Note this seems pretty weak and is sure to be dinged. We could at least 
-% comment on what we might observe with other inference methods? Would we 
-% recommend that people extend_edges() for their SINGER or Relate ARGs?
-A comparison across other ARG inference methods is left for future work.
+Indeed, \tsinfer{} provides inferred ARGS already containing unary nodes;
+to what degree are they taking advantage of this information already?
+Here, we study the behavior of \tsinfer{}, which produces unary nodes as a
+byproduct of its inference algorithm,
+and leave a comparison across other ARG inference methods for future work.
 We simulated ARGs containing unary-spanning haplotypes (as above), then
@@
 of \figref{fig:extending_diagram} would reduce the number of edges
-in the succinct tree sequence representation ARGs.
+in the succinct tree sequence representation of the ARG.
 This is essentially a recombination-based parsimony argument,
@@
 
-\paragraph{Ignorance and omission in an ARG}
-As motivation, we presented above a ``historical'' view of ARGs --
+\paragraph{Ignorance and omission in the ARG}
+As motivation, we presented above a ``historical'' view of the ARG --
 i.e., that each aspect of an inferred ARG is intended to represent
@@
 Furthermore, \figref[A\&B]{fig:conceptual} implicitly takes the position
-that relationships \emph{not} depicted in an ARG are implied to not exist.
+that relationships \emph{not} depicted in the ARG are implied to not exist.
 An alternative interpretation
@@
 through node 3.
-The ``simplification'' algorithm~\citep{kelleher2018efficient}
-and the Hudson algorithm for coalescent
-simulation~\citep{hudson1983properties,kelleher2016efficient}
+The ``simplification'' algorithm of \citet{kelleher2018efficient},
+or the Hudson algorithm for coalescent simulation
+as implemented in \msprime{} \citep{kelleher2016efficient},
 each specifically discard information about any such ``non-coalescent'' portions of ancestral haplotypes;

### 2024-11-29 paper/writeup.tex
Issues 10 and 14

@@
 at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
-One way of viewing the ARG is as a sequence of marginal trees,
+One way of viewing the ARG is as a sequence of local trees,
 i.e., the genealogical trees that describe how each portion of the genome
@@
 These two points of view are related because a single haplotype
-may extend over many marginal trees;
+may extend over many local trees;
 in other words, the internal nodes in the trees are labeled, and many of these labels
@@
 most methods that have been used so far to measure accuracy of inferred ARGs
-depend only on the sequence of marginal trees,
+depend only on the sequence of local trees,
 not on how ancestral haplotypes span across these trees.
@@
 should remain unchanged when possible.
-These extended portions of ancestral haplotypes manifest as unary nodes in the marginal trees.
+These extended portions of ancestral haplotypes manifest as unary nodes in the local trees.
 To quantify how accurate the new information is,
@@
         labeled $L$ and $R$;
-        \textbf{(C)} marginal trees with a unary node added,
+        \textbf{(C)} local trees with a unary node added,
         which produces \textbf{(D)} a more parsimonious haplotype inheritance pattern
@@
 Generalizing from \figref{fig:extending_diagram},
-we do this by identifying paths of inheritance that are shared across a sequence of marginal trees
+we do this by identifying paths of inheritance that are shared across a sequence of local trees
 but for which some of the intermediate nodes are missing.
@@
             \text{Set } $y_i = (p_i \neq NULL)
-                \& (p_i \notin T_O)
-                \& (t_i < t_o)$ \;
+                \ \&\ (p_i \notin T_O)
+                \ \&\ (t_i < t_o)$ \;
             \text{and } $y_o = (p_o \neq NULL)
-                \& (p_o \notin T_I)
-                \& (t_o < t_i)$ \;
+                \ \&\ (p_o \notin T_I)
+                \ \&\ (t_o < t_i)$ \;
             \If{not ($y_i$ or $y_o$)}{ \Break\; }
@@
 and let $s(\T,n)$ denote
-the total span that node $n$ is present in the marginal trees,
+the total span that node $n$ is present in the local trees,
 \begin{align*}
@@
 \begin{align*}
-    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
+    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists\ n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
 \end{align*}
@@
 \begin{align*}
-    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)}^N (a_k - a_{k-1})
+    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)} (a_k - a_{k-1})
 \end{align*}
@@
 
+Our new method \textit{extend haplotypes} is applicable to any ARG.
+However its accuracy depends on the overall structure of the ARG it is applied too. 
+Additionally, as \textit{ARG Robinson-Foulds}
+is given as a proportion of node spans
+it's value is also dependent on the pair of input ARGs. 
+Thus to understand how well our methods can infer ancestral haplotypes we 
+work with simulated tree sequences with a large variety in parameter choice.
 To do this, we simulate ARGs containing full haplotypes using \msprime{}
@@
 with the \texttt{coalescing\_segments\_only} option set to False.
-Although \msprime{} simulates many events that do not create a coalescence in some marginal tree,
+Although \msprime{} simulates many events that do not create a coalescence in some local tree,
 it only outputs information for nodes which contain a coalescence
@@
 on which the nodes are \emph{not} coalescent;
-i.e., where they are unary in the marginal trees.
+i.e., where they are unary in the local trees.
 Setting \texttt{coalescing\_segments\_only} to False includes just this information:
@@
 apply the simplification algorithm to reduce these so that there are no unary nodes
-(i.e., any node present in a marginal tree is a coalescent node or a sample),
+(i.e., any node present in a local tree is a coalescent node or a sample),
 and then apply \algorithmref{alg:extend} to the result
@@
         Accuracy and sensitivity of extended haplotypes
-        across a range of sample sizes, on a sequence of length $5\times10^7$ (top)
-        and sequence lengths, with 1,000 diploid samples (bottom).
+        across a range of sample sizes, on a sequence of length $5\times10^7$ (right)
+        and sequence lengths, with 1,000 diploid samples (left).
         For each, a simulated ARG containing unary haplotype spans
@@
 and that contain more complete information about true ancestral relationships.
-These extended ancestral haplotypes manifest as unary nodes in the marginal trees.
+These extended ancestral haplotypes manifest as unary nodes in the local trees.
 Although a number of ARG inference methods may be taking advantage of this information,
 it is our impression that this source of information is not widely appreciated.
-In fact, due to the field's focus on marginal trees rather than haplotypes,
+In fact, due to the field's focus on local trees rather than haplotypes,
 we had to develop a haplotype-aware measure of (dis)agreement between ARGs in order to study the accuracy of the proposed algorithm.
@@
 Extending ancestral haplotypes from one site into neighboring regions
-conceptually allows information from those marginal trees
+conceptually allows information from those local trees
 to inform age inference at that site as well.
@@
 the averaged Robinson-Foulds distance \citep{kelleher2019inferring}
-averages this distance across marginal trees, weighted by span along the genome.
+averages this distance across local trees, weighted by span along the genome.
 The method we present here for measuring dissimilarity between topologies of ARGs
@@
 in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
-however, it is still applied to ARGs as an average over marginal trees,
+however, it is still applied to ARGs as an average over local trees,
 without enforcement of identity across haplotypes;

### 2024-11-30 paper/writeup.tex
typos

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\paragraph{Metrics on ARGs}
+The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
+(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
+This is by design: in practice it is not possible to infer all aspects
+of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
+and so we wanted to quantify
+``How much of the true relationships does this ARG represent?''
+However, it is worth noting that the symmetrized version
+$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
+To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
+and view each ARG as a subset of the space $[0,L) \times N \times N$,
+where $N$ is the shared set of nodes.
+Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
+$|\T_1 \setminus \T_2|$,
+and so the symmetrized version is the measure of the symmetric difference
+$|\T_1\Delta\T_2|$,
+which is well-known to be a metric \citep{rudin1976principles}. 
+If the two ARGs have the same number of nodes,
+we can consider all bijections between their nodes.
+The symmetric difference between $\T_1$ and $\T_2$,
+related through each bijection is a metric.
+Then the minimum over all such metrics will still be a metric
+since the minimum over a finite number of metrics is also a metric.
+This also extends to two tree sequences with different numbers of nodes,
+$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
+
+The Robinson-Foulds distance \citep{robinson1981comparison}
+essentially counts the number of differing branches between two trees;
+the averaged Robinson-Foulds distance \citep{kelleher2019inferring}
+averages this distance across local trees, weighted by span along the genome.
+The method we present here for measuring dissimilarity between topologies of ARGs
+is a straightforward generalization
+that takes into account span along the genome of inferred ancestral haplotypes
+(and separates the metric into two pieces as discussed in the previous paragraph).
+However, the Robinson-Foulds metric has many undesirable properties --
+for instance, moving a single tip can result in a tree with maximum distance to the original --
+and there is a substantial literature giving more robust generalizations
+\citep[reviewed by][]{llabres2021generalized}.
+Many of these generalizations \citep[e.g.,][]{bocker2013generalized}
+relax the requirement that the match between subtended sample sets be exact,
+and weight matches in some way by the size of the dissimilarity.
+We considered such definitions as well, but kept to the simple case
+for computational tractability --
+the generalization of \citet{bocker2013generalized} is NP-hard to compute, even for a single tree.
+In the ARG literature, \citet{zhang2023biobankscale}
+defines a metric (called ``ARG total variation distance'') that includes branch lengths,
+in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
+however, it is still applied to ARGs as an average over local trees,
+without enforcement of identity across haplotypes;
+it would be useful to extend our dissimilarity to include branch lengths.
+
+%--------------------------------
 \subsection{Simulations}
@@
 Our new method \textit{extend haplotypes} is applicable to any ARG.
-However its accuracy depends on the overall structure of the ARG it is applied too. 
+However its accuracy depends on the overall structure of the ARG it is applied to. 
 Additionally, as \textit{ARG Robinson-Foulds}
 is given as a proportion of node spans
-it's value is also dependent on the pair of input ARGs. 
+its' value is also dependent on the pair of input ARGs. 
 Thus to understand how well our methods can infer ancestral haplotypes we 
@@
 
-\paragraph{Metrics on ARGs}
-The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
-(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
-This is by design: in practice it is not possible to infer all aspects
-of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
-and so we wanted to quantify
-``How much of the true relationships does this ARG represent?''
-However, it is worth noting that the symmetrized version
-$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
-To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
-and view each ARG as a subset of the space $[0,L) \times N \times N$,
-where $N$ is the shared set of nodes.
-Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
-$|\T_1 \setminus \T_2|$,
-and so the symmetrized version is the measure of the symmetric difference
-$|\T_1\Delta\T_2|$,
-which is well-known to be a metric \citep{rudin1976principles}. 
-If the two ARGs have the same number of nodes,
-we can consider all bijections between their nodes.
-The symmetric difference between $\T_1$ and $\T_2$,
-related through each bijection is a metric.
-Then the minimum over all such metrics will still be a metric
-since the minimum over a finite number of metrics is also a metric.
-This also extends to two tree sequences with different numbers of nodes,
-$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
-
-The Robinson-Foulds distance \citep{robinson1981comparison}
-essentially counts the number of differing branches between two trees;
-the averaged Robinson-Foulds distance \citep{kelleher2019inferring}
-averages this distance across local trees, weighted by span along the genome.
-The method we present here for measuring dissimilarity between topologies of ARGs
-is a straightforward generalization
-that takes into account span along the genome of inferred ancestral haplotypes
-(and separates the metric into two pieces as discussed in the previous paragraph).
-However, the Robinson-Foulds metric has many undesirable properties --
-for instance, moving a single tip can result in a tree with maximum distance to the original --
-and there is a substantial literature giving more robust generalizations
-\citep[reviewed by][]{llabres2021generalized}.
-Many of these generalizations \citep[e.g.,][]{bocker2013generalized}
-relax the requirement that the match between subtended sample sets be exact,
-and weight matches in some way by the size of the dissimilarity.
-We considered such definitions as well, but kept to the simple case
-for computational tractability --
-the generalization of \citet{bocker2013generalized} is NP-hard to compute, even for a single tree.
-In the ARG literature, \citet{zhang2023biobankscale}
-defines a metric (called ``ARG total variation distance'') that includes branch lengths,
-in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
-however, it is still applied to ARGs as an average over local trees,
-without enforcement of identity across haplotypes;
-it would be useful to extend our dissimilarity to include branch lengths.
-
 \paragraph{Parsimony}

### 2024-11-30 paper/writeup.tex

@@
     produced by recombination is inherited as a Mendelian marker.
+    % JK: do we though? The only application to inference we have is from
+    % tsinfer, where the results are pretty equivocal. I'd suggest rephrasing
+    % this to focus on the inherent information content of an ARG above and 
+    % beyond the local trees.
     In this paper, we describe how this recombination-mediated information
@@
 
-There has been substantial recent progress
-in the problem of ``ARG inference'',
-which seeks to infer (portions of) the ancestral recombination graph (or, ARG)
-that describes how a set of genotypes samples are related to each other
-at each position of the genome \citep{speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,lewanski2024introduction}.
-One way of viewing the ARG is as a sequence of local trees,
+Ancestral recombination graphs (ARGs) describe how a set of sampled sequences 
+are related to each other at each position of the genome in a recombining
+species~\citep{brandt2024promise,lewanski2024introduction,nielsen2024inference,
+wong2024general}, and there has been significant recent progress 
+on inference through a range of approaches~\citep{rasmussen2014genome,
+speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust}.
+One way of viewing ARGs is as a sequence of marginal trees,
 i.e., the genealogical trees that describe how each portion of the genome
@@
 
-Other authors \citep[e.g.,][]{kelleher2019inferring} have used
-an average Robinson-Foulds distance \citep{robinson1981comparison}
-as a measure of disagreement between tree sequences.
+A common measure of disagreement between ARGs, 
+first proposed by~\cite{kuhner2015assessing}, 
+is to use a weighted average Robinson-Foulds (RF) distance.
 This could be computed in a very similar way:
@@
 we simulate ARGs containing full haplotypes using \msprime{},
-apply the simplification algorithm to reduce these so that there are no unary nodes
+apply the simplification algorithm~\citep{kelleher2018efficient,wong2024general}
+to reduce these so that there are no unary nodes
 (i.e., any node present in a local tree is a coalescent node or a sample),
@@
 In this paper, we have shown that, for the most part, this missing information can be imputed.
+
+\paragraph{Metrics on ARGs}
+The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
+(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
+This is by design: in practice it is not possible to infer all aspects
+of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
+and so we wanted to quantify
+``How much of the true relationships does this ARG represent?''
+However, it is worth noting that the symmetrized version
+$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
+To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
+and view each ARG as a subset of the space $[0,L) \times N \times N$,
+where $N$ is the shared set of nodes.
+Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
+$|\T_1 \setminus \T_2|$,
+and so the symmetrized version is the measure of the symmetric difference
+$|\T_1\Delta\T_2|$,
+which is well-known to be a metric \citep{rudin1976principles}. 
+If the two ARGs have the same number of nodes,
+we can consider all bijections between their nodes.
+The symmetric difference between $\T_1$ and $\T_2$,
+related through each bijection is a metric.
+Then the minimum over all such metrics will still be a metric
+since the minimum over a finite number of metrics is also a metric.
+This also extends to two tree sequences with different numbers of nodes,
+$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
+
+The RF distance \citep{robinson1981comparison}
+essentially counts the number of differing branches between two trees;
+the averaged RF distance~\citep{kuhner2015assessing}
+averages this distance across marginal trees, weighted by span along the genome.
+The method we present here for measuring dissimilarity between topologies of ARGs
+is a straightforward generalization
+that takes into account span along the genome of inferred ancestral haplotypes
+(and separates the metric into two pieces as discussed in the previous paragraph).
+However, the RF metric has many undesirable properties --
+for instance, moving a single tip can result in a tree with maximum distance to the original --
+and there is a substantial literature giving more robust generalizations
+\citep[reviewed by][]{llabres2021generalized}.
+Many of these generalizations \citep[e.g.,][]{bocker2013generalized}
+relax the requirement that the match between subtended sample sets be exact,
+and weight matches in some way by the size of the dissimilarity.
+We considered such definitions as well, but kept to the simple case
+for computational tractability --
+the generalization of \citet{bocker2013generalized} is NP-hard to compute, even for a single tree.
+In the ARG literature, \citet{zhang2023biobankscale}
+defines a metric (called ``ARG total variation distance'') that includes branch lengths,
+in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
+however, it is still applied to ARGs as an average over marginal trees,
+without enforcement of identity across haplotypes;
+it would be useful to extend our dissimilarity to include branch lengths.
 

### 2024-11-30 paper/writeup.tex
Fixing final merge conflicts

@@
 
-The Robinson-Foulds distance \citep{robinson1981comparison}
+The RF distance \citep{robinson1981comparison}
 essentially counts the number of differing branches between two trees;
-the averaged Robinson-Foulds distance \citep{kelleher2019inferring}
-averages this distance across local trees, weighted by span along the genome.
+the averaged RF distance~\citep{kuhner2015assessing}
+averages this distance across marginal trees, weighted by span along the genome.
 The method we present here for measuring dissimilarity between topologies of ARGs
@@
 in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
-however, it is still applied to ARGs as an average over local trees,
+however, it is still applied to ARGs as an average over marginal trees,
 without enforcement of identity across haplotypes;
@@
 
-\paragraph{Metrics on ARGs}
-The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
-(i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
-This is by design: in practice it is not possible to infer all aspects
-of the true ancestry of a set of samples (i.e., all their genetic ancestors who ever lived),
-and so we wanted to quantify
-``How much of the true relationships does this ARG represent?''
-However, it is worth noting that the symmetrized version
-$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
-To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
-and view each ARG as a subset of the space $[0,L) \times N \times N$,
-where $N$ is the shared set of nodes.
-Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
-$|\T_1 \setminus \T_2|$,
-and so the symmetrized version is the measure of the symmetric difference
-$|\T_1\Delta\T_2|$,
-which is well-known to be a metric \citep{rudin1976principles}. 
-If the two ARGs have the same number of nodes,
-we can consider all bijections between their nodes.
-The symmetric difference between $\T_1$ and $\T_2$,
-related through each bijection is a metric.
-Then the minimum over all such metrics will still be a metric
-since the minimum over a finite number of metrics is also a metric.
-This also extends to two tree sequences with different numbers of nodes,
-$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
-
-The RF distance \citep{robinson1981comparison}
-essentially counts the number of differing branches between two trees;
-the averaged RF distance~\citep{kuhner2015assessing}
-averages this distance across marginal trees, weighted by span along the genome.
-The method we present here for measuring dissimilarity between topologies of ARGs
-is a straightforward generalization
-that takes into account span along the genome of inferred ancestral haplotypes
-(and separates the metric into two pieces as discussed in the previous paragraph).
-However, the RF metric has many undesirable properties --
-for instance, moving a single tip can result in a tree with maximum distance to the original --
-and there is a substantial literature giving more robust generalizations
-\citep[reviewed by][]{llabres2021generalized}.
-Many of these generalizations \citep[e.g.,][]{bocker2013generalized}
-relax the requirement that the match between subtended sample sets be exact,
-and weight matches in some way by the size of the dissimilarity.
-We considered such definitions as well, but kept to the simple case
-for computational tractability --
-the generalization of \citet{bocker2013generalized} is NP-hard to compute, even for a single tree.
-In the ARG literature, \citet{zhang2023biobankscale}
-defines a metric (called ``ARG total variation distance'') that includes branch lengths,
-in a way similar to \citet{robinson1979comparison} and \citet{kuhner1994simulation};
-however, it is still applied to ARGs as an average over marginal trees,
-without enforcement of identity across haplotypes;
-it would be useful to extend our dissimilarity to include branch lengths.
-
 \paragraph{Parsimony}

### 2024-11-30 paper/writeup.tex
some loose ends

@@
 % the OliveGreen bit below to get this to compile
-% \usepackage[dvipsnames]{xcolor}
-\usepackage[]{xcolor}
+\usepackage[dvipsnames]{xcolor}
+%\usepackage[]{xcolor}
 \usepackage{amsmath}
@@
     In a genealogical context, this information reflects the persistence
-    of ancestral haplotypes across marginal genealogical trees in which they
+    of ancestral haplotypes across local genealogical trees in which they
     do not represent coalescences.
@@
 speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust}.
-One way of viewing ARGs is as a sequence of marginal trees,
+One way of viewing ARGs is as a sequence of local trees,
 i.e., the genealogical trees that describe how each portion of the genome
@@
 However, the additional information provided by haplotype structure can be important:
-for instance, haplotypes that extend over many marginal genealogies
+for instance, haplotypes that extend over many local genealogies
 ``tie together'' those genealogies, allowing estimates of times of particular ancestors
@@
 could not be expected to produce anything different.
-However, the example becomes more complicated once we consider what these marginal genealogies imply about haplotype inheritance.
+However, the example becomes more complicated once we consider what these local genealogies imply about haplotype inheritance.
 \figref[B]{fig:conceptual} shows the implied inheritance of haplotypes,
@@
 This also extends to two tree sequences with different numbers of nodes,
-$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $T_1\to\T_2$ and $\T_2\to\T_1$.
+$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $\T_1\to\T_2$ and $\T_2\to\T_1$.
 
@@
 the averaged RF distance~\citep{kuhner2015assessing}
-averages this distance across marginal trees, weighted by span along the genome.
+averages this distance across local trees, weighted by span along the genome.
 The method we present here for measuring dissimilarity between topologies of ARGs

### 2024-12-09 paper/writeup.tex

@@
 However, we are also interested in whether \algorithmref{alg:extend}
-improves \emph{inferred} ARG.
+improves \emph{inferred} ARGs.
 Since we are not aware of any current methods for measuring (dis)agreement between ARGs
@@
 Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
-its' \emph{best matching node} in $\T_2$ as
+its \emph{best matching node} in $\T_2$ as
 \begin{align*}
@@
 is given as a proportion of node spans
-its' value is also dependent on the pair of input ARGs. 
+its value is also dependent on the pair of input ARGs. 
 Thus to understand how well our methods can infer ancestral haplotypes we 
@@
         For each, a simulated ARG containing unary haplotype spans
-        was (i) \emph{simplified}, removing the unary spans,
-        and (ii) \emph{inferred}, using \tsinfer{} on genotypes;
-        then each of these had its haplotypes \emph{extended}.
-        The inferred, then simplified, ARG is also shown.
+        was (i) \emph{simplified} (`S'), removing the unary spans,
+        and (ii) \emph{inferred} (`I'), using \tsinfer{} on genotypes;
+        then each of these had its haplotypes \emph{extended} (`SE', `IE').
+        The inferred, then simplified, ARG ('IS') and its subsequent extension (`ISE') are also shown.
         \textbf{Top row:} ARF, the dissimilarity to the true ARG,
@@
 
-\paragraph{Inference methods}
-Clearly, there is ample information in genomic data to infer the extent of ancestral haplotypes even where they are unary,
-and this has the potential to significantly inform downstream methods, especially dating.
-However, currently not all ARG inference methods return ARGs with unary nodes.
-For instance, we have seen that tsinfer returns ARGs to which the method presented here adds little --
-so, seems to be already taking advantage of this source of information.
-To our knowledge, methods based on threading (sequentially adding or removing haplotypes) such as
-SINGER \citep{deng2024robust} and ARGNeedle \citep{zhang2023biobankscale} return ARGs without any unary nodes,
-although there is no conceptual issue with incorporating unary haplotypes into threading algorithms,
-and at least SINGER contains information about unary nodes in its internal format.
-On the other hand, locally building trees separately in each segment of genome as is done by Relate \cite{speidel2019method}
-simply does not provide this sort of haplotype-based information (although it might be added using a method like ours).
+%\paragraph{Inference methods}
+%Clearly, there is ample information in genomic data to infer the extent of ancestral haplotypes even where they are unary,
+%and this has the potential to significantly inform downstream methods, especially dating.
+%However, currently not all ARG inference methods return ARGs with unary nodes.
+%For instance, we have seen that tsinfer returns ARGs to which the method presented here adds little --
+%so, seems to be already taking advantage of this source of information.
+%To our knowledge, methods based on threading (sequentially adding or removing haplotypes) such as
+%SINGER \citep{deng2024robust} and ARGNeedle \citep{zhang2023biobankscale} return ARGs without any unary nodes,
+%although there is no conceptual issue with incorporating unary haplotypes into threading algorithms,
+%and at least SINGER contains information about unary nodes in its internal format.
+%On the other hand, locally building trees separately in each segment of genome as is done by Relate \cite{speidel2019method}
+%simply does not provide this sort of haplotype-based information (although it might be added using a method like ours).
 

### 2024-12-09 paper/writeup.tex

--- 2024-11-30 paper/writeup.tex
@@
     do not represent coalescences.
-    We show how these non-coalescing haplotypes (``unary regions'') may be
+    We show how these non-coalescing haplotypes (``locally-unary nodes'') may be
     inserted into ancestral recombination graphs (ARGs), a compact
@@
     is nearly always correct where the initial ARG is correct.
-    We provide efficient algorithms to infer unary regions within existing ARGs,
+    We provide efficient algorithms to infer locally-unary nodes within existing ARGs,
     and explore some consequences for ARGs inferred from real data.
@@
 It is helpful to first describe what we compute in the simple case.
-We will first simulate tree sequences in such a way that
-the unary regions of coalescent haplotypes are retained.
+We will first simulate tree sequences where
+nodes that are unary in local trees between coalescent haplotypes are retained.
 Then, each node is present in both tree sequences,
@@
 
+\section{Acknowledgments}
+The authors would like to thank Dr. Yan Wong for his suggestions and insights for ARGs with locally-unary nodes.
+
 \bibliography{references}

### 2024-12-09 paper/writeup.tex
files management. figure styles

@@
     \begin{center}
-    \includegraphics[width=0.9\textwidth]{conceptual_figure.pdf}
+    \includegraphics[width=0.9\textwidth]{figures/conceptual_figure.pdf}
     \end{center}
@@
 \begin{center}
-	\includegraphics[width=5in]{edge_extend_method.pdf}
+	\includegraphics[width=5in]{figures/edge_extend_method.pdf}
 \end{center}
@@
 	\begin{center}
-\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
-\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
+\includegraphics[height=1.5in, width=3in]{figures/discrepancy_func_method_t1.pdf}
+\includegraphics[height=1.5in, width=3in]{figures/discrepancy_function_method_t2.pdf}
     \caption{
@@
 \begin{figure}[!bht]
-	\includegraphics{newplots_wo_ee/figure-4.pdf}
+	\includegraphics[width=\textwidth]{figures/Figure4-plt.pdf}
 	\caption{
@@
         nearly all have less than 5\% incorrectly added;
-        see Supplementary \figref{fig:incorrect_ratio}.
+        see Supplementary \figref{fig:incorrect_ratio}. Note the log scale on both the $x$ and $y$ axes.
         Plots \textbf{(C)} and \textbf{(D)} show total spans per node,
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{figure5.pdf}
+        \includegraphics[width=\textwidth]{figures/Figure5-plt.pdf}
 	\end{center}
@@
 \begin{figure}[!hbt]
-	\includegraphics{newplots_wo_ee/figure-4b-supplement}
+	\includegraphics[width=\textwidth]{figures/Figure4S-plt-nonzero.pdf}
 	\caption{Distribution of incorrectly added span percentages

### 2024-12-10 paper/writeup.tex

--- 2024-12-09 paper/writeup.tex
@@
 
+\section{Acknowledgments}
+The authors would like to thank Dr. Yan Wong for his helpful comments and suggestions.
+
 \bibliography{references}

### 2024-12-16 paper/writeup.tex
minor merge fix
IBD note; closes #26

@@
 
+\paragraph{IBD in ARGs}
+The term ``identity by descent'' (IBD) is used to mean many different (but related) things,
+and length distributions of shared IBD segments can be used for inference of recent demographic history
+\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate}.
+A commonly-used definition in the context of a given ARG says that
+the two genomes share an IBD segment if each has inherited the segment from their common ancestor
+along a single path \citep{ralph2013geography}.
+Largely for computational reasons, this is the definition that is used in \tskit's
+IBD-finding methods 
+(by G.~Tsambos \citet{tskit}; see also \citet{tsambos2023link}).
+However, many simulation and inference methods produce ARGs
+as shown in \figref[A\&B]{fig:conceptual},
+in which inheritance of a single segment is represented by more than one edge.
+This means that the \texttt{ibd\_segments} method of \tskit{} will return shorter segments than it ought to.
+However, applying our method of extending haplotypes
+will remove this issue, and so should make returned IBD segments
+more closely match those predicted by theory.
+
 %\paragraph{Inference methods}

### 2024-12-16 paper/writeup.tex
do something about parsimony: maybe closes #16

@@
 that relationships \emph{not} depicted in an ARG are implied to not exist.
-An alternative interpretation
+As discussed in \citet{wong2024general},
+an alternative interpretation
 of the ARG depicted in \figref[A\&B]{fig:conceptual} would be that we have no information as to 
@@
 mostly, scaled mutation rate and recombination rate
-\citep[e.g.,][]{hudson1985statistical,griffiths1996ancestral,kuhner2000maximum,stephens2000inference,fearnhead2001estimating}.
+(for instance, \citet{hudson1985statistical,griffiths1996ancestral,kuhner2000maximum,stephens2000inference,fearnhead2001estimating}).
 However, the space of possible ARGs for a given dataset is extremely large,
@@
 and searched for associations with traits in the resulting ensemble of ARGs.
+(See \citet{wong2024general} for more historical discussion.)
 Our approach for extending haplotypes follows the same logic,

### 2024-12-16 paper/writeup.tex
remove confusing inference thing; maybe closes #19?

@@
     A forest is more than its trees:
-    haplotypes and inferred ARGs
+    haplotypes and ancestral recombination graphs
 }
@@
 \author[$\ddagger$]{Jerome Kelleher}
-\author[$\ast$,$\dagger$,S]{Peter Ralph}
+\author[$\ast$,$\dagger$,\S]{Peter Ralph}
 
@@
     produced by recombination is inherited as a Mendelian marker.
-    % JK: do we though? The only application to inference we have is from
-    % tsinfer, where the results are pretty equivocal. I'd suggest rephrasing
-    % this to focus on the inherent information content of an ARG above and 
-    % beyond the local trees.
-    In this paper, we describe how this recombination-mediated information
-    can in many cases be recovered from inference based solely on polymorphic markers.
-    In a genealogical context, this information reflects the persistence
+    In a genealogical context, this recombination-mediated information reflects the persistence
     of ancestral haplotypes across local genealogical trees in which they

### 2024-12-17 paper/writeup.tex
one more citation
two more
final note

@@
 and length distributions of shared IBD segments can be used for inference of recent demographic history
-\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate}.
+\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate,yang2016rapidly,silcocks2023indigenous,huang2024estimating}.
 A commonly-used definition in the context of a given ARG says that
@@
 more closely match those predicted by theory.
+Further work is needed to understand how accurately IBD segments
+are inferred by various ARG inference methods.
 

### 2024-12-17 paper/writeup.tex
Update paper/writeup.tex

@@
 This means that the \texttt{ibd\_segments} method of \tskit{} will return shorter segments than it ought to.
-However, applying our method of extending haplotypes
-will remove this issue, and so should make returned IBD segments
-more closely match those predicted by theory.
+However, as we have shown above,
+our method of extending haplotypes will modify the tree sequence
+so that the inherited segment is represented by a single edge
+(as in \figref[C\&D]{fig:conceptual}).
+So, if our method (\texttt{extend_haplotypes}) is applied before finding IBD segments
+(with  \texttt{ibd\_segments}),
+then the resulting segments should much more accurately represent
+the IBD segments (in the ``path'' sense used here) implied by the tree sequence.
+(Whether the resulting segments better match those predicted by theory depends
+also on the quality of tree inference.)
 Further work is needed to understand how accurately IBD segments

### 2024-12-18 paper/writeup.tex
one more citation

--- 2024-12-17 paper/writeup.tex
@@
 and length distributions of shared IBD segments can be used for inference of recent demographic history
-\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate,yang2016rapidly,silcocks2023indigenous,huang2024estimating}.
+\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate,yang2016rapidly,silcocks2023indigenous}.
 A commonly-used definition in the context of a given ARG says that
@@
 This means that the \texttt{ibd\_segments} method of \tskit{} will return shorter segments than it ought to.
-However, applying our method of extending haplotypes
-will remove this issue, and so should make returned IBD segments
-more closely match those predicted by theory.
+However, as we have shown above,
+our method of extending haplotypes will modify the tree sequence
+so that the inherited segment is represented by a single edge
+(as in \figref[C\&D]{fig:conceptual}).
+So, if our method (\texttt{extend\_haplotypes}) is applied before finding IBD segments
+(with  \texttt{ibd\_segments}),
+then the resulting segments should much more accurately represent
+the IBD segments (in the ``path'' sense used here) implied by the tree sequence.
+Whether the resulting segments better match those predicted by theory depends
+also on the quality of tree inference.
+Note also that \citet{huang2024estimating} and \citet{guo2024strong}
+both provide methods for tree sequences to compute a different definition of IBD
+(segments on which the MRCA does not change),
+which is unaffected by this issue.
 Further work is needed to understand how accurately IBD segments

### 2025-02-04 paper/writeup.tex
removed repeat acknowledgements section
write up post tscompare changes

@@
 \usepackage{natbib}
+\usepackage{centernot}
 \usepackage[linesnumbered,ruled,vlined,algo2e]{algorithm2e}
@@
 \newcommand{\dis}{\operatorname{dis}}
-\newcommand{\similarity}{\operatorname{sim}}
+\newcommand{\match}{\underset{\longrightarrow}{\operatorname{match}}}
+\newcommand{\invmatch}{\underset{\longleftarrow}{\operatorname{match}}}
+\newcommand{\nmatch}{\underset{\centernot\longrightarrow}{\operatorname{match}}}
 \newcommand{\comment}[1]{{\color{violet} \it #1}}
@@
 that take into account haplotype identity,
-we define a measure of \emph{dissimilarity} to quantify this.
+we define a measure of \emph{matched span} to quantify this.
 The method is implemented in the \texttt{tscompare} package.
@@
 We would like to measure (a) how much of $\T_1$ is found in $\T_2$;
-(b) how much of $\T_1$ is \emph{not} found in $\T_2$; and
-(c) how much of $\T_2$ is not found in $\T_1$.
-(Think of these three quantities as the size of the intersection
-and two differences between the tree sequences,
+(b) how much of $\T_2$ is found in $\T_1$; and
+(c) how much of $\T_1$ is \emph{not} found in $\T_2$.
+(Think of these three quantities as the sizes of two relative intersections
+and difference between the tree sequences,
 thought of vaguely as sets.)
@@
 matches the samples below $n_2$ in $\T_2$.
-The \emph{similarity} between $\T_1$ and $\T_2$ is then defined to be
+The \emph{matched span} of $\T_1$ in $\T_2$ is then defined to be
 \begin{align*}
-    \similarity(\T_1, \T_2)
+    \match(\T_1, \T_2)
     =
@@
 \end{align*}
-where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
+where the maximum is over all mappings $\beta$ of nodes in $\T_1$ to nodes in $\T_2$, 
+and we require that samples in $\T_1$ are mapped to samples in $\T_2$.
 (Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
@@
 so that
-\begin{align*}
-    \similarity(\T_1, \T_2)
+\begin{align}
+    \match(\T_1, \T_2)
     =
     \sum_{n \in N_1} m(n, \alpha(n)) .
-\end{align*}
+\end{align}
 If the best-matching node is not unique, we define $\alpha(n_1)$ to be the node in $T_2$
@@
 (and if \emph{this} is not unique, we pick an arbitrary one) --
-however, this potential ambiguity does not affect the definition of $\similarity(\T_1, \T_2)$.
-Let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$,
-and let $s(\T,n)$ denote
-the total span that node $n$ is present in the local trees,
+however, this potential ambiguity does not affect the definition of $\match(\T_1, \T_2)$.
+Let $s(\T,n)$ denote the total span that node $n$ is present in the local trees,
 \begin{align*}
@@
 \end{align*}
-where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise).
-We then define the \emph{dissimilarity} of $\T_1$ from $\T_2$ by
+where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise),
+and let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
+We then define the \emph{non-matched span} of $\T_1$ in $\T_2$ by
 \begin{align*}
-    \dis(\T_1, \T_2)
+	\nmatch(\T_1,\T_2)
 	=
@@
     = 
-    \Vert \T_1\Vert -\similarity(\T_1,\T_2),
+    \Vert \T_1\Vert -\match(\T_1,\T_2),
 \end{align*}
 which is the total span for all nodes in $\T_1$
-over which their descendant samples do \emph{not} match those of their best match in $\T_2$.
+over which their descendant samples do \emph{not} match
+ those of their best match in $\T_2$.
+Contrarily, given a matching $\beta:\T_1\to\T_2$, 
+we want to quantify how much of $\T_2$ is represented in $\T_1$.
+To do this, we define the \emph{inverse matched span} of $\T_1$ in $\T_2$ as
+\begin{align}
+	\invmatch(\T_1,\T_2) = \sum_{n_2\in N_2} \max_{n_1\in \alpha^{-1}(n_2)} m(n_1,n_2)
+\end{align}
+where $\alpha^{-1}(n_2)$ is the set of all nodes $n_1\in \T_1$ 
+whose best match is $n_2$.
+Note that the inverse matched span is not equivalent
+ to the matched span of $\T_2$ in $\T_1$ 
+ as the inverse matched span maximizes over spans
+  which are still dependent on $\beta$, 
+  and, as previously stated, $\beta$ is not guaranteed to be bijective.
 
@@
 
-The dissimilarity $\dis(\T_1, \T_2)$ measures disagreement between \emph{topologies},
+The matched span, $\match(\T_1, \T_2)$, measures agreement between \emph{topologies},
 but not times.
@@
 additionally produces relative values:
-the ARG RF value (dissimilarity relative to $\T_1$)
-and true proportion represented (similarity relative to $\T_2$).
-The ARG RF (which we call ``ARF'') is defined to be the dissimilarity proportional
+the ARG RF value (non-matched span relative to $\T_1$)
+and true proportion represented (inverse matched span relative to $\T_2$).
+The ARG RF (which we call ``ARF'') is defined to be the matched span proportional
  to the total span of nodes in $\T_1$,
  \begin{align}\label{eqn:dissimilarity}
- 	\operatorname{ARF}(\T_1,\T_2)=\frac{\dis(\T_1,\T_2)}{\Vert \T_1 \Vert}
- 	=1-\frac{\similarity(\T_1,\T_2)}{\Vert \T_1\Vert} ,
+ 	\operatorname{ARF}(\T_1,\T_2)
+ 	=1-\frac{\match(\T_1,\T_2)}{\Vert \T_1\Vert} ,
 \end{align}
 and so if $\T_2$ represents the truth, is analogous to a false positive rate.
-The \emph{true proportion represented} (TPR) is similarity between two trees relative 
-to the total span of nodes in $\T_2$,
+The \emph{true proportion represented} (TPR) is the inverse matched span between 
+two trees relative to the total span of nodes in $\T_2$,
 \begin{align}\label{eqn:true-proportion}
-	\operatorname{TPR}(\T_1,\T_2) = \frac{\similarity(\T_1,\T_2)}
+	\operatorname{TPR}(\T_1,\T_2) = \frac{\invmatch(\T_1,\T_2)}
 	{\Vert \T_2\Vert} ,
@@
 The outputs framed as proportions relative to one of the given tree sequences is more easily understood
-for comparing pairs of tree sequences than the original dissimilarity and similarity, whose units are length of spans.
-We compute dissimilarity and ARF of two tree sequences in \figref{fig:conceptual_discrepancy}.
+for comparing pairs of tree sequences than the original matched span and inverse matched span, whose units are length of spans.
+We compute the ARF and TPR of two tree sequences in \figref{fig:conceptual_discrepancy}.
 
@@
         In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
-        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$.
-Thus the maximal mapping for 4 should be to itself.
+        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. Thus the maximal mapping for 4 should be to itself. 
         On the rest of the genome, all of nodes match with their identical counterpart.
-        This makes the dissimilarity $\dis(T1,T2)=3$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
+        This makes the matched span $\match(T1,T2)=43$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
+        Given the above matching, the inverse matching will match nodes $0$ through $7$, and node $9$ has no match 
+        since its only possible match ($4 \in T1$) was not the best match from $T1 \to T2$. 
+        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.
         \label{fig:conceptual_discrepancy}
@@
 \paragraph{Metrics on ARGs}
-The dissimilarity $\dis(\T_1,\T_2)$ is clearly not a metric in the mathematical sense
+Neither the matched span or non-matched span of $\T_1$ in $\T_2$ are not metrics in the mathematical sense
 (i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).
@@
 ``How much of the true relationships does this ARG represent?''
-However, it is worth noting that the symmetrized version
-$d(\T_1,\T_2) = \dis(\T_1,\T_2) + \dis(\T_2, \T_1)$ is a metric.
+However, it is worth noting that the symmetrized version of non-matched span
+$$\overline{\nmatch}(\T_1,\T_2) = \nmatch(\T_1,\T_2) + \nmatch(\T_2, \T_1),$$
+ is a metric.
 To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
@@
 
-These statistics are also reflected at the genomic scale, using measures of dissimilarity.
+These statistics are also reflected at the genomic scale, using measures of matched span.
 Line ``SE'' in \figref{fig:dissimilarity} shows total amounts of span removed by simplification
@@
 produced by extending haplotypes is very small ($\approx 1\%$).
-(Simplification does not produce dissimilarity.)
+(Simplification does not produce non-matched span.)
 The bottom row shows the proportion of the original ARG that is represented in the given ARG (TRP).
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{figures/Figure5-plt.pdf}
+        \includegraphics[width=\textwidth]{figures/figure6-ARF-TPR-plt.pdf}
 	\end{center}
@@
 we see that inclusion of these unary spans increases the amount of correctly inferred material
-by around 20\% (bottom panels),
-but that roughly 10\% of the unary spans in the inferred ARG are incorrect (top panels).
-Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span.
+by around 2\% (bottom panels),
+but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
+Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span ($\approx <1\%$).
 Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
@@
 
-\section{Acknowledgments}
-The authors would like to thank Dr. Yan Wong for his helpful comments and suggestions.
-
 \bibliography{references}

### 2025-02-22 paper/writeup.tex
figure-6changes (hopefully final)

@@
 \newcommand{\tsinfer}{\texttt{tsinfer}}
+\newcommand{\tsdate}{\texttt{tsdate}}
 \newcommand{\msprime}{\texttt{msprime}}
@@
  those of their best match in $\T_2$.
-Contrarily, given a matching $\beta:\T_1\to\T_2$, 
+Contrarily, given a matching $\alpha:\T_1\to\T_2$, 
 we want to quantify how much of $\T_2$ is represented in $\T_1$.
@@
 	\begin{center}
-        \includegraphics[width=\textwidth]{figures/figure6-ARF-TPR-plt.pdf}
+        \includegraphics[width=\textwidth]{figures/figure6-ARF-TPR-plt3.pdf}
 	\end{center}
@@
         was (i) \emph{simplified} (`S'), removing the unary spans,
-        and (ii) \emph{inferred} (`I'), using \tsinfer{} on genotypes;
+        and (ii) \emph{inferred} (`I'), using \tsinfer{} on genotypes
+        and dated using \tsdate{};
         then each of these had its haplotypes \emph{extended} (`SE', `IE').
@@
 on the results.
+Since our matched spans method breaks ties with time, 
+we date the re-inferred ARG using \tsdate.
 First, \figref{fig:dissimilarity} shows that \tsinfer{}
@@
 we see that inclusion of these unary spans increases the amount of correctly inferred material
-by around 2\% (bottom panels),
+by around 4\% (bottom panels),
 but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
@@
 of subtending samples.) 
+We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Figure-6}.
 
@@
 
+\begin{table}[!h]
+	\begin{center}
+		\begin{tabular}{{|c||c|c|c|c|c||c|c|c|c|c|}}
+			\hline
+			& \multicolumn{5}{|c||}{ARF} & \multicolumn{5}{|c|}{TPR}\\
+			\hline
+			\hline
+			Sample Size   & 10 & 50 & 100 & 500 & 1000 & 10 & 50 & 100 & 500 & 1000\\
+			\hline\hline
+			I & 0.3808 & .3490 & 0.3218 & 0.2734 & 0.2552 & 0.2904 & 0.2873 & 0.2908 & 0.3099 & 0.3299\\
+			\hline
+			IE & 0.4518 & 0.3932 & 0.3584 & 0.2932 & 0.2726 & \textbf{0.2917} & \textbf{0.2885} & \textbf{0.2916} & \textbf{0.3110} & \textbf{0.3309}\\
+			\hline
+			IS & \textbf{0.1993} & \textbf{0.1802} & \textbf{0.1701} & \textbf{0.1577} & \textbf{0.1516} & 0.2717 & 0.2703 & 0.2732 & 0.2909 & 0.3109\\
+			\hline
+			ISE & 0.3129 & 0.2779 & 0.2587 & 0.2332 & 0.2189 & 0.2850 & 0.2819 & 0.2848 & 0.3023 & 0.3228\\
+			\hline
+			\hline
+			Length   & $1\textnormal{e}6$ & $5\textnormal{e}6$ & $1\textnormal{e}7$ & $3\textnormal{e}7$ & $5\textnormal{e}7$ & $1\textnormal{e}6$ & $5\textnormal{e}6$ & $1\textnormal{e}7$ & $3\textnormal{e}7$ & $5\textnormal{e}7$\\
+			\hline\hline
+			I & 0.1383 & 0.2167 & 0.2324 & 0.2525 & 0.2595 & 0.5259 & 0.4478 & 0.4108 & 0.3540 & 0.3317\\
+			\hline
+			IE & 0.1474 & 0.2289 & 0.2469 & 0.2702 & 0.2754 & \textbf{0.5263} & \textbf{0.4485} & \textbf{0.4117} & \textbf{0.3551} & \textbf{0.3326}\\
+			\hline
+			IS & \textbf{0.0948} & \textbf{0.1359} & \textbf{0.1407} & \textbf{0.1485} & \textbf{0.1526} & 0.5172 & 0.4331 & 0.3939 & 0.3356 & 0.3118\\
+			\hline
+			ISE & 0.1193 & 0.1827 & 0.1960 & 0.2140 & 0.2216 & 0.5232 & 0.4432 & 0.4051 & 0.3481 & 0.3234\\
+			\hline
+		\end{tabular}
+		\caption{Values for 'I', 'IE', 'IS', 'ISE' in \figref{fig:dissimilarity}.}
+		\label{tab:Fig-6}
+	\end{center}
+\end{table}
+
 \end{document}

### 2025-02-22 create-data-fig6.py
data for Figure-6

@@ -2,28 +2,58 @@ import tskit
 import pandas as pd
 from remove_isolated_unary import remove_isolated_unary
 import tscompare
+import tsdate
+import msprime
+import tsinfer
 
 # Data with varying sample
 samplelist = [10, 50, 100, 500, 1000]
-sample_arf = pd.DataFrame(columns = samplelist, index=['S', 'SE', 'I', 'IE', 'IS', 'ISE'])
-sample_tpr = pd.DataFrame(columns = samplelist, index=['S', 'SE', 'I', 'IE', 'IS', 'ISE'])
+sample_arf = pd.DataFrame(columns = samplelist, index=['S', 'SE', 'I', 'IE', 'IS', 'ISE', 'ID', 'IDE', 'IDS', 'IDSE'])
+sample_tpr = pd.DataFrame(columns = samplelist, index=['S', 'SE', 'I', 'IE', 'IS', 'ISE', 'ID', 'IDE', 'IDS', 'IDSE'])
 
 for sample in samplelist:
     # generate true tree seqeunce ts and inferred tree sequence infer_ts
-    # ts = msprime.sim_ancestry(sample, population_size=1000, sequence_length=5e7,
-    #                         recombination_rate=1e-8, coalescing_segments_only=False)
-    # ts.dump(f'trees/ts_{sample}s_5e7')
-    # mutation_rate = 1e-7  # mutation rate for the inferred tree sequences
-    # ts = msprime.sim_mutations(ts, mutation_rate)
-    # infer_ts = tsinfer.infer(tsinfer.SampleData.from_tree_sequence(ts))
-    # t = infer_ts.tables
-    # t.compute_mutation_times()
-    # infer_ts = t.tree_sequence()
-    # infer_ts.dump(f'trees/infer_ts_{sample}s_5e7')
-    ts = tskit.load(f'trees/ts2_{sample}s_5e7_nounary')
-    infer_ts = tskit.load(f'trees/infer_ts2_{sample}s_5e7_nounary')
+    ts = msprime.sim_ancestry(sample, population_size=10000, sequence_length=5e7,
+                            recombination_rate=1e-8, coalescing_segments_only=False)
+    mu = 1.29e-8  # mutation rate for the inferred tree sequences
+    ts = msprime.sim_mutations(ts, mu)
+    infer_ts = tsinfer.infer(tsinfer.SampleData.from_tree_sequence(ts))
+    t = infer_ts.tables
[…]

### 2025-02-23 paper/writeup.tex

@@
 where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise),
-and let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total sum of nodes spans for nodes in $\T_1$.
+and let $\|\T_1\\|=\sum_{n\in N_1}s(\T_1,n)$ be the total span of all nodes in $\T_1$.
 We then define the \emph{non-matched span} of $\T_1$ in $\T_2$ by
@@
 whose best match is $n_2$.
-Note that the inverse matched span is not equivalent
- to the matched span of $\T_2$ in $\T_1$ 
- as the inverse matched span maximizes over spans
-  which are still dependent on $\beta$, 
-  and, as previously stated, $\beta$ is not guaranteed to be bijective.
+This differs from the matched span of $\T_2$ in $\T_1$
+because there may be more than one node in $\T_1$
+that is mapped to the same node in $\T_2$ --
+so, if nodes $n_1$ and $n_1'$ are both matched by $\alpha$ to the same node $n_2$,
+then both count towards $\match(\T_1,\T_2)$,
+but only the better match counts towards $\invmatch(\T_1,\T_2)$.
 
@@
 \paragraph{Metrics on ARGs}
-Neither the matched span or non-matched span of $\T_1$ in $\T_2$ are not metrics in the mathematical sense
+Neither the matched span or non-matched span of $\T_1$ in $\T_2$ are metrics in the mathematical sense
 (i.e., symmetric, nonzero distance between distinct points, and satisfying the triangle inequality).

### 2025-02-23 paper/writeup.tex

@@
 \newcommand{\tskit}{\texttt{tskit}}
+\newcommand{\tscompare}{\texttt{tscompare}}
 \newcommand{\tsinfer}{\texttt{tsinfer}}
@@
 on inference through a range of approaches~\citep{rasmussen2014genome,
-speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust}.
+speidel2019method,kelleher2019inferring,zhang2023biobankscale,deng2024robust,gunnarsson2024scalable}.
 One way of viewing ARGs is as a sequence of local trees,
@@
 % (as defined in the context of the ARG, \citep{XXX}).
-There has been some recent attention to haplotypes:
-\citet{ignatieva2024length} compared distributions of haplotype spans in true and inferred ARGs,
-as well as more sophisticated summaries of edges.
 Also, \citet{deng2021distribution} evaluated agreement of distributions of
@@
 that is the total variation distance between the induced distribution on genotypes;
-neither of which measure the sharing of haplotypes between adjacent trees.
-However, the additional information provided by haplotype structure can be important:
+however, neither of these measure the sharing of haplotypes between adjacent trees.
+An exception is
+\citet{ignatieva2024length}, who compared distributions of haplotype spans in true and inferred ARGs,
+as well as more sophisticated summaries of edges.
+The additional information provided by haplotype structure can be important:
 for instance, haplotypes that extend over many local genealogies
@@
     \begin{center}
-    \includegraphics[width=0.9\textwidth]{figures/conceptual_figure.pdf}
+    \includegraphics[width=0.8\textwidth]{figures/conceptual_figure.pdf}
     \end{center}
@@
 We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
-to take advantage of the extensive, stable, and well-documented set of open-source tools
-available in \tskit{} \citep{tskit}, 
+to take advantage of the % extensive, stable, and well-documented set of open-source 
+tools available in \tskit{} \citep{tskit}, 
 and our terminology and notation follows \citet{ralph2020efficiently}.
@@
 over a segment of genome $[\ell_e, r_e)$.
-(The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
+% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
 Suppose that the unique elements of the set of left and right edge endpoints
@@
 % JK note: earlier version said "relatively" - relative to what?
-An efficient algorithm to do this is described in \algorithmref{alg:extend}, 
-and implemented in \tskit{} as \texttt{extend\_haplotypes}.
+An efficient algorithm to do this is described in \algorithmref{alg:extend}.
+% and implemented in \tskit{} as \texttt{extend\_haplotypes}.
 The algorithm considers each tree transition from $T_k$ to $T_{k+1}$ in turn, updating its internal state
@@
 
-\algorithmref{alg:extend} simplifies the algorithm in \tskit{}
+\algorithmref{alg:extend} simplifies the full algorithm implemented in software % in \tskit{}
 in several ways for the sake of clarity --
@@
 need to be updated.
+
+\begin{algorithm2e}[!ht]
+	\SetStartEndCondition{ }{}{}%
+	\SetKwProg{Fn}{def}{\string:}{}
+	\SetKwFunction{Range}{range}%%
+	\SetKw{KwTo}{in}
+	\SetKwFor{For}{for}{\string:}{}%
+	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
+	\SetKwFor{While}{while}{:}{fintq}%
+	\SetKw{Break}{break}
+	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
+	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
+    \caption{
+        Extend haplotypes. Given an ARG $\T$ with $N$ trees $T_1, \ldots, T_N$
+        for which edges $O_k$ are removed to transition from $T_k$ to $T_{k+1}$,
+        identify and merge all mergeable paths (see text).
+        Each child node $c_e$ of each removed edge $e$ is checked to see if it
+        is at the base of two mergeable paths;
+        paths that add fewer new edges are merged first.
+        (The variables $m$, $M$, and $M'$ are to ensure this ordering
+        by number of new edges.)
+    }
+	\label{alg:extend}
+	\SetKwFunction{mergeable}{Mergeable}
+	\SetKwFunction{merge}{Merge}
+	\SetKwFunction{extend}{ExtendHaplotypes}
+	\DontPrintSemicolon
+    \Fn{\extend($\T$)}{
+        \For{$k$ in $1 \ldots N$}{
+            \text{Set } $M = 0$ \text{ and }
+            $M' = \infty$.\;
+            \While{$M < \infty$}{
+                \For{$e \in O_k$}{
+                    \text{Set } $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\;
+                    \eIf{$m < M$}{
+                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
+                    }{
+                        \text{Set } $M' = \min(m, M)$. \;
+                    }
+                }
+                \text{Set } $M = M'$ \text{ and }
+                $M' = \infty$ .\;
+            }
+        }
+    }
+\end{algorithm2e}
 
@@
 
-\begin{algorithm2e}[!ht]
-	\SetStartEndCondition{ }{}{}%
-	\SetKwProg{Fn}{def}{\string:}{}
-	\SetKwFunction{Range}{range}%%
-	\SetKw{KwTo}{in}
-	\SetKwFor{For}{for}{\string:}{}%
-	\SetKwIF{If}{ElseIf}{Else}{if}{:}{elif}{else:}{}%
-	\SetKwFor{While}{while}{:}{fintq}%
-	\SetKw{Break}{break}
-	\newcommand\forcond{$i$ \KwTo\Range{$n$}}
-	\AlgoDontDisplayBlockMarkers\SetAlgoNoEnd\SetAlgoNoLine%
-    \caption{Extend haplotypes.}
-	\label{alg:extend}
-	\SetKwFunction{mergeable}{Mergeable}
-	\SetKwFunction{merge}{Merge}
-	\SetKwFunction{extend}{ExtendHaplotypes}
-	\DontPrintSemicolon
-    \Fn{\extend($\T$)}{
-        \For{$k$ in $1 \ldots N$}{
-            \text{Set } $M = 0$ \text{ and }
-            $M' = \infty$.\;
-            \While{$M < \infty$}{
-                \For{$e \in O_k$}{
-                    \text{Set } $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\;
-                    \eIf{$m < M$}{
-                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
-                    }{
-                        \text{Set } $M' = \min(m, M)$. \;
-                    }
-                }
-                \text{Set } $M = M'$ \text{ and }
-                $M' = \infty$ .\;
-            }
-        }
-    }
-\end{algorithm2e}
-
 Because the algorithm needs to take multiple passes over the tree sequence
@@
 we define a measure of \emph{matched span} to quantify this.
-The method is implemented in the \texttt{tscompare} package.
+The method is implemented in the \tscompare{} package.
 
@@
 Call the two tree sequences $\T_1$ and $\T_2$, which should have the same genome length
-and the same set of sample nodes.
+and the same set of sample nodes;
+in what follows we think of $\T_2$ as the true ARG and $\T_1$ as an inferred ARG.
 We would like to measure (a) how much of $\T_1$ is found in $\T_2$;
@@
 and then compute for how much of their spans do their descendant samples agree (or not).
-
-Concretely, our method can be applied to any two tree sequences,
-as illustrated in \figref{fig:conceptual_discrepancy}.
-However, to simplify notation suppose that the two tree sequences have the same set of breakpoints
+An example of our method is illustrated in \figref{fig:conceptual_discrepancy}.
+
+The method works as follows.
+To simplify notation suppose that the two tree sequences have the same set of breakpoints
 between trees,
@@
 \end{align*}
-to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs.
-Further, we define
+to be the indices of all trees where $n_1$ and $n_2$ are ancestral to the same sample set in both ARGs,
+and
 \begin{align*}
@@
 (Note that multiple nodes in $\T_1$ may be mapped to the same node in $\T_2$,
-and that there some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
+and that some nodes in $\T_2$ may not be mapped to by any nodes in $\T_1$.)
 Since the maximum is independent over nodes, we may define for each node $n_1 \in \T_1$
@@
 where $\ind_{n \in T_k}$ is an indicator (i.e., it is 1 if $n \in T_k$ and 0 otherwise),
-and let $\|\T_1\\|=\sum_{n\in N_1}s(\T_1,n)$ be the total span of all nodes in $\T_1$.
+and let $\|\T_1\|=\sum_{n\in N_1}s(\T_1,n)$ be the total span of all nodes in $\T_1$.
 We then define the \emph{non-matched span} of $\T_1$ in $\T_2$ by
@@
 that is mapped to the same node in $\T_2$ --
-so, if nodes $n_1$ and $n_1'$ are both matched by $\alpha$ to the same node $n_2$,
+so, if nodes $n_1$ and $n_1'$ are both mapped by $\alpha$ to the same node $n_2$,
 then both count towards $\match(\T_1,\T_2)$,
@@
 \begin{align*}
-    \mathcal{M}'(\T_1, n; \T_2) = \left\{ k : \exists\ n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
+    \mathcal{M}'(n) = \left\{ k : \exists\ n_2  \;\text{for which}\; S\left(T^{(1)}_k,\ n\right) = S\left(T^{(2)}_k,\ n_2\right) \right\},
 \end{align*}
@@
 \begin{align*}
-    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(\T_1, n; \T_2)} (a_k - a_{k-1})
+    m'(n, \T_2) = \sum_{k \in \mathcal{M}'(n)} (a_k - a_{k-1})
 \end{align*}
@@
 but not times.
-If the ARG is dated~\citep{wohns2022unified}, 
+If the ARG is dated~\citep[e.g., as in][]{wohns2022unified,deng2024robust}, 
 we can naturally use the ``best match'' $\alpha$
@@
 
-The implementation of this method in \texttt{tscompare}
+The implementation of this method in \tscompare{}
 additionally produces relative values:
@@
 for comparing pairs of tree sequences than the original matched span and inverse matched span, whose units are length of spans.
-We compute the ARF and TPR of two tree sequences in \figref{fig:conceptual_discrepancy}.
+We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.
 
@@
 that takes into account span along the genome of inferred ancestral haplotypes
-(and separates the metric into two pieces as discussed in the previous paragraph).
+(and separates the metric into two pieces).
 However, the RF metric has many undesirable properties --
@@
 
-Our new method \textit{extend haplotypes} is applicable to any ARG.
+Our method for extending haplotypes is applicable to any ARG.
 However its accuracy depends on the overall structure of the ARG it is applied to. 
-Additionally, as \textit{ARG Robinson-Foulds}
-is given as a proportion of node spans
-its value is also dependent on the pair of input ARGs. 
+% Additionally, as \textit{ARG Robinson-Foulds} is given as a proportion of node spans
+% its value is also dependent on the pair of input ARGs. 
 Thus to understand how well our methods can infer ancestral haplotypes we 
-work with simulated tree sequences with a large variety in parameter choice.
+work with ARGs simulated across a range of parameter values.
 To do this, we simulate ARGs containing full haplotypes using \msprime{}
@@
 Although \msprime{} simulates many events that do not create a coalescence in some local tree,
-it only outputs information for nodes which contain a coalescence
+by default it only outputs information for nodes which contain a coalescence
 (i.e., are the MRCA of some pair of samples at some point on the genome).
@@
 (on which the node is unary).
-However, following this example, the first lineage might carry a segment $[x, y)$
+However, following this example, the first lineage might also carry a segment $[x, y)$
 that is disjoint from the segment $[a, d)$.
@@
 All jobs for which runtime was recorded were executed on an Intel Xeon Gold 6148 processor.
- Additionally, we compare accuracy between sequences modified from a ``true'' tree sequence
+Additionally, we compare accuracy between sequences modified from a ``true'' ARG
 using our dissimilarity methods. 
- These tree sequences have a population of 1,000 and recombination rate of $10^{-8}$ with samples ranging from 10 to 1,000 and genome length ranging from $10^6$ to $5\times 10^7$.
+These ARGs were simulated with an effective population size of 1,000
+and recombination rate of $10^{-8}$ with between 10 and 1,000 samples
+and a genome length between $10^6$ to $5\times 10^7$.
 
@@
 (except those pendant to the root and some of those adjacent to chromosome ends),
-leading to a reduction in number of edges by a third.
+leading to a reduction in number of edges by almost exactly one third.
 Experiments with an earlier version of the algorithm showed that
-if we only extend haplotypes on ``paths of length 1'', 
+if we only extend haplotypes on such ``paths of length 1'', 
 then the hypothesized reduction of 1/3 is achieved for long sequences.
@@
 running at chromosome scale in seconds to minutes for hundreds or thousands of samples,
-or minutes to hours for tens of thousands of samples
+or minutes to hours for tens of thousands of samples.
 
@@
         The effect of extending haplotypes on per-node spans
-        in an ARG simulated with $10^4$ diploid samples, $10^4$ population, and 
-        recombination rate of $10^{-8}$ on a sequence of length $10^8$.
+        in an ARG simulated with $10^4$ diploid samples in a population with $N_e=10^4$,
+        and recombination rate of $10^{-8}$ on a sequence of length $10^8$.
         \textbf{(A)}
@@
 (Lines labeled with ``I'' involve re-inference of the ARG; discussed next.)
-The top row shows the proportion of the given ARG that does not match the original (ARF),
+The top row shows the proportion of the given ARG that does not match the original (``ARF''),
 showing that the total amount of mis-matching span
 produced by extending haplotypes is very small ($\approx 1\%$).
-(Simplification does not produce non-matched span.)
-The bottom row shows the proportion of the original ARG that is represented in the given ARG (TRP).
+(Simplification does not produce non-matched span, so the ``S'' line is at zero.)
+The bottom row shows the proportion of the original ARG that is represented in the given ARG (``TRP'').
 This shows that a large proportion of an ARG can be removed by simplification,
@@
 (Since the \texttt{simplify} operations removes these unary portions of haplotypes,
-the ``simplified'' line (``S'') on the right of Figure~\ref{fig:dissimilarity} shows
+the ``simplified'' line (``S'') on the bottom two plots of Figure~\ref{fig:dissimilarity} shows
 the proportion of nodes' spans on which they are not unary.)
@@
 (the ``simplified--extended'' line; ``SE'').
-For instance, with 1,000 samples and a $5 \times 10^7$bp genome,
+For instance, the rightmost points show that
+with 1,000 samples and a $5 \times 10^7$bp genome,
 haplotypes are unary on about half their spans (on average),
-and extending haplotypes can infer more than half of this span from coalescent information only.
+and extending haplotypes can infer more than half of this missing unary span from coalescent information only.
 
@@
         Accuracy and sensitivity of extended haplotypes
-        across a range of sample sizes, on a sequence of length $5\times10^7$ (right)
-        and sequence lengths, with 1,000 diploid samples (left).
+        across a range of sample sizes, on a sequence of length $5\times10^7$ (right);
+        and a range of sequence lengths, with 1,000 diploid samples (left).
         For each, a simulated ARG containing unary haplotype spans
@@
 but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
-Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output with tsinfer adds relatively little span ($\approx <1\%$).
+Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output
+with tsinfer adds relatively little span (less than about $1\%$).
 Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
@@
 of subtending samples.) 
-We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Figure-6}.
+We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Fig-6}.
 
@@
 
+\section{Data availability}
+
+The method to extend haplotypes described here is available through the \tskit{}
+python and C APIs (\url{https://tskit.dev/tskit}) as \texttt{extend\_haplotypes};
+methods to compare ARGs are implemented in the \tscompare{} python package
+(\url{https://tskit.dev/tscompare}).
+\comment{Scripts used to produce the results in this paper are available at
+\url{https://github.com/XXX/TODO}.}
+
 \section{Discussion}
@@
 Further work is needed to determine how the balance of ``true and false positive rates''
-affects downstream uses.
-The efficient computational tools we have implemented (in \tskit{})
+affects downstream uses,
+and whether results would differ if the requirement that sets of subtended nodes match exactly
+was relaxed.
+The efficient computational tools we have implemented (in \tskit{} and \tscompare{})
 should facilitate this exploration.
@@
 which turns out to be NP-complete \citep{wang2001perfect}.
-For instance, \citet{minichiello2006mapping}
-used a heuristic algorithm to produce ``plausible'' ARGs
+So, the field turned to more heuristic methods --
+for instance, \citet{minichiello2006mapping}
+used an algorithm to produce ``plausible'' ARGs
 (i.e., those that explained the data with few mutations and recombinations),
@@
 the two genomes share an IBD segment if each has inherited the segment from their common ancestor
-along a single path \citep{ralph2013geography}.
+along a single path \citep[e.g.,][]{ralph2013geography}.
 Largely for computational reasons, this is the definition that is used in \tskit's
 IBD-finding methods 
-(by G.~Tsambos \citet{tskit}; see also \citet{tsambos2023link}).
+(by G.~Tsambos in \tskit{} \citep{tskit}; see also \citet{tsambos2023link}).
 However, many simulation and inference methods produce ARGs
@@
 \renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
 \setcounter{figure}{0}

### 2025-02-24 paper/writeup.tex
minor edits

@@
     \caption{
-        For two tree sequences $T1$ and $T2$ the \emph{dissimilarity}, $\dis(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
+        For two tree sequences $T1$ and $T2$ the \emph{matched span}, $\match(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
         based on identical sample sets.
@@
 Additionally, we compare accuracy between sequences modified from a ``true'' ARG
-using our dissimilarity methods. 
-These ARGs were simulated with an effective population size of 1,000
+using our matched span methods. 
+These ARGs were simulated with an effective population size of 10,000
 and recombination rate of $10^{-8}$ with between 10 and 1,000 samples
@@
 (\url{https://tskit.dev/tscompare}).
-\comment{Scripts used to produce the results in this paper are available at
-\url{https://github.com/XXX/TODO}.}
+Scripts used to produce the results in this paper are available at
+\url{https://github.com/hfr1tz3/haplotypes-and-ancestral-recombination-graphs}.
 

### 2025-05-08 paper/review-responses.tex
added review responses

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Halley Fritze  \\
  \today
\end{minipage}

\vskip 2em

\noindent
{\bf To the Editor(s) -- }

\vskip 1em

We are writing to submit a revised version of our manuscript,
GENETICS-2025-307917 
titled
`` A forest is more than its trees: haplotypes and ancestral recombination graphs''.
We appreciate the thoughtful comments by the reviewers.
We've done our best to address all the comments,
as detailed in our Response to Reviewers.

In addition to a Response to Reviewers (in which page/line numbers refer to the revised manuscript file),
we also provide a pdf with differences to the previously submitted version highlighted.

\vskip 2em

\noindent \hspace{4em}
\begin{minipage}{3in}
\noindent
{\bf Sincerely,}

\vskip 2em

{\bf
   Halley Fritze, Nathaniel Pope, Jerome Kelleher, and Peter Ralph
}\\
\end{minipage}

\vskip 4em

\pagebreak

%%%%%%%%%%%%%%
\reviewersection{1}

\begin{quote}
In this paper Fritze and colleagues take a haplotype-centric view of the ancestral recombination graph (ARG) and contrast this view with the more common simplification of the ARG as a sequence of marginal trees. In particular, internal nodes in the ARG represent ancestral haplotypes, but in the process of converting an ARG to a succinct tree sequence, some of these internal nodes are deleted (essentially if one of the lineages that descend from the node recombines and coalesces elsewhere, then that node would become unary and is not represented in the succinct tree sequence). As such, a given node in a succinct tree sequence may not span as many marginal trees as it ``should'', and this paper develops an algorithm to try to infer where these unary nodes should be added, which the authors refer to as extending haplotypes. To assess how well their algorithm can recover trees, the authors develop novel dissimilarity measures on ARGs, and use these on simulated true and inferred ARGs. Overall, the extension method works quite well on true ARGs, but less well on inferred ARGs. I think that this paper is an important contribution to the ARG literature -- this haplotype perspective is important and generally overlooked, and measures for assessing how well methods can recover ARGs in terms of this perspective will be useful for future method development. Overall the paper is very nice, well-written, interesting, and clear, but I do have a few comments for the authors in the hopes that they're useful. They're all minor, and the authors should feel free to just use what they find helpful.
\end{quote}

\begin{point}{}
 I really enjoyed the title of the paper!
\end{point}

\reply{
Thank you!
}

\begin{point}{\revref}
 I am a bit confused by the definition of weighted Robinson-Foulds distance at \revref.% on the bottom of page 6. 
I believe that this is more like a Robinson-Foulds \emph{similarity}, as I believe that as defined it should be maximized when the two ARGs are the same, and smaller for more dissimilar ARGs.
\end{point}

\reply{
}

\begin{point}{}
 Is the dissimilarity measure defined in equation \eqref{eqn:wrmse} used anywhere?
\end{point}

\reply{
}

\begin{point}{}
 What is ``dissimilarity'' as mentioned in the caption of \figref{fig:conceptual_discrepancy}?
\end{point}

\reply{
}

\begin{point}{\revref}
 I found the discussion around the metric version to be hard to follow and worth expanding.
For example, I believe that the edges of the ARG are what's getting embedded into
$[0, L) \times N \times N$, and the symmetric differences are taken in this space in the obvious way (i.e., without this mapping its unclear what the relative difference of two ARGs is).
There's also a mention of ``dissimilarity'' in this paragraph. I assume that that's the ``symmetrized version of non-matched span'' that was just defined, but it would be good to make that clear.
\end{point}

\reply{
}

\begin{point}{}
 The performance on inferred ARGs makes clear that this is a difficult problem, and that this haplotype view might be useful conceptually and for methods development, but that as of now we should probably not take these ancestral haplotypes at face value in inferred ARGs. It would be helpful if the authors could provide some intuition for what's going wrong in this case. Presumably the topologies are not inferred well, and many nodes can be extended, which often results in some true and false positives? Another use case I could see would be using this approach to simplify inferred ARGs to make various downstream computations faster (a la \figref{fig:speed_and_edges}). As such, I would be curious to see how the extension algorithm affects the number of edges in inferred ARGs.
\end{point}

\reply{
}

\begin{point}{}
 This is obviously a matter of taste, but I would have found the paper a bit easier to follow if some of the points discussed in the discussion were brought up earlier in the paper. For example, the ``ignorance and omission in the ARG'' section \llname{ignorance}
answered questions I had when reading the bits around \figref{fig:conceptual} -- upon initially reading the bit around \figref{fig:conceptual}
I was confused by the interpretation that the lack of a node implies that that individual is not represented in other parts of the ARG (as any point along any branch in an ARG represents \emph{some} individual ancestral to sample), and as a result I found the presented interpretation somewhat unnatural.
This was then eventually clarified (very eloquently) in the discussion.
Similarly, I would have appreciated the connection to parsimony earlier.
It was not completely obvious to me that recovering the unary nodes from the succinct tree sequence might not be possible in all cases, and as such that the algorithm is an attempt at inference, and furthermore, that it's optimizing some reasonable objective. Again, this point was addressed very well, but I had to wait until the discussion.
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{2}

\begin{quote}
The authors present interesting results in a clear and well-written manuscript, providing a deeper understanding of Ancestral Recombination Graphs and developing improved inference methods. The key idea is to take into account the additional information provided by haplotype structure which has rarely been considered before. On the way, the authors also develop a haplotype-aware measure of (dis)agreement between ARGs. In particular, the achieved goal is to infer information about the intermediate (unary) nodes in the local trees and to show how accurate the inference is.
\end{quote}

\begin{point}{\revref} % p.2 l.53
 ``Notation and terminology'': it would be good to clarify at this point that parents and children in the graph correspond to ancestors and descendants in the population.
\end{point}

\reply{
}

\begin{point}{\revref} % p.4 l.10
 ``in practice we do repeated passes in both directions until no changes can be made'': to someone who does not have hands-on experience with the implementation of this algorithm, the reason why the algorithm needs to do several passes in both directions might not be very clear. If at each tree transition all pair of mergeable paths are merged, it would be good to explain why, after one pass in each direction, additional mergeable paths are found. Perhaps explaining when/why additional mergeable paths are found/created could also give further insight on how many passes will be needed?
\end{point}

\reply{
}

\begin{point}{\revref} % p.8 l.25
 ``we simulate ARGs containing full haplotypes using msprime'': since msprime can simulate ARGs under various models, please specify here which model is used (SMC, SMC'... ?).
\end{point}

\reply{
Good suggestion; we've done this (we used ``hudson''). \revref
}

\begin{point}{}
In general, it comes natural to wonder whether the algorithms can be corrected to avoid incorrectly added spans
(e.g. \revref). % p.10 l. 5).
It would be worth to mention whether an attempt of correction has been made and to explain the difficulties/limitations involved.
\end{point}

\reply{
}

\begin{point}{\revref} % p.12 l.9
 ``Data availability'': the link https:// github.com/ XXX/ TODO should be updated.
\end{point}

\reply{
}

%%%%%%%%%%%%%%
\reviewersection{3}

\begin{quote}
The fundamental idea of this paper is important and the worked-out
algorithm appears quite valuable. I want to see it published. However
the current version is so difficult to read that it is likely to
have a very limited audience, and appears to have some errors.
\end{quote}

\begin{quote}
I do not feel I have understood the entire paper despite several
hours' work. Some of my comments may be misunderstandings, but if
so, the text I misunderstood could probably use clarification.
\end{quote}

\begin{point}{} % (1)
The description of \algorithmref{alg:extend} and \algorithmref{alg:mergeable} is confusing.
Naming them this way implies that they are separate algorithms but
in fact ``Algorithm 2'' is the definition of a function needed by
Algorithm 1.
\end{point}

\reply{
}

\begin{point}{}
\algorithmref{alg:mergeable} is mentioned in an explanation a full page before it
is introduced.
\end{point}

\reply{
}

\begin{point}{}
\algorithmref{alg:extend} seems to be named ExtendHaplotypes and \algorithmref{alg:mergeable} seems
to be named Mergeable: having two names for each one is not helpful.
\end{point}

\reply{
}

\begin{point}{}
``Mergeable'' is an unfortunate name for a function which does not
return a yes or no answer but instead returns the number of
new edges to be added. Also the text needs a comment about why
infinity is sometimes returned.
\end{point}

\reply{
}

\begin{point}{}
In \algorithmref{alg:extend} the variable $m$ means the number of new edges to be
added. In \algorithmref{alg:mergeable} that quantity is called $m_e$ and $m$ means the
number of extended haplotypes. This is confusing. Please use the
same name for a given concept in all pseudocode.
\end{point}

\reply{
}

\begin{point}{}
Comments in the pseudocode would make it easier to follow, as could
use of more meaningful variable names (I particularly dislike $M$, $m$, $M'$
in the same algorithm).
\end{point}

\reply{
}

\begin{point}{}
Capital $I$ is an unfortunate subscript as it is almost indistinguishable
from digit 1.
\end{point}

\reply{
}

\begin{quote} % (2)
\textbf{Issues in algorithms.}
\end{quote}

\begin{point}{}
\algorithmref{alg:extend} appears to have an error leading to an infinite
loop. Variable $M$ is initially set to 0. It is compared to variable $m$,
which is a number of new edges to be added and must be positive or zero.
Thus, $m$ will never be lower than $M$, so the action taken is to set $M'$ to
the minimum of $m$ and $M$, which will always be zero. $M$ is then set to
$M'$, that is, to zero. Since this is inside a while loop with terminating
condition $M$ == infinity, it does not terminate. (I coded it to check.)

The logic problem is probably around line 10. We have just tested
whether $m < M$ and this was false, so $m \ge M$. It is therefore
strange to take the minimum of $m$ and $M,$ as this will always be
$M$. However I can't work out the correct code.
\end{point}

\reply{
Good catch -- and, many apologies --
that is indeed a bug; the correction is changing line 7 to
``if $m \le M$''.
(To be clear, the bug was in the paper, not the implemented code;
if the reviewer is interested they can see our implementation
in the tskit library at
\href{https://github.com/tskit-dev/tskit/blob/c6b06b0c4f90b5098aedddca7dd0e3fbbad964d0/c/tskit/trees.c#L9078}{c/tskit/trees.c:9078} in the current release.)
}

\begin{point}{}
The text description of \algorithmref{alg:extend} (ExtendHaplotypes) says that
paths with the fewest new edges are added first. The pseudocode
given does not appear to do that. It appears (bearing in mind that
it has at least one bug, so I can't be positive) to be willing to
execute a Merge before examining all children, so I think it will
Merge the first child with a better score than previous children,
regardless of the fact that an even better child may remain to be
found.
\end{point}

\reply{
We apologize again for the bug in the previous point that caused this misunderstanding.
Fixed, it should perform as advertised
(and, we've added a bit clarifying why it works
to the description of \algorithmref{alg:extend}).
}

\begin{point}{}
After considerable work I have been unable to code \algorithmref{alg:mergeable}.
When is a node NULL? Does this mean the node doesn't exist (for
example, the parent of the root)?
\end{point}

\reply{
}

\begin{point}{} % (3)
I recommend that the example partial tree in \figref{fig:conceptual} be
extended to a full tree (it only needs one more node) and used in
the practical examples in the text.
\end{point}

\reply{
Good idea, but on consideration we think that adding the rest of the tree
here would make the figure too confusing
(not so much for the trees in A\&C, but more so in the haplotype diagrams B\&D).
We do now reference the figure in the text,
at \revref{} and \llname{ref_conceptual2}.
}

\begin{point}{} % (4) 
The concrete example under ``An algorithm to extend haplotypes'' is
difficult to visualize. Can it be related to \figref{fig:conceptual}? When I try to
do this, though, I have difficulty because in \figref{fig:conceptual} the right-hand
tree does not \emph{have} $v_1 \ldots v_n$, it jumps straight from $p$ to $c.$ Is
this example not mergeable, or is the description of mergeablility
missing any indication that some of the nodes it relies on may not
exist?
\end{point}

\reply{
}

\begin{point}{} % (5) 
The direction in which time is measured on the tree needs to be stated
as unfortunately the literature has both conventions. Also, the text
implies at several points that this can be used for trees without
time information, but it cannot as times are used in the algorithm.
\end{point}

\reply{
}

\begin{point}{}
Following on from this, in \algorithmref{alg:extend}, why can no node in $u$ be at the
same time as a node in $v?$ At first I thought we were assuming that no
two nodes, other than samples, are ever at the same time (a fairly
common assumption) but then we would not need this rule. Is this just
a convenience to avoid putting two nodes on the same branch at the same
timepoint? What happens if this algorithm is attempted with inferred
ARGs that discretize time, as some inference algorithms do, so that ties
become common?
\end{point}

\reply{
}

\begin{point}{} % (6) 
``In what follows we think of $T_2$ as the true ARG and $T_1$ as an
inferred ARG.'' This is not strongly enough stated. The measure is
asymmetrical with regards to $T_1$ and $T_2$, and will give wrong answers
if they are reversed. It is also not suitable for comparing two
ARGs where neither is the true ARG.
\end{point}

\reply{
}

\begin{point}{}
In general the section on comparison of true and inferred ARGs is
difficult to follow. \figref{fig:conceptual_discrepancy} helps but appears too late in the
paper.
\end{point}

\reply{
}

\begin{point}{} % (7) 
In \figref{fig:conceptual_discrepancy} it is unfortunate to use to use an example where
the forward and inverse matches are the same. I would also appreciate
a brief explanation of why it's 46 and 47 in the denominator--I figured
it out but it's less than obvious. Why do sample nodes count, as they
cannot vary among trees? Doesn't this just add bookkeeping effort?
It's otherwise a very helpful figure though.
\end{point}

\reply{
}

\begin{point}{} % (8) 
I cannot clearly distinguish the three purplish colors in \figref{fig:speed_and_edges}.
\end{point}

\reply{
}

\begin{point}{} % (9) 
The interpretation of \figref{fig:node-spans} is completely unclear to me. I
think it needs more explanatory text. I have no idea what point is being
made by these diagrams.
\end{point}

\reply{
}

\begin{point}{} % (10) 
I remind the authors to replace the script availability placeholder
URL with the real thing! 
\end{point}

\reply{
}

### 2025-05-08 paper/writeup.tex
funding
added review responses

@@
 
-\SetKwInput{input}{Input}
-\SetKwInput{output}{Output}
+% Responses to reviews
+\input{review-response-commands}
+% set this to show line numbers and include responses to reviews or not
+\newif\ifreviewresponses
+\reviewresponsestrue  % include them
+% \reviewresponsesfalse  % don't include them
+\newcommand{\responsefile}{review-responses}  % name of the review responses file
+
+% redefining \input{} messes things up (and seems like a bad idea)
+% \SetKwInput{input}{Input}
+% \SetKwInput{output}{Output}
 
@@
 % Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
-Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$,
+Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$, \revpoint{2}{1}
 over a segment of genome $[\ell_e, r_e)$.
@@
 and conversely the nodes $\{v_j\}$ should lie on that path in tree $T_k$.
+For instance, in \figref{fig:conceptual} take $p=4$ and $c=1$,
+so that $u_1=3$ (and $m=1)$ and there are no $v$ (so $n=0$); \revpoint{3}{11}
+and so as shown in \figref[B\&D]{fig:conceptual},
+we might extend node 3 over this entire segment.
 Of course, this does not always make sense --
@@
 and $I$ defines a sub-forest $F_I$ of $T_{k+1}$.
+In \figref{fig:conceptual} (in the portion of the tree shown),
+the removed edges that make up $F_O$ \llabel{ref_conceptual2}
+are $4 \to 3$, $3 \to 1$, and $3 \to 2$,
+and the added edges that make up $F_I$ are $4 \to 1$ and the new edge leading to 2.
 The key step in the algorithm is to determine whether the pair of paths
@@
 Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
-in practice we do repeated passes in both directions until no changes can be made.
+in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
 The main step that is omitted is a description of the \texttt{merge} operation,
@@
         is at the base of two mergeable paths;
-        paths that add fewer new edges are merged first.
-        (The variables $m$, $M$, and $M'$ are to ensure this ordering
-        by number of new edges.)
+        paths that add fewer new edges are merged first
+        because \texttt{Mergeable} returns the number of new edges required
+        to merge the two paths, and $M$ is always less than or equal to
+        the minimum number of new edges across all nodes in the current tree transition.
     }
@@
                     \text{Set } $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\;
-                    \eIf{$m < M$}{
+                    \eIf{$m \le M$}{
                         \merge($c_e$, $T_k$, $T_{k+1}$)\;
@@
 the total span over which $n$ finds a match.
-Then the average RF distance (averaged over locations in the genome)
+Then the average RF distance (averaged over locations in the genome) \revpoint{1}{2}
 is
@@
 
-\paragraph{Metrics on ARGs}
+\paragraph{Metrics on ARGs} \revpoint{1}{5}
 Neither the matched span or non-matched span of $\T_1$ in $\T_2$ are metrics in the mathematical sense
@@
 work with ARGs simulated across a range of parameter values.
-To do this, we simulate ARGs containing full haplotypes using \msprime{}
+To do this, we simulate ARGs containing full haplotypes using 
+Hudson's algorithm as implemented in \msprime{} \revpoint{2}{3}
 \citep{kelleher2016efficient,baumdicker2021efficient},
@@
 To do this, 
-we simulate ARGs containing full haplotypes using \msprime{},
+we simulate ARGs containing full haplotypes using Hudson's algorithm in \msprime{},
 apply the simplification algorithm~\citep{kelleher2018efficient,wong2024general}
@@
 99\% of all added span is correct;
-95\% of nodes have no incorrectly added span;
+95\% of nodes have no incorrectly added span; \revpoint{2}{4}
 and those incorrectly added spans are nearly always a small fraction of the original span.
@@
 
+\revpoint{2}{5}
 The method to extend haplotypes described here is available through the \tskit{}
@@
 
-\paragraph{Ignorance and omission in an ARG}
+\paragraph{Ignorance and omission in an ARG} \llabel{ignorance}
 As motivation, we presented above a ``historical'' view of ARGs --
@@
 The authors would like to thank Dr. Yan Wong for his helpful comments and suggestions.
+JK acknowledges EPSRC (research grant EP/X024881/1)
+and the Robertson Foundation.
+JK and PR acknowledge support from 
+the NIH NHGRI (research grant HG012473).
 
@@
 
+%%%%%%%%%%%%%%%
+\includereviews
+
 \end{document}

### 2025-05-15 benchmarks/plot_results.py
color change figures

@@ -42,7 +42,7 @@ length_vals = np.unique(df.length)
 length_cols = np.array([colors[n] for n in ['blue', 'green', 'lblue', 'lgreen', 'gold']])
 ns_vals = np.unique(df.num_samples)
 # ns_cols = plt.colormaps["plasma"](np.linspace(0, 1, len(ns_vals)))
-ns_cols = np.array([colors[n] for n in ['wine', 'magenta', 'red', 'gold', 'blue']])
+ns_cols = np.array([colors[n] for n in ['wine', 'red', 'gold', 'lgreen', 'blue']])
 
 # combined number of edges & runtime for Tajimas D
 fname = f"{basename}_absolute_values.pdf"

### 2025-05-16 paper/review-responses.tex
reviews (mostly) complete

@@
 
-\reply{
+\reply{ For Peter.
 }
@@
 
-\reply{
+\reply{No, the weighted mean square error defined in \eqref{eqn:wrmse} is not shown for our later experiments shown in \figref{fig:dissimilarity}, however it is always computed when using our implementation is \texttt{tscompare}.
+\revref
 }
@@
 
-\reply{
+\reply{``dissimilarity'' was our original language to describe the non-matched spans, and we must've missed one such reference to the old name.
+We have now changed the caption in \figref{fig:conceptual_discrepancy} to reflect that change.
+\revref
 }
@@
 
-\reply{
+\reply{Good point.
+Again ``dissimilarity'' is the original name we meant to remove, so it has been correctly changed to \emph{non-matched spans}.
+We have also added more explanation to our argument to improve its clarity.\revref
 }
@@
 
-\reply{
+\reply{Thanks for this feedback.
+Yes, if an inferred ARG is incorrect for certain nodes, our method of extend haplotypes exacerbates this issue as extending an incorrectly inferred node extends this incorrect node span even further.
+Based on our tests in \figref{fig:speed_and_edges} and \tabref{tab:edge-counts}, these results would additionally hold for the inferred ARGs.
+For this reason, we didn't include additional figures for edge counts and computation time on inferred ARGs.
 }
@@
 
-\reply{
+\reply{This is a valid point, however we think our discussion of these points was best put in the discussion section.
+With respect to the motivation, we were likely heavily relying on the phrase of a \emph{hypothetical ARG}, but we find this acceptable.
 }
@@
 
-\reply{
+\reply{Good observation. We have made this change. \revref
 }
@@
 
-\reply{
+\reply{Noted.
+We have written some follow up for that sentence at your suggestion.
+The creation of new mergable paths after a number of passes is something we thought about quite a bit, however could not come up with a good theoretical conclusion as to if there was a number of required passes before termination.
+We instead provide some empirical evidence to the number of required passes later in the section and \tabref{tab:edge-counts}.
+\revref
 }
@@
 
-\reply{
+\reply{This is a good point. It is certainly something we have thought about but we omitted additional information that would improve accuracy in place
+	of simplicity of the construction. We also though our empirical results were strong without additional correction.
+	We have made note of this point. \revref
 }
@@
 
-\reply{
+\reply{Now updated. \revref
 }
@@
 
-\reply{
+\reply{Yes, this is the case. Originally we had both \algorithmref{alg:extend} and \algorithmref{alg:mergeable} within 
+	one algorithm environment. However, we discuss both the \texttt{Mergeable} and \texttt{ExtendHaplotypes} functions
+	in enough detail that we thought it poignant to separate them into their own algorithms. We additionally decided to 
+	make \algorithmref{alg:extend} the first algorithm as it is the main construction and the most important of the two.
 }
 
-\begin{point}{}
+\begin{point}{} %(2)
 \algorithmref{alg:mergeable} is mentioned in an explanation a full page before it
@@
 
-\reply{
+\reply{That is true, but with our additional figures, the earliest the environment will appear is the next page following the previous figure.
+	Upon consideration we think an introduction about the algorithm's conceptual construction in the section 
+	and then properly defining is more easily read than the other way around.
 }
 
-\begin{point}{}
+\begin{point}{} %(3)
 \algorithmref{alg:extend} seems to be named ExtendHaplotypes and \algorithmref{alg:mergeable} seems
@@
 
-\reply{
+\reply{We like to think of ExtendHaplotypes and Mergable as the names of the functions contained in \algorithmref{alg:extend} and \algorithmref{alg:mergeable}, respectively.
 }
 
-\begin{point}{}
+\begin{point}{} %(4)
 ``Mergeable'' is an unfortunate name for a function which does not
@@
 
-\reply{
+\reply{This is a credible point, however we require counting new edges that would be created by a merged path so that we can determine, out of all possible mergeable paths between trees, which to merge based on minimizing the number of newly created edges.
+	For clarity, we have included a note about returning infinity in the description of \algorithmref{alg:mergeable}. \revref
 }
 
-\begin{point}{}
+\begin{point}{} %(5)
 In \algorithmref{alg:extend} the variable $m$ means the number of new edges to be
@@
 
-\reply{
+\reply{Good point, this change has been made. \revref
 }
 
-\begin{point}{}
+\begin{point}{} %(6)
 Comments in the pseudocode would make it easier to follow, as could
@@
 
-\reply{
+\reply{Good point.
+We have included some comments for clarity, however since $M$, $M'$ and $m_e$ are used simply for book keeping a while loop we don't think it's necessary to use distinct variables.
+\revref
 }
@@
 
-\reply{
+\reply{While we agree, within the context of our algorithm we are interested paths of inheritance 
+	between the edges removed from a local tree (the out-forest) and the edges just added into the next local tree (the in-forest). 
+	Because of our language use of out and in, we feel like our choice of $O$ and $I$ is reasonable.
 }
@@
 
-\reply{
+\reply{A node is NULL when it is the root of a local tree.
 }
@@
 
-\reply{
+\reply{Good point. We have connected our concrete construction in that section
+	 to reference \figref{fig:conceptual} to improve clarity. \revref
 }
@@
 
-\reply{
+\reply{We state how time is measured here \revref.
+We additionally do use time in our algorithm, but mainly as a way to break ties, as is the case with \eqnref{eqn:matched-spans}.
+This tie break will default to the first node if the ARG does not contain time information.
 }
@@
 
-\reply{
+\reply{We do not exactly assume that all nodes in an ARG occur at unique times, however we do assume that
+	any two nodes with the same descendants should be unique. 
+	Additionally, in our example if $u$ and $v$ occurred at the same time and had descendant $c$, they would be nodes with the same index
+	and the paths for $c$ in $T_O$ and $T_I$ would be from $c\to u=v$.
+	In the case of discritized time of inferred ARGs, this is an interesting question we have not yet thought about it depth. 
+	Peter do you have extra thoughts about this? 
 }
@@
 \reply{
+	We think that while $\match(\T_2,\T_1)$ might not be what the user intends for our method, it can still give some valuable information between $\T_1$ and $\T_2$.
+We additionally discuss this when comparing inverse matched spans between $(\T_1,\T_2)$ and matched spans between $(\T_2, \T_1)$ \revref.
+	While this method is developed for measuring the correctness of an inferred ARG to the true ARG, we also think of this as a function to compare two ARGs whether or not they are truly related.
+Also our theoretical results on this method being a metric could allow for distances in the space of all ARGs which we think is an interesting generalization within the field.
 }
@@
 
-\reply{
+\reply{We feel like it is optimal to keep the example that computes $\operatorname{ARF}$ and $\operatorname{TPR}$
+	until after the are defined, and hence why it is at the end of the section. We have however, heard your point and
+	so we have expanded our example to give more clarity. \revref
 }
@@
 
-\reply{
+\reply{Good point.
+It is unfortunate how the math works out, but we did construct the example as such to discern that the matching for node 4 requires a max to be taken over shared spans of its two possible matches 4 and 9.
+To show some variety, we additionally will compute $\match(T2,T1)$ to show some additional variety and hopefully give more clarity to the methods themselves \revref.
+Additionally we note that we are matching spans for all nodes in a pair of ARGs including the samples however they trivially match to themselves by construction (this is not a hassle for bookkeeping).
+We also must have that the two ARGs being compared has the same number of sample nodes.
 }
@@
 
-\reply{
+\reply{Thank you for this feedback.
+We have changed \figref{fig:speed_and_edges} and the subsequent supplementary figures (\figref{fig:speed_and_edges_constant}, \figref{fig:speed_and_edges_other}, \figref{fig:speed_and_edges_other_constant}) to have more color variety.
 }
@@
 
-\reply{
+\reply{Thanks for the feedback. We have added some additional feedback \revref{} and the caption of \figref{fig:node-spans}.
 }
@@
 
-\reply{
+\reply{Noted and now changed, thank you!\revref
 }

### 2025-05-16 paper/writeup.tex
reviews (mostly) complete

@@
 
-Consider the (small portion) of a hypothetical ARG in \figref[A]{fig:conceptual}.
+Consider the (small portion of a) hypothetical ARG in \figref[A]{fig:conceptual}.
 On the first portion of the genome (left-hand tree), the sample nodes (labeled 0, 1, and 2)
@@
 Each node $n \in N$ has a \emph{time} $t_n$,
-which is the amount of time in the past that the individual who carried that segment of genome lived.
+which is the amount of time in the past that the individual who carried that segment of genome lived. \revpoint{3}{13}
 Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
 % Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
-Each edge $e \in E$ describes inheritance between a parent node $p_e$ and child node $c_e$, \revpoint{2}{1}
+Each edge $e \in E$ describes inheritance between ancestors and their descendants within the population.
+The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
 over a segment of genome $[\ell_e, r_e)$.
@@
 For instance, in \figref{fig:conceptual} take $p=4$ and $c=1$,
-so that $u_1=3$ (and $m=1)$ and there are no $v$ (so $n=0$); \revpoint{3}{11}
+so that $u_1=3$ (and $m=1)$ and there are no $v$ (so $n=0$); \revpoint{3}{11}\revpoint{3}{12}
 and so as shown in \figref[B\&D]{fig:conceptual},
@@
 in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
-The main step that is omitted is a description of the \texttt{merge} operation,
+We require repeated passes due to the fringe cases of some paths not being mergeable until 
+additional structure is added by a future pass of the algorithm. 
+We suspect that most of these cases occur for paths with large time differences between the parent and child
+nodes, or when a lot of nodes in a path are coalescent, however it is not something we thoroughly investigated
+in the theoretical setting.
+We instead give some empirical results on the number of passes for the algorithm below.
+
+The main step that is omitted is a description of the \texttt{Merge} operation,
 which performs the actual extending of haplotypes.
-This algorithm is essentially the same as \texttt{mergeable} in \algorithmref{alg:mergeable},
+This algorithm is essentially the same as \texttt{Mergeable} in \algorithmref{alg:mergeable},
 except with additional bookkeeping.
@@
     \Fn{\extend($\T$)}{
+    	\tcc{For pairs of trees in the ARG, find mergeable paths.}
         \For{$k$ in $1 \ldots N$}{
@@
                 \For{$e \in O_k$}{
-                    \text{Set } $m$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\;
-                    \eIf{$m \le M$}{
-                        \merge($c_e$, $T_k$, $T_{k+1}$)\;
+                    \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Count the number of new edges to be merged.}
+                    \eIf{$m_e \le M$}{
+                        \merge($c_e$, $T_k$, $T_{k+1}$)\tcc*[r]{Merge extended haplotypes from $T_k$ into $T_{k+1}$.}
                     }{
-                        \text{Set } $M' = \min(m, M)$. \;
+                        \text{Set } $M' = \min(m_e, M)$.\tcc*[r]{Bookkeep $M$ to determine an ordering for path merging.}
                     }
@@
 \end{algorithm2e}
-
+\revpoint{3}{5}
+\revpoint{3}{4}
+\revpoint{3}{6}
 \begin{algorithm2e}[!ht]
@@
         and $m$ will record the number of extended haplotypes.
+        Note that if no haplotypes can be extended ($m=0$), we return $m_e$ as $\infty$
+        so that the path is not merged in \algorithmref{alg:extend}.
     }
@@
     \Fn{\mergeable($c$, $T_O$, $T_I$)}{
+    	\tcc{We construct paths beginning at $c$ and traverse through inheritance in $T_O$ and $T_I$.}
         \text{Let } $p_i = P_I[c]$, 
@@
         \While{\tn{True}}{
+        	\tcc{States to determine if paths can still be constructed backward in time.}
             \text{Set } $y_i = (p_i \neq NULL)
@@
                 \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
-                    \text{Set } $m_e = m_e + 1$.\;
+                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_i$ is not in $T_O$ so we add a new edge.}
                 }
@@
                 \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
-                    \text{Set } $m_e = m_e + 1$.\;
+                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_o$ is not in $T_I$ so we add a new edge.}
                 }
@@
                 $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
-                \text{and } $m = m + 1$ .\;
+                \text{and } $m = m + 1$ . \tcc*[r]{We extend haplotype from $p_o$ into $T_I$.}
             }
@@
         \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
-            \text{Set } $m_e = \infty$. \;
+            \text{Set } $m_e = \infty$. \tcc*[r]{The paths constructed at $c$ cannot be merged.}
         }
@@
 we define a measure of \emph{matched span} to quantify this.
-The method is implemented in the \tscompare{} package.
+The method (including \eqnref{eqn:matched-spans} - \eqnref{eqn:true-proportion}) \revpoint{1}{3}
+ is implemented in the \tscompare{} package.
 
@@
 so that
-\begin{align}
+\begin{align}\label{eqn:matched-spans}
     \match(\T_1, \T_2)
@@
 To do this, we define the \emph{inverse matched span} of $\T_1$ in $\T_2$ as
-\begin{align}
+\begin{align}\label{eqn:non-matched-spans}
 	\invmatch(\T_1,\T_2) = \sum_{n_2\in N_2} \max_{n_1\in \alpha^{-1}(n_2)} m(n_1,n_2)
@@
 where $\alpha^{-1}(n_2)$ is the set of all nodes $n_1\in \T_1$ 
-whose best match is $n_2$.
+whose best match is $n_2$.  \revpoint{3}{15}
 This differs from the matched span of $\T_2$ in $\T_1$
@@
 for comparing pairs of tree sequences than the original matched span and inverse matched span, whose units are length of spans.
-We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.
-
+We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.\revpoint{1}{4}
+
+\revpoint{3}{16}\revpoint{3}{17}
 \begin{figure}[!ht]
@@
         For two tree sequences $T1$ and $T2$ the \emph{matched span}, $\match(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
-        based on identical sample sets.
-        In this example, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
-        Node $4$ has no match on $[4,5)$ and matches with node $9$ on $[5,6)$. Thus the maximal mapping for 4 should be to itself. 
-        On the rest of the genome, all of nodes match with their identical counterpart.
+        based on identical sample sets. We first can compute the total span of nodes in $T1$ and $T2$ as $\Vert T1\Vert = 46$ and $\Vert T2\Vert = 47$.
+        First, we always require that sample nodes match to their identical counterpart. Then we match intermediate nodes by comparing their
+        descendent samples across pairs of local trees over the genome. Here we have distinct local tree pairings on the genome segments:
+        $[0,1), [1,2), [2,4), [4,5),$ and $[5,6)$.
+        Node $4$ has no match on $[4,5)$, matches with node $9$ on $[5,6)$, and matches with node $4$ in $T2$ everywhere else. 
+        Thus the maximal mapping for 4 should be to itself since $m(4,4) > m(4,9)$. 
+        Nodes $5, 6,$ and $7$ all match with their identical counterpart.
+        Lastly, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
         This makes the matched span $\match(T1,T2)=43$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
@@
         since its only possible match ($4 \in T1$) was not the best match from $T1 \to T2$. 
-        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.
+        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.\newline
+        Additionally, let's compute $\operatorname{ARF}(T2,T1)$ and $\operatorname{TPR}(T2,T1)$.
+        For matched spans, we again see that sample nodes, root 6, and nodes 4, 5, and 7 will match to their identical counterpart
+        and have matched nodes spans 6 (for each sample), 6, 4, 4, and 2, respectively.
+        Node 9 matches to node 4 with a matched node span of 1. Thus we compute $\operatorname{ARF}(T2,T1)=\frac{41}{47}$. 
+        The inverse matched span will match nodes 5, 6, 7 and sample nodes to their identical counterparts. 
+        Node 8 in $T1$ had no match from $T2$, so it has no inverse match. Node 4 had two best matches from $T2$: nodes 4 and 9.
+        Since 4 shares more span with its identical counter part the inverse match for 4 in $T1$ is node 4 in $T2$. Therefore we compute 
+        $\operatorname{TPR}(T2,T1)=\frac{40}{46}$. 
         \label{fig:conceptual_discrepancy}
@@
  is a metric.
-To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
-and view each ARG as a subset of the space $[0,L) \times N \times N$,
-where $N$ is the shared set of nodes.
-Then, dissimilarity is the Lebesgue measure of the relative difference of the two sets:
+ To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$, called $\varphi\colon N\to N$, where $N$ is the shared set of nodes.
+ View both ARGs as a subset of the space $N\times N\times [0,L)$ so that elements in the ARG are tuples $(n, \varphi(n), A)$ where $A=\cup_k [a_{k-1}, a_k)$ represents the portion of the genome where $n$ exists in the ARG.
+ Then for each node $n$ in $N$ its representatives in $\T_1$ and $\T_2$ can differ with respect to the portion of the genome where $n$ exists, ie. $A_{\T_1}$ and $A_{\T_2}$ may not be equal.
+ Then the relative difference (as sets) of the two ARGs, $\T_1\setminus \T_2$, is the collection of tuples $(n, \varphi(n), A_{\T_1}\setminus A_{\T_2})$.
+This is the portion of \emph{non-matched spans} between $n$ and $\varphi(n)$.
+%To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
+%and view each ARG as a subset of the space $[0,L) \times N \times N$,
+%where $N$ is the shared set of nodes.
+Thus, the measure of non-matched spans
+%dissimilarity 
+is the Lebesgue measure of the relative difference of the two sets:
 $|\T_1 \setminus \T_2|$,
@@
 95\% of nodes have no incorrectly added span; \revpoint{2}{4}
-and those incorrectly added spans are nearly always a small fraction of the original span.
+and those incorrectly added spans are nearly always a small fraction of the original span (\figref[B]{fig:node-spans}).
 The added information is significant:
 the algorithm typically increases spans (i.e., lengths of ancestral haplotypes)
-by around 50\%.
+by around 50\% (\figref[C, D]{fig:node-spans}).\revpoint{3}{19}
 
@@
         ordered by total span in the original ARG (which includes unary nodes).
-        Dark blue dots in both show the spans in this original ARG.
-        In \textbf{(C)}, lighter green dots show
+        Dark blue dots in both show the span of each node in this original ARG.
+        In \textbf{(C)}, lighter green dots show node
         span after removing unary spans using \textit{simplify},
         while in \textbf{(D)},
-        lighter red dots show span after extending haplotypes,
+        lighter red dots show node span after extending haplotypes,
         i.e., applying \algorithmref{alg:extend} to the simplified ARG.
@@
 (Since the \texttt{simplify} operations removes these unary portions of haplotypes,
-the ``simplified'' line (``S'') on the bottom two plots of Figure~\ref{fig:dissimilarity} shows
+the ``simplified'' line (``S'') on the bottom two plots of \figref{fig:dissimilarity} shows
 the proportion of nodes' spans on which they are not unary.)
@@
 and extending haplotypes can infer more than half of this missing unary span from coalescent information only.
+
+While our results are accurate at both the individual and genomic scale, 
+we suspect that corrections can be made to further increase accuracy. 
+\algorithmref{alg:extend} relies on only the structure of the ARG as a tree sequence (its nodes, edges, and local trees), 
+and does not use any other parameters for inference.
+However, it is possible that information including the recombination rate or gene conversion rate of an ARG could improve 
+accuracy and reduce the amount of incorrectly added span for individuals.
+We decided to omit incorporating this information for simplicity and efficiency of our construction.
 
@@
 \revpoint{2}{5}
+\revpoint{3}{20}
 The method to extend haplotypes described here is available through the \tskit{}

### 2025-05-21 paper/writeup.tex
.

@@
 \newcommand{\pluseq}{\mathrel{{+}{=}}}
+\newcommand{\arf}\operatorname{ARF}
+\newcommand{\tpr}\operatorname{TPR}
 
@@
 \begin{align*}
-    \frac{1}{L} \left( \sum_{n_1 \in N_1} m'(n_1, \T_2)  + \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
+    \frac{1}{L} \left( \|\T_1\| + \|\T_2\|
+        - \sum_{n_1 \in N_1} m'(n_1, \T_2) - \sum_{n_2 \in N_2} m'(n_2, \T_1) \right).
 \end{align*}
+(This holds because the average number of nodes per local tree in $\T_1$ is $\|\T_1\|/L$,
+so the difference gives the average number of non-matching nodes.)
 In other words, we require a node in $\T_1$ to match the \emph{same} node in $\T_2$
@@
 was defined by \citet{robinson1981comparison}
-% JK - change edge->branch here to avoid confusion with tskit/ARG edge
 to be the minimum number of branch contraction/expansion operations needed to move
 from one tree to the other
-(which they then show is equal to the number of edges 
+(which they then show is equal to the number of branches 
 that induce different splits on the labels).
@@
 for comparing pairs of tree sequences than the original matched span and inverse matched span, whose units are length of spans.
-We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.\revpoint{1}{4}
+We compute the ARF and TPR of in a simple example in \figref{fig:conceptual_discrepancy}.
 
@@
 ``How much of the true relationships does this ARG represent?''
+Indeed, even a symmetrized version of non-matched span is not a metric:
+for instance, if $\T_2$ is produced from $\T_1$ by adding a new, entirely unary node,
+then $\nmatch(\T_1,\T_2) = \nmatch(\T_2,\T_1) = 0$.
+This is again by design: in general, edges in ARGs represent transmission of genetic material
+through many ancestors, and so addition of entirely unary ancestors on an edge
+is (arguably) not wrong.
+
 However, it is worth noting that the symmetrized version of non-matched span
-$$\overline{\nmatch}(\T_1,\T_2) = \nmatch(\T_1,\T_2) + \nmatch(\T_2, \T_1),$$
+$$\overline{\nmatch}(\T_1,\T_2) = \nmatch(\T_1,\T_2) + \nmatch(\T_2, \T_1)$$
 is a metric.
- To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$, called $\varphi\colon N\to N$, where $N$ is the shared set of nodes.
- View both ARGs as a subset of the space $N\times N\times [0,L)$ so that elements in the ARG are tuples $(n, \varphi(n), A)$ where $A=\cup_k [a_{k-1}, a_k)$ represents the portion of the genome where $n$ exists in the ARG.
- Then for each node $n$ in $N$ its representatives in $\T_1$ and $\T_2$ can differ with respect to the portion of the genome where $n$ exists, ie. $A_{\T_1}$ and $A_{\T_2}$ may not be equal.
- Then the relative difference (as sets) of the two ARGs, $\T_1\setminus \T_2$, is the collection of tuples $(n, \varphi(n), A_{\T_1}\setminus A_{\T_2})$.
-This is the portion of \emph{non-matched spans} between $n$ and $\varphi(n)$.
-%To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
-%and view each ARG as a subset of the space $[0,L) \times N \times N$,
-%where $N$ is the shared set of nodes.
-Thus, the measure of non-matched spans
-%dissimilarity 
+To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
+and so can treat them as having the same set of samples $S$ and nodes $N$
+(with $S \subset N$).
+To each ARG $\T_i$ corresponds a mapping $\mathcal{D}_i : [0, L) \times N \to \{0,1\}^S$,
+such that $\mathcal{D}_i(u,n)$ gives the set of samples that descend from $n$ at position $u$
+along the genome,
+and hence a set $P_i = \{(u,n,\mathcal{D}_i(u,n)): u \in [0,L), n \in N\}$.
+The non-matched span $\nmatch(\T_1,\T_2)$
 is the Lebesgue measure of the relative difference of the two sets:
-$|\T_1 \setminus \T_2|$,
+$|P_1 \setminus P_2|$,
 and so the symmetrized version is the measure of the symmetric difference
@@
 This also extends to two tree sequences with different numbers of nodes,
-$|N_1|\neq|N_2|$, as we can take the minimum over all possible matchings $\T_1\to\T_2$ and $\T_2\to\T_1$.
+since we can take the minimum over all possible matchings $\T_1\to\T_2$ and $\T_2\to\T_1$.
 

### 2025-05-25 paper/review-responses.tex
pass through reviews

--- 2025-05-08 paper/review-responses.tex
@@
 \reply{
+    Oh dear, that was an embarrassing error
+    (that happily doesn't affect anything else in the paper).
+    This has been corrected.
 }
@@
 
-\reply{
+\reply{No, the weighted mean square error defined in \eqref{eqn:wrmse} is not shown for our later experiments shown in \figref{fig:dissimilarity}, however it is always computed when using our implementation is \texttt{tscompare};
+we now explicitly say this \revref.
 }
@@
 
-\reply{
+\reply{``dissimilarity'' was our original language to describe the non-matched spans, and we missed this use of the old name.
+We have now changed the caption in \figref{fig:conceptual_discrepancy} to reflect that change.
 }
@@
 \reply{
+    Thanks again to the reviewer for carefully thinking about what we wrote,
+    which on further examination has turned out to be wrong in several ways (apologies!).
+    Please see the updated version that we believe to be correct
+    and thus hopefully less confusing. \revref
 }
@@
 \reply{
+    We agree on all counts -- in particular,
+    it would be helpful to dig into why haplotypes are being inferred wrongly;
+    however, this more detailed investigation is beyond the scope of this paper
+    (since it's more an investigation into the inference methods).
+    More generally: if an inferred ARG is incorrect for certain nodes, our method of extend haplotypes exacerbates this issue as extending an incorrectly inferred node extends this incorrect node span even further; we've added a note to this effect \revref.
+    \comment{TODO: add number of edges?}
+    Based on our tests in \figref{fig:speed_and_edges} and \tabref{tab:edge-counts}, these results would additionally hold for the inferred ARGs.
+    For this reason, we didn't include additional figures for edge counts and computation time on inferred ARGs.
 }
@@
 \reply{
+    This is a good suggestion;
+    rather than break up the flow (for less sophisticated readers)
+    we've inserted a pointer to these parts of the discussion, at \revref.
 }
@@
 \reply{
+    Good suggestion; we've added a note about this. \llname{bad_spans}
 }
@@
 
+We're glad you think it's important, and apologies for the errors.
+
 \begin{quote}
@@
 \end{quote}
+
+As you'll see, we've tried to improve clarity throughout;
+hopefully it is more digestable now.
 
@@
 \reply{
+    We think it's reasonably standard for one algorithm to make use of another
+    (for instance, many algorithms apply a \texttt{sort} at some point).
+    However, it's a good point that referring abstractly to ``\algorithmref{alg:extend}'' and 
+    ``\algorithmref{alg:mergeable}'' is a little abstract and hard to remember.
+    So, we've gone back to places where the text just said ``\algorithmref{alg:extend}'',
+    and where we thought helpful, changed it to something like
+    ``extending haplotypes with \algorithmref{alg:extend}''.
+    For more background:
+    originally we had both \algorithmref{alg:extend} and \algorithmref{alg:mergeable} within 
+	one algorithm environment.
+    However, the \texttt{Mergeable} is a conceptually separate operation
+    that by separating it out makes understanding \texttt{ExtendHaplotypes} easier.
+	\algorithmref{alg:extend} is the first algorithm as it is the main construction and the most important of the two.
 }
 
-\begin{point}{}
+\begin{point}{} %(2)
 \algorithmref{alg:mergeable} is mentioned in an explanation a full page before it
@@
 \reply{
+    This is a consequence of the typesetting: \algorithmref{alg:mergeable} 
+    is in a floating environment, and
+    the earliest the environment will appear is the next page following the previous figure.
+    Rest assured that Genetics will typeset this appropriately.
 }
 
-\begin{point}{}
+\begin{point}{} %(3)
 \algorithmref{alg:extend} seems to be named ExtendHaplotypes and \algorithmref{alg:mergeable} seems
@@
 \reply{
+    This point overlaps with point 3.1 -- see our response above for how we've hopefully
+    improved the situation.
+    We like to think of ExtendHaplotypes and Mergable as the names of the functions contained in \algorithmref{alg:extend} and \algorithmref{alg:mergeable}, respectively.
 }
 
-\begin{point}{}
+\begin{point}{} %(4)
 ``Mergeable'' is an unfortunate name for a function which does not
@@
 \reply{
+    A reasonable point; we've changed the name to \texttt{MergeNum}.
+	For clarity, we have also included a note about returning infinity in the description of \algorithmref{alg:mergeable}.
 }
 
-\begin{point}{}
+\begin{point}{} %(5)
 In \algorithmref{alg:extend} the variable $m$ means the number of new edges to be
@@
 
-\reply{
+\reply{Good point, this change has been made.
 }
 
-\begin{point}{}
+\begin{point}{} %(6)
 Comments in the pseudocode would make it easier to follow, as could
@@
 
-\reply{
+\reply{Good point. We have included some comments for clarity.
+However, we're going to respectfully differ on stylistic choice of concise variable names.
 }
@@
 
-\reply{
+\reply{While we agree in general,
+within the context of our algorithm we are interested paths of inheritance 
+	between the edges removed from a local tree (the out-forest) and the edges just added into the next local tree (the in-forest). 
+	Because of our language use of out and in, we feel like our choice of $O$ and $I$ is reasonable.
 }
@@
 Good catch -- and, many apologies --
-that is indeed a bug; the correction is changing line 7 to
-``if $m \le M$''.
+that is indeed a bug; the correction changes line 7 to
+``if $m \le M$'' and line 10 to ``Set $M' = \min(m_e, M')$.''.
 (To be clear, the bug was in the paper, not the implemented code;
@@
 \reply{
+    A node is NULL when it has no parent; this is now explained in the description
+    of the algorithm.
+    Apologies, this is part of the conventions in tskit.
 }
@@
 \reply{
-Good idea, but on consideration we think that adding the rest of the tree
+Good idea,
+we now reference the figure in the text,
+at \revref{} and \llname{ref_conceptual2}.
+On further consideration we think that adding the rest of the tree
 here would make the figure too confusing
 (not so much for the trees in A\&C, but more so in the haplotype diagrams B\&D).
-We do now reference the figure in the text,
-at \revref{} and \llname{ref_conceptual2}.
 }
@@
 
-\reply{
+\reply{We state how time is measured here \revref;
+we've added a reminder at \llname{time_reminder}.
+We do use time in our algorithm, but mainly as a way to break ties, as is the case with \eqnref{eqn:matched-spans}.
+This tie break will default to the first node if the ARG does not contain time information.
+We've also added a note about undated ARGs \llname{undated}.
 }
@@
 \reply{
+    This is exactly right:
+    we don't assume that node times are unique,
+    as it's common (e.g., from discrete-time simulators)
+    for this not to be true;
+    so the check is to avoid producing an illegal situation
+    (parent and child at the same time).
+    We now explain this at \llname{undated}.
 }
 
-\begin{point}{} % (6) 
+\begin{point}{\llname{truth}}
 ``In what follows we think of $T_2$ as the true ARG and $T_1$ as an
@@
 \reply{
+    We've added extra emphasis to the point that the quantities are not symmetric
+    \llname{asymmetric};
+    furthermore the additional discussion in the ``Metrics on ARGs'' section
+    should help reinforce this point.
+    As for reversing $\T_1$ and $\T_2$,
+	we think that while $\match(\T_2,\T_1)$ might not be the most useful quantity,
+    it can still give some valuable information between inferred $\T_1$ and true $\T_2$.
+    We discuss this when comparing inverse matched spans between $(\T_1,\T_2)$ and matched spans between $(\T_2, \T_1)$ \revref.
 }
@@
 \reply{
+    We have expanded our example to hopefully give more clarity (see additional text
+    in caption).
+    The first reference to \figref{fig:conceptual_discrepancy} occurs at the end
+    of the second paragraph in the section on comparison;
+    it would be hard to include the example earlier
+    (and, we're not sure how helpful it will be until
+    $\arf$ and $\tpr$ are defined).
 }
@@
 \reply{
+    We sympathize with the reviewer that working through even a simple example like this
+    is tricky and confusing (but, we also find it very helpful).
+    Good point about the same forward and inverse matches.
+    Note that we did construct this example so that matching for node 4 requires a max to be taken over shared spans of its two possible matches 4 and 9.
+To show some variety, we have added computation of $\match(\T_2,\T_1)$ to show some additional variety and hopefully give more clarity to the methods themselves (see caption).
+The samples are included because in practice they do \emph{not} necessarily always match:
+consider a case where some of the samples are ancestors of others
+(ancient samples; trios; etcetera).
 }
@@
 
-\reply{
+\reply{Thanks for the feedback. We have added substantial additional explanation at \revref{} and the caption of \figref{fig:node-spans}.
 }

### 2025-05-25 paper/writeup.tex
pass through reviews

@@
 \newcommand{\pluseq}{\mathrel{{+}{=}}}
-\newcommand{\arf}\operatorname{ARF}
-\newcommand{\tpr}\operatorname{TPR}
+\newcommand{\arf}{\operatorname{ARF}}
+\newcommand{\tpr}{\operatorname{TPR}}
 
@@
 How can we do this, and how accurate is the resulting inference?
+See the discussion for more on what this question assumes,
+and the connection to parsimony. \revpoint{1}{7}
 
@@
 % Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
-Each edge $e \in E$ describes inheritance between ancestors and their descendants within the population.
+Each edge $e \in E$ describes inheritance between an ancestor and a descendant.
 The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
@@
 the algorithm traverses up from $c$ in both $F_O$ and $F_I$;
-terminating if a node in the other tree is found
+the variables $y_i$ and $y_o$ indicate whether the next node upwards in the traversals
+(i.e., the next older node, as recall $t_n$ is amount of time in the past that $n$ lived) \llabel{time_reminder}
+is in $F_I$ or $F_O$.
+The traversals terminate if a node in the other tree is found
 (i.e., if the node traversed in $F_O$ is in $T_{k+1}$ or if the node traversed in $F_I$ is in $T_k$)
@@
 until no more are found.
+
+Note that the algorithm could be applied to an undated ARG, \llabel{undated}
+with some adjustments.
+The algorithm only uses node time for two reasons:
+convenience, when iterating jointly up the two paths;
+and to avoid illegal conditions: if we tried to merge two paths on which
+some $u_i$ had the same time as some $v_j$,
+then we would have a parent and child with the same time, which is not allowed.
 
@@
 which performs the actual extending of haplotypes.
-This algorithm is essentially the same as \texttt{Mergeable} in \algorithmref{alg:mergeable},
+This algorithm is essentially the same as \texttt{MergeNum} in \algorithmref{alg:mergeable},
 except with additional bookkeeping.
@@
         paths that add fewer new edges are merged first
-        because \texttt{Mergeable} returns the number of new edges required
+        because \texttt{MergeNum} returns the number of new edges required
         to merge the two paths, and $M$ is always less than or equal to
@@
 	\label{alg:extend}
-	\SetKwFunction{mergeable}{Mergeable}
+	\SetKwFunction{mergeable}{MergeNum}
 	\SetKwFunction{merge}{Merge}
@@
                 \For{$e \in O_k$}{
-                    \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Count the number of new edges to be merged.}
+                    \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Find the number of new edges required.}
                     \eIf{$m_e \le M$}{
-                        \merge($c_e$, $T_k$, $T_{k+1}$)\tcc*[r]{Merge extended haplotypes from $T_k$ into $T_{k+1}$.}
+                        \merge($c_e$, $T_k$, $T_{k+1}$)\tcc*[r]{Extended haplotypes from $T_k$ into $T_{k+1}$.}
                     }{
-                        \text{Set } $M' = \min(m_e, M)$.\tcc*[r]{Bookkeep $M$ to determine an ordering for path merging.}
+                        \text{Set } $M' = \min(m_e, M')$.
                     }
@@
 \end{algorithm2e}
-\revpoint{3}{5}
-\revpoint{3}{4}
-\revpoint{3}{6}
+
 \begin{algorithm2e}[!ht]
@@
         in the set of edges to be removed and added, respectively
-        (i.e., in $F_O$ and $F_I$).
+        (i.e., in $F_O$ and $F_I$);
+        these are NULL if $n$ has no parent.
         The variable $m_e$ will record the number of new edges to be added,
@@
 	\label{alg:mergeable}
-	\SetKwFunction{mergeable}{Mergeable}
+	\SetKwFunction{mergeable}{MergeNum}
 	\DontPrintSemicolon
     \Fn{\mergeable($c$, $T_O$, $T_I$)}{
-    	\tcc{We construct paths beginning at $c$ and traverse through inheritance in $T_O$ and $T_I$.}
+    	 %\tcc{We construct paths beginning at $c$ and traverse through inheritance in $T_O$ and $T_I$.}
         \text{Let } $p_i = P_I[c]$, 
@@
         \While{\tn{True}}{
-        	\tcc{States to determine if paths can still be constructed backward in time.}
             \text{Set } $y_i = (p_i \neq NULL)
                 \ \&\ (p_i \notin T_O)
-                \ \&\ (t_i < t_o)$ \;
+                \ \&\ (t_i < t_o)$ \; \tcc*{Next node is from $T_I$.}
             \text{and } $y_o = (p_o \neq NULL)
                 \ \&\ (p_o \notin T_I)
-                \ \&\ (t_o < t_i)$ \;
+                \ \&\ (t_o < t_i)$ \; \tcc*{Next node is from $T_O$.}
             \If{not ($y_i$ or $y_o$)}{ \Break\; }
@@
                 \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
-                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_i$ is not in $T_O$ so we add a new edge.}
+                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_i$ is not in $T_O$}
                 }
@@
                 \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
-                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_o$ is not in $T_I$ so we add a new edge.}
+                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_o$ is not in $T_I$}
                 }
@@
                 $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
-                \text{and } $m = m + 1$ . \tcc*[r]{We extend haplotype from $p_o$ into $T_I$.}
+                \text{and } $m = m + 1$ . \tcc*[r]{$p_o$ would be extended into $T_I$.}
             }
@@
         \If{$m = 0$ or $p_i \neq p_o$ or $p_i = NULL$}{
-            \text{Set } $m_e = \infty$. \tcc*[r]{The paths constructed at $c$ cannot be merged.}
+            \text{Set } $m_e = \infty$. \tcc*[r]{The paths cannot be merged.}
         }
@@
 it is straightforward to remove the portions of each node's span on which it is unary,
-apply \algorithmref{alg:extend},
+extend haplotypes (\algorithmref{alg:extend}),
 and quantify how much node span was correctly or incorrectly added.
-However, we are also interested in whether \algorithmref{alg:extend}
+However, we are also interested in whether extending haplotypes
 improves \emph{inferred} ARGs.
@@
 and the same set of sample nodes;
-in what follows we think of $\T_2$ as the true ARG and $\T_1$ as an inferred ARG.
+in what follows we think of $\T_2$ as the true ARG and $\T_1$ as an inferred ARG. \llabel{truth}
 We would like to measure (a) how much of $\T_1$ is found in $\T_2$;
@@
 and difference between the tree sequences,
-thought of vaguely as sets.)
+thought of vaguely as sets;
+so note in particular that switching the roles of $\T_1$ and $\T_2$ gets 
+an entirely different set of quantities.) \llabel{asymmetric}
 Roughly speaking, we first identify matching nodes
@@
 
-\revpoint{3}{16}\revpoint{3}{17}
 \begin{figure}[!ht]
@@
 for instance, if $\T_2$ is produced from $\T_1$ by adding a new, entirely unary node,
-then $\nmatch(\T_1,\T_2) = \nmatch(\T_2,\T_1) = 0$.
+then $\arf(\T_1,\T_2) = \arf(\T_2,\T_1) = 0$.
 This is again by design: in general, edges in ARGs represent transmission of genetic material
@@
 is (arguably) not wrong.
-
-However, it is worth noting that the symmetrized version of non-matched span
-$$\overline{\nmatch}(\T_1,\T_2) = \nmatch(\T_1,\T_2) + \nmatch(\T_2, \T_1)$$
-is a metric.
-To see this, first suppose that we have a bijection between the nodes of $\T_1$ and $\T_2$,
-and so can treat them as having the same set of samples $S$ and nodes $N$
-(with $S \subset N$).
-To each ARG $\T_i$ corresponds a mapping $\mathcal{D}_i : [0, L) \times N \to \{0,1\}^S$,
-such that $\mathcal{D}_i(u,n)$ gives the set of samples that descend from $n$ at position $u$
-along the genome,
-and hence a set $P_i = \{(u,n,\mathcal{D}_i(u,n)): u \in [0,L), n \in N\}$.
-The non-matched span $\nmatch(\T_1,\T_2)$
-is the Lebesgue measure of the relative difference of the two sets:
-$|P_1 \setminus P_2|$,
-and so the symmetrized version is the measure of the symmetric difference
-$|\T_1\Delta\T_2|$,
-which is well-known to be a metric \citep{rudin1976principles}. 
-If the two ARGs have the same number of nodes,
-we can consider all bijections between their nodes.
-The symmetric difference between $\T_1$ and $\T_2$,
-related through each bijection is a metric.
-Then the minimum over all such metrics will still be a metric
-since the minimum over a finite number of metrics is also a metric.
-This also extends to two tree sequences with different numbers of nodes,
-since we can take the minimum over all possible matchings $\T_1\to\T_2$ and $\T_2\to\T_1$.
+(Note that our metrics do distinguish these two ARGs: in this example $\tpr(\T_1,\T_2) < 1$.)
+
+A related (but not identical) way to construct a metric on (undated) ARGs
+modifies the definitions above so that $\alpha$ is a bijection.
+To do this, write $N_1$ and $N_2$ for the nodes of $\T_1$ and $\T_2$ respectively,
+and suppose that $\varphi$ is a bijection from $N_1$ to $N_2$.
+Since we want to allow nodes to remain unpaired,
+for notational convenience suppose that both $N_1$ and $N_2$ have appended an arbitrary
+number of ``unrelated'' nodes (since these are not ancestral to any samples they do not affect the metrics).
+Then, define $d_\varphi(\T_1,\T_2)$ to be the total span of nodes $n_1 \in N_1$
+on which $n_1$ and $\varphi(n_1)$ do not match (i.e., are not ancestral to the same set of samples),
+plus the total span for nodes $n_2 \in N_2$ on which $n_2$ and $\varphi^{-1}(n_2)$ do not match.
+The quantity $\min_\varphi d_\varphi(\T_1,\T_2)$,
+where the minimum is over bijections that preserve the partial ordering induced by the ARG,
+defines a metric on undated ARGs.
+(Actually, as written it defines a metric on the equivalence classes of undated ARGs
+that are obtained by removing any structure not ancestral to any samples.)
+To see this, note that
+relative ordering and descendant sample sets uniquely determine the (equivalence class of an) ARG,
+and that the triangle inequality is satisfied because
+$d_{\varphi_{12}}(\T_1,\T_2) + d_{\varphi_{23}}(\T_2,\T_3) \le d_{\varphi_{12}\circ\varphi_{23}}(\T_1,\T_3)$.
+
+This differs from the definitions above because here
+we require a strict one-to-one match between nodes in $\T_1$ and $\T_2$
+(i.e., $\varphi$ is bijective, while there may be more than one node in $\T_1$
+that $\alpha$ matches to the same node in $\T_2$).
+We did not require this property above for two reasons:
+first, the arguable non-wrongness of additional unary nodes,
+and second, this definition makes computation much more difficult.
 
@@
 then the hypothesized reduction of 1/3 is achieved for long sequences.
-It is possible for \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
+It is possible for extending haplotypes with \algorithmref{alg:extend} to add edges, as in \figref{fig:extending_diagram},
 but we still expect the number of edges to decrease by more than 1/3.
@@
 95\% of nodes have no incorrectly added span; \revpoint{2}{4}
-and those incorrectly added spans are nearly always a small fraction of the original span (\figref[B]{fig:node-spans}).
+and those incorrectly added spans are nearly always a small fraction of the original span
+(\figref[B]{fig:node-spans} shows that incorrectly added span is often a factor of
+10--100 smaller than the added span).
 The added information is significant:
 the algorithm typically increases spans (i.e., lengths of ancestral haplotypes)
-by around 50\% (\figref[C, D]{fig:node-spans}).\revpoint{3}{19}
+by around 50\%.
+For instance, \figref[A]{fig:node-spans} \revpoint{3}{19}
+shows that tens of megabases have been added to the spans of many nodes;
+while comparing \figref[C]{fig:node-spans} to \figref[D]{fig:node-spans} 
+shows that much of the span removed by simplify has been replaced --
+in those figures, each point is one node;
+span after extending (points in D) are much closer to the original full span
+(solid blue points forming a line)
+than is span after simpmlifying (points in C).
+Some of the incorrectly added spans may be because of violations of the parsimony assumption,
+while others are due to situations with ambiguous information
+(and hence more than one way to extend a haplotype).
+It may be possible to better deal with the latter situation,
+at the cost of a more complex algorithm. \llabel{bad_spans}
 
@@
         \textbf{(A)}
-        Distribution of total amount of span added across nodes by \algorithmref{alg:extend};
+        Distribution of total amount of span added across nodes by 
+        extending haplotypes with \algorithmref{alg:extend};
         note the log scale on the $y$ axis.
@@
 Line ``SE'' in \figref{fig:dissimilarity} shows total amounts of span removed by simplification
-and re-inferred by \algorithmref{alg:extend} (correctly and incorrectly).
+and re-inferred by extending haplotypes with \algorithmref{alg:extend} (correctly and incorrectly).
 (Lines labeled with ``I'' involve re-inference of the ARG; discussed next.)
@@
 with tsinfer adds relatively little span (less than about $1\%$).
-Applying \algorithmref{alg:extend} to the inferred-and-then-simplified ARG
+Extending haplotypes with \algorithmref{alg:extend} of the inferred-and-then-simplified ARG
 (``ISE''; solid green line)
@@
 but also introduces incorrect spans.
+Presumably, this occurs because both correctly and incorrectly inferred haplotypes are extended. \revpoint{1}{6}
 Further work is needed to determine how the balance of ``true and false positive rates''

### 2025-05-26 paper/review-responses.tex
final pass through reviews

@@
 we also provide a pdf with differences to the previously submitted version highlighted.
+
+The main changes we've made are:
+(1) additional detail and description to guide the reader through some of the sections;
+(2) a fix to a bug in the algorithm
+(thanks, Reviewer 3; note this did not affect the code, just the paper);
+and (3) more description of how extending haplotypes affects efficiency on inferred ARGs.
 
@@
     More generally: if an inferred ARG is incorrect for certain nodes, our method of extend haplotypes exacerbates this issue as extending an incorrectly inferred node extends this incorrect node span even further; we've added a note to this effect \revref.
-    \comment{TODO: add number of edges?}
-    Based on our tests in \figref{fig:speed_and_edges} and \tabref{tab:edge-counts}, these results would additionally hold for the inferred ARGs.
-    For this reason, we didn't include additional figures for edge counts and computation time on inferred ARGs.
+    We've also added \figref{fig:inferred_runtime} that shows numbers of edges and runtime
+    for the inferred ARGs used in our tests,
+    and additional discussion of this point \llname{inferred_runtime}.
+    The results are interesting, but detailed investigation is outside the scope of this paper --
+    here we are introducing concepts and tools (and, there's plenty to digest);
+    while further investigation could depend strongly on the inference method used.
 }

### 2025-05-26 paper/writeup.tex

@@
 How can we do this, and how accurate is the resulting inference?
-See the discussion for more on what this question assumes,
+See the Discussion for more on what this question assumes,
 and the connection to parsimony. \revpoint{1}{7}
@@
 % Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
-Each edge $e \in E$ describes inheritance between an ancestor and a descendant.
-The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
+Each edge $e \in E$ describes inheritance between an ancestor and a descendant
 over a segment of genome $[\ell_e, r_e)$.
+The ancestor and descendant are represented, respectively, by the parent node $p_e$ and child node $c_e$ of the edge. \revpoint{2}{1}
 % (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
@@
 in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
-We require repeated passes due to the fringe cases of some paths not being mergeable until 
-additional structure is added by a future pass of the algorithm. 
-We suspect that most of these cases occur for paths with large time differences between the parent and child
-nodes, or when a lot of nodes in a path are coalescent, however it is not something we thoroughly investigated
-in the theoretical setting.
-We instead give some empirical results on the number of passes for the algorithm below.
+We require repeated passes because the additional structure added by a given path of the algorithm
+may introduce more mergeable paths.
+Most of these cases seem to occur for paths
+with large time differences between the parent and child nodes,
+or when many nodes in a path are coalescent. 
+% however it is not something we thoroughly investigated in the theoretical setting.
+The number of required passes is in practice small:
+see empirical results below.
 
@@
 so note in particular that switching the roles of $\T_1$ and $\T_2$ gets 
-an entirely different set of quantities.) \llabel{asymmetric}
+an different set of quantities.) \llabel{asymmetric}
 Roughly speaking, we first identify matching nodes
@@
         Additionally, let's compute $\operatorname{ARF}(T2,T1)$ and $\operatorname{TPR}(T2,T1)$.
-        For matched spans, we again see that sample nodes, root 6, and nodes 4, 5, and 7 will match to their identical counterpart
+        Again, the sample nodes, root 6, and nodes 4, 5, and 7 will match to their identical counterpart
         and have matched nodes spans 6 (for each sample), 6, 4, 4, and 2, respectively.
-        Node 9 matches to node 4 with a matched node span of 1. Thus we compute $\operatorname{ARF}(T2,T1)=\frac{41}{47}$. 
+        Node 9 matches to node 4 with a matched node span of 1.
+        Thus, $\operatorname{ARF}(T2,T1)=\frac{41}{47}$. 
         The inverse matched span will match nodes 5, 6, 7 and sample nodes to their identical counterparts. 
-        Node 8 in $T1$ had no match from $T2$, so it has no inverse match. Node 4 had two best matches from $T2$: nodes 4 and 9.
-        Since 4 shares more span with its identical counter part the inverse match for 4 in $T1$ is node 4 in $T2$. Therefore we compute 
+        Node 8 in $T1$ had no match from $T2$, so it has no inverse match, while
+        node 4 had two best matches from $T2$: nodes 4 and 9.
+        Since 4 in $T1$ shares more span with the same-numbered node in $T2$ than with 9,
+        the inverse match for 4 in $T1$ is node 4 in $T2$. Therefore,
         $\operatorname{TPR}(T2,T1)=\frac{40}{46}$. 
@@
 and that the triangle inequality is satisfied because
-$d_{\varphi_{12}}(\T_1,\T_2) + d_{\varphi_{23}}(\T_2,\T_3) \le d_{\varphi_{12}\circ\varphi_{23}}(\T_1,\T_3)$.
-
-This differs from the definitions above because here
+$d_{\varphi_{12}}(\T_1,\T_2) + d_{\varphi_{23}}(\T_2,\T_3) \ge d_{\varphi_{23}\circ\varphi_{12}}(\T_1,\T_3)$.
+
+This metric differs from the definitions above because here
 we require a strict one-to-one match between nodes in $\T_1$ and $\T_2$
@@
 first, the arguable non-wrongness of additional unary nodes,
-and second, this definition makes computation much more difficult.
+and second, requiring a bijection makes computation much more difficult.
 
@@
 while comparing \figref[C]{fig:node-spans} to \figref[D]{fig:node-spans} 
-shows that much of the span removed by simplify has been replaced --
+shows that much of the span removed by simplification has been replaced --
 in those figures, each point is one node;
@@
 (solid blue points forming a line)
-than is span after simpmlifying (points in C).
+than is span after simplifying (points in C).
 Some of the incorrectly added spans may be because of violations of the parsimony assumption,
@@
 (and hence more than one way to extend a haplotype).
-It may be possible to better deal with the latter situation,
+It may be possible to improve the behavior of the algorithm in the latter situation,
 at the cost of a more complex algorithm. \llabel{bad_spans}
@@
 with tsinfer adds relatively little span (less than about $1\%$).
-Extending haplotypes with \algorithmref{alg:extend} of the inferred-and-then-simplified ARG
+Extending haplotypes (with \algorithmref{alg:extend}) of the inferred-and-then-simplified ARG
 (``ISE''; solid green line)
@@
 of subtending samples.) 
-We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Fig-6}.
+We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Fig-6},
+and numbers of edges and runtime for Tajima's D
+in \figref{fig:inferred_runtime}.
 
@@
 should facilitate this exploration.
+
+Another consideration is storage and computational efficiency: \llabel{inferred_runtime}
+extending haplotypes reduces the number of edges,
+and thus filesize and (usually) runtime for computation.
+\figref{fig:speed_and_edges} shows that the case is clear for simulated ARGs,
+substantially reducing both size and runtime.
+However, \figref{fig:inferred_runtime}
+shows the situation is more complex for inferred ARGs:
+extending haplotypes on a \tsinfer-produced ARG actually increases runtime somewhat,
+but simplifying-then-extending reduces runtime dramatically
+while keeping the number of edges roughly the same.
+More work is needed to understand how generalizable this is and what the source of these effects are.
 
@@
 
+\begin{figure}[!hbt]
+    \includegraphics{figures/fig_s5}
+	\caption{
+        Numbers of edges and runtimes for computing Tajima's D
+        for the ARGs displayed in \figref{fig:dissimilarity}.
+        Grey lines connect each group of points that derive from the same original simulated ARG
+        (i.e., each unique combination of sample size and sequence length).
+        The label ``orig'' marks the originally simulated ARG.
+    }
+	\label{fig:inferred_runtime}
+\end{figure}
+
 %%%%%%%%%%%%%%%

### 2025-05-27 paper/writeup.tex
algorithm tweaks

@@
 then the edges of the two paths must lie in $O$ and $I$, respectively.
-Suppose an edge in $O$ has child $c$.
+Suppose an edge in $I$ has child $c$.
 To see if $c$ is the base of a pair of mergeable paths,
@@
         Extend haplotypes. Given an ARG $\T$ with $N$ trees $T_1, \ldots, T_N$
-        for which edges $O_k$ are removed to transition from $T_k$ to $T_{k+1}$,
+        for which edges $I_k$ are added to transition from $T_k$ to $T_{k+1}$,
         identify and merge all mergeable paths (see text).
@@
     	\tcc{For pairs of trees in the ARG, find mergeable paths.}
-        \For{$k$ in $1 \ldots N$}{
+        \For{$k$ in $1 \ldots N-1$}{
             \text{Set } $M = 0$ \text{ and }
@@
             \While{$M < \infty$}{
-                \For{$e \in O_k$}{
+                \For{$e \in I_k$}{
                     \text{Set } $m_e$ = \mergeable($c_e$, $T_k$, $T_{k+1}$).\tcc*[r]{Find the number of new edges required.}
@@
                 \ \&\ (p_i \notin T_O)
-                \ \&\ (t_i < t_o)$ \; \tcc*{Next node is from $T_I$.}
+                \ \&\ (t_i < t_o)$ \tcc*{Next node is from $T_I$.}
             \text{and } $y_o = (p_o \neq NULL)
                 \ \&\ (p_o \notin T_I)
-                \ \&\ (t_o < t_i)$ \; \tcc*{Next node is from $T_O$.}
+                \ \&\ (t_o < t_i)$ \tcc*{Next node is from $T_O$.}
             \If{not ($y_i$ or $y_o$)}{ \Break\; }
             \eIf{$y_i$}{
-                \If{$P_I[c] \neq p_i$ and $P_O[c] \neq p_i$}{
-                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_i$ is not in $T_O$}
+                \If{$P_I[c] \neq p_i$}{
+                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$c \to p_i$ is a new edge.}
                 }
@@
             }{
-                \If{$P_I[c] \neq p_o$ and $P_O[c] \neq p_o$}{
-                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$p_o$ is not in $T_I$}
+                \If{$P_O[c] \neq p_o$}{
+                    \text{Set } $m_e = m_e + 1$. \tcc*[r]{$c \to p_o$ is a new edge.}
                 }
@@
                 $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
-                \text{and } $m = m + 1$ . \tcc*[r]{$p_o$ would be extended into $T_I$.}
+                \text{and } $m = m + 1$ . \tcc*[r]{$c \to p_o$ would be extended.}
             }

### 2025-05-27 paper/writeup.tex
t[NULL] fixup

@@
         (i.e., in $F_O$ and $F_I$);
-        these are NULL if $n$ has no parent.
+        these are NULL if $n$ has no parent,
+        and we take $t[\text{NULL}] = \infty$.
         The variable $m_e$ will record the number of new edges to be added,
@@
                 \text{Set } $c = p_i$,
-                $p_i = P_I[p_i]$, \text{ and}\;
-                $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
+                $p_i = P_I[p_i]$, \text{ and }
+                $t_i = t[p_i]$ .\;
+                % $t_i = \text{ if }(p_i = NULL)\text{ then }\infty\text{ else }t[p_i]$ .\; 
             }{
@@
                 \text{Set }$c = p_o$,
-                $p_o = P_O[p_o]$, \;
-                $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
+                $p_o = P_O[p_o]$,
+                $t_o = t[p_o]$ \;
+                % $t_o = \text{ if }(p_o = NULL)\text{ then }\infty\text{ else }t[p_o]$ ,\;
                 \text{and } $m = m + 1$ . \tcc*[r]{$c \to p_o$ would be extended.}
